### 2023-05-15 paper/markdown/main_manuscript.Rmd
Initial commit
Complete analysis code, figures, supplementary figures and supplementary tables

---
title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

## Data and code availability

Anonymized patient data will be made available upon request to the corresponding author. 
The INCOV cohort data are publicly available at https://data.mendeley.com/datasets/96v329bg7g/1.
The analysis pipeline code is available at https://github.com/PiotrTymoszuk/simmun_revision.

\newpage

# Tables

```{r tab-stigma-cohort, tab.cap = 'Characteristic of the local SIMMUN cohort. Numeric variables are presented as medians with interquartile ranges. Categorical variables are presented as percentages and counts within complete observation sets.'}

flextable::flextable(tabs$stigma_cohort) %>% 
  width(1, width = 4, unit = 'cm') %>% 
  width(2:3, width = 4, unit = 'cm') %>% 
  width(4, width = 2.5, unit = 'cm') %>% 
  width(5:6, width = 2.7, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('HADS: hospital anxiety and depression scale; PSS-4: perceived stress scale, 4 item; RBD: receptor binding domain; IgG: immunoglobulin gamma; AU: arbitrary urnits.'),
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5, 
           value = as_paragraph('corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(4, 1, 
           value = as_paragraph('overweight: body mass index (BMI) 25 - 30 kg/m\u00B2; obesity: BMI > 30 kg/m\u00B2'), 
           part = 'body', 
           ref_symbols = 'c') %>% 
  theme_vanilla

```

\newpage

```{r tab-incov-cohort, tab.cap = 'Characteristic of the external INCOV cohort. Numeric variables are presented as medians with interquartile ranges. Categorical variables are presented as percentages and counts within complete observation sets.'}

flextable::flextable(tabs$incov_cohort) %>% 
  width(1, width = 4, unit = 'cm') %>% 
  width(2:3, width = 4, unit = 'cm') %>% 
  width(4, width = 2.5, unit = 'cm') %>% 
  width(5:6, width = 2.7, unit = 'cm') %>% 
  footnote(3, 1, 
           value = as_paragraph('overweight: body mass index (BMI) 25 - 30 kg/mm\u00B2, obesity: BMI > 30 kg/mm\u00B2'), 
           part = 'body', 
           ref_symbols = 'a') %>% 
  footnote(5, 1, 
           value = as_paragraph('WHO ordinal scale for clinical improvement; mild: 1 - 2, moderate: 3 - 4, severe: 5 - 6, critical: 7'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

# Figures

```{r fig-inclusion, fig.width = figur::convert(figs$inclusion, to = 'in')$w, fig.height = figur::convert(figs$inclusion, to = 'in')$h, fig.cap = 'Inclusion scheme for the SIMMUN and INCOV cohorts, and analysis strategy.'}

figs$inclusion$plot

```

__Figure \@ref(fig:fig-inclusion). Inclusion scheme for the SIMMUN and INCOV cohorts, and analysis strategy.__

_TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine - tryptophan ratio; PHE/TYR: phenylalanine - tyrosine ratio; QUIN: quinolinate; 5-HT: serotonin; DA: dopamine; Ab: antibody; IL6: interleukin-6; IL10: interleukin-10; TNF: tumor-necrosis factor alpha; IFNG: interferon gamma._

\newpage

```{r fig-modeling, fig.width = figur::convert(figs$modeling, to = 'in')$w, fig.height = figur::convert(figs$modeling, to = 'in')$h, fig.cap = 'Results of multi-parameter linear modeling of tryptophan, kynurenine, kynurenine/tryptophan ratio, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.'}

figs$modeling$plot

```

__Figure \@ref(fig:fig-modeling). Results of multi-parameter linear modeling of tryptophan, kynurenine, kynurenine/tryptophan ratio, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.__

_Effects of systemic inflammation markers (neopterin [NEO], neutrophil-lymphocyte ratio [NLR]), SARS-CoV-2 infection status, titre of immunoglobulin gamma against the receptor binding domain of the S1 SARS-CoV-2 protein (anti-RBD IgG), scores of anxiety, depression (hospital depression and anxiety scale [HADS]) and mental stress (perceived stress scale, 4 item [PSS4]), age and sex was investigated by multi-parameter linear regression with backward elimination of non-significant terms. Overall model validity was assessed by likelihood-ratio test (LRT). Significant model coefficient estimates with 95% confidence intervals are presented in Forest plots. LRT results (F values, degrees of freedom and p values) and total numbers of complete observations are indicated in the plot captions. Model intercepts (baseline) are indicated in the X axis titles._

_TRP: tryptophan; KYN: kynurenin; KYN/TRP: kynurenine - tryptophan ratio; PHE: phenylalanine; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio._

\newpage

```{r fig-uni-trp, fig.width = figur::convert(figs$uni_trp, to = 'in')$w, fig.height = figur::convert(figs$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}

figs$uni_trp$plot

```

__Figure \@ref(fig:fig-uni-trp). Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__

_Age, systemic inflammatory markers: neopterin (NEO), neutrophil - lymphocyte ratio (NLR), SARS-CoV-2 infection status, depression signs (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of systemic levels of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP). Their association with serum concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._

_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method. (A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold. (B) Scatter plots of serum concentrations of NEO, TRP, KYN and KYN/TRP. Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons. Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._

_(C) Levels of the metabolite levels were compared between infected and uninfected participants, and between participants with and without depression signs by two-tailed T test with Cohen's d effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points. Effect sizes and p values are displayed in the plot captions. Numbers of observations in the strata are indicated in the X axes._

\newpage

```{r fig-uni-tyr, fig.width = figur::convert(figs$uni_tyr, to = 'in')$w, fig.height = figur::convert(figs$uni_tyr, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.'}

figs$uni_tyr$plot

```

__Figure \@ref(fig:fig-uni-tyr). Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.__

_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of systemic levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR). Their association with serum concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._

_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method. (A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold. (B) Scatter plots of patient's age, PHE, TYR and PHE/TYR. Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons. Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._

_(C) Levels of the metabolite levels were compared between infected and uninfected participants by two-tailed T test with Cohen's d effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points. Effect sizes and p values are displayed in the plot captions. Numbers of observations in the strata are indicated in the X axes._

\newpage

```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of serum levels of serotonin and dopamine sulfate in the INCOV cohort.'}

figs$incov_neuro$plot

```

__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of serum levels of serotonin and dopamine sulfate in the INCOV cohort.__

_Effects of patient's age, timepoint after SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection as compared with healthy controls), serum levels of metabolites related to neurotransmitter biosynthesis and competitor pathways (tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE] and tyrosine [TYR]), and serum concentrations of inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on serum concentrations of serotonin (5-hydroxy tryptophan [5-HT]) and dopamine sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._

_(A) Model fit error (root mean square error [RMSE]) and fraction of explained variances (R-squared) of the robust linear models assessed in the genuine training dataset and 10-fold cross-validation (CV)._

_(B) Estimated of model coefficients with 95% confidence intervals presented in Forest plots. Numbers of complete observations are indicated in the plot captions. Model intercepts (baseline) are indicated in the X axis titles._

\newpage

```{r fig-time-ourse, fig.width = figur::convert(figs$time_course, to = 'in')$w, fig.height = figur::convert(figs$time_course, to = 'in')$h, fig.cap = 'Time course of inflammatory cytokines, serotonin, dopamine sulfate and the neurotransmitter biosynthesis precursors and competitor pathway products during SARS-CoV infection and recovery in the INCOV cohort.'}

figs$time_course$plot

```

__Figure \@ref(fig:fig-time-ourse). Time course of inflammatory cytokines, serotonin, dopamine sulfate and the neurotransmitter biosynthesis precursors and competitor pathway products during SARS-CoV infection and recovery in the INCOV cohort.__

_Time courses of normalized log~2~-transformed serum concentrations of inflammatory cytokines (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF] and interferon gamma [INFG]), metabolites implicated in serotonin synthesis (B, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN] and serotonin/5-hydroxy tryptophan [5-HT]) and metabolites implicated in dopamine turnover (C, phenylalanine [PHE], tyrosine [TYR] and dopamine sulfate [DA sulfate]). Differences between uninfected controls and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after infection, respectively) were modeled by robust linear regression with the uninfected controls as baseline. Mean normalized concentrations with 2 standard errors of the mean (2 $\times$ SEM) are depicted as solid lines with tinted ribbons (plots on the left). Robust linear model coefficient estimates with 95% confidence intervals are visualized in Forest plots (right). Numbers of complete observations per timepoint are indicated in the X axes._

\newpage

```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of serum levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.'}

figs$incov_corr$plot

```

__Figure \@ref(fig:fig-incov-corr). Correlation of serum levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.__

_Serum levels of metabolites implicated in serotonin synthesis (A, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin [5-hydroxy tryptophan, 5-HT]), dopamine turnover (B, phenylalanine [PHE], tyrosine [TYR], dopamine sulfate [DA sulfate]) and inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in uninfected and SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection). Correlation coefficient matrices for at least moderate effects ($\rho$ > 0.2) were visualized as force-directed network plots. Node color codes for the parameter type (gray: inflammatory cytokines, orange: serotonin, green: dopamine turnover), edge width and color codes for the value and sign of the correlation coefficient._

\newpage

# Acknowledgements

The scheme of report findings was created with bioicons by Simon Duerr. 
The SARS-CoV-2 virus image was created by Hanna Vega and licensed under the terms of a Creative Commons CC-BY SA 4.0.

# References

### 2023-05-15 paper/markdown/methods_figures_tables.Rmd
Initial commit
Complete analysis code, figures, supplementary figures and supplementary tables

---
title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
subtitle: "Manuscript parts"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Methods

## Bioinformatic and statistical analysis

R version 4.2.0 was employed for the data analysis.

Normalized, age- and sex-adjusted, log~2~-transformed serum protein and metabolite levels and clinical data for the INCOV cohort were extracted from the report by Su at al. [@Su2022].

Normality and homogeneity of variances was investigated by Shapiro-Wilk and Levene test, respectively. 
Non-parametric tests or normality/homogeneity-improving variable transformations (logarithm and square root) were used in cases of non-normal distribution. 
Comparison of serum metabolite and cytokine levels between uninfected controls and COVID-19 individuals at consecutive timepoints after infection was done by Kruskal-Wallis test with Benjamini-Hochberg-corrected Mann-Whitney post-hoc test [@Benjamini1995]. 
Correlations of metabolite and cytokine serum levels and of mental disorder scoring with metabolite and cytokine levels was investigated by Spearman test. 
Correlation of SIMMUN dataset metabolite levels with age was accomplished by Pearson test. 
Significance of differences in SIMMUN cohort metabolite and inflammatory marker serum concentrations between participants stratified by gender or SARS-CoV-2 was assessed by two-tailed T test.

In multi-parameter linear modeling of aminoacid neurotransmitter precursors and their decay products, neopterin (representative inflammation marker), SARS-CoV-2 infection status, scores of anxiety (HADS), depression (HADS) and stress (PSS-4), age and gender served as candidate explanatory variables. 
Modeling responses and numeric explanatory variables were normalized. 
The models models with the complete explanatory variable set were optimized by Bayesian information criterion (BIC) driven backwards elimination of non-significant terms. 
The normality and homogeneity model residual assumptions were by Shapiro-Wilk and Levene test, respectively, and additionally visually inspected in standard diagnostic plots (residuals versus fitted, quantile-quantile plots). 
Reproducibility of the optimized multi-parameter models was investigated by repeated cross-validation (50 repeats, 10 folds) and by comparison of the RMSE and $R^2$ statistics obtained with the training dataset and in cross-validation.

Details of bioinformatic and statistical analysis are provided in __Supplementary Methods__.

\newpage

# Results

\newpage

# Discussion

\newpage

## Data and code availability

Anonymized patient data will be made available upon request to the corresponding author. 
The INCOV cohort data are publicly available.
The study pipeline code is available at https://github.com/PiotrTymoszuk/stigma_validation.

\newpage

# Tables

```{r tab-stigma-cohort, tab.cap = 'Characteristic of the local SIMMUN cohort stratified by SARS-CoV-2 infection status. Numeric variables are presented as medians with interquartile ranges. Categorical variables are presented as percentages and counts within complete observations.'}

flextable::flextable(tabs$stigma_cohort) %>% 
  width(1, width = 4, unit = 'cm') %>% 
  width(2:3, width = 4, unit = 'cm') %>% 
  width(4:5, width = 2.8, unit = 'cm') %>% 
  footnote(4, 1, 
           value = as_paragraph('normal: body mass index [BMI] < 25 kg/mm\u00B2, overweight: BMI 25 - 30 kg/mm\u00B2, obesity: BMI > 30 kg/mm\u00B2'), 
           part = 'body', 
           ref_symbols = 'a') %>% 
  footnote(7:9, 1, 
           value = as_paragraph('HADS: hospital anxiety and depression scale'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  footnote(10, 1, 
           value = as_paragraph('PSS-4: 4 item perceived stress scale'),
           part = 'body', 
           ref_symbols = 'c') %>%
  footnote(11, 1, 
           value = as_paragraph('immunoglobulin gamma anti receptor binding domain of the SARS-CoV-2 S1 protein, arbitrary units (AU)'), 
           part = 'body', 
           ref_symbols = 'd') %>% 
  footnote(12, 1, 
           value = as_paragraph('mild: ambulatory care, moderate: hospitalized, normal ward, severe: intensive care unit'), 
           part = 'body', 
           ref_symbols = 'e') %>% 
  theme_vanilla

```

\newpage

```{r tab-incov-cohort, tab.cap = 'Characteristic of the external INCOV cohort. Numeric variables are presented as medians with interquartile ranges. Categorical variables are presented as percentages and counts within complete observations.'}

flextable::flextable(tabs$incov_cohort) %>% 
  width(1, width = 4, unit = 'cm') %>% 
  width(2:3, width = 4, unit = 'cm') %>% 
  width(4:5, width = 2.8, unit = 'cm') %>% 
  footnote(3, 1, 
           value = as_paragraph('normal: body mass index [BMI] < 25 kg/mm\u00B2, overweight: BMI 25 - 30 kg/mm\u00B2, obesity: BMI > 30 kg/mm\u00B2'), 
           part = 'body', 
           ref_symbols = 'a') %>% 
  footnote(5, 1, 
           value = as_paragraph('WHO ordinal scale for clinical improvement; mild: 1 - 2, moderate: 3 - 4, severe: 5 - 6, critical: 7'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

# Figures

```{r fig-incov-neuro-tc, fig.width = mm_inch(figs$incov_neuro_tc$w), fig.height = mm_inch(figs$incov_neuro_tc$h), fig.cap = 'Serum levels of neurotransmitter precursor and their decay products in course of COVID-19 and recovery, INCOV cohort.'}

figs$incov_neuro_tc$plot

```

__Figure \@ref(fig:fig-incov-neuro-tc). Serum levels of neurotransmitter precursor and their decay products in course of COVID-19 and recovery, INCOV cohort.__

_Serum levels of tryptophan (TRP), kynurenine (KYN), quinolinic acid (QUIN), phenylalanine (PHE) and tyrosine (TYR) in serum of uninfected controls and COVID-19 individuals during acute, sub-acute and recovery phase of the disease in the INCOV cohort. Statistical significance was determined by Kruskal-Wallis test with Benjamini-Hochberg-corrected Mann-Whitney U test. Normalized serum level concentrations are presented in box plots. Points represent single samples. The Kruskal-Wallis test results are indicated in the plot captions. Results of the post-hoc tests are indicated in the plots. Numbers of complete observations are displayed under the plots._

\newpage

```{r fig-met-infl, fig.width = mm_inch(figs$met_infl$w), fig.height = mm_inch(figs$met_infl$h), fig.cap = 'Correlation of serum levels of aminoacid precursors of neurotransmitters and their decay products with markers of inflammation.'}

figs$met_infl$plot

```

__Figure \@ref(fig:fig-met-infl). Correlation of serum levels of aminoacid precursors of neurotransmitters and their decay products with markers of inflammation.__

_Serum levels of tryptophan (TRP), kynurenine (KYN), quinolinic acid (QUIN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TYR) were correlated with serum levels of inflammatory markers interleukin 6 (IL6), interleukin 10 (IL10), tumor necrosis factor-alpha (TNF), interferon-gamma (INFG), C-reactive protein (CRP), neopterin (NEO) and neutrophil/lymphocyte ratio (NLR) in uninfected and SARS-CoV-2-infected individuals from the SIMMUN (A) and INCOV cohort (B). Statistical significance was assessed by Spearman test. Correlation coefficients are presented in bubble plots. Point sizes correspond to absolute values of correlation coefficient. Point color corresponds to the correlation coefficient value. Correlation coefficients for significant effects are highlighted in bold._

\newpage

```{r fig-cov, fig.width = mm_inch(figs$cov$w), fig.height = mm_inch(figs$cov$h), fig.cap = 'Association of neurotransmitter precursor aminoacids and their decay products with SARS-CoV-2 infection and anxiety, depression and stress scoring in the SIMMUN cohort.'}

figs$cov$plot

```

__Figure \@ref(fig:fig-cov). Association of neurotransmitter precursor aminoacids and their decay products with SARS-CoV-2 infection and anxiety, depression and stress scoring in the SIMMUN cohort.__

_(A) Serum levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TYR) in the SIMMUN study participants stratified by SARS-CoV-2 infections status. Statistical significance was determined by two-tailed T test. Serum levels are shown in box plots. Points represent single observations. Significance p values are displayed in the plot captions, numbers of complete observations are indicated in the plot axes._

_(B) Serum levels of TRP, KYN, KYN/TRP, PHE, TYR and PHE/TYR were correlated with anxiety, depression and stress scores in the SIMMUN study participants (n = `r nrow(stigma$data)`). Statistical significance was assessed by Spearman test. Correlation coefficients are presented in bubble plots. Point sizes correspond to absolute values of correlation coefficient. Point color corresponds to the correlation coefficient value. Correlation coefficients for significant effects are highlighted in bold._

\newpage

```{r fig-multi-model, fig.width = mm_inch(figs$multi_model$w), fig.height = mm_inch(figs$multi_model$h), fig.cap = 'Results of multi-parameter modeling of aminoacid neurotransmitter precursor and their decay products.'}

figs$multi_model$plot

```

__Figure \@ref(fig:fig-multi-model). Results of multi-parameter modeling of aminoacid neurotransmitter precursor and their decay products.__

_Effects of systemic inflammation (neopterin, NEO), SARS-CoV-2 infection status, titre of immunoglobulin gamma against the receptor binding domain of the S1 SARS-CoV-2 protein (anti-RBD IgG), scores of anxiety, depression and stress, age and sex was investigated by multi-parameter linear regression with backward elimination of non-significant terms. Overall model validity was assessed by likelihood-ratio test (LRT). Significant model coefficient estimates with 95% confidence intervals are presented in Forest plots. LRT results (F values, degrees of freedom and p values) and total numbers of complete observations are indicated in the plot captions. Model intercepts (baseline) are indicated in the X axis titles._

\newpage

```{r fig-summary, fig.width = mm_inch(figs$summary$w), fig.height = mm_inch(figs$summary$h), fig.cap = 'Schematic representation of effects of inflammation, SARS-CoV-2 and depression/anxiety symptoms on indolamine and catecholamine neurotransmitter precursor metabolism.'}

figs$summary$plot

```

__Figure \@ref(fig:fig-summary). Schematic representation of effects of inflammation, SARS-CoV-2 and depression/anxiety symptoms on indolamine neurotransmitter precursor metabolism.__

_Effects on metabolism of tryptophan (TRP), the serotonin precursor. SARS-CoV-2-dependent and -independent inflammation reduces systemic TRP levels, stimulates IDO1-mediated conversion of TRP to kynurenin (KYN) and quinolinic acid (QUIN). Independently of inflammation, during recovery from SARS-CoV-2 infection, TRP/KYN/QUIN conversion is stimulated as well. As a result, less substrate for production of the anti-depressive and anxiolytic neurotransmitter serotonin is available, which may predispose to depression or anxiety. Depression and anxiety symptoms on their own are paralleled by decrease in circulating TRP._

\newpage

# Acknowledgements

The scheme of report findings was created with bioicons by Simon Duerr. 
The SARS-CoV-2 virus image was created by Hanna Vega and licensed under the terms of a Creative Commons CC-BY SA 4.0

# References

### 2023-05-15 paper/markdown/supplementary_material.Rmd
Initial commit
Complete analysis code, figures, supplementary figures and supplementary tables

---
title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
subtitle: "Supplementary Material"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Supplementary Methods

## Software

Proteome and metabolome data were analyzed with R version 4.2.0. 
General data transformation tasks were accomplished with the _tidyverse_ package bundle [@Wickham2019], _rlang_ [@Henry2022] and the development package _trafo_ (https://github.com/PiotrTymoszuk/trafo). 
Statistical data testing was done with the packages _rstatix_ [@Kassambara2021], _ggpubr_ [@Kassambara2020] along with the development package _ExDA_ (https://github.com/PiotrTymoszuk/ExDA). 
In linear modeling, base R functions, the development packages _lmqc_ (https://github.com/PiotrTymoszuk/lmqc) and _caretExtra_ (https://github.com/PiotrTymoszuk/caretExtra), and the package _caret_ [@Kuhn2008] were utilized.

Results were visualized with tools provided by the packages _ggplot2_ [@Wickham2016], _cowplot_ [@Wilke2019] and _ExDA_. 
Manuscript and supplementary tables were created with _flextable_ [@Gohel2022]. 
Supplementary Material file was written in the _rmarkdown_ environment [@Allaire2022] and rendered with the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages.

\newpage

# Supplementary Tables

```{r tab-study-vars, tab.cap = 'Study variables in the SIMMUN cohort.'}

flextable::flextable(suppl_tabs$study_vars) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 2.7, unit = 'cm') %>% 
  width(4, width = 2.5, unit = 'cm') %>% 
  width(5, width = 2.5, unit = 'cm') %>% 
  width(6, width = 3.3, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine - tryptophan ratio; PHE/TYR: phenylalanine/tyrosine ratio; HADS: hospital anxiety and depression score; PSS-4: perceived stress scale, 4 item; NEO: neopterin; NLR: neutrophil - lymphocyte ratio; anti-RBD IgG: anti-receptor-binding domain S1/S2 immunoglobulin.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(12, 2, 
           value = as_paragraph('normal: body mass index (BMI) < 25 kg/m\u00B2, overweight: BMI 25 - 30 kg/m\u00B2, obesity: BMI > 30 kg/m\u00B2.'), 
           part = 'body', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r tab-excluded, tab.cap = 'Significant differences between participant of the SIMMUN study included in the analysis and SIMMUN participants excluded due to data missingness. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(suppl_tabs$excluded) %>% 
  width(1, width = 4, unit = 'cm') %>% 
  width(2:3, width = 3.9, unit = 'cm') %>% 
  width(4, width = 2, unit = 'cm') %>% 
  width(5, width = 2.8, unit = 'cm') %>% 
  width(6, width = 2.6, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('HADS: hospital anxiety and depression scale; PSS-4: perceived stress scale, 4 item.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5, 
           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-incov-study-vars, tab.cap = 'Study variables in the INCOV cohort.'}

flextable::flextable(suppl_tabs$incov_study_vars) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 6, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinate; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r tab-samples, tab.cap = 'Number of available samples and sampling timepoints in the INCOV cohort.'}

flextable::flextable(suppl_tabs$samples) %>% 
  width(1:3, width = 3.5, unit = 'cm') %>% 
  theme_vanilla

```

\newpage

```{r tab-modeling-simmun, tab.cap = 'Results of multi-parameter linear modeling of serum concentrations of tryptophan, kynurenine, tyrosine, and kynurenine - tryptophan, and phenylalanine - tyrosine ratios in the SIMMUN cohort.'}

flextable::flextable(suppl_tabs$modeling_simmun) %>% 
  width(1, width = 2, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 2.7, unit = 'cm') %>% 
  width(4, width = 2.7, unit = 'cm') %>% 
  width(5, width = 3.9, unit = 'cm') %>% 
  width(6, width = 2.7, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine - tryptophan ratio; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  footnote(1, 2, 
           value = as_paragraph('NEO: neopterin; HADS: hospital anxiety and depression scale; NLR: neutrophil - lymphocyte ratio; PSS-4: perceived stress scale, 4 item.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-univariate-correlation, tab.cap = "Correlation of age, stress scoring blood neopterin levels and neutrophils - leukocyte ratio with serum metabolite levels in the SIMMUN cohort investigated by Pearson's test."}

flextable::flextable(suppl_tabs$univariate_correlation) %>% 
  width(1, width = 3.7, unit = 'cm') %>% 
  width(2, width = 3.7, unit = 'cm') %>% 
  width(3, width = 2, unit = 'cm') %>% 
  width(4, width = 3.5, unit = 'cm') %>% 
  width(5, width = 2.8, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine - tryptophanm ratio; PHE: phenylalanine; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  footnote(1, 2, 
           value = as_paragraph('PSS-4: perceived stress scale, 4 item; NEO: neopterin; NLR: neutrophil - lymphocyte ratio.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 5, 
           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

```

\newpage

```{r tab-univariate-complarison, tab.cap = "Comparison of serum metabolite levels in SIMMUN study participants stratified by depression signs and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}

flextable::flextable(suppl_tabs$univariate_complarison) %>% 
  width(1, width = 2.6, unit = 'cm') %>% 
  width(2, width = 3.7, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4, width = 2.7, unit = 'cm') %>% 
  width(5, width = 3.7, unit = 'cm') %>% 
  width(6, width = 2.7, unit = 'cm') %>% 
  width(7, width = 2.6, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine - tryptophan ratio; PHE: phenylalanine; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('HADS: hospital anxiety and depression scale.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 3, 
           value = as_paragraph('Median with interquartile range (IQR) and range.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 5, 
           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
  theme_vanilla

```

\newpage

```{r tab-modeling-incov, tab.cap = 'Results of multi-parameter robust linear modeling of serum concentrations of serotonin, and dopamine sulfate in the INCOV cohort.'}

flextable::flextable(suppl_tabs$modeling_incov) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 3.3, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4, width = 2.7, unit = 'cm') %>% 
  width(5, width = 3.7, unit = 'cm') %>% 
  width(6, width = 2.7, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('5-HT: 5-hydroxy tryptophan/serotonine; DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  footnote(1, 2, 
           value = as_paragraph('TRP: tryptophan; KYN: kynuernine; QUIN: quinolinate; IL6: interleukin-6, IL10: interleukin-10; TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; PHE: phenylalanine; TYR: tyrosine.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-time-modeling, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokines, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.'}

flextable::flextable(suppl_tabs$time_modeling) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 3.7, unit = 'cm') %>% 
  width(3, width = 3.7, unit = 'cm') %>% 
  width(4, width = 2.9, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinate; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

# Supplementary Figures

```{r fig-consistency, fig.width = figur::convert(suppl$consistency, to = 'in')$w, fig.height = figur::convert(suppl$consistency, to = 'in')$h, fig.cap = 'Consistency of the PSS4 stress HADS depression and HADS anxiety psychometric tools in the SIMMUN cohort.'}

suppl$consistency$plot

```

__Figure \@ref(fig:fig-consistency). Consistency of the PSS4 stress HADS depression and HADS anxiety psychometric tools in the SIMMUN cohort.__

_(A) Assessment of tau-equivalence of the HADS (hospital anxiety and depression scale) depression and anxiety tool by three-dimensional factor analysis. Loadings for the first two major factors are presented. Each point represents a single HADS item. Percentages of total dataset variance associated with the factors are indicated in the plot axes. Note: tau equivalence implicates equal loadings of each HADS items for all factors. In case of tau equivalence, proximity of the items to the diagonal of the plot (orange dashed line) is expected._

_(B) Consistency of the HADS depression, HADS anxiety and PSS-4 (perceived stress scale, 4 item) psychometric tools determined by McDonald's $\omega$ metric._

\newpage

```{r fig-fit-stats, fig.width = figur::convert(suppl$fit_stats, to = 'in')$w, fig.height = figur::convert(suppl$fit_stats, to = 'in')$h, fig.cap = 'Root mean square error and R-squared statistics for multi-parameter linear models of tryptophan, tyrosine and their metabolites in the SIMMUN cohort.'}

suppl$fit_stats$plot

```

__Figure \@ref(fig:fig-fit-stats). Root mean square error and R-squared statistics for multi-parameter linear models of tryptophan, tyrosine and their metabolites in the SIMMUN cohort.__

_Multi-parameter linear regression models of serum levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by repeated cross-validation (CV, 10 folds). Values of root mean square error (RMSE, A) and $R^2$ (B) in the training data set and CV are plotted._

\newpage

```{r fig-corr-trp, fig.width = figur::convert(suppl$corr_trp, to = 'in')$w, fig.height = figur::convert(suppl$corr_trp, to = 'in')$h, fig.cap = 'Correlation of tryptophan and its metabolites with age, serum levels of neopterin, neutrophil - lymphocyte ratio, and mental stress scoring in the SIMMUN cohort.'}

suppl$corr_trp$plot

```

__Figure \@ref(fig:fig-corr-trp). Correlation of tryptophan and its metabolites with age, serum levels of neopterin, neutrophil - lymphocyte ratio, and mental stress scoring in the SIMMUN cohort.__

_Correlation of age (A), the inflammatory marker neutrophil-leukocyte ratio (B, NLR), and stress scoring (C, perceived stress scale, 4-item [PSS-4]) with serum concentrations of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP) was assessed by Pearson's test corrected for multiple testing with the false discovery rate method (n = `r nrow(stigma$analysis_tbl)`. Single observations are depicted as points, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons. Values of correlation coefficients (r) with 95% confidence intervals and p values are displayed in the plot captions._

\newpage

```{r fig-corr-tyr, fig.width = figur::convert(suppl$corr_tyr, to = 'in')$w, fig.height = figur::convert(suppl$corr_tyr, to = 'in')$h, fig.cap = 'Correlation of phenylalanine and tyrosine with serum levels of neopterin in the SIMMUN cohort.'}

suppl$corr_tyr$plot

```

__Figure \@ref(fig:fig-corr-tyr). Correlation of phenylalanine and tyrosine with serum levels of neopterin in the SIMMUN cohort.__

_Correlation of the inflammatory marker neopterin (NEO) with serum concentrations of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR) was was assessed by Pearson's test corrected for multiple testing with the false discovery rate method (n = `r nrow(stigma$analysis_tbl)`. Single observations are depicted as points, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons. Values of correlation coefficients (r) with 95% confidence intervals and p values are displayed in the plot captions._

\newpage

# References

### 2023-05-15 paper/markdown/supplementary_material_legacy.Rmd
Initial commit
Complete analysis code, figures, supplementary figures and supplementary tables

---
title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
subtitle: "Supplementary Material"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Supplementary Methods

## Software

Proteome and metabolome data were analyzed with R version 4.2.0. 
General data transformation tasks were accomplished with the _tidyverse_ package bundle [@Wickham2019], _rlang_ [@Henry2022] and the development package _trafo_ (https://github.com/PiotrTymoszuk/trafo). 
Statistical data testing was done with the packages _rstatix_ [@Kassambara2021], _ggpubr_ [@Kassambara2020] along with the development package _ExDA_ (https://github.com/PiotrTymoszuk/ExDA). 
In linear modeling, base R functions, the development packages _lmqc_ (https://github.com/PiotrTymoszuk/lmqc) and _caretExtra_ (https://github.com/PiotrTymoszuk/caretExtra), and the package _caret_ [@Kuhn2008] were utilized.

Results were visualized with tools provided by the packages _ggplot2_ [@Wickham2016], _cowplot_ [@Wilke2019] and _ExDA_. 
Manuscript and supplementary tables were created with _flextable_ [@Gohel2022]. 
Supplementary Material file was written in the _rmarkdown_ environment [@Allaire2022] and rendered with the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages.

## INCOV cohort data import and transformation

Proteome and metabolome data in form of normalized, age- and sex-adjusted, log~2~-transformed serum levels as well as clinical information (sex, SARS-CoV-2 infection status, COVID-19 severity, timepoint, post-COVID-19 syndrome status and particular symptoms) for the INCOV cohort were extracted from supplementary tables accompanying the report by Su and colleagues [@Su2022].

Characteristic of the INCOV cohorts is presented in __Table 2__. 
Numbers of available INCOV cohort samples and the sampling timepoints are shown in __Supplementary Table S\@ref(tab:tab-samples)__.

## Variable distribution and transformation

Distribution normality and variance homogeneity of normalized cytokine and metabolite serum levels was assessed by Shapiro-Wilk and Levene test, respectively. 
The distribution testing revealed substantial deviations from normality for multiple study parameters. 
For this reason statistical hypothesis testing in the INCOV dataset was done with non-parametric tests. 
For SIMMUN cohort metabolite and inflammation marker levels (KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio, NEO, CRP, IL6, anti-RBD IgG) used in analysis of correlation with age and mental disorder scoring, comparison between participants stratified by SARS-CoV-2 infection status or gender as well as for linear modeling, logarithm and square root transformations was used, which improved both normality and variance homogeneity.

## Statistical hypothesis testing

Comparison of normalized serum cytokine and metabolite levels between uninfected controls and COVID-19 individuals from the INCOV cohort at consecutive timepoints after symptom onset was done with Kruskal-Wallis test. 
Differences between SARS-CoV-2-negative controls and consecutive timepoints of COVID-19 were investigated with Mann-Whitney post-hoc U test corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995].

Correlation of serum cytokine and metabolite levels in uninfected controls and COVID-19 individuals at consecutive timepoints in the INCOV and SIMMUN datasets was analyzed by Spearman test. 
Correlation of serum cytokine and metabolite levels with scores of anxiety, depression and stress was assessed by Spearman test. 
Correlation of SIMMUN dataset metabolite levels with age was assessed by Pearson test.

Comparison of SIMMUN dataset metabolite and inflammatory marker levels between participants stratified by SARS-CoV-2 infection status or gender was done with two-tailed T test.

## Multi-parameter linear regression with backwards elimination

Effects of the representative inflammation marker (NEO), SARS-CoV-2 infection status, titre of immunoglobulin gamma against the receptor binding domain of the S1 SARS-CoV-2 protein (anti-RBD IgG), anxiety (HADS), depression (HADS) and stress scoring (PSS-4), age and gender on systemic levels of aminoacid neurotransmitter precursor and their decay products (TRP, KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio) was assessed by multi-parameter linear regression with backwards elimination. 
Modeling responses and numeric explanatory variables were transformed with log or squared root functions to improve normality and normalized.

Full models including the complete set of explanatory variables listed above were constructed (function `make_lm()`, package _lmqc_) and optimized by Bayesian information criterion (AIC) driven backwards elimination of non-significant terms (method `step()`, package _lmqc_). 
Normality and homogeneity of distribution of the model residuals was checked by Shapiro-Wilk and Levene test, respectively (method `summary(type = 'assumptions')`, package _lmqc_) and visually inspected in standard diagnostic plots of model residuals (residuals vs fitted, quantile-quantile plots, method `plot()` called for the model objects). 
Fit stats ($R^2$ and root mean squared error [RMSE]) were retrieved from the model objects by calling `summary(type = 'fit')` (package _lmqc_). 
Validity of the optimized models was determined by likelihood-ratio test (LRT) versus the respective null models (method `anova()`). 
Reproducibility and proper parameterization of the optimized multi-parameter models was investigated by repeated cross-validation (50 repeats, 10 folds, function `train(method = 'lm')`, package _caret_) and by comparison of the RMSE and $R^2$ statistics obtained with in the training dataset and in cross-validation (method `summary()`, package _caretExtra_). 
As presented in __Supplementary Figure S\@ref(fig:fig-fit-stats)A__, similar values of error fit in the training and cross-validation data suggest good reproducibility of the optimized models and lack of over-parameterization.

## Data and code availability

Anonymized local cohort data will be made available upon request to the corresponding author. 
The INCOV cohort data are publicly available.
The study pipeline code is available at https://github.com/PiotrTymoszuk/SIMMUN_validation.

\newpage

# Supplementary Tables

```{r tab-samples, tab.cap = 'Number of available samples and sampling timepoints in the INCOV cohort.'}

flextable::flextable(suppl_tabs$samples) %>% 
  width(1, width = 4, unit = 'cm') %>% 
  width(2, width = 6, unit = 'cm') %>% 
  width(3, width = 4, unit = 'cm') %>% 
  footnote(1, 2, value = as_paragraph('Median with interquartile range, days after symptom onset.'), ref_symbols = 'a', part = 'header') %>% 
  theme_vanilla

```

\newpage

# Supplementary Figures

```{r fig-incov-neuro-tc, fig.width = mm_inch(suppl$incov_infl_tc$w), fig.height = mm_inch(suppl$incov_infl_tc$h), fig.cap = 'Serum levels of inflammatory cytokines in course of COVID-19 and recovery, INCOV cohort.'}

suppl$incov_infl_tc$plot

```

__Supplementary Figure S\@ref(fig:fig-incov-neuro-tc). Serum levels of inflammatory cytokines in course of COVID-19 and recovery, INCOV cohort.__

_Serum levels of interleukin 6 (IL6), interleukin 10 (IL10), tumor necrosis factor alpha (TNF) and interferon-gamma (IFN) in serum of uninfected controls and COVID-19 individuals during acute, sub-acute and recovery phase of the disease in the INCOV cohort. Statistical significance was determined by Kruskal-Wallis test with Benjamini-Hochberg-corrected Mann-Whitney U test. Normalized serum level concentrations are presented in box plots. Points represent single samples. The Kruskal-Wallis test results are indicated in the plot captions. Results of the post-hoc tests are indicated in the plots. Numbers of complete observations are displayed under the plots. _

\newpage

```{r fig-hads-cov, fig.width = mm_inch(suppl$hads_cov$w), fig.height = mm_inch(suppl$hads_cov$h), fig.cap = 'Association of inflammatory markers with SARS-CoV-2 infection convalescence in the SIMMUN cohort. Association of neurotransmitter precursor aminoacids and their decay products with anti-SARS-CoV-2 antibody response in the SIMMUN cohort.'}

suppl$hads_cov$plot

```

__Supplementary Figure S\@ref(fig:fig-hads-cov). Association of inflammatory markers with SARS-CoV-2 infection convalescence in the SIMMUN cohort. Association of neurotransmitter precursor aminoacids and their decay products with anti-SARS-CoV-2 antibody response in the SIMMUN cohort.__

_(A) Serum levels of interleukin 6 (IL6), C-reactive protein (CRP), neopterin (NEO), and neutrophil-leukocyte ratio (NLR) in the SIMMUN study participants stratified by SARS-CoV-2 infections status. Statistical significance was determined by two-tailed T test. Serum levels are shown in box plots. Points represent single observations. Significance p values are displayed in the plot captions, numbers of complete observations are indicated in the plot X axes._

_(B) Correlation of TRP, KYN, KYN/TRP, PHE, TYR and PHE/TRP with the titre of immunoglobulin gamma to receptor binding domain of the SARS-CoV-2 S1 protein (anti-RBD IgG, arbitrary units [AU]) in the SIMMUN cohort (n = `r nrow(stigma$data)`) was investigated by Spearman test. Points represent single observations, blue lines with gray ribbons depict fitted generalized additive model (GAM) trends with 95% confidence intervals. Values of correlation coefficients with 95% confidence intervals and significance are indicated in the plot captions._

\newpage

```{r fig-age, fig.width = mm_inch(suppl$age$w), fig.height = mm_inch(suppl$age$h), fig.cap = 'Correlation of aminoacid neurotransmitter precursors and their decay products with age.'}

suppl$age$plot

```

__Supplementary Figure S\@ref(fig:fig-age). Correlation of aminoacid neurotransmitter precursors and their decay products with age.__

_Correlation of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) with age in the SIMMUN cohort was investigated by Pearson test. Points represent single observations, blue lines with gray ribbons depict fitted linear trends with 95% confidence intervals. Values of correlation coefficients with 95% confidence intervals, significance and numbers of complete observations are indicated in the plot captions._

\newpage

```{r fig-gender, fig.width = mm_inch(suppl$gender$w), fig.height = mm_inch(suppl$gender$h), fig.cap = 'Levels of aminoacid neurotransmitter precursors and their decay products in females and males.'}

suppl$gender$plot

```

__Supplementary Figure S\@ref(fig:fig-gender). Levels of aminoacid neurotransmitter precursors and their decay products in females and males.__

_Serum concentrations of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in female and male participants of the SIMMUN study. Statistical significance was assessed by two-tailed T test. Serum levels are shown in box plots. Points represent single observations. Significance p values are displayed in the plot captions. Numbers of complete observation are indicated in the plot X axes._

\newpage

```{r fig-fit-stats, fig.width = mm_inch(suppl$fit_stats$w), fig.height = mm_inch(suppl$fit_stats$h), fig.cap = 'Root mean square error and R-squared statistics for multi-parameter linear models of aminoacid neurotransmitter precursors and their decay products in the SIMMUN cohort'}

suppl$fit_stats$plot

```

__Supplementary Figure S\@ref(fig:fig-fit-stats). Root mean square error and R-squared statistics for multi-parameter linear models of aminoacid neurotransmitter precursors and their decay products in the SIMMUN cohort.__

_Multi-parameter linear regression models of serum levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by repeated cross-validation (CV, 50 repeats, 10 folds). Values of root mean square error (RMSE, A) and $R^2$ (B) in the training data set and CV are plotted._

\newpage

# References

### 2023-05-15 exec.R

@@ -1,0 +1,12 @@
+# Launches the entire pipeline
+
+  library(soucer)
+
+  print(source_all(c('import.R',
+                     'exploration.R',
+                     'modeling.R',
+                     'testing.R',
+                     'incov_analyses.R',
+                     'paper.R'),
+                   message = TRUE,
+                   crash = TRUE))

### 2023-05-15 exploration scripts/cohort.R

@@ -1,0 +1,210 @@
+# Characteristic of the local STIGMA and INCOV study cohorts
+
+  insert_head()
+
+# container ------
+
+  cohort <- list()
+
+# Globals -------
+
+  insert_msg('Globals')
+
+  ## variable lexicon
+
+  cohort$var_lexicon <-
+    c(sex = 'Sex',
+      age = 'Age, years',
+      bmi = 'BMI, kg/m\u00B2',
+      bmi_class = 'Body mass class',
+      ethnics = 'Ethnics',
+      somatic_comorb = 'Somatic comorbidity',
+      psych_comorb = 'Psychiatric comorbidity',
+      hads_anx_score = 'HADS anxiety score',
+      hads_dpr_score = 'HADS depression score',
+      hads_signs = 'Depression or anxiety signs, HADS \u2265 8',
+      pss_stress_score = 'PSS-4 stress score',
+      cov = 'Infection',
+      anti_rbd = 'anti-RBD SARS-CoV-2, IgG, AU',
+      severity = 'COVID-19 severity') %>%
+    compress(names_to = 'variable',
+             values_to = 'label')
+
+  ## analysis tables: STIGMA
+
+  cohort$analysis_tbl$stigma <- stigma$data %>%
+    select(cov,
+           patient_id,
+           any_of(cohort$var_lexicon$variable)) %>%
+    filter(patient_id %in% stigma$complete_ids)
+
+  ## analysis tables: INCOV
+
+  cohort$analysis_tbl$incov <- incov$clinic %>%
+    filter(patient_id %in% incov$complete_ids) %>%
+    filter(!duplicated(patient_id)) %>%
+    mutate(long_cov = ifelse(long_cov == 'healthy',
+                             'no', as.character(long_cov)),
+           long_cov = factor(long_cov),
+           psy_long_cov = ifelse(psy_long_cov %in% c('healthy',
+                                                     'recovered'),
+                                 'no', as.character(psy_long_cov)),
+           psy_long_cov = factor(psy_long_cov),
+           neuro_long_cov = ifelse(neuro_long_cov %in% c('healthy',
+                                                         'recovered'),
+                                   'no', as.character(neuro_long_cov)),
+           neuro_long_cov = factor(neuro_long_cov),
+           gastro_long_cov = ifelse(gastro_long_cov %in% c('healthy',
+                                                           'recovered'),
+                                    'no', as.character(gastro_long_cov)),
+           gastro_long_cov = factor(gastro_long_cov),
[…]

### 2023-05-15 exploration scripts/consistency.R

@@ -1,0 +1,102 @@
+# Checking the consistency of the psychometric tools
+
+  insert_head()
+
+# container ----
+
+  cons <- list()
+
+# analysis globals -----
+
+  insert_msg('Analysis globals')
+
+  cons$analysis_tbl <-
+    stigma[c("pss4", "hads_depression", "hads_anxiety")] %>%
+    map(filter, patient_id  %in% stigma$complete_ids) %>%
+    map(column_to_rownames, 'patient_id') %>%
+    map(select, -ends_with('results')) %>%
+    map(~filter(.x, complete.cases(.x)))
+
+# Factor analysis -----
+
+  insert_msg('Factor analysis')
+
+  ## to check the tau equivalence
+
+  cons$fa_objects <-
+    list(data = cons$analysis_tbl,
+         kdim = c(1, 3, 2)) %>%
+    pmap(reduce_data,
+         red_fun = 'fa')
+
+  ## anxiety nearly tau equivalent! (all items in the vicinity of diagonal)
+
+  cons$fa_loading_plots <-
+    list(x = cons$fa_objects[c("hads_depression", "hads_anxiety")]) %>%
+    pmap(plot,
+         type = 'loadings',
+         cust_theme = globals$common_theme) %>%
+    map2(.,
+         c('Depression, HADS',
+           'Anxiety, HADS'),
+         ~.x +
+           labs(title = .y,
+                subtitle = 'Factor analysis loadings') +
+           geom_vline(xintercept = 0,
+                      linetype = 'dashed') +
+           geom_hline(yintercept = 0,
+                      linetype = 'dashed') +
+           geom_abline(slope = 1,
+                       intercept = 0,
+                       color = 'coral3',
+                       linetype = 'dashed'))
+
+# McDonald's omega ------
+
+  insert_msg('McDonald omega')
+
+  set.seed(1234)
+
+  cons$omega_obj <-
+    list(m = cons$analysis_tbl,
+         nfactors = c(1, 3, 2)) %>%
+    pmap(omega,
+         fm = 'ml')
+
+  cons$omega_stats <- cons$omega_obj %>%
+    map(~.x$omega.group) %>%
+    map(as.data.frame) %>%
+    map(rownames_to_column, 'factor') %>%
+    map(~.x[c('factor', 'total')]) %>%
+    compress(names_to = 'tool') %>%
+    as_tibble
+
+# Plotting total omegas -----
+
+  insert_msg('Plotting total omegas')
+
+  cons$omega_plot <- cons$omega_stats %>%
[…]

### 2023-05-15 exploration scripts/correlation.R

@@ -1,0 +1,95 @@
+# Inter-correlations of each of the outcomes, inflammatory markers
+# and psychometric scores
+
+  insert_head()
+
+# container ------
+
+  corr <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  corr$variables <-
+    list(outcomes = c('trp', 'log_kyn', 'log_kyn_trp',
+                      'log_phe', 'log_tyr', 'sqrt_phe_tyr'),
+         inflammation = c('log_neo', 'log_nlr'),
+         psychometry = c('pss_stress_score',
+                         'hads_anx_score',
+                         'hads_dpr_score'))
+
+  corr$pairs <- corr$variables %>%
+    map(~combn(.x, m = 2, simplify = FALSE))
+
+# Correlation -----
+
+  insert_msg('Correlation')
+
+  corr$test <-
+    list(x = corr$pairs,
+         y = c('pearson', 'pearson', 'spearman')) %>%
+    pmap(function(x, y) x %>%
+           map_dfr(~correlate_variables(stigma$analysis_tbl,
+                                        variables = .x,
+                                        what = 'correlation',
+                                        type = y,
+                                        pub_styled = FALSE))) %>%
+    compress(names_to = 'task') %>%
+    mutate(task = factor(task, names(corr$variables))) %>%
+    re_adjust %>%
+    mutate(significant = ifelse(p_adjusted < 0.05,
+                                'yes', 'no'),
+           fontface = ifelse(significant == 'yes',
+                             'bold', 'plain')) %>%
+    blast(task)
+
+# Correlograms ------
+
+  insert_msg('Correlograms')
+
+  corr$plots <-
+    list(x = corr$test,
+         y = c('Metabolites',
+               'Inflammatory markers',
+               'Psychometry'),
+         z = corr$variables,
+         v = c('Pearson correlation',
+               'Pearson correlation',
+               'Spearman correlation')) %>%
+    pmap(function(x, y, z, v) x %>%
+           ggplot(aes(x = variable1,
+                      y = variable2,
+                      fill = estimate,
+                      size = abs(estimate)))  +
+           geom_point(shape = 21) +
+           geom_text(aes(label = signif(estimate, 2),
+                         fontface = fontface,
+                         alpha = significant),
+                     size = 2.75,
+                     vjust = -1.6,
+                     hjust = 0.5) +
+           scale_alpha_manual(values = c(yes = 1,
[…]

### 2023-05-15 exploration scripts/distribution.R

@@ -1,0 +1,102 @@
+# Normality of distribution and equality of variances
+#
+# Split by the the study group (SARS-CoV2 status/HADS, STIGMA)
+# and timepoint (INCOV)
+
+  insert_head()
+
+# container -----
+
+  distr <- list()
+
+# globals: analysis tables -------
+
+  insert_msg('Analysis tables')
+
+  ## variables, in the SIMMUN/STIGMA cohort, CRP and IL6
+  ## are excluded from testing, they are handled as factors in modeling
+
+  distr$variables <- map(globals[c('stigma_lexicon',
+                                   'incov_lexicon')],
+                         ~.x$variable) %>%
+    set_names(c('stigma', 'incov'))
+
+  distr$variables$stigma <-
+    distr$variables$stigma[!distr$variables$stigma %in% c('crp', 'il6', 'infection')]
+
+  distr$variables$stigma <-
+    distr$variables$stigma[map_lgl(stigma$data[distr$variables$stigma],
+                                   is.numeric)]
+
+  ## analysis tables for both cohorts
+
+  distr$analysis_tbl$stigma <- stigma$data
+
+  distr$analysis_tbl$incov <- incov[c('proteome',
+                                      'metabolome')] %>%
+    map(select,
+        sample_id,
+        patient_id,
+        timepoint,
+        any_of(distr$variables$incov))
+
+  ## reducing the INCOV dataset to a single data frame
+
+  distr$analysis_tbl$incov <- distr$analysis_tbl$incov %>%
+    reduce(left_join,
+           by = c('sample_id', 'patient_id', 'timepoint'))
+
+# Normality: Shapiro-Wilk tests ------
+
+  insert_msg('Normality: Shapiro-Wilk test')
+
+  distr$shapiro_test <-
+    list(data = distr$analysis_tbl,
+         variables = distr$variables,
+         split_factor = c('study_group', 'timepoint')) %>%
+    pmap(explore,
+         what = 'normality',
+         pub_styled = FALSE)
+
+  ## indicating the best transformation of the STIGMA variables
+  ## in terms of normality
+
+  distr$stigma_normality_trans <-
+    distr$shapiro_test$stigma %>%
+    map(mutate,
+        variable = factor(variable, distr$variables$stigma),
+        base_variable = stri_replace(variable,
+                                     regex = '(log_)|(sqrt_)',
+                                     replacement = ''),
+        transformation = ifelse(stri_detect(variable, fixed = 'log'),
[…]

### 2023-05-15 exploration scripts/excluded.R

@@ -1,0 +1,102 @@
+# Comparison of the included and excluded SIMMUN/STIGMA participants
+
+  insert_head()
+
+# container ----
+
+  excl <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variable lexicon
+
+  excl$var_lexicon <-
+    c(sex = 'Sex',
+      age = 'Age, years',
+      bmi_class = 'Body mass class',
+      somatic_comorb = 'Somatic comorbidity',
+      psych_comorb = 'Psychiatric comorbidity',
+      hads_anx_score = 'HADS anxiety score',
+      hads_dpr_score = 'HADS depression score',
+      hads_signs = 'Depression or anxiety signs, HADS \u2265 8',
+      pss_stress_score = 'PSS-4 stress score',
+      cov = 'Infection',
+      anti_rbd = 'anti-RBD SARS-CoV-2, IgG, AU',
+      severity = 'COVID-19 severity',
+      trp = 'TRP, µmol/L',
+      kyn = 'KYN, µmol/L',
+      kyn_trp = 'KYN/TRP',
+      tyr = 'TYR, µmol/L',
+      phe = 'PHE, µmol/L',
+      phe_tyr = 'PHE/TYR',
+      neo = 'NEO, nmol/L',
+      nlr = 'NLR') %>%
+    compress(names_to = 'variable',
+             values_to = 'label')
+
+  excl$var_lexicon$test_type <- stigma$data[excl$var_lexicon$variable] %>%
+    map_lgl(is.numeric) %>%
+    ifelse('wilcoxon_r', 'cramer_v')
+
+  ## analysis table
+
+  excl$analysis_tbl <- stigma$data %>%
+    mutate(analysis_status = ifelse(patient_id %in% stigma$complete_ids,
+                                    'analyzed', 'excluded'),
+           analysis_status = factor(analysis_status,
+                                    c('analyzed', 'excluded'))) %>%
+    select(patient_id,
+           analysis_status,
+           all_of(excl$var_lexicon$variable))
+
+# Descriptive stats -----
+
+  insert_msg('Descriptive stats')
+
+  excl$stats <- excl$analysis_tbl %>%
+    explore(split_factor = 'analysis_status',
+            variables = excl$var_lexicon$variable,
+            what = 'table',
+            pub_styled = TRUE) %>%
+    reduce(left_join, by = 'variable') %>%
[…]

### 2023-05-15 exploration.R

@@ -1,0 +1,41 @@
+# Explorative data analysis
+#
+# characteristic of the cohorts
+#
+# Testing for the normality and EOV for the study variables (STIGMA and INCOV)
+#
+# Comparing included and excluded participants of the SIMMUN study
+
+# tools --------
+
+  library(plyr)
+  library(tidyverse)
+  library(trafo)
+  library(stringi)
+  library(exda)
+  library(soucer)
+  library(psych)
+  library(clustTools)
+
+  select <- dplyr::select
+  reduce <- purrr::reduce
+
+  insert_head()
+
+  source_all('./tools/tools.R',
+             message = TRUE, crash = TRUE)
+
+# analysis scripts -------
+
+  insert_msg('Analysis scripts')
+
+  c('./exploration scripts/cohort.R',
+    './exploration scripts/distribution.R',
+    './exploration scripts/excluded.R',
+    './exploration scripts/consistency.R',
+    './exploration scripts/correlation.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# END -----
+
+  insert_tail()

### 2023-05-15 import scripts/incov.R

@@ -1,0 +1,296 @@
+# Import of the INCOV cohort proteomic and metabolomic data
+
+  insert_head()
+
+# container ----
+
+  incov <- list()
+
+# importing the metabolomics and proteomics data ------
+
+  insert_msg('Metabolomics and proteomics data')
+
+  incov[c('proteome',
+          'metabolome')] <- c('S2.1 Proteomics',
+                              'S2.2 Metabolomics') %>%
+    map(~read_excel(path = './data/INCOV/Table S2.xlsx',
+                    sheet = .x))
+
+  incov[c('proteome',
+          'metabolome')] <- incov[c('proteome',
+                                    'metabolome')] %>%
+    map(mutate,
+        patient_id = `Patient Subject ID`,
+        sample_id = ifelse(`Healthy or INCOV` == 'Healthy',
+                           patient_id,
+                           `Blood Draw`),
+        .before = `Patient Subject ID`) %>%
+    map(select,
+        - `Patient Subject ID`,
+        - `Blood Draw`,
+        - `Healthy or INCOV`,
+        - age,
+        - sex,
+        - BMI)
+
+# Annotation tables -----
+
+  insert_msg('Annotation tables')
+
+  incov$annotation_proteome <-
+    tibble(variable = names(incov$proteome)[!names(incov$proteome) %in% c('patient_id', 'sample_id')]) %>%
+    mutate(label = stri_replace(variable, regex = '_.*$', replacement = ''),
+           unit = 'Z-score of log\u2082 concentration',
+           axis_lab = paste(label, unit, sep = ', '))
+
+  incov$annotation_metabolome <-
+    tibble(variable = names(incov$metabolome)[!names(incov$metabolome) %in% c('patient_id', 'sample_id')]) %>%
+    mutate(label = variable,
+           label = car::recode(label,
+                               "'tryptophan' = 'TRP';
+                               'kynurenine' = 'KYN';
+                               'phenylalanine' = 'PHE';
+                               'tyrosine' = 'TYR';
+                               'dopamine 3-O-sulfate' = 'DA sulfate';
+                               'serotonin' = '5-HT';
+                               'quinolinate' = 'QUIN'"),
+           unit = 'Z-score of log\u2082 concentration',
+           axis_lab = paste(label, unit, sep = ', '))
+
+# Reading the clinical data for CoV patients -------
+
+  insert_msg('Reading the clinical data for CoV patients')
+
[…]

### 2023-05-15 import scripts/local.R

@@ -1,0 +1,196 @@
+
+
+  insert_head()
+
+# container -----
+
+  stigma <- list()
+
+# Reading the dataset from SPSS -----
+
+  insert_msg('Reading the raw SPSS dataset')
+
+  ## raw data for the selected variables
+
+  stigma$raw <- read.spss('./data/local/01.06.2022.sav',
+                          to.data.frame = TRUE)
+
+  stigma$stress <- read.spss('./data/local/27.10.2021.sav',
+                             to.data.frame = TRUE)
+
+# Wrangling, NA handling, calculation of the ratios ------
+
+  insert_msg('Wrangling')
+
+  ## subscores for the stress and HADS depression/anxiety
+  ## used later for testing of the psychometric tool consistency
+
+  stigma$pss4 <- stigma$stress %>%
+    mutate(patient_id = stri_replace(studien_id,
+                                     regex = '\\s+$',
+                                     replacement = '')) %>%
+    select(patient_id,
+           all_of(c('pss_4_1', 'pss_4_2', 'pss_4_3', 'pss_4_4')))
+
+  stigma$pss4[c('pss_4_2', 'pss_4_3')] <-
+    stigma$pss4[c('pss_4_2', 'pss_4_3')] %>%
+    map_dfc(function(x) 4 - x)
+
+  stigma[c('hads_depression',
+           'hads_anxiety')] <-
+    list(c('hads_depression_results',
+           'hads_02', 'hads_04', 'hads_06', 'hads_08',
+           'hads_10', 'hads_12', 'hads_14'),
+         c('hads_anxiety_results',
+           'hads_01', 'hads_03', 'hads_05', 'hads_07',
+           'hads_09', 'hads_11', 'hads_13')) %>%
+    map(~select(stigma$raw,
+                studien_id,
+                all_of(.x))) %>%
+    map(mutate,
+        patient_id = stri_replace(studien_id,
+                                  regex = '\\s+$',
+                                  replacement = '')) %>%
+    map(select, -studien_id)
+
+  ## stress, depression, final scores
+
+  stigma$stress <- stigma$stress %>%
+    as_tibble %>%
+    transmute(patient_id = stri_replace(studien_id,
+                                        regex = '\\s+$',
+                                        replacement = ''),
+              phq_dpr_score = phq_results,
+              log_phq_dpr_score = log(phq_dpr_score + 1),
+              sqrt_phq_dpr_score = sqrt(phq_dpr_score),
[…]

### 2023-05-15 import.R

@@ -1,0 +1,139 @@
+# Import and wrangling of the manuscript data
+#
+# The local STIGMA cohort dataset encompasses readouts of inflammation
+# and metabolic markers of neurotransmitter precursors along with the
+# explanatory variables (anxiety/depression signs and persistent symptoms)
+# and confounders (age and sex)
+#
+# The INCOV validation dataset contains selected metabolites of the TRP decay
+# pathway (so called TRYCATS pathway) and key inflammatory cytokines measured
+# in blood of healthy individuals and COVID-19 participants at consecutive time
+# points (acute, sub-acute and recovery) after clnical onset.
+
+
+# tools --------
+
+  library(plyr)
+  library(tidyverse)
+  library(soucer)
+  library(trafo)
+  library(stringi)
+  library(readxl)
+  library(foreign)
+
+  insert_head()
+
+  c('./tools/globals.R',
+    './tools/tools.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# Sourcing the import scripts for single studies ------
+
+  insert_msg('Sourcing the import scripts')
+
+  source_all(c('./import scripts/local.R',
+               './import scripts/incov.R'),
+             message = TRUE, crash = TRUE)
+
+# Updating the project globals ------
+
+  insert_msg('Project globals')
+
+  globals$incov_lexicon <-
+    rbind(incov$annotation_proteome %>%
+            filter(variable %in% globals$incov_proteins),
+          incov$annotation_metabolome %>%
+            filter(variable %in% globals$incov_metabolites))
+
+# SIMMUN, modeling and analysis dataset -----
+
+  insert_msg('Modeling SIMUN dataset')
+
+  ## modeling responses, the optimal transformations defined
+  ## during explorative analysis
+
+  stigma$responses <-
+    c('trp' = 'trp',
+      'kyn' = 'log_kyn',
+      'kyn_trp' = 'log_kyn_trp',
+      'phe' = 'log_phe',
+      'tyr' = 'log_tyr',
+      'phe_tyr' = 'sqrt_phe_tyr')
+
+  ## explanatory variables
+
+  stigma$expl_lexicon <-
+    c(## infection-related features
+      ## there are to few non-ambulatory COV cases
+      ## for modeling
+      'infection' = 'SARS-CoV-2',
+      'anti_rbd_class' = 'anti-RBD IgG',
+      #'severity' = 'COVID-19 severity',
[…]

### 2023-05-15 incov scripts/correlation.R

@@ -1,0 +1,171 @@
+# Correlation analysis results to be presented in the paper
+# Levels of 5-HT and DA sulfate are correlated ith normalized serum
+# concentrations of the relevant metabolites and inflammatory cytokines
+
+  insert_head()
+
+# container -----
+
+  corr <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variables
+
+  corr$responses <- c(serotonin = 'serotonin',
+                      dopamine = 'dopamine 3-O-sulfate')
+
+  corr$variables <-
+    list(serotonin = c('tryptophan',
+                       'kynurenine',
+                       'quinolinate',
+                       'IL6_INF',
+                       'IL10_INF',
+                       'TNF_INF',
+                       'IFNG_INF'),
+         dopamine = c('phenylalanine',
+                      'tyrosine',
+                      'IL6_INF',
+                      'IL10_INF',
+                      'TNF_INF',
+                      'IFNG_INF'))
+
+  ## analysis tables
+
+  corr$analysis_tbl <- incov$analysis_tbl %>%
+    select(timepoint,
+           all_of(unname(corr$responses)),
+           all_of(reduce(corr$variables, union))) %>%
+    set_names(make.names(names(.))) %>%
+    mutate(timepoint = car::recode(timepoint,
+                                   "'healthy' = 'uninfected'"),
+           timepoint = factor(timepoint,
+                              c('uninfected', 'acute',
+                                'sub-acute', 'recovery'))) %>%
+    blast(timepoint)
+
+  ## lexicon
+
+  corr$lexicon <- globals$incov_lexicon %>%
+    mutate(variable = make.names(variable),
+           class = ifelse(stri_detect(variable, regex = '_INF$'),
+                          'inflammation',
+                          'metabolites'))
+
+  corr$responses <- make.names(corr$responses) %>%
+    set_names(names(corr$responses))
+
+  corr$variables <- corr$variables %>%
+    map(make.names)
+
+  ## variable pairs
+
+  corr$pairs <-
+    map2(corr$responses,
+         corr$variables,
+         function(x, y) map(y, ~c(x, .x))) %>%
+    unlist(recursive = FALSE)
+
+  corr$pairs <- corr$pairs %>%
+    set_names(map(corr$pairs, paste, collapse = '|'))
+
+  ## n numbers per timepoint
+
[…]

### 2023-05-15 incov scripts/cytokine_correlation.R

@@ -1,0 +1,120 @@
+# cyt_correlation of inflammatory and metabolic INCOV variables (Spearman)
+
+  insert_head()
+
+# container -----
+
+  cyt_corr <- list()
+
+# globals ------
+
+  insert_msg('Globals')
+
+  ## analysis table
+
+  cyt_corr$analysis_tbl <- incov$analysis_tbl %>%
+    select( patient_id,
+            timepoint,
+            any_of(globals$incov_lexicon$variable)) %>%
+    blast(timepoint)
+
+  ## variable pairs
+
+  cyt_corr$pairs <- globals$incov_proteins %>%
+    map(function(prot) globals$incov_metabolites[!globals$incov_metabolites %in% c('serotonin', 'dopamine 3-O-sulfate')] %>%
+          map(~c(prot, .x))) %>%
+    unlist(recursive = FALSE)
+
+# cyt_correlation analysis ------
+
+  insert_msg('correlation')
+
+  cyt_corr$test <- cyt_corr$analysis_tbl %>%
+    map(function(dat) cyt_corr$pairs %>%
+          map_dfr(~correlate_variables(dat,
+                                       variables = .x,
+                                       what = 'correlation',
+                                       type = 'spearman',
+                                       ci = TRUE)))
+  ## FDR cyt_correction, plot elements
+
+  cyt_corr$test <- cyt_corr$test %>%
+    compress(names_to = 'timepoint') %>%
+    re_adjust %>%
+    mutate(timepoint = factor(timepoint, names(cyt_corr$test)),
+           plot_lab = paste0(signif(estimate, 2),
+                             ' [', signif(lower_ci, 2),
+                             ' - ', signif(upper_ci, 2), ']'),
+           plot_cap = paste0(plot_lab, ', ', significance, ', n = ', n),
+           x_lab = exchange(variable1,
+                            dict = globals$incov_lexicon),
+           x_lab = paste0(x_lab, ', rank'),
+           y_lab = exchange(variable2,
+                            dict = globals$incov_lexicon),
+           y_lab = paste0(y_lab, ', rank')) %>%
+    blast(timepoint)
+
+# Bubble plots -------
+
+  insert_msg('Bubble plots')
+
+  cyt_corr$bubbles <-
+    list(data = cyt_corr$test,
+         plot_title =  c('uninfected, INCOV',
+                         'acute CoV, INCOV',
[…]

### 2023-05-15 incov scripts/mds.R

@@ -1,0 +1,166 @@
+# Visualizing cytokine and metabolite levels via multi-dimensional scaling
+# Separate analyses for each timepoint and metabolite system
+#
+# In addition, correlations are visualized as a network plot
+
+  insert_head()
+
+# container ------
+
+  incov_mds <- list()
+
+# analysis tables ------
+
+  insert_msg('Analysis tables')
+
+  ## variables
+
+  incov_mds$variables <-
+    list(serotonin = c('serotonin',
+                       'tryptophan',
+                       'kynurenine',
+                       'quinolinate',
+                       'IL6_INF',
+                       'IL10_INF',
+                       'TNF_INF',
+                       'IFNG_INF'),
+         dopamine = c('dopamine.3.O.sulfate',
+                      'phenylalanine',
+                      'tyrosine',
+                      'IL6_INF',
+                      'IL10_INF',
+                      'TNF_INF',
+                      'IFNG_INF'))
+
+  ## analysis tables with ranks
+
+  incov_mds$analysis_tbl <- incov$analysis_tbl %>%
+    select(patient_id,
+           timepoint,
+           all_of(globals$incov_lexicon$variable)) %>%
+    set_names(make.names(names(.)))
+
+  incov_mds$analysis_tbl <- incov_mds$variables %>%
+    map(~select(incov_mds$analysis_tbl,
+                patient_id,
+                timepoint,
+                all_of(.x))) %>%
+    map(blast, timepoint) %>%
+    unlist(recursive = FALSE) %>%
+    map(select, -timepoint) %>%
+    map(~map_dfc(.x, function(x) if(is.numeric(x)) rank(x) else x)) %>%
+    map(column_to_rownames, 'patient_id') %>%
+    map(t) %>%
+    map(as.data.frame)
+
+  ## variable lexicon
+
+  incov_mds$lexicon <- globals$incov_lexicon %>%
+    mutate(variable = make.names(variable))
+
+# Calculation of distance - Spearman's correlation -----
+
+  insert_msg('Scaled Spearman distance')
+
+  ## is nothing else as a squared Euclidean distance between ranks
+  ## scaling the results (min/max)
+
+  incov_mds$dists <- incov_mds$analysis_tbl %>%
+    map(calculate_dist, method = 'sumofsquares')
+
+# MDS objects ------
+
+  insert_msg('MDS objects')
+
[…]

### 2023-05-15 incov scripts/metabolite_correlation.R

@@ -1,0 +1,125 @@
+# Inter-correlation of neurotransmitter-related metabolites
+# In the INCOV cohort
+
+  insert_head()
+
+# container ------
+
+  met_corr <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variables
+
+  met_corr$variables <-
+    c('tryptophan',
+      'kynurenine',
+      'quinolinate',
+      'serotonin',
+      'phenylalanine',
+      'tyrosine',
+      'dopamine 3-O-sulfate')
+
+  met_corr$lexicon <- globals$incov_lexicon %>%
+    filter(variable %in% met_corr$variables) %>%
+    mutate(variable = make.names(variable))
+
+  ## analysis tables
+
+  met_corr$analysis_tbl <- incov$analysis_tbl %>%
+    select(timepoint,
+           all_of(met_corr$variables)) %>%
+    set_names(make.names(names(.))) %>%
+    blast(timepoint) %>%
+    map(select, -timepoint)
+
+  met_corr$variables <- make.names(met_corr$variables)
+
+  ## variable pairs
+
+  met_corr$pairs <- met_corr$variables %>%
+    combn(m = 2, simplify = FALSE)
+
+# Correlation analysis: Spearman's rank test -----
+
+  insert_msg('Correlation')
+
+  met_corr$test <- met_corr$analysis_tbl %>%
+    map(function(data) met_corr$pairs %>%
+          map_dfr(~correlate_variables(data,
+                                       variables = .x,
+                                       what = 'correlation',
+                                       type = 'spearman',
+                                       ci = TRUE,
+                                       pub_styled = FALSE))) %>%
+    compress(names_to = 'timepoint') %>%
+    re_adjust %>%
+    mutate(timepoint = factor(timepoint, names(met_corr$analysis_tbl)),
+           est_lab = paste0(signif(estimate, 2), ' [',
+                            signif(lower_ci, 2), ' - ',
+                            signif(upper_ci, 2), ']'),
+           plot_cap = paste(est_lab, significance, sep = ', '),
+           x_lab = exchange(variable1,
+                            dict = globals$incov_lexicon),
+           x_lab = paste0(x_lab, ', rank'),
+           y_lab = exchange(variable2,
+                            dict = globals$incov_lexicon),
+           y_lab = paste0(y_lab, ', rank')) %>%
+    blast(timepoint)
+
+# Correlograms ------
[…]

### 2023-05-15 incov scripts/modeling.R

@@ -1,0 +1,185 @@
+# Robust linear modeling of metabolites as a function of infection status
+# and levels of inflammatory cytokines
+
+  insert_head()
+
+# container ------
+
+  incov_mod <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## responses and explanatory variables
+  ## the explanatory model selection is based on
+  ## the results of SIMMUN modeling by backward elimination LM
+  ## and include age, infection status and inflammatory parameters
+
+  incov_mod$response_lexicon <-
+    c('tryptophan',
+      'kynurenine',
+      'quinolinate',
+      'serotonin',
+      'phenylalanine',
+      'tyrosine',
+      'dopamine 3-O-sulfate') %>%
+    exchange(dict = globals$incov_lexicon) %>%
+    compress(names_to = 'variable',
+             values_to = 'label')
+
+  incov_mod$variable_lexicon <-
+    c('timepoint' = 'SARS-CoV-2',
+      'age' = 'Age',
+      'IL6_INF' = 'IL6',
+      'IL10_INF' = 'IL10',
+      'TNF_INF' = 'TNF',
+      'IFNG_INF' = 'IFNG') %>%
+    compress(names_to = 'variable',
+             values_to = 'label')
+
+  ## analysis table
+  ## normalizing numeric variables
+
+  incov_mod$analysis_tbl <- incov$analysis_tbl %>%
+    select(all_of(incov_mod$response_lexicon$variable),
+           all_of(incov_mod$variable_lexicon$variable)) %>%
+    map_dfc(function(x) if(is.numeric(x)) scale(x)[, 1] else x) %>%
+    set_names(make.names(names(.)))
+
+  incov_mod[c("response_lexicon", "variable_lexicon")] <-
+    incov_mod[c("response_lexicon", "variable_lexicon")] %>%
+    map(mutate,
+        variable = make.names(variable))
+
+  ## model formulas
+
+  incov_mod$formulas <- incov_mod$response_lexicon$variable %>%
+    map(~paste(.x,
+               paste(incov_mod$variable_lexicon$variable, collapse = ' + '),
+               sep = ' ~ ')) %>%
+    map(as.formula) %>%
+    set_names(incov_mod$response_lexicon$variable)
+
+# Construction of models -----
+
+  insert_msg('Construction of models')
+
+  incov_mod$models <- incov_mod$formulas %>%
+    map(~make_lm(data = incov_mod$analysis_tbl,
[…]

### 2023-05-15 incov scripts/neuro_modeling.R

@@ -1,0 +1,188 @@
+# Modeling of neurotransmitter levels as a function of treir precursors,
+# competitive decay products, infection status and inflammatory cytokines
+
+  insert_head()
+
+# container ------
+
+  incov_neuro <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## responses and explanatory variables
+  ## the explanatory model selection is based on
+  ## the results of SIMMUN modeling by backward elimination LM
+  ## and include age, infection status and inflammatory parameters
+
+  incov_neuro$response_lexicon <-
+    c('serotonin',
+      'dopamine 3-O-sulfate') %>%
+    exchange(dict = globals$incov_lexicon) %>%
+    compress(names_to = 'variable',
+             values_to = 'label')
+
+  incov_neuro$variable_lexicons <-
+    list(serotonin = c('tryptophan' = 'TRP',
+                       'kynurenine' = 'KYN',
+                       'quinolinate' = 'QUIN'),
+         dopamine = c('phenylalanine' = 'PHE',
+                      'tyrosine' = 'TYR')) %>%
+    map(~c(.x,
+           c('timepoint' = 'SARS-CoV-2',
+             'age' = 'Age',
+             'IL6_INF' = 'IL6',
+             'IL10_INF' = 'IL10',
+             'TNF_INF' = 'TNF',
+             'IFNG_INF' = 'IFNG'))) %>%
+    map(compress,
+        names_to = 'variable',
+        values_to = 'label')
+
+  ## analysis table, expressing age in decades
+  ## normalizing numeric variables
+
+  incov_neuro$analysis_tbl <- incov$analysis_tbl %>%
+    select(all_of(incov_neuro$response_lexicon$variable),
+           all_of(union(incov_neuro$variable_lexicons$serotonin$variable,
+                        incov_neuro$variable_lexicons$dopamine$variable))) %>%
+    mutate(age = age/10) %>%
+    map_dfc(function(x) if(is.numeric(x)) scale(x)[, 1] else x) %>%
+    set_names(make.names(names(.)))
+
+  incov_neuro$response_lexicon <- incov_neuro$response_lexicon %>%
+    mutate(variable = make.names(variable))
+
+  ## model formulas
+
+  incov_neuro$formulas <-
+    map2(incov_neuro$response_lexicon$variable,
+         incov_neuro$variable_lexicons,
+         ~paste(.x,
[…]

### 2023-05-15 incov scripts/time_course.R

@@ -1,0 +1,114 @@
+# Comparison of time_course and the timepoints
+# Comparison via Kruskal-Wallis test, since
+# multiple variables are non-normally distributed
+# and violations of EOV were detected by the Levene test
+
+  insert_head()
+
+# container -----
+
+  time_course <- list()
+
+# analysis table and variables -----
+
+  insert_msg('Globals')
+
+  time_course$analysis_tbl <- incov$analysis_tbl %>%
+    mutate(timepoint = car::recode(timepoint,
+                                   "'healthy' = 'uninfected'"),
+           timepoint = factor(timepoint,
+                              c('uninfected', 'acute', 'sub-acute', 'recovery')))
+
+# Descriptive statistics --------
+
+  insert_msg('Descriptive stats')
+
+  time_course$stats <- time_course$analysis_tbl %>%
+    explore(variables = globals$incov_lexicon$variable,
+            split_factor = 'timepoint',
+            what = 'table',
+            pub_styled = TRUE)
+
+# Post-hoc Mann-Whitney tests ------
+
+  insert_msg('Post-hoc Mann-Whitney tests')
+
+  ## the variable name 'dopamine 3-O-sulfate' is not compatible with the
+  ## compare_means() function and need to be substituted by 'DA'
+
+  time_course$post_hoc_formulas <-
+    c(globals$incov_lexicon$variable[globals$incov_lexicon$variable != 'dopamine 3-O-sulfate'],
+      'DA') %>%
+    paste('~ timepoint') %>%
+    map(as.formula) %>%
+    set_names(globals$incov_lexicon$variable)
+
+  time_course$post_hoc_test <- time_course$post_hoc_formulas %>%
+    map(compare_means,
+        data = time_course$analysis_tbl %>%
+          mutate(DA = `dopamine 3-O-sulfate`),
+        method = 'wilcox.test',
+        #p.adjust.method = 'BH',
+        ref.group = 'uninfected')
+
+  ## global FDR adjustment
+
+  time_course$post_hoc_test <- time_course$post_hoc_test %>%
+    compress(name = 'metabolite') %>%
+    mutate(p.adj = p.adjust(p, 'BH'),
+           metabolite = factor(metabolite,
+                               names(time_course$post_hoc_test))) %>%
+    blast(metabolite)
+
+# Testing by Kruskal-Wallis test --------
+
+  insert_msg('Testing for differences between the timepoints')
+
+  time_course$test <- time_course$analysis_tbl %>%
[…]

### 2023-05-15 incov scripts/time_modeling.R

@@ -1,0 +1,202 @@
+# Additional analysis with robust linear modeling as requested by a reviewer
+
+  insert_head()
+
+# container ----
+
+  time_rlm <- list()
+
+# analysis globals ------
+
+  insert_msg('Globals')
+
+  ## variable lexicon
+
+  time_rlm$lexicon <- globals$incov_lexicon %>%
+    mutate(label_long = variable,
+           variable = make.names(variable))
+
+  ## analysis tables
+
+  time_rlm$analysis_tbl <- incov$analysis_tbl %>%
+    select(patient_id,
+           timepoint,
+           any_of(set_names(time_rlm$lexicon$label_long,
+                            time_rlm$lexicon$variable))) %>%
+    mutate(timepoint = car::recode(timepoint,
+                                   "'healthy' = 'uninfected'"))
+
+  ## setting the baselines
+
+  time_rlm$analysis_tbl <-
+    list(uninfected = c('uninfected', 'acute', 'sub-acute', 'recovery'),
+         acute = c('acute', 'uninfected', 'sub-acute', 'recovery')) %>%
+    map(~mutate(time_rlm$analysis_tbl,
+                timepoint = factor(timepoint, levels = .x)))
+
+  ## model formulas and null model formulas
+
+  time_rlm$formulas <- time_rlm$lexicon$variable %>%
+    map(~paste0('`', .x, '` ~ timepoint')) %>%
+    map(as.formula) %>%
+    set_names(time_rlm$lexicon$variable)
+
+  time_rlm$null_formulas <- time_rlm$lexicon$variable %>%
+    map(~paste0('`', .x, '` ~ 1')) %>%
+    map(as.formula) %>%
+    set_names(time_rlm$lexicon$variable)
+
+  ## colors for plots
+
+  time_rlm$plot_colors <-
+    list(inflammatory = c('IL6_INF' = 'orangered2',
+                          'IL10_INF' = 'orangered4',
+                          'TNF_INF' = 'gray60',
+                          'IFNG_INF' = 'black'),
+         trp = c('tryptophan' = 'coral',
+                 'kynurenine' = 'indianred2',
+                 'quinolinate' = 'indianred4',
+                 'serotonin' = 'steelblue'),
+         tyr = c('phenylalanine' = 'darkolivegreen3',
+                 'tyrosine' = 'darkolivegreen4',
+                 'dopamine.3.O.sulfate' = 'steelblue'))
+
[…]

### 2023-05-15 incov_analyses.R

@@ -1,0 +1,83 @@
+# analyses in the INCOV cohort
+
+# tools ------
+
+  library(plyr)
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+
+  library(soucer)
+  library(exda)
+  library(ggpubr)
+
+  library(MASS)
+  library(lmqc)
+  library(caret)
+  library(clustTools)
+
+  library(doParallel)
+  library(furrr)
+
+  library(igraph)
+  library(ggnetwork)
+
+  library(ggrepel)
+
+  library(soucer)
+
+  select <- dplyr::select
+  reduce <- purrr::reduce
+  explore <- exda::explore
+
+  insert_head()
+
+  c('./tools/tools.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# analysis scripts ------
+
+  insert_msg('Analysis scripts')
+
+  ## modeling of serum levels of dopamine sulfate and serotonin
+
+  if(file.exists('./cache/incov_neuro.RData')) {
+
+    insert_msg('Loading cached INCOV modeling results for serotonin and dopamine')
+
+    load('./cache/incov_neuro.RData')
+
+  } else {
+
+    source_all('./incov scripts/neuro_modeling.R')
+
+  }
+
+  ## modeling for particular metabolites
+
+  if(file.exists('./cache/incov_mod.RData')) {
+
+    insert_msg('Loading cached INCOV modeling results')
+
+    load('./cache/incov_mod.RData')
+
+  } else {
+
+    source_all('./incov scripts/modeling.R',
+               message = TRUE, crash = TRUE)
+
+  }
+
+  ## univariable analyses
+
+  c('./incov scripts/metabolite_correlation.R',
+    './incov scripts/cytokine_correlation.R',
+    './incov scripts/correlation.R',
+    './incov scripts/mds.R',
+    './incov scripts/time_course.R',
+    './incov scripts/time_modeling.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# END -----
+
+  insert_tail()

### 2023-05-15 modeling scripts/cforest.R

@@ -1,0 +1,172 @@
+# Modeling metabolite responses with ranger's Random forest
+
+  insert_head()
+
+# container ----
+
+  mod_crf <- list()
+
+# analysis globals -----
+
+  insert_msg('Analysis globals')
+
+  ## since there's no normality assumption, variables
+  ## are fed into the model without normality-stabilizing transformations
+
+  mod_crf$lexicon <- stigma$expl_lexicon %>%
+    mutate(variable = stri_replace(variable,
+                                   regex = '^(log|sqrt)_',
+                                   replacement = ''))
+
+  mod_crf$responses <- stigma$responses %>%
+    stri_replace(regex = '^(log|sqrt)_',
+                 replacement = '') %>%
+    set_names(names(stigma$responses))
+
+  mod_crf$analysis_tbl <- stigma$data %>%
+    mutate(infection = car::recode(cov,
+                                   "'healthy' = 'uninfected';
+                             'SARS-CoV-2' = 'recovery'"),
+           infection = factor(infection,
+                              c('uninfected', 'recovery')),
+           age = age/10) %>%
+    select(patient_id,
+           all_of(mod_crf$responses),
+           all_of(mod_crf$lexicon$variable)) %>%
+    filter(complete.cases(.))
+
+  ## normalizing the explanatory variables
+
+  mod_crf$analysis_tbl[mod_crf$lexicon$variable] <-
+    mod_crf$analysis_tbl[mod_crf$lexicon$variable] %>%
+    map_dfc(function(x) if(is.numeric(x)) scale(x)[, 1] else x)
+
+  mod_crf$analysis_tbl <- mod_crf$analysis_tbl %>%
+    column_to_rownames('patient_id')
+
+  ## model formulas
+
+  mod_crf$formulas <- mod_crf$responses %>%
+    map(~paste(.x,
+               paste(mod_crf$lexicon$variable, collapse = ' + '),
+               sep = ' ~ ')) %>%
+    map(as.formula)
+
+
+  ## tuning grids
+
+  mod_crf$tune_grid <-
+    expand.grid(mtry = c(2, 4, 6, 12))
+
+# Caret models -----
+
+  insert_msg('Caret models')
+
+  set.seed(123456)
+
+  registerDoParallel(cores = 7)
+
+  mod_crf$caret_models <- mod_crf$formulas %>%
+    map(~train(form = .x,
+               data = mod_crf$analysis_tbl,
+               method = 'cforest',
+               metric = 'RMSE',
[…]

### 2023-05-15 modeling scripts/elimination.R

@@ -1,0 +1,223 @@
+# Modeling of the metabolite responses by backward elimination
+
+  insert_head()
+
+# container ----
+
+  mod_eli <- list()
+
+# Construction of the null and full models -------
+
+  insert_msg('construction of the null and full models')
+
+  mod_eli$null_models <- list(formula = mod_globals$null_formulas) %>%
+    pmap(make_lm,
+         data = mod_globals$analysis_tbl,
+         mod_fun = lm,
+         family = NULL)
+
+  mod_eli$full_models <-
+    list(formula = mod_globals$full_formulas) %>%
+    pmap(make_lm,
+         data = mod_globals$analysis_tbl,
+         mod_fun = lm,
+         family = NULL)
+
+# Backward elimination via step() -----
+
+  insert_msg('Backward elimination')
+
+  mod_eli$step_models <- mod_eli$full_models %>%
+    map(step,
+        step_fun = MASS::stepAIC,
+        direction = 'backward',
+        k = floor(log(nrow(stigma$analysis_tbl))))
+
+# Checking assumptions of the optimized models -----
+
+  insert_msg('Assumptions of the optimized models')
+
+  ## normality and EOV of the residuals
+
+  mod_eli$assumptions <- mod_eli$step_models %>%
+    map(summary,
+        type = 'assumptions',
+        type.predict = 'response')
+
+  ## diagnostic plots of the residuals
+
+  mod_eli$resid_plots <- mod_eli$step_models %>%
+    map(plot,
+        type = 'residuals',
+        cust_theme = globals$common_theme,
+        type.predict = 'response')
+
+  ## response - explanatory variable relationship
+
+  mod_eli$relation_plots <- mod_eli$step_models %>%
+    map(plot,
+        type = 'relationship',
+        cust_theme = globals$common_theme,
+        type.predict = 'response')
+
+# LRT versus the null model ------
+
+  insert_msg('LRT versus the null model')
+
+  mod_eli$lrt <- map2(mod_eli$step_models,
+                        mod_eli$null_models,
+                        ~anova(.x$model, .y$model))
+
+  ## LRT summaries
+
+  mod_eli$lrt_summary <- mod_eli$lrt %>%
+    map(as.data.frame) %>%
+    map(~.x[2, ]) %>%
+    compress(names_to = 'response') %>%
+    mutate(significance = ifelse(`Pr(>F)` < 0.05,
[…]

### 2023-05-15 modeling scripts/glmnet.R

@@ -1,0 +1,280 @@
+# Modeling of the metabolite responses with Elastic Net regression
+
+  insert_head()
+
+# container ------
+
+  mod_net <- list()
+
+# analysis globals -----
+
+  insert_msg('Analysis globals')
+
+  ## model matrices
+
+  mod_net$y <- mod_globals$analysis_tbl[stigma$responses] %>%
+    as.list %>%
+    set_names(names(stigma$responses))
+
+  mod_net$x <-
+    mod_globals$analysis_tbl[c('patient_id', stigma$expl_lexicon$variable)] %>%
+    column_to_rownames('patient_id') %>%
+    model.matrix(~., data = .)
+
+  ## model constants
+
+  mod_net$alpha <- 0.5
+
+  mod_net$repetitions <- 1:100 %>%
+    set_names(paste0('rep_', 1:100))
+
+  ## CV folds
+
+  set.seed(1234)
+
+  mod_net$cv_folds <- mod_net$repetitions %>%
+    map(function(x) createFolds(y = stigma$analysis_tbl$patient_id,
+                                k = 100,
+                                list = FALSE,
+                                returnTrain = TRUE))
+
+# Tuning of the glmnet model -----
+
+  insert_msg('Finding the optimal lambda')
+
+  ## models
+
+  plan('multisession')
+
+  mod_net$tuning$models <- mod_net$y %>%
+    map(function(resp) mod_net$cv_folds %>%
+          future_map(~cv.glmnet(x = mod_net$x,
+                                y = resp,
+                                alpha = mod_net$alpha,
+                                type.measure = 'deviance',
+                                foldid = .x,
+                                family = 'gaussian'),
+                     .options = furrr_options(seed = TRUE)))
+
+  plan('sequential')
+
+  ## lambdas
+
+  mod_net$tuning$stats <- mod_net$tuning$models %>%
+    map(~map(.x,
+             ~filter(as_tibble(.x[c('lambda', 'nzero', 'cvm', 'cvlo', 'cvup')]),
+                     lambda == .x[['lambda.min']]))) %>%
+    map(compress,
+        names_to = 'repetition') %>%
+    map(arrange, cvm) %>%
+    map(mutate,
+        optimum = ifelse(cvm == min(cvm), 'yes', 'no'),
+        plot_lab = ifelse(optimum == 'yes',
+                          signif(lambda, 3),
+                          NA))
+
[…]

### 2023-05-15 modeling scripts/ranger.R

@@ -1,0 +1,165 @@
+# Modeling metabolite responses with ranger's Random forest
+
+  insert_head()
+
+# container ----
+
+  mod_ranger <- list()
+
+# analysis globals -----
+
+  insert_msg('Analysis globals')
+
+  ## since there's no normality assumption, variables
+  ## are fed into the model without normality-stabilizing transformations
+
+  mod_ranger$lexicon <- stigma$expl_lexicon %>%
+    mutate(variable = stri_replace(variable,
+                                   regex = '^(log|sqrt)_',
+                                   replacement = ''))
+
+  mod_ranger$responses <- stigma$responses %>%
+    stri_replace(regex = '^(log|sqrt)_',
+                 replacement = '') %>%
+    set_names(names(stigma$responses))
+
+  mod_ranger$analysis_tbl <- stigma$data %>%
+    mutate(infection = car::recode(cov,
+                                   "'healthy' = 'uninfected';
+                             'SARS-CoV-2' = 'recovery'"),
+           infection = factor(infection,
+                              c('uninfected', 'recovery')),
+           age = age/10) %>%
+    select(patient_id,
+           all_of(mod_ranger$responses),
+           all_of(mod_ranger$lexicon$variable)) %>%
+    filter(complete.cases(.))
+
+  ## normalizing the explanatory variables
+
+  mod_ranger$analysis_tbl[mod_ranger$lexicon$variable] <-
+    mod_ranger$analysis_tbl[mod_ranger$lexicon$variable] %>%
+    map_dfc(function(x) if(is.numeric(x)) scale(x)[, 1] else x)
+
+  mod_ranger$analysis_tbl <- mod_ranger$analysis_tbl %>%
+    column_to_rownames('patient_id')
+
+  ## model formulas
+
+  mod_ranger$formulas <- mod_ranger$responses %>%
+    map(~paste(.x,
+               paste(mod_ranger$lexicon$variable, collapse = ' + '),
+               sep = ' ~ ')) %>%
+    map(as.formula)
+
+
+  ## tuning grids
+
+  mod_ranger$tune_grid <-
+    expand.grid(mtry = c(2, 4, 6, 12),
+                splitrule = c('variance', 'extratrees', 'maxstat'),
+                min.node.size = c(3, 5))
+
+# Caret models -----
+
+  insert_msg('Caret models')
+
+  set.seed(123456)
+
+  registerDoParallel(cores = 7)
+
[…]

### 2023-05-15 modeling.R

@@ -1,0 +1,74 @@
+# Modeling of the TRP, KYN, KYN/TRP, PHE, TYR and PHE/TYR as a function
+# of inflammatory, CoV-related and psychometric explanatory variables in
+# the SIMMUN/STIGMA cohort
+#
+# Ramerkable over-fitting in LASSO!
+
+# tools -------
+
+  library(plyr)
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+
+  library(lmqc)
+  library(MASS)
+  library(glmnet)
+  library(ranger)
+  library(party)
+
+  library(caret)
+  library(caretExtra)
+
+  library(exda)
+  library(ggrepel)
+  library(ggtext)
+
+  library(doParallel)
+  library(furrr)
+
+  library(soucer)
+
+  train <- caret::train
+  explore <- exda::explore
+  reduce <- purrr::reduce
+
+  c('./tools/tools.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# Analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  mod_globals <- list()
+
+  ## analysis table with normalized numeric variables
+
+  mod_globals$analysis_tbl <- stigma$analysis_tbl %>%
+    map_dfc(function(x) if(is.numeric(x)) scale(x)[, 1] else x)
+
+  ## full and null model formulas
+
+  mod_globals$full_formulas <- stigma$responses %>%
+    map(~paste(.x,
+               paste(stigma$expl_lexicon$variable, collapse = ' + '),
+               sep = ' ~ ')) %>%
+    map(as.formula)
+
+  mod_globals$null_formulas <- stigma$responses %>%
+    map(~paste(.x, '~ 1')) %>%
+    map(as.formula)
+
+# Analysis scripts ------
+
+  insert_msg('Analysis scripts')
+
+  c('./modeling scripts/elimination.R',
+    './modeling scripts/glmnet.R',
+    './modeling scripts/ranger.R',
+    './modeling scripts/cforest.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# END ------
+
+  insert_tail()

### 2023-05-15 paper scripts/figures.R

@@ -1,0 +1,339 @@
+# Main figures
+
+  insert_head()
+
+# container ------
+
+  figs <- list()
+
+# Figure 1: CONSORT -----
+
+  insert_msg('Figure 1: study inclusion diagram')
+
+  figs$inclusion <-
+    plot_grid(ggdraw() +
+                draw_image('./schemes/inclusion.png')) %>%
+    as_figure(label = 'figure_1_study_inclusion_scheme',
+              ref_name = 'inclusion',
+              caption = paste('Inclusion scheme for the SIMMUN',
+                              'and INCOV cohorts, and analysis strategy.'),
+              w = 180,
+              h = 180 * 2832/4044)
+
+# Figure 2: linear modeling -----
+
+  insert_msg('Figure 2: linear modeling')
+
+  figs$modeling <- mod_eli$forest_plots %>%
+    map(~.x + theme(legend.position = 'none'))
+
+  figs$modeling$kyn <- figs$modeling$kyn +
+    expand_limits(x = -0.15)
+
+  figs$modeling$tyr <- figs$modeling$tyr +
+    expand_limits(x = -0.13)
+
+  figs$modeling <-
+    c(figs$modeling["trp"],
+      list(get_legend(figs$modeling$kyn_trp +
+                        theme(legend.title = element_blank()))),
+      figs$modeling[c("kyn", "kyn_trp", "tyr", "phe_tyr")]) %>%
+    plot_grid(plotlist = .,
+              ncol = 2,
+              align = 'hv',
+              axis = 'tblr',
+              rel_heights = c(1, 1.35, 1),
+              labels = c('A', '',
+                         '', '',
+                         'B', ''),
+              label_size = 10) %>%
+    as_figure(label = 'figure_2_linear_modeling',
+              ref_name = 'modeling',
+              caption = paste('Results of multi-parameter linear modeling',
+                              'of tryptophan,',
+                              'kynurenine, kynurenine/tryptophan ratio,',
+                              'tyrosine and phenylalanine/tyrosine ratio',
+                              'in the SIMMUN cohort.'),
+              w = 180,
+              h = 180)
+
+# Figure 3: univariate analysis for TRP/KYN ------
+
+  insert_msg('Figure 3: Univariate analysis for TRP/KYN')
+
+  ## upper panel: correlogram
+
+  figs$uni_trp$upper <-
+    plot_grid(tst_trp$correlation$correlogram +
+                theme(plot.subtitle = element_text(size = 7.4),
+                      legend.position = 'none') +
[…]

### 2023-05-15 paper scripts/render.R

@@ -1,0 +1,31 @@
+# Renders the manuscript and the supplementary material
+
+  insert_head()
+
+# bibliography ------
+
+  insert_msg('Bibliography')
+
+  cov_biblio <- read_bib('./paper/markdown/infl_biblio.bib')
+
+# supplementary material ------
+
+  insert_msg('Rendering the supplements')
+
+  render('./paper/markdown/supplementary_material.Rmd',
+         output_format = word_document2(number_sections = FALSE,
+                                        reference_docx = 'ms_template.docx'),
+         output_dir = './paper')
+
+# paper -----
+
+  insert_msg('Rendering the paper')
+
+  render('./paper/markdown/main_manuscript.Rmd',
+         output_format = word_document2(number_sections = FALSE,
+                                        reference_docx = 'ms_template.docx'),
+         output_dir = './paper')
+
+# END -----
+
+insert_tail()

### 2023-05-15 paper scripts/supplement.R

@@ -1,0 +1,122 @@
+# Supplementary Figures
+
+  insert_head()
+
+# container ------
+
+  suppl <- list()
+
+# Figure S1: consistency of psychometric tools ------
+
+  insert_msg('Figure S1: consistency of psychometric tools')
+
+  suppl$consistency <- cons$fa_loading_plots %>%
+    map(~.x + theme(plot.tag = element_blank())) %>%
+    plot_grid(plotlist = .,
+              ncol = 2,
+              align = 'hv',
+              axis = 'tblr') %>%
+    plot_grid(plot_grid(cons$omega_plot,
+                        ncol = 2),
+              nrow = 2,
+              rel_heights = c(1, 0.7),
+              labels = LETTERS,
+              label_size = 10) %>%
+    as_figure(label = 'figure_s1_psycho_consistency',
+              ref_name = 'consistency',
+              caption = paste('Consistency of the PSS4 stress',
+                              'HADS depression and HADS anxiety psychometric',
+                              'tools in the SIMMUN cohort.'),
+              w = 180,
+              h = 140)
+
+# Figure S2: error and R-squared, multiple linear regression ------
+
+  insert_msg('Figure S2: Fits stats, multiple regression')
+
+  suppl$fit_stats <- mod_eli$fit_stat_plots %>%
+    map2(c('pseudo_log', 'identity'),
+         ~.x +
+           scale_x_continuous(trans = .y) +
+           theme(legend.position = 'none')) %>%
+    plot_grid(plotlist = .,
+              ncol = 2,
+              align = 'hv',
+              labels = LETTERS,
+              label_size = 10) %>%
+    plot_grid(get_legend(mod_eli$fit_stat_plots[[1]] +
+                           theme(legend.position = 'bottom')),
+              nrow = 2,
+              rel_heights = c(0.9, 0.1)) %>%
+    as_figure(label = 'figure_s2_fit_stats',
+              ref_name = 'fit_stats',
+              caption = paste('Root mean square error and R-squared',
+                              'statistics for multi-parameter linear models',
+                              'of trayptophan, tyrosine and their metabolites',
+                              'in the SIMMUN cohort.'),
+              w = 180,
+              h = 110)
+
+# Figure S3: scatter plots for TRP, KYN and KYN/TRP -----
+
+  insert_msg('Figure S3: scatter plots for TRP, KYN and KYN/TRP')
+
+  suppl$corr_trp <-
+    tst_trp$correlation$plots[c("age|trp", "age|log_kyn", "age|log_kyn_trp",
[…]

### 2023-05-15 paper scripts/tables.R

@@ -1,0 +1,233 @@
+# manuscript tables
+
+  insert_head()
+
+# container -----
+
+  tabs <- list()
+  suppl_tabs <- list()
+
+# Tables 1 and 2: characteristic of the STIGMA and INCOV collectives -------
+
+  insert_msg('Table 1 and 2: characteristic of the STIGMA and INCOV collective')
+
+  tabs[c('stigma_cohort',
+         'incov_cohort')] <-
+    cohort$feat_tables %>%
+    map(filter,!stri_detect(Variable, regex = '(post-COVID)|(BMI)')) %>%
+    map(~map_dfc(.x,
+                 stri_replace,
+                 regex = '^HADS.*\\nHADS\\+:\\s{1}',
+                 replacement = '')) %>%
+    map(set_names,
+        c('Variable', 'Uninfected',
+          'SARS-CoV-2', 'Test', 'Significance', 'Effect size'))
+
+  ## skipping the somatic symptoms
+
+  tabs$stigma_cohort <- tabs$stigma_cohort %>%
+    filter(Variable %in% c('Participants, n',
+                           'Sex',
+                           'Age, years',
+                           'Body mass class',
+                           'Somatic comorbidity',
+                           'Psychiatric comorbidity',
+                           'HADS anxiety score',
+                           'HADS depression score',
+                           'Depression or anxiety signs, HADS ≥ 8',
+                           'PSS-4 stress score',
+                           'anti-RBD SARS-CoV-2, IgG, AU',
+                           'COVID-19 severity'))
+
+  ## the ready to use tables
+
+  tabs[c('stigma_cohort',
+         'incov_cohort')] <- tabs[c('stigma_cohort',
+                                    'incov_cohort')] %>%
+    list(x = .,
+         label = c('table_1_stigma_cohort',
+                   'table_2_incov_cohort'),
+         ref_name = c('stigma_cohort',
+                      'incov_cohort'),
+         caption = c(paste('Characteristic of the local SIMMUN cohort.',
+                           'Numeric variables are presented as medians with',
+                           'interquartile ranges. Categorical variables are',
+                           'presented as percentages and counts within',
+                           'complete observation sets.'),
+                     paste('Characteristic of the external INCOV cohort.',
+                           'Numeric variables are presented as medians with',
+                           'interquartile ranges. Categorical variables are',
+                           'presented as percentages and counts within',
+                           'complete observation sets.'))) %>%
+    pmap(mdtable)
+
[…]

### 2023-05-15 testing scripts/result_tables.R

@@ -1,0 +1,62 @@
+# Tables with results of the univariate analysis
+
+  insert_head()
+
+# container ------
+
+  tst_results <- list()
+
+# correlation results -----
+
+  insert_msg('Correlation result table')
+
+  tst_results$correlation <- list(trp = tst_trp,
+                                  tyr = tst_tyr) %>%
+    map(~.x$correlation$test) %>%
+    compress(names_to = 'metabolic_subsystem') %>%
+    transmute(Metabolite = exchange(variable2,
+                                    dict = globals$stigma_lexicon),
+              `Explanatory variable` = exchange(variable1,
+                                                dict = globals$stigma_lexicon),
+              n = n,
+              `Correlation coefficient, 95% CI` = est_lab,
+              Significance = significance)
+
+# Comparison results -----
+
+  insert_msg('Comparison results')
+
+  tst_results$comparison <- list(trp = tst_trp,
+                                 tyr = tst_tyr) %>%
+    map(~left_join(.x$comparison$stats[c('variable', 'split_factor',
+                                         'level', 'statistic')],
+                   .x$comparison$test[c('variable', 'split_factor',
+                                        'significance', 'eff_size')],
+                   by = c('variable', 'split_factor'))) %>%
+    compress(names_to = 'metabolic_subsystem') %>%
+    format_tbl(dict = globals$stigma_lexicon) %>%
+    mutate(split_factor = exchange(split_factor, dict = globals$stigma_lexicon),
+           n = stri_extract(statistic, regex = '\\d+$'),
+           statistic = stri_replace(statistic, regex = '\\ncomplete.*$', replacement = ''),
+           variable = factor(variable,
+                             exchange(c(tst_trp$responses, tst_tyr$responses),
+                                      dict = globals$stigma_lexicon))) %>%
+    arrange(variable, split_factor) %>%
+    select(variable,
+           split_factor,
+           level,
+           n,
+           statistic,
+           significance,
+           eff_size) %>%
+    set_names(c('Metabolite',
+                'Explanatory variable',
+                'Stratum',
+                'Observations/stratum',
+                'Statistic',
+                'Significance',
+                'Effect size'))
+
+# END ------
+
+  insert_tail()

### 2023-05-15 testing scripts/trp_system.R

@@ -1,0 +1,209 @@
+# Testing for significant association of single factors for the TRP/KYN system
+
+  insert_head()
+
+# container -----
+
+  tst_trp <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variables
+
+  tst_trp$lexicon <- globals$stigma_lexicon %>%
+    filter(variable %in% test_globals$variables$trp)
+
+  tst_trp$responses <- stigma$responses[c('trp', 'kyn', 'kyn_trp')]
+
+  ## analysis tables with non-normalized variable values
+
+  tst_trp$analysis_tbl <- stigma$data %>%
+    mutate(infection = car::recode(cov,
+                                   "'healthy' = 'uninfected';
+                                    'SARS-CoV-2' = 'recovery'"),
+           infection = factor(infection,
+                              c('uninfected', 'recovery'))) %>%
+    select(patient_id,
+           all_of(unname(tst_trp$responses)),
+           all_of(tst_trp$lexicon$variable)) %>%
+    filter(patient_id %in% stigma$complete_ids)
+
+  ## analysis types
+
+  tst_trp$lexicon$analysis_type <-
+    tst_trp$analysis_tbl[c(tst_trp$lexicon$variable)] %>%
+    map_lgl(is.numeric) %>%
+    ifelse('correlation', 'comparison')
+
+  ## variable types
+
+  tst_trp$var_types <- tst_trp$lexicon %>%
+    blast(analysis_type) %>%
+    map(~set_names(.x$variable,
+                   .x$variable))
+
+# Comparison ------
+
+  insert_msg('Comparison')
+
+  ## descriptive stats
+
+  tst_trp$comparison$stats <- tst_trp$var_types$comparison %>%
+    map(~explore(tst_trp$analysis_tbl,
+                 split_factor = .x,
+                 variables = tst_trp$responses,
+                 what = 'table',
+                 pub_styled = TRUE)) %>%
+    map(compress, names_to = 'level') %>%
+    compress(names_to = 'split_factor')
+
+  ## testing for differences: T test
+
+  tst_trp$comparison$test <- tst_trp$var_types$comparison %>%
+    map(~compare_variables(tst_trp$analysis_tbl,
+                           variables = tst_trp$responses,
+                           split_factor = .x,
+                           what = 'eff_size',
+                           types = 'cohen_d',
+                           ci = FALSE,
+                           pub_styled = TRUE)) %>%
+    compress(names_to = 'split_factor') %>%
[…]

### 2023-05-15 testing scripts/tyr_system.R

@@ -1,0 +1,207 @@
+# Testing for significant association of single factors for the TRP/KYN system
+
+  insert_head()
+
+# container -----
+
+  tst_tyr <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variables
+
+  tst_tyr$lexicon <- globals$stigma_lexicon %>%
+    filter(variable %in% test_globals$variables$tyr)
+
+  tst_tyr$responses <- stigma$responses[c('phe', 'tyr', 'phe_tyr')]
+
+  ## analysis tables with non-normalized variable values
+
+  tst_tyr$analysis_tbl <- stigma$data %>%
+    mutate(infection = car::recode(cov,
+                                   "'healthy' = 'uninfected';
+                                    'SARS-CoV-2' = 'recovery'"),
+           infection = factor(infection,
+                              c('uninfected', 'recovery'))) %>%
+    select(patient_id,
+           all_of(unname(tst_tyr$responses)),
+           all_of(tst_tyr$lexicon$variable)) %>%
+    filter(patient_id %in% stigma$complete_ids)
+
+  ## analysis types
+
+  tst_tyr$lexicon$analysis_type <-
+    tst_tyr$analysis_tbl[c(tst_tyr$lexicon$variable)] %>%
+    map_lgl(is.numeric) %>%
+    ifelse('correlation', 'comparison')
+
+  ## variable types
+
+  tst_tyr$var_types <- tst_tyr$lexicon %>%
+    blast(analysis_type) %>%
+    map(~set_names(.x$variable,
+                   .x$variable))
+
+# Comparison ------
+
+  insert_msg('Comparison')
+
+  ## descriptive stats
+
+  tst_tyr$comparison$stats <- tst_tyr$var_types$comparison %>%
+    map(~explore(tst_tyr$analysis_tbl,
+                 split_factor = .x,
+                 variables = tst_tyr$responses,
+                 what = 'table',
+                 pub_styled = TRUE)) %>%
+    map(compress, names_to = 'level') %>%
+    compress(names_to = 'split_factor')
+
+  ## testing for differences: T test
+
+  tst_tyr$comparison$test <- tst_tyr$var_types$comparison %>%
+    map(~compare_variables(tst_tyr$analysis_tbl,
+                           variables = tst_tyr$responses,
+                           split_factor = .x,
+                           what = 'eff_size',
+                           types = 'cohen_d',
+                           ci = FALSE,
+                           pub_styled = TRUE)) %>%
[…]

### 2023-05-15 testing.R

@@ -1,0 +1,48 @@
+# Univariable testing and correlation for factors found to be significantly
+# associated with the TRP/KN and the PHE/TYR systems as indentified by
+# linear modeling with backward elimination
+
+# tools ------
+
+  library(plyr)
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+
+  library(exda)
+
+  library(soucer)
+
+  insert_head()
+
+  c('./tools/tools.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  test_globals <- list()
+
+  test_globals$variables <-
+    list(trp = mod_eli$inference[c("trp", "kyn", "kyn_trp")],
+         tyr = mod_eli$inference[c("tyr", "phe_tyr")]) %>%
+    map(~map(.x,
+             filter,
+             p_value < 0.05,
+             variable != 'Intercept')) %>%
+    map(~map(.x, ~.x$variable)) %>%
+    map(reduce, union)
+
+# analysis scripts -----
+
+  insert_msg('Analysis scripts')
+
+  c('./testing scripts/trp_system.R',
+    './testing scripts/tyr_system.R',
+    './testing scripts/result_tables.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# END -----
+
+  insert_tail()

### 2023-05-15 tools/globals.R

@@ -1,0 +1,169 @@
+# This script contains project globals
+
+# libraries ----
+
+  library(plyr)
+  library(tidyverse)
+  library(stringi)
+
+# Data container ------
+
+  globals <- list()
+
+# graphics -----
+
+  globals$common_text <- element_text(size = 8,
+                                      face = 'plain',
+                                      color = 'black')
+
+  globals$common_margin <- ggplot2::margin(t = 5,
+                                           l = 4,
+                                           r = 2,
+                                           unit = 'mm')
+
+  globals$common_theme <-
+    theme_classic() +
+    theme(axis.text = globals$common_text,
+          axis.title = globals$common_text,
+          plot.title = element_text(size = 8,
+                                    face = 'bold'),
+          plot.subtitle = globals$common_text,
+          plot.tag = element_text(size = 8,
+                                  face = 'plain',
+                                  color = 'black',
+                                  hjust = 0,
+                                  vjust = 1),
+          plot.tag.position = 'bottom',
+          legend.text = globals$common_text,
+          legend.title = globals$common_text,
+          strip.text = globals$common_text,
+          strip.background = element_rect(fill = 'gray95',
+                                          color = 'gray80'),
+          plot.margin = globals$common_margin,
+          panel.grid.major = element_line(color = 'gray90'))
+
+# Variable lexicon, the local cohort ------
+
+  insert_msg('Variable lexicon, the local cohort')
+
+  globals$stigma_lexicon <-
+    list(age = c('age', 'years', NA),
+         log_age = c('log age', 'years', NA),
+         sqrt_age = c('sqrt age', 'years', NA),
+         sex = c('sex', NA, NA),
+         psych_comorb = c('mental illness', NA, NA),
+         somatic_comorb = c('somatic illness', NA, NA),
+         infection = c('SARS-CoV2', NA, NA),
+         bmi_class = c('body mass', NA, NA),
+         smoking = c('smoking', NA, NA),
+         alcohol = c('alcohol', NA, NA),
[…]

### 2023-05-15 tools/tools.R

@@ -1,0 +1,752 @@
+# This script provides tools for project data analyses
+
+# tools ----
+
+  library(plyr)
+  library(tidyverse)
+  library(furrr)
+  library(stringi)
+  library(trafo)
+  library(sciplot)
+
+# Descriptive stats -----
+
+  format_tbl <-function(data,
+                        rm_complete = FALSE,
+                        dict = NULL, ...) {
+
+    data <- data %>%
+      map_dfc(stri_replace,
+              regex = '^no.*\\nyes:\\s{1}',
+              replacement = '') %>%
+      map_dfc(stri_replace,
+              regex = '^Mean.*\\nMedian\\s{1}=\\s{1}',
+              replacement = '') %>%
+      map_dfc(stri_replace,
+              fixed = 'Range',
+              replacement = 'range') %>%
+      map_dfc(stri_replace,
+              fixed = 'Complete',
+              replacement = 'complete')
+
+    if(!is.null(dict)) {
+
+      data <- data %>%
+        exchange('variable',
+                 dict = dict, ...)
+
+    }
+
+    if(rm_complete) {
+
+      data <- data %>%
+        map_dfc(stri_replace,
+                regex = '\\ncomplete.*$',
+                replacement = '')
+
+    }
+
+    data
+
+  }
+
+# modeling ------
+
+  extract_caret_stats <- function(caret_list) {
+
+    caret_list %>%
+      future_map(summary.caretx,
+                 .options = furrr_options(seed = TRUE)) %>%
+      map(~map(.x, ~.x[c(3, 4), c('statistic', 'estimate')]) %>%
+            compress(names_to = 'data_type')) %>%
+      compress(names_to = 'response') %>%
+      mutate(statistic = factor(statistic, c('RMSE', 'rsq')),
+             data_type = factor(data_type, c('train', 'cv')))
+
+  }
+
+  plot_caret_stats <- function(caret_stats,
+                               titles = c('Fit error, SIMMUN', 'Explained variance, SIMMUN'),
+                               x_labs = list('RMSE', expression('R'^2)),
+                               subtitle = NULL) {
+
+    ## plots R-squared and RMSE for the training data and cross-validation
+
+    list(data = blast(caret_stats, statistic),
+         title = titles,
+         x_lab = x_labs) %>%
+      pmap(function(data, title, x_lab) data %>%
+             ggplot(aes(x = estimate,
+                        y = response,
+                        fill = data_type)) +
+             geom_bar(stat = 'identity',
+                      position = position_dodge(0.9),
[…]

### 2023-05-16 paper/markdown/main_manuscript.Rmd
Methods and Supplementary Methods
In addition, fitting the robust models with the MM algorithm

@@
 
-## Data and code availability
+# Introduction
+
+\newpage
+
+# Materials and Methods
+
+## Ethics statement
+
+The SIMMUN study was conducted in accordance with the Declaration of Helsinki and European Data Policy. 
+All participants gave written informed consent prior to enrollment. 
+Participants' data were stored and analyzed in anonymized form.
+This study was approved by the ethics committee of the Medical University Innsbruck, Austria (approval number: 1132/2020). 
+No approval by the ethics committee was required for analysis of the published anonymized INCOV data set [@Su2022].
+
+## Study cohorts
+
+### SIMMUN study
+
+Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck Medical University (Innsbruck, Austria) and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR screening were invited to participate in the SIMMUN study.
+The study enrollment was conducted between 10. June 2020 (first patient enrolled) and 27.May 2021 (last study visit). 
+The inclusion criteria were age of 18 - 70 years, proficiency in German language, residence in the study region (Tyrol, Austria), results of a SARS-CoV-2 PCR test conducted at the University Hospital of Innsbruck.
+The exclusion criteria were active SARS-CoV-2 infection (< 14 days following a positive test), pregnancy, active malignancies, organ transplantation, prior surgery in the past 3 months, or acute or chronic inflammatory illness and treatment with oral corticosteroids.
+The analysis inclusion criterion was the complete study variable dataset (__Supplementary Table S1__). 
+A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were included in the analysis (__Figure \@ref(fig:fig-inclusion)__, __Table \@ref(tab:tab-stigma-cohort)__). 
+Significant differences between the analyzed participants and participants excluded due to data missingness are listed in __Supplementary Table S2__.
+
+### INCOV study
+
+Proteome and metabolome data and clinical information for the INCOV cohort were extracted from supplementary tables accompanying the report by Su and colleagues [@Su2022]. 
+In the current analysis, `r nrow(incov$analysis_tbl)` INCOV study participants with the complete age information and dataset of metabolites related to serotonin and dopamine turnover and major inflammatory cytokines were included (__Figure 1__, __Table \@ref(tab:tab-incov-cohort)__, __Supplementary Table S3__).
+
+## Procedures
+
+Details on study procedures are provided in __Supplementary Methods__.
+
+### SIMMUN study
+
+The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
+The SIMMUN study data were gathered during in-person visits including a physician assessment, supervised completion of self-rating questionnaire and a blood sample collection.
+
+Demographic and clinical history variables: age, sex, body mass index (BMI), professionally diagnosed mental illness, self-reported chronic somatic conditions, smoking and alcohol consumption history, result and date of the SARS-CoV-2 PCR test were surveyed during the study visit or extracted from electronic patient records.
+
+Inflammatory markers: plasma neopterin concentrations (NEO) and neutrophil - lymphocyte ratio (NLR) were determined by the certified clinical routine laboratory at the University Hospital of Innsbruck.
+Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, and the KYN/TRP and PHE/TYR ratios calculated [@Capuron2011; @Widner1997; @Neurauter2008].
+Plasma titer of immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) were quantified by ELISA as described before [@Deisenhammer2021]. 
+Laboratory measurements at and beyond the detection limits were substituted with the lower or upper detection limit value, respectively.
+
+Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
+Anxiety and depression signs were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
+The total possible score range for each subscale is 0 to 21, with higher scores indicating more intense symptoms of anxiety/depression. 
+Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points [@Bjelland2002; @Zigmond1983].
+
+In order to improve normality of some numeric study variables prior to linear modeling and statistical hypothesis testing with parametric tools, logarithm or square root transformations were applied (__Supplementary Table S1__)
+
+### INCOV cohort
+
+Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and ultra-high-performance liquid chromatography/tandem accurate mass spectrometry (Metabolon, USA) [@Su2022; @Su2020].
+The proteome, metabolome data in form of normalized, age- and sex-adjusted, log~2~-transformed plasma concentrations as well as clinical information were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) of the report by Su et al.
+[@Su2022].
+In the current report plasma levels of metabolites implicated in biosynthesis and turnover, and competitor pathway products for serotonin (serotonin [5-HT], TRP, KYN, quinolinate [QUIN]) and dopamine (PHE, TYR and dopamine 3-O-sulfate [DA sulfate]), as well as major inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) were analyzed (__Supplementary Table S3__).
+
+Plasma metabolites and proteome were measured in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
+acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['acute']` days), 
+sub-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['sub-acute']`) and 
+recovery (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['recovery']` after diagnosis) (__Supplementary Table S3__).
+
+## Study endpoints
+
+The primary analysis endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing plasma levels of serotonin and dopamine precursors and products of competitor metabolic pathways (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR) in a real world collective of medical service users during the SARS-CoV-2 pandemics.
+This endpoint was addressed by multi-parameter modeling and univariable statistical hypothesis testing in the SIMMUN cohort. 
+The secondary analysis endpoint was to investigate how the readouts of systemic serotonin and dopamine availability, plasma 5-HT and DA sulfate, are influenced by age, inflammatory cytokines, timepoint of the SARS-CoV-2 infection and serum levels of their precursors (TRP, PHE, TYR) and competitor pathway products (KYN, QUIN).
+This endpoint was addressed by multi-parameter robust linear modeling, univariate time course modeling and correlation analysis in the INCOV cohort (__Figure \@ref(fig:fig-inclusion)__).
+
+## Statistical analysis
+
+Details of data transformation and analysis are provided in __Supplementary Methods__.
+
+R version 4.2.3 was used for statistical analysis.
+
+Numeric variables were presented in the text and tables as medians with interquartile ranges and ranges. 
+Categorical variables are presented as percentages and counts within the complete observation set. 
+Distribution normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively. 
+In the SIMMUN cohort, non-normally distributed numeric variables were transformed with the logarithm or square root functions prior to modeling and analyses with parametric tests (__Supplementary Table S1__).
+Since most of the INCOV study variables were non-normally distributed, robust linear modeling and non-parametric testing were employed in the analysis of those data. 
+Uniformity and consistency of psychometric tools was investigated by factor analysis and global McDonald's $\omega$ statistic [@McDonald1999; @Bartlett1937; @Revelle2022]. 
+Except for results of multi-parameter modeling, p values were corrected for multiple testing with the false discovery rate method [@Benjamini1995] separately for each analysis task. 
+Effects with p < 0.05 were considered significant.
+
+Correlation was assessed by Spearman's rank or Pearson test. 
+Significance of comparisons of numeric variables between two groups was determined by Mann-Whitney test with r effect size statistic or by two-tailed T test with Cohen's d effect size metric. 
+Comparisons of categorical variable distributions were evaluated by $\chi^2$ test with Cramer V effect size statistic. 
+Correlation was assessed by Pearson's or Spearman's rank test [@Cohen2013; @Kassambara2021].
+
+In the SIMMUN cohort, effects of age, sex, body mass class, somatic and metal conditions, body mass class, smoking and alcohol consumption history, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, signs of depression and anxiety (HADS), mental stress scoring (PSS-4) on systemic levels of aminoacid neurotransmitter precursor and competitor pathway products (TRP, KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio) was assessed by multi-parameter linear regression with backward elimination.
+Modeling responses and explanatory variables were subjected to normality-stabilizing transformations (__Supplementary Table S1__) and normalized prior to modeling in the SIMMUN cohort. 
+The normality and homogeneity of linear model residuals were evaluated by Shapiro-Wilk and Levene test, respectively, and visual inspection (residuals versus fitted and quantile-quantile plots). 
+In the INCOV cohort, effects of age, plasma levels of inflammatory cytokines (IL6, IL10, TNF, IFNG), timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control) and plasma levels of neurotransmitter precursors and competitor pathway products (TRP, KYN, QUIN, PHE, TYR) on plasma levels of 5-HT and DA sulfate were modeled by multi-parameter robust linear regression with the MM algorithm and Huber psi function [@Huber2011; @Ripley2022].
+Reproducibility and proper parameterization of the the linear and robust multi-parameter models was investigated by comparison of the RMSE and $R^2$ statistics in with the training dataset and in 10-fold cross-validation [@Kuhn2008].
+Significance of the model estimates was assessed by two-tailed T test.
+
+Differences in normalized blood concentrations of inflammatory cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery in the INCOV collective were investigated with robust linear modeling (MM algorithm, Huber's psi function) with uninfected subset serving as baseline.
+
+Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of inflammatory cytokines and metabolites in the INCOV cohort. 
+The correlation matrices were subsequently scaled into the [0, 1] range and converted to undirected force directed graphs [@Csardi2006]. 
+The graphs were visualized as two-dimensional network plots with the node proximity determined by the $\rho$ value and random repulsion, and edge color and width coding for the $\rho$ value and sign [@Briatte2021].
+
+\newpage
+
+# Results
+
+\newpage
+
+# Discussion
+
+\newpage
+
+# Data and code availability
 
@@
 The INCOV cohort data are publicly available at https://data.mendeley.com/datasets/96v329bg7g/1.
-The analysis pipeline code is available at https://github.com/PiotrTymoszuk/simmun_revision.
+The analysis pipeline code is available at https://github.com/PiotrTymoszuk/simmun.
 
@@
 
-```{r fig-uni-trp, fig.width = figur::convert(figs$uni_trp, to = 'in')$w, fig.height = figur::convert(figs$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}
+```{r fig-uni-trp, fig.width = figur::convert(figs$uni_trp, to = 'in')$w, fig.height = figur::convert(figs$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, plasma inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}
 
@@
 __Figure \@ref(fig:fig-uni-trp).
-Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__
+Effects of age, plasma inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__
 
 _Age, systemic inflammatory markers: neopterin (NEO), neutrophil - lymphocyte ratio (NLR), SARS-CoV-2 infection status, depression signs (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of systemic levels of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP).
-Their association with serum concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
+Their association with plasma concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
 
@@
 (A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
-(B) Scatter plots of serum concentrations of NEO, TRP, KYN and KYN/TRP.
+(B) Scatter plots of plasma concentrations of NEO, TRP, KYN and KYN/TRP.
 Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
@@
 
-```{r fig-uni-tyr, fig.width = figur::convert(figs$uni_tyr, to = 'in')$w, fig.height = figur::convert(figs$uni_tyr, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.'}
+```{r fig-uni-tyr, fig.width = figur::convert(figs$uni_tyr, to = 'in')$w, fig.height = figur::convert(figs$uni_tyr, to = 'in')$h, fig.cap = 'Effects of age, plasma inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-uni-tyr). Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.__
+__Figure \@ref(fig:fig-uni-tyr). Effects of age, plasma inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.__
 
 _Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of systemic levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR).
-Their association with serum concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
+Their association with plasma concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
 
@@
 
-```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of serum levels of serotonin and dopamine sulfate in the INCOV cohort.'}
+```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine sulfate in the INCOV cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of serum levels of serotonin and dopamine sulfate in the INCOV cohort.__
+__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine sulfate in the INCOV cohort.__
 
-_Effects of patient's age, timepoint after SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection as compared with healthy controls), serum levels of metabolites related to neurotransmitter biosynthesis and competitor pathways (tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE] and tyrosine [TYR]), and serum concentrations of inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on serum concentrations of serotonin (5-hydroxy tryptophan [5-HT]) and dopamine sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
+_Effects of patient's age, timepoint after SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection as compared with healthy controls), plasma levels of metabolites related to neurotransmitter biosynthesis and competitor pathways (tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE] and tyrosine [TYR]), and plasma concentrations of inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin (5-hydroxy tryptophan [5-HT]) and dopamine sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
 
@@
 
-_Time courses of normalized log~2~-transformed serum concentrations of inflammatory cytokines (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF] and interferon gamma [INFG]), metabolites implicated in serotonin synthesis (B, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN] and serotonin/5-hydroxy tryptophan [5-HT]) and metabolites implicated in dopamine turnover (C, phenylalanine [PHE], tyrosine [TYR] and dopamine sulfate [DA sulfate]).
+_Time courses of normalized log~2~-transformed plasma concentrations of inflammatory cytokines (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF] and interferon gamma [INFG]), metabolites implicated in serotonin synthesis (B, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN] and serotonin/5-hydroxy tryptophan [5-HT]) and metabolites implicated in dopamine turnover (C, phenylalanine [PHE], tyrosine [TYR] and dopamine sulfate [DA sulfate]).
 Differences between uninfected controls and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after infection, respectively) were modeled by robust linear regression with the uninfected controls as baseline.
@@
 
-```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of serum levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.'}
+```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of plasma levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.'}
 
@@
 __Figure \@ref(fig:fig-incov-corr).
-Correlation of serum levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.__
+Correlation of plasma levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.__
 
-_Serum levels of metabolites implicated in serotonin synthesis (A, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin [5-hydroxy tryptophan, 5-HT]), dopamine turnover (B, phenylalanine [PHE], tyrosine [TYR], dopamine sulfate [DA sulfate]) and inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in uninfected and SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection).
+_plasma levels of metabolites implicated in serotonin synthesis (A, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin [5-hydroxy tryptophan, 5-HT]), dopamine turnover (B, phenylalanine [PHE], tyrosine [TYR], dopamine sulfate [DA sulfate]) and inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in uninfected and SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection).
 Correlation coefficient matrices for at least moderate effects ($\rho$ > 0.2) were visualized as force-directed network plots.

### 2023-05-16 paper/markdown/supplementary_material.Rmd
Methods and Supplementary Methods
In addition, fitting the robust models with the MM algorithm

@@
 
-Proteome and metabolome data were analyzed with R version 4.2.0. 
-General data transformation tasks were accomplished with the _tidyverse_ package bundle [@Wickham2019], _rlang_ [@Henry2022] and the development package _trafo_ (https://github.com/PiotrTymoszuk/trafo).
-Statistical data testing was done with the packages _rstatix_ [@Kassambara2021], _ggpubr_ [@Kassambara2020] along with the development package _ExDA_ (https://github.com/PiotrTymoszuk/ExDA). 
-In linear modeling, base R functions, the development packages _lmqc_ (https://github.com/PiotrTymoszuk/lmqc) and _caretExtra_ (https://github.com/PiotrTymoszuk/caretExtra), and the package _caret_ [@Kuhn2008] were utilized.
+Study data were analyzed with R version 4.2.3. 
+Import and general data transformation tasks were accomplished with _foreign_ [@RCoreTeam], _readxl_ [@Wickham2022], the _tidyverse_ package bundle [@Wickham2019], _rlang_ [@Henry2022] and the development package [_trafo_](https://github.com/PiotrTymoszuk/trafo).
+Text data was handled with _stringi_ [@Gagolewski2021].
 
-Results were visualized with tools provided by the packages _ggplot2_ [@Wickham2016], _cowplot_ [@Wilke2019] and _ExDA_. 
-Manuscript and supplementary tables were created with _flextable_ [@Gohel2022]. 
-Supplementary Material file was written in the _rmarkdown_ environment [@Allaire2022] and rendered with the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages.
+Descriptive statistics, statistical hypothesis testing and correlation analysis was done with the packages _rstatix_ [@Kassambara2021] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA). 
+For linear modeling and robust linear modeling, base R functions, the packages _stats_, _MASS_ [@Ripley2022], _caret_ [@Kuhn2008], [_lmqc_](https://github.com/PiotrTymoszuk/lmqc) and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) were employed.
+Correlation matrices were converted to force-directed graphs with _igraph_ [@Csardi2006]. 
+For factor analysis and assessment of psychometric tool consistency, _stats_, _psych_ [@Revelle2022] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) were used.
+
+Results were visualized with tools provided by the packages _ggplot2_ [@Wickham2016] (ribbon and bubble plots), [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (scatter and box plots), [_lmqc_](https://github.com/PiotrTymoszuk/lmqc) (Forest plots) and _ggnetwork_ [@Briatte2021] (force-directed graphs).
+Figures and tables were created with _cowplot_ [@Wilke2019] and _flextable_ [@Gohel2022], respectively. 
+Parts of the manuscript and Supplementary Material were written in the _rmarkdown_ environment [@Allaire2022] and rendered with the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages. 
+Management of figures and tables in the rmarkdown documents was accomplished with [_figur_](https://github.com/PiotrTymoszuk/figur).
+
+## SIMMUN cohort dataset
+
+Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck Medical University (Innsbruck, Austria), as well as inpatients and outpatients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR screening were invited to participate in the SIMMUN study.
+The study enrollment was initiated on 10. June 2020. 
+The inclusion criteria for the SIMMUN study were age of 18 - 70 years, proficiency in German language, residence in the study region (Tyrol, Austria), results of a SARS-CoV-2 PCR test conducted at the University Hospital of Innsbruck.
+The exclusion criteria were active SARS-CoV-2 infection (< 14 days following a positive test), pregnancy, active malignancies, organ transplantation, prior surgery in the past 3 months, or acute or chronic inflammatory illness and treatment with oral corticosteroids.
+The analysis inclusion criterion was the complete study variable dataset consisting of 
+basic demographic and medical history variables, 
+SARS-CoV-2 infection status and titer of anti-receptor-binding domain S12/S2 protein immunoglobulin gamma (anti-RBD IgG), 
+psychometric scoring of depression, anxiety and stress as well as blood levels of inflammatory markers and metabolites related to serotonin and dopamine biosynthesis. 
+A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were analyzed here. 
+The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
+The SIMMUN study data were gathered during in-person visits at the study center conducted between 17.June 2020 and 27.May 2021. 
+The SIMMUN analysis inclusion scheme is presented in __Figure 1__. 
+Significant differences between participants included in the analysis and excluded due to data missingness are shown in __Supplementary Table S\@ref(tab:tab-excluded)__. 
+Characteristic of the analyzed SIMMUN collective split by the SARS-CoV-2 infection status is presented in __Table 1__.
+
+Demographic and clinical history variables: age, sex, body mass index (BMI), professionally diagnosed mental illness, self-reported chronic somatic conditions, smoking and alcohol consumption history were surveyed during the study visit or extracted from electronic patient records.
+Body mass classes were defined as follows: normal for BMI < 25 kg/m^2^, overweight for BMI 25 - 30 kg/m^2^ and obesity for BMI > 30 kg/m^2^. 
+Result and date of the SARS-CoV-2 PCR test were retrieved from electronic patient records.
+
+The inflammatory markers included in the analysis: plasma concentration of neopterin (NEO) and neutrophil - lymphocyte ratio (NLR) were determined by the certified clinical routine laboratory at the University Hospital of Innsbruck.
+C-reactive protein (CRP) and interleukin-6 (IL6) were determined for the SIMMUN cohort as well but excluded from the analysis, since their levels were within the reference range in the great majority of participants.
+Elevated CRP > 0.5 mg/dL was detected in 
+`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
+elevated IL6 > 7 pg/mL observed in 
+`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of the analyzed SIMMUN collective. 
+Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, as described elsewhere [@Widner1997; @Neurauter2008].
+Kynurenine - tryptophan and phenylalanine - tyrosine ratios (KYN/TRP and PHE/TYR) serving as readouts of systemic indoleamine 2,3-dioxygenase (IDO) and phenylalanine hydroxylase (PAH) were computed [@Capuron2011].
+Plasma titer of anti-RBD IgG were quantified by ELISA as described before [@Deisenhammer2021]. 
+Strongly non-normally distributed anti-RBD IgG levels were expressed as arbitrary units (AU) were stratified with the $\leq$ 1, 1 - 16.3 and > 16.3 AU cutpoints defined by the maximal titer in the uninfected participants (1 AU) and median titer in the SARS-CoV-2 subset (16.3 AU).
+Laboratory measurements at and below the lower detection limit were substituted with the lower detection limit. 
+Laboratory measurements at and above the upper detection limit were substituted with the upper detection limit.
+
+Mental stress was gauged with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
+Anxiety and depression signs were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
+The total possible score range for each subscale is 0 to 21, with higher scores indicating more intense symptoms of anxiety/depression. 
+Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points at the respective subscale as proposed in the seminal HADS paper and a recent metaanalysis [@Bjelland2002; @Zigmond1983].
+PSS-4, HADS anxiety and depression scales displayed good-to-excellent internal consistency as measured by the McDonald $\omega$ statistic [@McDonald1999] (__Supplementary Figure S\@ref(fig:fig-consistency)__).
+
+In order to improve normality of some numeric SIMMUN study variables prior to linear modeling and statistical hypothesis testing with parametric tools, 
+logarithm 
+(`r suppl_tabs$study_vars %>% filter(Transformation == 'logarithm') %>% .[['Variable label']] %>% paste(collapse = ', ')`) or square root transformations 
+(`r suppl_tabs$study_vars %>% filter(Transformation == 'square root') %>% .[['Variable label']] %>% paste(collapse = ', ')`) were applied. 
+Strongly non-normally distributed HADS scores and anti-RBD IgG were stratified with the cutoffs described above. 
+Transformation and stratification schemes of study variable are provided in __Supplementary Table S\@ref(tab:tab-study-vars)__.
+
+SIMMUN cohort data were stored as SPSS files and imported in R with the `read.spss()` function (package _foreign_). 
+Transformation, stratification of numeric study variables and coding of categorical features was done with an in-hose developed script.
+
+## INCOV cohort dataset
+
+Plasma proteome measurements in the INCOV cohort were obtained by proximity extension assay (PEA, Olink, Sweden). 
+Plasma metabolome was quantified by ultra-high-performance liquid chromatography/tandem accurate mass spectrometry (Metabolon, USA) [@Su2022; @Su2020]. 
+Proteome and metabolome data in form of normalized, age- and sex-adjusted, log~2~-transformed plasma levels as well as clinical information (sex, SARS-CoV-2 infection status, COVID-19 severity, timepoint) for the INCOV cohort were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) accompanying the report by Su and colleagues [@Su2022] (function `read_xlsx`, package _readxl_) [@Wickham2022].
+In the current analysis, `r nrow(incov$analysis_tbl)` INCOV study participants with the complete dataset of metabolites of interest (serotonin [5-HT], tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE], tyrosine [TYR] and dopamine 3-O-sulfate [DA sulfate]), inflammatory cytokines of interest (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and age at enrollment were included in the current analysis.
+The INCOV study variables are listed in __Supplementary Table S\@ref(tab:tab-incov-study-vars)__. 
+The analysis inclusion scheme for the INCOV cohort is presented in __Figure 1__. 
+Characteristic of the INCOV collective is shown in __Table 2__.
+Concerning the INCOV study design, plasma metabolites and proteome were determined in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
+acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['acute']` days), 
+sub-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['sub-acute']`) and 
+recovery (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['recovery']` after diagnosis). 
+The measurements were only partially matched by participant (i.e. multiple longitudinal measurements for an individual) hence precluding a classical repeated measurement analysis approach. 
+Numbers of available INCOV cohort samples and the sampling timepoints are shown in __Supplementary Table S\@ref(tab:tab-samples)__.
+
+## Descriptive statistic and variable distribution
+
+Numeric variables were presented in the text and tables as medians with interquartile ranges and ranges. 
+Categorical variables are presented as percentages and counts within the complete observation set. 
+Descriptive statistics were computed with the function `explore()` from the [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package.
+
+Distribution normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively (functions `explore()` and `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+The distribution testing revealed substantial deviations from normality for some SIMMUN study parameters. 
+For this reason, logarithm and square root transformations were applied prior to modeling and statistical hypothesis testing with parametric tools as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__.
+Most of the INCOV study variables were non-normally distributed. 
+For this reason, non-parametric Spearman rank test was chosen for correlation analysis and robust linear regression was used as a modeling approach.
+
+## Statistical significance, effect size
+
+Except for results of multi-parameter linear modeling and multi-parameter robust linear regression, p values were corrected for multiple testing with the false discovery rate method (FDR) [@Benjamini1995] separately for each analysis task.
+Effects with p < 0.05 were considered significant. 
+The following effect size measures were used [@Cohen2013]: 
+Cramer's V for comparison of distribution of categorical variables (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5), 
+Cohen's d for two group comparisons of normally distributed variables (weak: < 0.5, moderate: 0.5 - 0.8, large effect: $\geq$ 0.8), 
+r for two group comparisons of non-normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5), 
+$\rho$ for correlation of non-normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, strong correlation: $\geq$ 0.5), 
+r for correlation of normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, strong correlation: $\geq$ 0.5), 
+$R^2$ as a measure of explained variance in modeling (weak: < 0.13, moderate: 0.13 - 0.26, substantial: $\geq$ 0.26).
+
+## Statistical hypothesis testing
+
+Differences in characteristic of SIMMUN participants included in the analysis and excluded due to data missingness were assessed by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Supplementary Table S\@ref(tab:tab-excluded)__).
+Differences in characteristic of SIMMUN participants or INCOV study participants stratified by the SARS-CoV-2 infection status were investigated by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Tables 1__ and __2__).
+Correlation of metabolite concentrations with age, stress scoring (PSS-4) and inflammation measures (NEO and NLR) in the SIMMUN cohort was done by Pearson's test (__Supplementary Table S\@ref(tab:tab-univariate-correlation)__).
+Differences in metabolite concentrations and metabolite ratios in the SIMMUN cohort stratified by the SARS-CoV-2 infection status or presence of depression symptoms (HADS) were assessed by tow-tailed T test with Cohen's d effect size metric (__Supplementary Table S\@ref(tab:tab-univariate-complarison)__).
+
+Statistical hypothesis testing was accomplished with the functions `compare_variables()` and `correlate_variables()` from the development package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) for comparison and correlation, respectively.
+
+## Multi-parameter linear regression with backwards elimination in the SIMMUN cohort
+
+Effects of age, sex, body mass class, somatic and metal conditions, body mass class, smoking and alcohol consumption history, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, signs of depression and anxiety (HADS), mental stress scoring (PSS-4) on systemic levels of aminoacid neurotransmitter precursor and products of competitor pathways (TRP, KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio) was assessed by multi-parameter linear regression with backward elimination.
+Responses and numeric explanatory variables were transformed with log or squared root functions to improve normality as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__ and normalized prior to modeling.
+
+Full models including the complete set of explanatory variables listed above were constructed (function `make_lm()`, package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) and optimized by Bayesian information criterion (BIC) driven backwards elimination of non-significant terms (method `step()`, [_lmqc_](https://github.com/PiotrTymoszuk/lmqc), employing internally `stepAIC()` from the _MASS_ package) [@Ripley2022].
+Normality and homogeneity of distribution of the model residuals was checked by Shapiro-Wilk and Levene test, respectively (method `summary(type = 'assumptions')`, package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) and visually inspected in standard diagnostic plots of model residuals (residuals vs fitted, quantile-quantile plots, method `plot()` called for the model objects).
+Fit stats ($R^2$ and root mean squared error [RMSE]) were retrieved from the model objects by calling `summary(type = 'fit')` (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)). 
+Validity of the optimized models was determined by likelihood-ratio test (LRT) versus the respective null models (method `anova()`). 
+Reproducibility and proper parameterization of the optimized multi-parameter models was investigated by repeated cross-validation (10 folds, function `train(method = 'lm')`, package _caret_)[@Kuhn2008] and by comparison of the RMSE and $R^2$ statistics obtained with in the training dataset and in cross-validation (method `summary()`, package _caretExtra_).
+As presented in __Supplementary Figure S\@ref(fig:fig-fit-stats)A__, similar values of error fit in the training and cross-validation data suggest good reproducibility of the optimized models and lack of over-parameterization.
+
+Inference statistics for the model estimates ($\beta$, expected values and 95% confidence intervals) were extracted with the `summary()` method. 
+Significance of the estimates ($p(\beta\neq0)$) was determined by two-tailed T test. 
+Results of the linear modeling in the INCOV cohort are presented in __Figure 2__ and __Supplementary Table S\@ref(tab:tab-modeling-simmun)__.
+
+## Multi-paramater robust linear modeling in the INCOV cohort
+
+Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on systemic neurotransmitter availability were subsequently tested in the publicly available INCOV collective [@Su2022].
+To this end we modeled normalized plasma levels of serotonin (5-HT) as the end product of serotonin biosynthesis associated with mental disorders [@Holck2019; @Tateishi2021; @Sun2020; @Kheirouri2016; @Paul-Savoie2011; @Ertugrul2007], and the major product of systemic dopamine decay dopamine 3-O-sulfate (DA-sulfate) [@Goldstein1999; @Meiser2013].
+The explanatory variables were age, plasma levels of inflammatory cytokines (IL6, IL10, TNF, IFNG), plasma concentrations of neurotransmitter precursors and competitor pathway products (for 5-HT: TRP, KYN and quinolinate [QUIN], for DA sulfate: PHE and TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control).
+Since most numeric features were non-normally distributed, MM algorithm robust linear regression with Huber's psi function was chosen [@Huber2011; @Ripley2022] as a modeling approach.
+
+The robust linear models were built with the `make_lm()` wrapper ([_lmqc_](https://github.com/PiotrTymoszuk/lmqc) package) around the MASS package `rlm()` function [@Ripley2022]. 
+The parameterization and reproducibility of the models were tested by 10-fold cross-validation with the _caret_ package [@Kuhn2008] as described for multi-parameter linear regression. 
+As presented in __Figure 5A__, similar values of error fit in the training and cross-validation data suggest good reproducibility of the robust linear models and proper model parameterization. 
+Inference statistics were extracted from the models with `summary()` method. 
+Significance of the model estimates ($p(\beta\neq0)$) was determined by two-tailed T test, 95% confidence intervals were computed based on theoretical t distribution. 
+The robust linear modeling results are presented in __Figure 5__ and __Supplementary Table S\@ref(tab:tab-modeling-incov)__.
+
+## Time course of inflammatory cytokines and metabolites in the INCOV cohort
+
+Differences in normalized blood concentrations of inflammatory cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset serving as baseline.
+Models were constructed with the `make_lm()` function (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) employing internally the MM`rlm()` algorithm with Huber's psi function (package _MASS_) [@Huber2011; @Ripley2022].
+Inference statistics were computed as described for multi-parameter robust linear modeling. 
+The time course modeling results are presented in __Figure 6__ and __Supplementary Table S\@ref(tab:tab-time-modeling)__.
+
+## Correlation and network analysis in the INCOV cohort
+
+Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of inflammatory cytokines (IL6, IL10, TNF and IFNG), and metabolite involved in serotonin (5-HT, TRP, KYN and QUIN) and dopamine turnover (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
+The correlation matrices were subsequently scaled into the [0, 1] range with the $0.5\times(x + 1)$ function and converted to undirected force directed graphs with the `graph_from_adjacency_matrix()` function from the _igraph_ package [@Csardi2006].
+The graphs were visualized with tools provided by the _ggnetwork_ package [@Briatte2021] as two-dimensional network plots with the node proximity determined by the $\rho$ value and random repulsion, and edge color and width coding for the $\rho$ value and sign (__Figure 7__).
+
+## Data and code availability
+
+Anonymized patient data will be made available upon request to the corresponding author. 
+The INCOV cohort data are publicly available at https://data.mendeley.com/datasets/96v329bg7g/1.
+The analysis pipeline code is available at https://github.com/PiotrTymoszuk/simmun.
 
@@
   width(1, width = 3, unit = 'cm') %>% 
-  width(2, width = 3, unit = 'cm') %>% 
+  width(2, width = 3.7, unit = 'cm') %>% 
   width(3, width = 2.7, unit = 'cm') %>% 
@@
   width(5, width = 2.5, unit = 'cm') %>% 
-  width(6, width = 3.3, unit = 'cm') %>% 
+  width(6, width = 3.7, unit = 'cm') %>% 
   merge_v(1) %>% 
@@
            part = 'body', 
-           ref_symbols = 'a') %>% 
+           ref_symbols = 'b') %>% 
   theme_vanilla
@@
 
-```{r tab-modeling-simmun, tab.cap = 'Results of multi-parameter linear modeling of serum concentrations of tryptophan, kynurenine, tyrosine, and kynurenine - tryptophan, and phenylalanine - tyrosine ratios in the SIMMUN cohort.'}
+```{r tab-modeling-simmun, tab.cap = 'Results of multi-parameter linear modeling of plasma concentrations of tryptophan, kynurenine, tyrosine, and kynurenine - tryptophan, and phenylalanine - tyrosine ratios in the SIMMUN cohort.'}
 
 flextable::flextable(suppl_tabs$modeling_simmun) %>% 
-  width(1, width = 2, unit = 'cm') %>% 
-  width(2, width = 3, unit = 'cm') %>% 
+  width(1, width = 2.2, unit = 'cm') %>% 
+  width(2, width = 4, unit = 'cm') %>% 
   width(3, width = 2.7, unit = 'cm') %>% 
@@
 
-```{r tab-univariate-correlation, tab.cap = "Correlation of age, stress scoring blood neopterin levels and neutrophils - leukocyte ratio with serum metabolite levels in the SIMMUN cohort investigated by Pearson's test."}
+```{r tab-univariate-correlation, tab.cap = "Correlation of age, stress scoring blood neopterin levels and neutrophils - leukocyte ratio with plasma metabolite levels in the SIMMUN cohort investigated by Pearson's test."}
 
@@
   width(3, width = 2, unit = 'cm') %>% 
-  width(4, width = 3.5, unit = 'cm') %>% 
+  width(4, width = 3.9, unit = 'cm') %>% 
   width(5, width = 2.8, unit = 'cm') %>% 
   footnote(1, 1, 
-           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine - tryptophanm ratio; PHE: phenylalanine; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio.'), 
+           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine - tryptophan ratio; PHE: phenylalanine; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio.'), 
            part = 'header', 
@@
 
-```{r tab-univariate-complarison, tab.cap = "Comparison of serum metabolite levels in SIMMUN study participants stratified by depression signs and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}
+```{r tab-univariate-complarison, tab.cap = "Comparison of plasma metabolite levels in SIMMUN study participants stratified by depression signs and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}
 
@@
   width(1, width = 2.6, unit = 'cm') %>% 
-  width(2, width = 3.7, unit = 'cm') %>% 
-  width(3, width = 3, unit = 'cm') %>% 
-  width(4, width = 2.7, unit = 'cm') %>% 
+  width(2, width = 4.2, unit = 'cm') %>% 
+  width(3, width = 2.7, unit = 'cm') %>% 
+  width(4, width = 2.5, unit = 'cm') %>% 
   width(5, width = 3.7, unit = 'cm') %>% 
   width(6, width = 2.7, unit = 'cm') %>% 
-  width(7, width = 2.6, unit = 'cm') %>% 
+  width(7, width = 2.4, unit = 'cm') %>% 
   footnote(1, 1, 
@@
 
-```{r tab-modeling-incov, tab.cap = 'Results of multi-parameter robust linear modeling of serum concentrations of serotonin, and dopamine sulfate in the INCOV cohort.'}
+```{r tab-modeling-incov, tab.cap = 'Results of multi-parameter robust linear modeling of plasma concentrations of serotonin, and dopamine sulfate in the INCOV cohort.'}
 
@@
   footnote(1, 1, 
-           value = as_paragraph('5-HT: 5-hydroxy tryptophan/serotonine; DA sulfate: dopamine 3-O-sulfate.'), 
+           value = as_paragraph('5-HT: 5-hydroxy tryptophan/serotonin; DA sulfate: dopamine 3-O-sulfate.'), 
            part = 'header', 
@@
   footnote(1, 2, 
-           value = as_paragraph('TRP: tryptophan; KYN: kynuernine; QUIN: quinolinate; IL6: interleukin-6, IL10: interleukin-10; TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; PHE: phenylalanine; TYR: tyrosine.'),
+           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; QUIN: quinolinate; IL6: interleukin-6, IL10: interleukin-10; TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; PHE: phenylalanine; TYR: tyrosine.'),
            part = 'header', 
@@
 
-```{r tab-time-modeling, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokines, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.'}
+```{r tab-time-modeling, tab.cap = 'Results of robust linear modeling of plasma levels of inflammatory cytokines, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-consistency). Consistency of the PSS4 stress HADS depression and HADS anxiety psychometric tools in the SIMMUN cohort.__
+__Supplementary Figure S\@ref(fig:fig-consistency). Consistency of the PSS4 stress HADS depression and HADS anxiety psychometric tools in the SIMMUN cohort.__
 
@@
 
-__Figure \@ref(fig:fig-fit-stats). Root mean square error and R-squared statistics for multi-parameter linear models of tryptophan, tyrosine and their metabolites in the SIMMUN cohort.__
+__Supplementary Figure S\@ref(fig:fig-fit-stats).
+Root mean square error and R-squared statistics for multi-parameter linear models of tryptophan, tyrosine and their metabolites in the SIMMUN cohort.__
 
-_Multi-parameter linear regression models of serum levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by repeated cross-validation (CV, 10 folds).
+_Multi-parameter linear regression models of plasma levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by repeated cross-validation (CV, 10 folds).
 Values of root mean square error (RMSE, A) and $R^2$ (B) in the training data set and CV are plotted._
@@
 
-```{r fig-corr-trp, fig.width = figur::convert(suppl$corr_trp, to = 'in')$w, fig.height = figur::convert(suppl$corr_trp, to = 'in')$h, fig.cap = 'Correlation of tryptophan and its metabolites with age, serum levels of neopterin, neutrophil - lymphocyte ratio, and mental stress scoring in the SIMMUN cohort.'}
+```{r fig-corr-trp, fig.width = figur::convert(suppl$corr_trp, to = 'in')$w, fig.height = figur::convert(suppl$corr_trp, to = 'in')$h, fig.cap = 'Correlation of tryptophan and its metabolites with age, plasma levels of neopterin, neutrophil - lymphocyte ratio, and mental stress scoring in the SIMMUN cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-corr-trp). Correlation of tryptophan and its metabolites with age, serum levels of neopterin, neutrophil - lymphocyte ratio, and mental stress scoring in the SIMMUN cohort.__
+__Supplementary Figure S\@ref(fig:fig-corr-trp).
+Correlation of tryptophan and its metabolites with age, plasma levels of neopterin, neutrophil - lymphocyte ratio, and mental stress scoring in the SIMMUN cohort.__
 
-_Correlation of age (A), the inflammatory marker neutrophil-leukocyte ratio (B, NLR), and stress scoring (C, perceived stress scale, 4-item [PSS-4]) with serum concentrations of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP) was assessed by Pearson's test corrected for multiple testing with the false discovery rate method (n = `r nrow(stigma$analysis_tbl)`.
+_Correlation of age (A), the inflammatory marker neutrophil-leukocyte ratio (B, NLR), and stress scoring (C, perceived stress scale, 4-item [PSS-4]) with plasma concentrations of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP) was assessed by Pearson's test corrected for multiple testing with the false discovery rate method (n = `r nrow(stigma$analysis_tbl)`.
 Single observations are depicted as points, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
@@
 
-```{r fig-corr-tyr, fig.width = figur::convert(suppl$corr_tyr, to = 'in')$w, fig.height = figur::convert(suppl$corr_tyr, to = 'in')$h, fig.cap = 'Correlation of phenylalanine and tyrosine with serum levels of neopterin in the SIMMUN cohort.'}
+```{r fig-corr-tyr, fig.width = figur::convert(suppl$corr_tyr, to = 'in')$w, fig.height = figur::convert(suppl$corr_tyr, to = 'in')$h, fig.cap = 'Correlation of phenylalanine and tyrosine with plasma levels of neopterin in the SIMMUN cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-corr-tyr). Correlation of phenylalanine and tyrosine with serum levels of neopterin in the SIMMUN cohort.__
+__Supplementary Figure S\@ref(fig:fig-corr-tyr). Correlation of phenylalanine and tyrosine with plasma levels of neopterin in the SIMMUN cohort.__
 
-_Correlation of the inflammatory marker neopterin (NEO) with serum concentrations of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR) was was assessed by Pearson's test corrected for multiple testing with the false discovery rate method (n = `r nrow(stigma$analysis_tbl)`.
+_Correlation of the inflammatory marker neopterin (NEO) with plasma concentrations of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR) was was assessed by Pearson's test corrected for multiple testing with the false discovery rate method (n = `r nrow(stigma$analysis_tbl)`.
 Single observations are depicted as points, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.

### 2023-05-16 incov scripts/modeling.R

@@ -70,7 +70,18 @@
                  formula = .x,
                  mod_fun = rlm,
                  family = NULL,
+                 method = 'MM',
                  psi = psi.huber))
+
+# Custom model list for caret -------
+
+  insert_msg('A custom list for caret')
+
+  ## merely a copy-paste from the output of getModelInfo('rlm')
+  ## but specifying the method "MM" explicitly
+
+  incov_mod$caret_list <-
+
 
 # cross-validation via caret ------
 
@@ -85,7 +96,7 @@
                data = incov_mod$analysis_tbl %>%
                  mutate(timepoint = car::recode(timepoint,
                                                 "'sub-acute' = 'sub'")),
-               method = 'rlm',
+               method = mm_rlm,
                metric = 'RMSE',
                trControl = trainControl(method = 'repeatedcv',
                                         number = 10,

### 2023-05-16 incov scripts/neuro_modeling.R

@@ -74,9 +74,10 @@
                  formula = .x,
                  mod_fun = rlm,
                  family = NULL,
-                 psi = psi.huber))
+                 psi = psi.huber,
+                 method = 'MM'))
 
-  # cross-validation via caret ------
+# cross-validation via caret ------
 
   insert_msg('Cross-validation via caret')
 
@@ -89,7 +90,7 @@
                data = incov_neuro$analysis_tbl %>%
                  mutate(timepoint = car::recode(timepoint,
                                                 "'sub-acute' = 'sub'")),
-               method = 'rlm',
+               method = mm_rlm,
                metric = 'RMSE',
                trControl = trainControl(method = 'repeatedcv',
                                         number = 10,

### 2023-05-16 incov scripts/time_modeling.R

@@ -73,6 +73,7 @@
                        formula = .x,
                        mod_fun = rlm,
                        family = NULL,
+                       method = 'MM',
                        psi = psi.huber)))
 
 # Inference --------

### 2023-05-16 paper scripts/links.R

@@ -1,0 +1,30 @@
+# Handling of common links in the Rmarkdown project
+
+  insert_head()
+
+# container ------
+
+  proj_links <- list()
+
+# Links to development packages --------
+
+  insert_msg('Links to development packages')
+
+  proj_links <-
+    list('ExDA' = 'https://github.com/PiotrTymoszuk/ExDA',
+         'trafo' = 'https://github.com/PiotrTymoszuk/trafo',
+         'figur' = 'https://github.com/PiotrTymoszuk/figur',
+         'clustTools' = 'https://github.com/PiotrTymoszuk/clustTools',
+         'lmqc' = 'https://github.com/PiotrTymoszuk/lmqc',
+         'caretExtra' = 'https://github.com/PiotrTymoszuk/caretExtra') %>%
+    compress(names_to = 'obj_name',
+             values_to = 'x') %>%
+    mutate(ref_name = paste0('_', obj_name, '_'))
+
+  proj_links <- proj_links[c('ref_name', 'x')] %>%
+    pmap(mdlink) %>%
+    set_names(proj_links$obj_name)
+
+# END ------
+
+  insert_tail()

### 2023-05-16 paper.R

@@ -27,6 +27,7 @@
   c('./paper scripts/tables.R',
     './paper scripts/figures.R',
     './paper scripts/supplement.R',
+    './paper scripts/links.R',
     './paper scripts/render.R') %>%
     source_all(message = TRUE, crash = TRUE)
 

### 2023-05-16 tools/tools.R

@@ -749,4 +749,53 @@
 
   }
 
+# custom caret method for MM robust linear regression -----
+
+  mm_rlm <-
+    list(label = 'MM robust linear regression',
+         library = 'MASS',
+         type = 'Regression',
+         parameters = data.frame(parameter = c('intercept', 'psi'),
+                                 class = c('logical', 'character'),
+                                 label = c('intercept', 'psi')),
+         grid = function(x, y, search = "grid") {
+
+           expand.grid(intercept = c(TRUE, FALSE), psi = c("psi.huber",
+                                                           "psi.hampel", "psi.bisquare"))
+
+         },
+         fit = function (x, y, wts, param, lev, last, classProbs, ...) {
+           dat <- if (is.data.frame(x))
+             x
+           else as.data.frame(x, stringsAsFactors = TRUE)
+           dat$.outcome <- y
+           psi <- MASS::psi.huber
+           if (param$psi == "psi.bisquare")
+             psi <- MASS::psi.bisquare
+           else if (param$psi == "psi.hampel")
+             psi <- MASS::psi.hampel
+           if (!is.null(wts)) {
+             if (param$intercept)
+               out <- MASS::rlm(.outcome ~ ., data = dat, weights = wts,
+                                psi = psi, method = 'MM', ...)
+             else out <- MASS::rlm(.outcome ~ 0 + ., data = dat, weights = wts,
+                                   psi = psi, method = 'MM', ...)
+           }
+           else {
+             if (param$intercept)
+               out <- MASS::rlm(.outcome ~ ., data = dat, psi = psi,
+                                ...)
+             else out <- MASS::rlm(.outcome ~ 0 + ., data = dat, psi = psi,
+                                   ...)
+           }
+           out
+         },
+         predict = function (modelFit, newdata, submodels = NULL) {
+           if (!is.data.frame(newdata))
+             newdata <- as.data.frame(newdata, stringsAsFactors = TRUE)
+           predict(modelFit, newdata)
+         },
+         prob = NULL,
+         sor = function(x) x)
+
 # END ------

### 2023-05-18 paper/markdown/main_manuscript.Rmd
Results, new figure set

--- 2023-05-15 paper/markdown/main_manuscript.Rmd
@@
 ---
-title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
+title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic levels of aminoacid neurotransmitter precursors"
 subtitle: "Manuscript"
@@
 
-## Data and code availability
+# Keywords
+
+SARS-CoV-2, inflammation, serotonin, dopamine, tryptophan, kynurenine, mental health
+
+\newpage
+
+# Abstract
+
+__Background:__ Serotonin and dopamine metabolism poses a junction of mental and somatic health. 
+We aimed to identify demographic, clinical, mental health- and SARS-CoV-2 infection-related factors affecting systemic serotonin and dopamine turnover.
+
+__Methods:__ The cross-sectional SIMMUN (n = `r nrow(stigma$analysis_tbl)`, Austria) and longitudinal INCOV cohort (n = `r  length(unique(incov$complete_ids))`, Su et al. 2022) were investigated. 
+Explanatory variables encompassed age, sex, clinical and inflammatory parameters, SARS-CoV-2 infection, and anxiety, depression (HADS) and mental stress scoring (PSS-4). 
+Their effects on plasma concentrations of serotonin (serotonin, tryptophan, kynurenine) and dopamine availability markers (dopamine sulfate, phenylalanine, tyrosine) were assessed by linear modeling, correlation analysis and two-tailed T tests.
+
+__Results:__ In the SIMMUN collective, the inflammatory marker neopterin, SARS-CoV-2 infection, age, depression symptoms and mental stress independently stimulated catabolism of the serotonin precursor tryptophan to kynurenine.
+Inflammation and age were also suppressed the first step of dopamine biosynthesis, phenylalanine - tyrosine conversion. 
+In the INCOV cohort, inflammation was associated with lowered serotonin (IL6: $\beta$ = `r incov_neuro$inference$serotonin %>% filter(variable == 'IL6_INF') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`) and the dopamine metabolite, dopamine sulfate (interferon-gamma: $\beta$ = `r incov_neuro$inference$dopamine %>% filter(variable == 'IFNG_INF') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`).
+Serotonin correlated with tryptophan levels, especially in acute and sub-acute infection. 
+Serotonin ($\beta$ = `r incov_neuro$inference$serotonin %>% filter(variable == 'timepoint', level == 'recovery') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`)
+and dopamine sulfate ($\beta$ = `r incov_neuro$inference$dopamine %>% filter(variable == 'timepoint', level == 'recovery') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`) were significantly increased during SARS-CoV-2 infection recovery.
+
+__Conclusion:__ 
+Inflammation limits systemic serotonin and dopamine by activating the competitor kynurenine pathway and inhibiting the phenylalanine - tyrosine conversion, respectively, and may link SARS-CoV-2 infection with mental health problems.
+Advanced age and mental disorders can additionally suppress neurotransmitter synthesis.
+
+\newpage
+
+# Introduction
+
+_Katharina, Sophia: your part_
+
+The impact of persistent and SARS-CoV-2-related inflammation and immunity, SARS-CoV-2 recovery on indolamine and katecholamine neurotransmitter metabolism as well as its mutual interaction with mental disorder symptoms are still incompletely resolved.
+To address that, we explored effects of demographic and clinical factors, inflammation, SARS-CoV-2 infection, anti-SARS-CoV-2 humoral response and mental disorders on the tryptophan - kynurenine and phenylalanine axes as readouts of systemic availability of serotonin and dopamine in the SIMMUN cohort, a cross-sectional medical service user collective.
+Furthermore, association of circulating serotonin [@Paul-Savoie2011; @Sun2020; @Tateishi2021; @Holck2019] and of the major dopamine decay product, dopamine 3-O-sulfate [@Meiser2013; @Goldstein1999; @Hashizume1987] with systemic levels of inflammatory cytokines, SARS-CoV-2 course, age and availability of biosynthesis precursors and competition pathway metabolites were investigated in the published longitudinal INCOV collective [@Su2022].
+
+\newpage
+
+# Materials and Methods
+
+## Ethics statement
+
+The SIMMUN study was conducted in accordance with the Declaration of Helsinki and European Data Policy. 
+All participants gave written informed consent prior to enrollment. 
+Participants' data were stored and analyzed in anonymized form.
+This study was approved by the ethics committee of the Medical University Innsbruck, Austria (approval number: 1132/2020). 
+No approval by the ethics committee was required for analysis of the published INCOV data set [@Su2022].
+
+## Study cohorts
+
+Details on study cohorts, procedures and analses are provided in __Supplementary Methods__.
+
+### SIMMUN study
+
+Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR were invited to participate in the SIMMUN study.
+The study was conducted between 10. June 2020 and 27.May 2021. 
+The inclusion criteria were age of 18 - 70 years, proficiency in German, residence in the study region (Tyrol, Austria), and a SARS-CoV-2 PCR test conducted at the study site. 
+The exclusion criteria were active SARS-CoV-2 infection (< 14 days after diagnosis), pregnancy, active malignancy, organ transplantation, surgery in the past 3 months, inflammatory illness and oral corticosteroid treatment.
+The analysis inclusion criterion was the complete study variable dataset (__Figure \@ref(fig:fig-inclusion)__, __Supplementary Table S1__). 
+Significant differences between the analyzed and excluded participants are listed in __Supplementary Table S2__.
+
+### INCOV study
+
+Proteome and metabolome data and clinical information for the INCOV cohort were extracted from supplementary tables accompanying the report by Su and colleagues [@Su2022]. 
+In the current analysis, samples obtained for individuals with the complete age information and complete study variable dataset (__Figure 1__, __Table \@ref(tab:tab-incov-cohort)__, __Supplementary Table S3__).
+
+## Procedures
+
+### SIMMUN study
+
+The SIMMUN study data were gathered during in-person visits including a physician assessment, supervised completion of self-rating questionnaire and a blood sample collection (__Supplementary Table S1__).
+
+Demographic and clinical variables: age, sex, body mass index, professionally diagnosed mental illness, self-reported chronic somatic conditions, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR test were surveyed during the study visit or extracted from electronic patient records.
+
+Inflammatory markers: plasma neopterin concentrations (NEO) and neutrophil - lymphocyte ratio (NLR) were determined by the certified clinical routine laboratory at the University Hospital of Innsbruck.
+Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, and the KYN/TRP and PHE/TYR ratios calculated [@Capuron2011; @Widner1997; @Neurauter2008].
+Plasma titer of immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) were quantified by ELISA [@Deisenhammer2021]. 
+Laboratory measurements at and beyond the detection limits were substituted with the lower or upper detection limit value, respectively.
+
+Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
+Anxiety and depression signs were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
+The total possible score range for each subscale is 0 to 21, with higher scores indicating more intense symptoms of anxiety/depression. 
+Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points [@Bjelland2002; @Zigmond1983].
+
+In order to improve normality of some numeric study variables prior to linear modeling and statistical hypothesis testing with parametric tools, logarithm or square root transformations were applied (__Supplementary Table S1__)
+
+### INCOV cohort
+
+Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and ultra-high-performance liquid chromatography/tandem accurate mass spectrometry (Metabolon, USA) [@Su2022; @Su2020].
+Normalized, age- and sex-adjusted, log~2~-transformed plasma concentrations of metabolites and cytokines, and clinical information were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) of the report by Su et al.
+[@Su2022].
+Plasma levels of metabolites implicated in systemic turnover of serotonin (serotonin, TRP, KYN, quinolinate [QUIN]) and dopamine (PHE, TYR and dopamine 3-O-sulfate [DA sulfate]), major inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and age were analyzed (__Supplementary Table S3__).
+
+Plasma features were sampled in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
+acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['acute']` days), 
+sub-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['sub-acute']`) and 
+recovery (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['recovery']` after diagnosis) (__Supplementary Table S3__).
+
+## Study endpoints
+
+Our analysis pursued two endpoints. 
+The first endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing plasma levels of serotonin and dopamine precursors and products of competitor metabolic pathways (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR).
+This endpoint was addressed by multi-parameter modeling in the SIMMUN cohort. 
+The second endpoint was to investigate how the readouts of systemic serotonin and dopamine availability, plasma serotonin [@Paul-Savoie2011; @Sun2020; @Tateishi2021; @Holck2019] and DA sulfate [@Meiser2013; @Hashizume1987; @Goldstein1999], are influenced by age, inflammatory cytokines, timepoint of the SARS-CoV-2 infection, their precursors (TRP, PHE, TYR) and competitor pathway products (KYN, QUIN).
+This endpoint was addressed by multi-parameter robust linear modeling, time course modeling and correlation analysis in the INCOV cohort (__Figure \@ref(fig:fig-inclusion)__).
+
+## Statistical analysis
+
+R version 4.2.3 was used for statistical analysis.
+
+Numeric variables were presented as medians with interquartile ranges and ranges. 
+Categorical variables are presented as percentages and counts within the complete observation set. 
+Distribution normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively. 
+In the SIMMUN cohort, non-normally distributed numeric variables were transformed with the logarithm or square root prior to modeling and analyses with parametric tests (__Supplementary Table S1__). 
+Since most of the INCOV study variables were non-normally distributed, robust linear modeling and non-parametric testing were employed. 
+Since the tau-equivalence assumption investigated by factor analysis for the HADS scales was violated, consistency of psychometric tools was investigated by global McDonald's $\omega$ [@McDonald1999; @Bartlett1937; @Revelle2022].
+Except for multi-parameter modeling, p values were corrected for multiple testing with the false discovery rate method [@Benjamini1995] separately for each analysis task. 
+Effects with p < 0.05 were considered significant.
+
+Correlation was assessed by Spearman's rank or Pearson test. 
+Significance of comparisons of numeric variables between two groups was determined by Mann-Whitney test with r effect size statistic or by two-tailed T test with Cohen's d effect size metric. 
+Comparisons of categorical variable distributions were evaluated by $\chi^2$ test with Cramer V effect size statistic. 
+Correlation was assessed by Pearson's or Spearman's rank test [@Cohen2013; @Kassambara2021].
+
+In the SIMMUN cohort, effects of age, sex, body mass class, somatic and metal conditions, body mass class, smoking and alcohol consumption history, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, signs of depression and anxiety (HADS), mental stress scoring (PSS-4) on systemic levels of aminoacid neurotransmitter precursor and competitor pathway products (TRP, KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio) was assessed by multi-parameter linear regression with backward elimination.
+Modeling responses and explanatory variables were subjected to normality-stabilizing transformations (__Supplementary Table S1__) and normalized prior to modeling in the SIMMUN cohort. 
+Normality and homogeneity of linear model residuals were evaluated by Shapiro-Wilk and Levene test, respectively, and visual inspection (residuals versus fitted and quantile-quantile plots). 
+In the INCOV cohort, effects of age, plasma levels of inflammatory cytokines (IL6, IL10, TNF, IFNG), timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control) and plasma levels of neurotransmitter precursors and competitor pathway products (TRP, KYN, QUIN, PHE, TYR) on plasma serotonin and DA sulfate were modeled by multi-parameter robust linear regression with the MM algorithm and Huber psi function [@Huber2011; @Ripley2022].
+Reproducibility and proper parameterization of the multi-parameter linear and robust models was investigated by RMSE and $R^2$ statistics in 10-fold cross-validation [@Kuhn2008]. 
+Significance of the model estimates was assessed by two-tailed T test.
+
+Differences in cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery in the INCOV collective were investigated by robust linear modeling (MM algorithm, Huber's psi function) with uninfected subset or acute infection serving as baselines.
+
+Pairwise Spearman's correlation coefficients $\rho$ were calculated for cytokines and metabolites in the INCOV cohort. 
+The correlation matrices were subsequently scaled into the [0, 1] range and converted to undirected force directed graphs [@Csardi2006]. 
+The graphs were visualized as two-dimensional network plots with the node proximity determined by the $\rho$ value and distance-dependent repulsion, and edge color and width coding for the $\rho$ value and sign [@Briatte2021].
+
+\newpage
+
+# Results
+
+## Characteristic of the study cohorts
+
+Herein, two collectives of uninfected controls and COVID-19 convalescents were analyzed. 
+The SIMMUN cohort, included adult healthy individuals, patients of the University Hospital of Innsbruck and patients of the local psychiatry clinic (Innsbruck, Austria), who underwent routine SARS-CoV-2 PCR screening at the study site between June 2020 and May 2021.
+Out of `r nrow(stigma$data)` individuals enrolled, `r nrow(stigma$analysis_tbl)` SIMMUN participants with a study variable dataset were analyzed (__Figure \@ref(fig:fig-inclusion)__, __Supplementary Table S1__).
+The excluded individuals were characterized by more frequent psychiatric conditions, depression and anxiety signs, elevated mental stress scoring and less frequent SARS-CoV-2 infections as compared with the analyzed participants (__Supplementary Table S2__).
+SARS-CoV-2-positive individuals accounted for `r signif(table(stigma$analysis_tbl$infection)[2]/nrow(stigma$analysis_tbl), 2) * 100`% of the SIMMUN cohort. 
+None of the participants had received anti-SARS-CoV-2 vaccination prior to enrollment. 
+Males represented 
+`r signif(table(stigma$analysis_tbl$sex)[2]/nrow(stigma$analysis_tbl), 2) * 100`% of the cohort and the median age was `r median(cohort$analysis_tbl$stigma$age)` years. 
+The gender and age structure of the SARS-CoV-2-negative and -positive subsets was comparable. 
+Roughly half of participants were overweight or obese and suffered from somatic comorbidities; these figures were similar for SARS-CoV-2-negative and -positive individuals. 
+The psychiatric conditions affected `r signif(table(stigma$analysis_tbl$psych_comorb)[2]/nrow(stigma$analysis_tbl), 2) * 100`% of the SIMMUN collective and were more significantly more common in SARS-CoV-2-negative participants.
+Signs of depression and anxiety, and intensity of mental stress were gauged with the HADS depression, HADS anxiety and PSS-4 tools, respectively [@Zigmond1983; @Cohen1983]. 
+The psychometric tools displayed good-to-excellent internal consistency ($\omega$ = `r cons$omega_stats %>% filter(factor == 'g') %>% .$total %>% range %>% signif(2) %>% paste(collapse = ' - ')`) [@McDonald1999] (__Supplementary Figure S1__).
+Percentages of clinically relevant anxiety signs (HADS $\geq$ 8) [@Bjelland2002; @Zigmond1983] were significantly higher in the uninfected subset. 
+Depression signs tended to be more common in SASR-CoV-2-negative participants; scores of mental stress were comparable in both SARS-CoV-2 strata. 
+As expected, titer of antibodies against the S1/S2 SARS-CoV-2 pathogen (anti-RBD IgG) was significantly higher in the convalescent subset. 
+In 73% of SARS-CoV-2-infected SIMMUN study participants, the infection was mild and ambulatory (__Table \@ref(tab:tab-stigma-cohort)__). 
+In the SIMMUN study, inflammatory markers and metabolites were recorded at a single timepoint. 
+The median SARS-CoV-2 test - sampling interval was 138.5 days (IQR: 119 - 157.25).
+
+The analyzed subset of the genuine INCOV cohort [@Su2020; @Su2022] encompassed participants with complete age information, serotonin- and dopamine-related metabolites and major inflammatory cytokines (__Figure \@ref(fig:fig-inclusion)__, __Supplementary Table S3__).
+The metabolites and cytokines were sampled in uninfected individuals and at three timepoints of SARS-CoV-2 infection: 
+acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['acute']` days), 
+sub-acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['sub-acute']` days) and 
+recovery (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['recovery']` days after diagnosis) (__Supplementary Table S4__). 
+Out of 
+`r full_join(incov$proteome, incov$metabolome, by = 'patient_id') %>% filter(!duplicated(patient_id)) %>% nrow` initially enrolled participants, 
+`r length(unique(incov$complete_ids))` individuals with `r length(incov$complete_ids)` samples were included in the analyzed INCOV subset (__Figure \@ref(fig:fig-inclusion)__). 
+SARS-CoV-2-positive individuals comprised `r signif(table(cohort$analysis_tbl$incov$cov)[2]/nrow(cohort$analysis_tbl$incov), 2) * 100`% of the analyzed INCOV collective. 
+Males constituted 
+`r signif(table(cohort$analysis_tbl$incov$sex)[2]/nrow(cohort$analysis_tbl$incov), 2) * 100`% of participants, the median age was `r signif(median(cohort$analysis_tbl$incov$age), 2)` years. 
+SARS-CoV-2-positive participants were significantly older than the uninfected subset. 
+Shares of overweight or obese individuals tended to be higher in the SARS-COV-2 group. 
+Nearly all (97%) of SARS-CoV-2-positive INCOV individuals were hospitalized due to a moderate-to-critical infection (__Table \@ref(tab:tab-incov-cohort)__).
+
+As compared with the SIMMUN collective, INCOV study participants were characterized by a significantly higher percentages of males, more advanced age, elevated rates of overweight or obesity, higher share of SARS-CoV-2 cases and more severe infection course (__Supplementary Table S5__).
+
+## Inflammation, SARS-CoV-2 infection, age, mental stress and depression influence systemic levels of neutrotransmitter precursors
+
+In an initial search for predictors of neurotransmitter turnover, we modeled plasma TRP, KYN and KYN/TRP as readouts of systemic serotonin [@Sun2020; @Capuron2011], and PHE, TYR and PHE/TYR as markers of dopamine availability [@Meiser2013; @Neurauter2008] in the INCOV collective.
+The candidate explanatory variables were age, sex, body mass, somatic and mental conditions, alcohol and tobacco consumption, SARS-CoV-2 infection recovery, anti-SARS-CoV-2 antibody titer, NEO and NLR as inflammatory markers, as well as mental stress, depression and anxiety signs (__Supplementary Table S1__).
+
+Full multi-parameter linear models were optimized by backwards elimination of non-significant terms [@Ripley2022] (__Supplementary Table S6__). 
+Meaningful models could be established for TRP, KYN, KYN/TRP, TYR and PHE/TYR. 
+The PHE model had a negligible explanatory performance, proved non-significant as compared with the intercept-only null model and was hence not further analyzed. 
+The remaining models were characterized by good reproducibility and proper parameterization as indicated by comparable fit errors in the genuine dataset and cross-validation. 
+The KYN and KYN/TRP ratio models had the best explanatory performance measured by cross-validated $R^2$ of 
+`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn') %>% .$estimate %>% signif(2)` and 
+`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn_trp') %>% .$estimate %>% signif(2)`, respectively. 
+The TRP, TYR and PHE/TYR models could explain between 
+`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[1] * 100`% and 
+`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[2] * 100`% of the cross-validated response variances ($R^2$, __Supplementary Figure S2__).
+
+The inflammatory marker NEO and depression signs were independently associated with reduced plasma TRP. 
+Increased KYN and KYN/TRP were significantly associated with SARS-CoV-2 infection, elevated NEO and age. 
+Additionally, high mental stress was linked to increased KYN/TRP.  
+Low NLR, in turn, was related to increased KYN (__Figure \@ref(fig:fig-modeling)A__). 
+Plasma TYR concentrations were negatively regulated by inflammation gauged by NEO and rose significantly with age. 
+SARS-CoV infection and age were independently linked to lower PHE/TYR (__Figure \@ref(fig:fig-modeling)B__).
+
+In, univariable analysis of the INCOV dataset (__Supplementary Tables S7 - S8__), we could corroborate significant, positive, moderate-to-strong correlations of age and NEO with increased KYN and  higher KYN/TRP (__Supplementary Figure S3__).
+Participant's age was paralleled by elevated circulating TYR and lowered PHE/TYR in univariable correlation analysis (__Supplementary Figure S4__).
+
+Collectively, in the SIMMUN cohort, systemic inflammation reflected by NEO, SARS-CoV-2 infection and age were the key factors, which possibly brake serotonin synthesis by depletion of its precursor TRP via the KYN pathway.
+This mechanism may be further perpetuated by signs of depression and mental stress. 
+Age and SARS-CoV-2 infection are proposed as regulators of the dopamine precursors, PHE and TYR.
+
+## Inflammatory cytokines, SARS-CoV-2 infection course, age and availability of biosynthesis precursors regulate systemic serotonine and dopamine turnover
+
+Next, we validated the effects of inflammation, SARS-CoV-2 infection and age on serotonin and dopamine metabolism in the INCOV cohort [@Su2022]. 
+The outcome parameters in such analyses were plasma levels of serotonin [@Paul-Savoie2011; @Sun2020; @Tateishi2021; @Holck2019], and DA sulfate, a the main circulating dopamine catabolite [@Meiser2013; @Hashizume1987; @Goldstein1999].
+The explanatory variables were age, infection timepoint, major inflammatory cytokines (IL6, IL10, TNG, IFNG), precursors and competitor metabolites of serotonin (TRP, KYN, QUIN), and precursors of dopamine (PHE, TYR) (__Supplementary Table S3__).
+
+Multi-parameter robust regression models could explain 
+`r (incov_neuro$cv_stats %>% filter(data_type == 'cv', response == 'serotonin', statistic == 'rsq') %>% .$estimate %>% signif(2)) * 100`% and 
+`r (incov_neuro$cv_stats %>% filter(data_type == 'cv', response == 'dopamine', statistic == 'rsq') %>% .$estimate %>% signif(2)) * 100`% of cross-validated variance of serotonin and DA sulfate. 
+Comparable model errors in the training dataset and cross-validation indicated good reproducibility and proper parameterization of the models (__Figure \@ref(fig:fig-incov-neuro)A__). 
+In the INCOV collective, 
+`r incov_neuro$inference$serotonin %>% filter(variable != 'Intercept', p_value < 0.05) %>% filter(!duplicated(variable)) %>% nrow` predictors of plasma serotonin were identified. 
+Age and IL6 was associated with significantly lower serotonin levels, whereas TRP and the SARS-CoV-2 recovery timepoint, i.e.
+`r cohort$sampling_time %>% filter(timepoint == 'recovery') %>% .$median %>% signif(2)` were predictors of increased serotonin.
+The SARS-CoV-2 recovery timepoint and TNF predicted higher DA sulfate levels, and IFNG was associated with a significantly decreased DA sulfate (__Figure \@ref(fig:fig-incov-neuro)__, __Supplementary Table S9__).
+
+These findings were confirmed by time course modeling. 
+The peak inflammatory cytokine levels during acute infection were paralleled by a significant decrease in the serotonin precursor TRP and upregulated TRP decay products KYN and QUIN [@Capuron2011]. 
+The inflammatory milieu of acute infection correlated with a significant rise in PHE, suggestive of an inhibited PHE - TYR conversion [@Neurauter2008]. 
+In turn, resolution of systemic inflammation during sub-acute SARS-CoV-2 infection and recovery, was associated with significantly decreased KYN and QUIN, paralleled by elevated TRP and serotonin. 
+This suggests re-routing of the TRP from IDO-mediated decay to the serotonin biosynthesis [@Capuron2011]. 
+Similarly, infection recovery was associated with significantly decreased PHE and elevated DA sulfate suggestive of an efficient PH - TYR conversion, dopamine synthesis and subsequent dopamine sulfonylation [@Meiser2013; @Hashizume1987; @Neurauter2008; @Goldstein1999].
+
+In correlation analysis, plasma serotonin TRP were negatively associated with inflammatory cytokines with at least moderate strength ($\rho$ > 0.2) during acute and sub-acute infection. 
+By contrast, the TRP decay products KYN and QUIN correlated positively with circulating IL6, IL10, TNF and IFNG. 
+Serotonin levels correlated also TRP during acute and sub-acute infection with moderate effect size (acute: $\rho$ = `r (incov_mds$analysis_tbl$serotonin.acute[c("serotonin", "tryptophan"), ] %>% t %>% cor(method = 'spearman'))[2, 1] %>% signif(2)`,
+sub-acute: $\rho$ = `r (incov_mds$analysis_tbl[['serotonin.sub-acute']][c("serotonin", "tryptophan"), ] %>% t %>% cor(method = 'spearman'))[2, 1] %>% signif(2)`). 
+This may indicate that systemic availability of tryptophan may be particularly crucial for serotonin synthesis at the peak of the infection-related inflammation (__Figure \@ref(fig:fig-incov-corr)A__).
+Concerning the dopamine turnover metabolites, moderate-to-strong positive correlations of PHE with inflammatory cytokines and negative association of TYR with inflammatory markers could be observed in acute and sub-acute infection.
+For these timepoints, moderate, negative correlation of IFNG and IL6 with plasma levels of DA sulfate were identified. 
+This regulatory pattern of dopamine-related metabolite regulation during infection fits well into the proposed inhibition of PHE-TYR conversion and dopamine biosynthesis by inflammation [@Neurauter2008].
+
+In sum, the analyses of the longitudinal INCOV collective suggest the SARS-CoV-2-elicited inflammation bolsters the TRP catabolism via the KYN pathway and downregulates the PHE - TYR conversion, which may limit serotonin and dopamine formation during acute infection and early recovery.
+
+\newpage
+
+# Discussion
+
+With two cohorts consisting of uninfected and SARS-CoV-2-positive individuals, we explored the effects of demographic, clinical, psychometric, infection-, inflammation- and SARS-CoV-2-related factors on systemic turnover of serotonin and dopamine.
+Our results underline the pivotal role of SARS-CoV-2-dependent and -independent, e.g. chronic low grade inflammation on systemic availability of serotonin and dopamine. 
+Furthermore, they put forward plasma TRP levels as a one of limiting factors for systemic serotonin biosynthesis during SARS-CoV-2 infection. 
+Mechanistically, we corroborate the stimulatory effect of inflammation on IDO, which catalyzes the decay of TRP to KYN and hence suppresses serotonin synthesis by depletion of its precursor. 
+Advanced age, mental stress and depression were proposed as infection-independent factors contributing to reduced TRP and, consequently, reduced serotonin formation. 
+Our data suggest similar inhibitory effects of inflammation and age on PHE - TYR conversion, which is the initial step of dopamine formation.
+
+_Points to discuss_
+
+* Role of inflammation and SARS-CoV-2 in activity of the TRP -> KYN pathway and its implication for systemic serotonin synthesis
+
+* Role of inflammation and SARS-CoV-2 in the PHE - TYR conversion and subsequent dopamine synthesis
+
+* Effects of mental health-related factors on serotonin and dopamine availability, mutual relationship?
+
+* Systemic turnover of serotonin and dopamine - is it relevant for the central nervous system? Pro and contra!
+
+# Limitations
+
+Our study has limitations. 
+First, circulating serotonin, dopamine or any abundant dopamine metabolites (e.g.
+DA sulfate) [@Meiser2013; @Hashizume1987; @Goldstein1999], i.e.
+more direct readouts of neurotransmitter turnover, were not measured in the SIMMUN cohort.
+Additionally, the metabolite were measured in the SIMMUN cohort at a single time point with a highly variable infection - sampling time interval. 
+This precluded time course analyses like those conducted in the INCOV cohort. 
+For the INCOV collective, KYN/TRP and PHE/TYR could not be computed with normalized metabolome data, and measurements of stress, depression and anxiety were not available. 
+This incompatibility of the datasets precluded development of comprehensive multi-parameter models in one of the cohorts and subsequent direct validation in the other. 
+Second, relevance of circulating serotonin- and dopamine-related metabolites for the neurotransmitter metabolism in the central nervous system and mental health disorders is controversial, as discussed above.
+Third, the SIMMUN cohort suffered from a selection bias due to enrichment in psychiatric and hospital patients with a high rate of mental and somatic conditions. 
+Hence, the SIMMUN collective is not representative for the entire pandemic population. 
+Fourth, both the SIMMUN and INCOV cohorts were recruited during initial phases of the pandemic and do not include any (SIMMUN) or systematically vaccinated patients (INCOV). 
+Similarly, the analyzed cohorts were exposed to wild-type-like SARS-CoV-2 variants and do not allow to assess effects of repeated, seasonal infections with highly transmissible but far less virulent omicron pathogens.
+For these reasons, studies with recent, real world post-pandemic collectives are urgently needed to validate our findings. 
+Finally, the cross-sectional SIMMUN and INCOV cohorts encompassed uninfected controls and SARS-CoV-2 infections ranging from asymptomatic to critical disease. 
+In particular, the shares of pathogen-positive individuals and the infection severity differed significantly between the analyzed collectives. 
+The more severe infection in INCOV study participants can explain stronger effects of inflammation and infection recovery on the metabolites of interest that those observed in the SIMMUN collective.
+
+# Conclusions
+
+SARS-CoV-2-dependent and -independent inflammation can lower systemic availability of serotonin and dopamine by depletion of the tryptophan via the competitive kynurenine pathway and inhibition of the phenylalanine - tyrosine suppression, respectively.
+Those effects can be further amplified by advanced age, mental stress and depression.  
+It remains to be investigated, if and how this mechanism may contribute to neurotransmitter metabolism in the central nervous system and, consequently, to mental disorders following SARS-CoV-2 infection.
+
+\newpage
+
+# Acknowledgements
+
+We thank all participants and patients for the participation in the study.
+
+# Funding
+
+The study was supported by the Science Fund of the Land Tirol (grant number GZ71134 to Katharina Hüfner).
+
+# Data and code availability
 
@@
 
-```{r fig-uni-trp, fig.width = figur::convert(figs$uni_trp, to = 'in')$w, fig.height = figur::convert(figs$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}
-
-figs$uni_trp$plot
-
-```
-
-__Figure \@ref(fig:fig-uni-trp).
-Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__
-
-_Age, systemic inflammatory markers: neopterin (NEO), neutrophil - lymphocyte ratio (NLR), SARS-CoV-2 infection status, depression signs (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of systemic levels of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP).
-Their association with serum concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
-
-_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method.
-(A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
-(B) Scatter plots of serum concentrations of NEO, TRP, KYN and KYN/TRP.
-Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
-Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._
-
-_(C) Levels of the metabolite levels were compared between infected and uninfected participants, and between participants with and without depression signs by two-tailed T test with Cohen's d effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points.
-Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the strata are indicated in the X axes._
-
-\newpage
-
-```{r fig-uni-tyr, fig.width = figur::convert(figs$uni_tyr, to = 'in')$w, fig.height = figur::convert(figs$uni_tyr, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.'}
-
-figs$uni_tyr$plot
-
-```
-
-__Figure \@ref(fig:fig-uni-tyr). Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.__
-
-_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of systemic levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR).
-Their association with serum concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
-
-_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method.
-(A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
-(B) Scatter plots of patient's age, PHE, TYR and PHE/TYR.
-Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
-Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._
-
-_(C) Levels of the metabolite levels were compared between infected and uninfected participants by two-tailed T test with Cohen's d effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points.
-Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the strata are indicated in the X axes._
-
-\newpage
-
-```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of serum levels of serotonin and dopamine sulfate in the INCOV cohort.'}
+```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine sulfate in the INCOV cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of serum levels of serotonin and dopamine sulfate in the INCOV cohort.__
+__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine sulfate in the INCOV cohort.__
 
-_Effects of patient's age, timepoint after SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection as compared with healthy controls), serum levels of metabolites related to neurotransmitter biosynthesis and competitor pathways (tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE] and tyrosine [TYR]), and serum concentrations of inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on serum concentrations of serotonin (5-hydroxy tryptophan [5-HT]) and dopamine sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
+_Effects of patient's age, timepoint after SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection as compared with healthy controls), plasma levels of metabolites related to neurotransmitter biosynthesis and competitor pathways (tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE] and tyrosine [TYR]), and plasma concentrations of inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin (5-hydroxy tryptamine [5-HT]) and dopamine sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
 
@@
 
-_Time courses of normalized log~2~-transformed serum concentrations of inflammatory cytokines (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF] and interferon gamma [INFG]), metabolites implicated in serotonin synthesis (B, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN] and serotonin/5-hydroxy tryptophan [5-HT]) and metabolites implicated in dopamine turnover (C, phenylalanine [PHE], tyrosine [TYR] and dopamine sulfate [DA sulfate]).
+_Time courses of normalized log~2~-transformed plasma concentrations of inflammatory cytokines (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF] and interferon gamma [INFG]), metabolites implicated in serotonin synthesis (B, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN] and serotonin/5-hydroxy tryptamine [5-HT]) and metabolites implicated in dopamine turnover (C, phenylalanine [PHE], tyrosine [TYR] and dopamine sulfate [DA sulfate]).
 Differences between uninfected controls and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after infection, respectively) were modeled by robust linear regression with the uninfected controls as baseline.
@@
 __Figure \@ref(fig:fig-incov-corr).
-Correlation of serum levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.__
+Correlation of plasma levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.__
 
-_Serum levels of metabolites implicated in serotonin synthesis (A, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin [5-hydroxy tryptophan, 5-HT]), dopamine turnover (B, phenylalanine [PHE], tyrosine [TYR], dopamine sulfate [DA sulfate]) and inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in uninfected and SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection).
+_plasma levels of metabolites implicated in serotonin synthesis (A, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin [5-hydroxy tryptamine, 5-HT]), dopamine turnover (B, phenylalanine [PHE], tyrosine [TYR], dopamine sulfate [DA sulfate]) and inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in uninfected and SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection).
 Correlation coefficient matrices for at least moderate effects ($\rho$ > 0.2) were visualized as force-directed network plots.

### 2023-05-18 paper/markdown/supplementary_material.Rmd
Results, new figure set

@@
 ---
-title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
+title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic levels of aminoacid neurotransmitter precursors"
 subtitle: "Supplementary Material"
@@
 Proteome and metabolome data in form of normalized, age- and sex-adjusted, log~2~-transformed plasma levels as well as clinical information (sex, SARS-CoV-2 infection status, COVID-19 severity, timepoint) for the INCOV cohort were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) accompanying the report by Su and colleagues [@Su2022] (function `read_xlsx`, package _readxl_) [@Wickham2022].
-In the current analysis, `r nrow(incov$analysis_tbl)` INCOV study participants with the complete dataset of metabolites of interest (serotonin [5-HT], tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE], tyrosine [TYR] and dopamine 3-O-sulfate [DA sulfate]), inflammatory cytokines of interest (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and age at enrollment were included in the current analysis.
+In the current analysis, `r nrow(incov$analysis_tbl)` INCOV study samples obtained for 
+`r length(unique(incov$complete_ids))` individuals with the complete dataset of metabolites of interest (serotonin, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE], tyrosine [TYR] and dopamine 3-O-sulfate [DA sulfate]), inflammatory cytokines of interest (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and complete information of age at enrollment were included.
 The INCOV study variables are listed in __Supplementary Table S\@ref(tab:tab-incov-study-vars)__. 
@@
 Concerning the INCOV study design, plasma metabolites and proteome were determined in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['acute']` days), 
-sub-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['sub-acute']`) and 
-recovery (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['recovery']` after diagnosis). 
+acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['acute']` days), 
+sub-acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['sub-acute']` days) and 
+recovery (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['recovery']` days after diagnosis). 
 The measurements were only partially matched by participant (i.e. multiple longitudinal measurements for an individual) hence precluding a classical repeated measurement analysis approach. 
@@
 
-## Descriptive statistic and variable distribution
+## Descriptive statistic, variable distribution and psychometric tool consistency
 
@@
 For this reason, non-parametric Spearman rank test was chosen for correlation analysis and robust linear regression was used as a modeling approach.
+
+The tau-equivalence assumption was clearly violated for both the HADS depression and HADS anxiety scale in the SIMMUN collective as investigated by factor analysis (function `reduce_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), employing internally `factanal()` from the _stats_ package).
+For this reason, global McDonald's $\omega$ statistic was used as a consistency measure for all psychometric tools in the SIMMUN cohort (function `omega()`, package _psych_) [@McDonald1999; @Bartlett1937; @Revelle2022] (__Supplementary Figure S\@ref(fig:fig-consistency)__).
 
@@
 Correlation of metabolite concentrations with age, stress scoring (PSS-4) and inflammation measures (NEO and NLR) in the SIMMUN cohort was done by Pearson's test (__Supplementary Table S\@ref(tab:tab-univariate-correlation)__).
-Differences in metabolite concentrations and metabolite ratios in the SIMMUN cohort stratified by the SARS-CoV-2 infection status or presence of depression symptoms (HADS) were assessed by tow-tailed T test with Cohen's d effect size metric (__Supplementary Table S\@ref(tab:tab-univariate-complarison)__).
+Differences in metabolite concentrations and metabolite ratios in the SIMMUN cohort stratified by the SARS-CoV-2 infection status or presence of depression symptoms (HADS) were assessed by two-tailed T test with Cohen's d effect size metric (__Supplementary Table S\@ref(tab:tab-univariate-complarison)__).
 
@@
 Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on systemic neurotransmitter availability were subsequently tested in the publicly available INCOV collective [@Su2022].
-To this end we modeled normalized plasma levels of serotonin (5-HT) as the end product of serotonin biosynthesis associated with mental disorders [@Holck2019; @Tateishi2021; @Sun2020; @Kheirouri2016; @Paul-Savoie2011; @Ertugrul2007], and the major product of systemic dopamine decay dopamine 3-O-sulfate (DA-sulfate) [@Goldstein1999; @Meiser2013].
-The explanatory variables were age, plasma levels of inflammatory cytokines (IL6, IL10, TNF, IFNG), plasma concentrations of neurotransmitter precursors and competitor pathway products (for 5-HT: TRP, KYN and quinolinate [QUIN], for DA sulfate: PHE and TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control).
+To this end we modeled normalized plasma levels of serotonin as the end product of serotonin biosynthesis associated with mental disorders [@Holck2019; @Tateishi2021; @Sun2020; @Kheirouri2016; @Paul-Savoie2011; @Ertugrul2007], and the major product of systemic dopamine decay dopamine 3-O-sulfate (DA-sulfate) [@Goldstein1999; @Meiser2013].
+The explanatory variables were age, plasma levels of inflammatory cytokines (IL6, IL10, TNF, IFNG), plasma concentrations of neurotransmitter precursors and competitor pathway products (for serotonin: TRP, KYN and quinolinate [QUIN], for DA sulfate: PHE and TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control).
 Since most numeric features were non-normally distributed, MM algorithm robust linear regression with Huber's psi function was chosen [@Huber2011; @Ripley2022] as a modeling approach.
@@
 Significance of the model estimates ($p(\beta\neq0)$) was determined by two-tailed T test, 95% confidence intervals were computed based on theoretical t distribution. 
-The robust linear modeling results are presented in __Figure 5__ and __Supplementary Table S\@ref(tab:tab-modeling-incov)__.
+The robust linear modeling results are presented in __Figure 3__ and __Supplementary Table S\@ref(tab:tab-modeling-incov)__.
 
@@
 
-Differences in normalized blood concentrations of inflammatory cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset serving as baseline.
+Differences in normalized blood concentrations of inflammatory cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset or acute infection serving as baselines.
 Models were constructed with the `make_lm()` function (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) employing internally the MM`rlm()` algorithm with Huber's psi function (package _MASS_) [@Huber2011; @Ripley2022].
 Inference statistics were computed as described for multi-parameter robust linear modeling. 
-The time course modeling results are presented in __Figure 6__ and __Supplementary Table S\@ref(tab:tab-time-modeling)__.
+The time course modeling results are presented in __Figure 4__, __Supplementary Table S\@ref(tab:tab-time-uninfected)__ - __S\@ref(tab:tab-time-acute)__.
 
@@
 
-Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of inflammatory cytokines (IL6, IL10, TNF and IFNG), and metabolite involved in serotonin (5-HT, TRP, KYN and QUIN) and dopamine turnover (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
+Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of inflammatory cytokines (IL6, IL10, TNF and IFNG), and metabolite involved in serotonin (serotonin, TRP, KYN and QUIN) and dopamine turnover (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
 The correlation matrices were subsequently scaled into the [0, 1] range with the $0.5\times(x + 1)$ function and converted to undirected force directed graphs with the `graph_from_adjacency_matrix()` function from the _igraph_ package [@Csardi2006].
-The graphs were visualized with tools provided by the _ggnetwork_ package [@Briatte2021] as two-dimensional network plots with the node proximity determined by the $\rho$ value and random repulsion, and edge color and width coding for the $\rho$ value and sign (__Figure 7__).
+The graphs were visualized with tools provided by the _ggnetwork_ package [@Briatte2021] as two-dimensional network plots with the node proximity determined by the $\rho$ value and distance-dependent repulsion, and edge color and width coding for the $\rho$ value and sign (__Figure 5__).
 
@@
   width(1:3, width = 3.5, unit = 'cm') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-cohort-comp, tab.cap = 'Comparison of the SIMMUN and INCOV cohorts.
+Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+flextable::flextable(suppl_tabs$cohort_comp) %>% 
+  width(1, width = 3.9, unit = 'cm') %>% 
+  width(2:3, width = 4, unit = 'cm') %>% 
+  width(4, width = 2, unit = 'cm') %>% 
+  width(5, width = 2.7, unit = 'cm') %>% 
+  width(6, width = 2.5, unit = 'cm') %>% 
+  footnote(4, 1, 
+           value = as_paragraph('overweight: body mass index (BMI) 25 - 30 kg/mm\u00B2, obesity: BMI > 30 kg/mm\u00B2'), 
+           part = 'body', 
+           ref_symbols = 'a') %>% 
   theme_vanilla
@@
   footnote(1, 1, 
-           value = as_paragraph('5-HT: 5-hydroxy tryptophan/serotonin; DA sulfate: dopamine 3-O-sulfate.'), 
+           value = as_paragraph('5-HT: 5-hydroxy tryptamine/serotonin; DA sulfate: dopamine 3-O-sulfate.'), 
            part = 'header', 
@@
 
-```{r tab-time-modeling, tab.cap = 'Results of robust linear modeling of plasma levels of inflammatory cytokines, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.'}
+```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokines, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.
+The uninfected subset served as a baseline.'}
 
-flextable::flextable(suppl_tabs$time_modeling) %>% 
+flextable::flextable(suppl_tabs$time_uninfected) %>% 
   width(1, width = 3, unit = 'cm') %>% 
-  width(2, width = 3.7, unit = 'cm') %>% 
+  width(2, width = 3, unit = 'cm') %>% 
   width(3, width = 3.7, unit = 'cm') %>% 
-  width(4, width = 2.9, unit = 'cm') %>% 
+  width(4, width = 3.7, unit = 'cm') %>% 
+  width(5, width = 2.9, unit = 'cm') %>% 
   merge_v(1) %>% 
-  footnote(1, 1, 
+  footnote(1, 2, 
+           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinate; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 4, 
+           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokines, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.
+The acute SARS-CoV-2 infection subset served as a baseline.'}
+
+flextable::flextable(suppl_tabs$time_acute) %>% 
+  width(1, width = 3, unit = 'cm') %>% 
+  width(2, width = 3, unit = 'cm') %>% 
+  width(3, width = 3.7, unit = 'cm') %>% 
+  width(4, width = 3.7, unit = 'cm') %>% 
+  width(5, width = 2.9, unit = 'cm') %>% 
+  merge_v(1) %>% 
+  footnote(1, 2, 
            value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinate; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'),
@@
 
-```{r fig-corr-trp, fig.width = figur::convert(suppl$corr_trp, to = 'in')$w, fig.height = figur::convert(suppl$corr_trp, to = 'in')$h, fig.cap = 'Correlation of tryptophan and its metabolites with age, plasma levels of neopterin, neutrophil - lymphocyte ratio, and mental stress scoring in the SIMMUN cohort.'}
+```{r fig-uni-trp, fig.width = figur::convert(suppl$uni_trp, to = 'in')$w, fig.height = figur::convert(suppl$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}
 
-suppl$corr_trp$plot
+suppl$uni_trp$plot
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-corr-trp).
-Correlation of tryptophan and its metabolites with age, plasma levels of neopterin, neutrophil - lymphocyte ratio, and mental stress scoring in the SIMMUN cohort.__
+__Supplementary Figure S\@ref(fig:fig-uni-trp).
+Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__
 
-_Correlation of age (A), the inflammatory marker neutrophil-leukocyte ratio (B, NLR), and stress scoring (C, perceived stress scale, 4-item [PSS-4]) with plasma concentrations of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP) was assessed by Pearson's test corrected for multiple testing with the false discovery rate method (n = `r nrow(stigma$analysis_tbl)`.
-Single observations are depicted as points, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
-Values of correlation coefficients (r) with 95% confidence intervals and p values are displayed in the plot captions._
+_Age, systemic inflammatory markers: neopterin (NEO), neutrophil - lymphocyte ratio (NLR), SARS-CoV-2 infection status, depression signs (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of systemic levels of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP).
+Their association with plasma concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
+
+_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method.
+(A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
+(B) Scatter plots of plasma concentrations of NEO, TRP, KYN and KYN/TRP.
+Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
+Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._
+
+_(C) Levels of the metabolite levels were compared between infected and uninfected participants, and between participants with and without depression signs by two-tailed T test with Cohen's d effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of observations in the strata are indicated in the X axes._
 
@@
 
-```{r fig-corr-tyr, fig.width = figur::convert(suppl$corr_tyr, to = 'in')$w, fig.height = figur::convert(suppl$corr_tyr, to = 'in')$h, fig.cap = 'Correlation of phenylalanine and tyrosine with plasma levels of neopterin in the SIMMUN cohort.'}
+```{r fig-uni-tyr, fig.width = figur::convert(suppl$uni_tyr, to = 'in')$w, fig.height = figur::convert(suppl$uni_tyr, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.'}
 
-suppl$corr_tyr$plot
+suppl$uni_tyr$plot
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-corr-tyr). Correlation of phenylalanine and tyrosine with plasma levels of neopterin in the SIMMUN cohort.__
+__Supplementary Figure S\@ref(fig:fig-uni-tyr).
+Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.__
 
-_Correlation of the inflammatory marker neopterin (NEO) with plasma concentrations of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR) was was assessed by Pearson's test corrected for multiple testing with the false discovery rate method (n = `r nrow(stigma$analysis_tbl)`.
-Single observations are depicted as points, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
-Values of correlation coefficients (r) with 95% confidence intervals and p values are displayed in the plot captions._
+_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of systemic levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR).
+Their association with plasma concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
+
+_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method.
+(A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
+(B) Scatter plots of patient's age, PHE, TYR and PHE/TYR.
+Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
+Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._
+
+_(C) Levels of the metabolite levels were compared between infected and uninfected participants by two-tailed T test with Cohen's d effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of observations in the strata are indicated in the X axes._
 

### 2023-05-18 exploration scripts/cohort_comparison.R

@@ -1,0 +1,105 @@
+# Comparison of the SIMMUN and INCOV cohort
+
+  insert_head()
+
+# container ------
+
+  comp <- list()
+
+# analysis globals ----
+
+  insert_msg('Analysis globals')
+
+  ## common variables
+
+  comp$lexicon <-
+    c('sex' = 'Sex',
+      'age' = 'Age, years',
+      'bmi_class' = 'Body mass',
+      'cov' = 'SARS-CoV-2',
+      'severity' = 'SARS-CoV-2 hospitalization') %>%
+    compress(names_to = 'variable',
+             values_to = 'label')
+
+  ## analysis tables
+
+  comp$analysis_tbl <-
+    list(SIMMUN = stigma$data %>%
+           filter(patient_id %in% stigma$complete_ids),
+         INCOV = incov$clinic %>%
+           filter(patient_id %in% incov$complete_ids) %>%
+           mutate(severity = who_severity)) %>%
+    map(filter, !duplicated(patient_id)) %>%
+    map(select, all_of(comp$lexicon$variable))
+
+  comp$analysis_tbl$INCOV <-
+    comp$analysis_tbl$INCOV %>%
+    mutate(severity = ifelse(cov == 'healthy',
+                             'healthy',
+                             ifelse(stri_detect(severity,
+                                                regex = '(1|2)$'),
+                                    'ambulatory', 'hospitalized')),
+           severity = factor(severity, c('healthy', 'ambulatory', 'hospitalized')))
+
+  comp$analysis_tbl <- comp$analysis_tbl %>%
+    compress(names_to = 'cohort') %>%
+    mutate(cohort = factor(cohort, c('SIMMUN', 'INCOV')))
+
+  ## test type
+
+  comp$lexicon$test_type <-
+    comp$analysis_tbl[comp$lexicon$variable] %>%
+    map_lgl(is.numeric) %>%
+    ifelse('wilcoxon_r', 'cramer_v')
+
+# Descriptive stats ----
+
+  insert_msg('Descriptive stats')
+
+  comp$stats <- comp$analysis_tbl %>%
+    explore(variables = comp$lexicon$variable,
+            split_factor = 'cohort',
+            what = 'table',
+            pub_styled = TRUE) %>%
+    reduce(left_join, by = 'variable') %>%
+    set_names(c('variable', levels(comp$analysis_tbl$cohort)))
+
+# Testing ------
+
+  insert_msg('Testing')
+
[…]

### 2023-05-18 exploration.R

@@ -30,6 +30,7 @@
   insert_msg('Analysis scripts')
 
   c('./exploration scripts/cohort.R',
+    './exploration scripts/cohort_comparison.R',
     './exploration scripts/distribution.R',
     './exploration scripts/excluded.R',
     './exploration scripts/consistency.R',

### 2023-05-18 incov scripts/time_modeling.R

@@ -161,42 +161,60 @@
   insert_msg('Forest plots')
 
   time_rlm$forest_plots <-
-    list(vars = list(inflammatory = c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF'),
-                     trp = c('tryptophan', 'kynurenine', 'quinolinate', 'serotonin'),
-                     tyr = c('phenylalanine', 'tyrosine', 'dopamine.3.O.sulfate')),
-         plot_title = paste(c('Inflammatory cytokines',
-                              'TRP decay and 5-HT',
-                              'PHE/TYR and DA'))) %>%
-    pmap(plot_tc_beta,
-         data = time_rlm$inference$uninfected,
-         plot_subtitle = 'Robust linear modeling',
-         hide_baseline = FALSE,
-         show_connector = FALSE) %>%
-    map2(., time_rlm$plot_colors ,
-         ~.x +
-           scale_color_manual(values = .y,
-                              labels = function(x) exchange(x, dict = time_rlm$lexicon)) +
-           scale_fill_manual(values = .y,
-                             labels = function(x) exchange(x, dict = time_rlm$lexicon)))
+    list(x = time_rlm$inference,
+         y = paste('Robust linear modeling,',
+                   c('uninfected', 'acute')),
+         z = c('uninfected', 'acute')) %>%
+    pmap(function(x, y, z) list(vars = list(inflammatory = c('IL6_INF',
+                                                             'IL10_INF',
+                                                             'TNF_INF',
+                                                             'IFNG_INF'),
+                                            trp = c('tryptophan',
+                                                    'kynurenine',
+                                                    'quinolinate',
+                                                    'serotonin'),
+                                            tyr = c('phenylalanine',
+                                                    'tyrosine',
+                                                    'dopamine.3.O.sulfate')),
+                                plot_title = paste(c('Inflammatory cytokines',
+                                                     'TRP decay and 5-HT',
+                                                     'PHE/TYR and DA'))) %>%
+           pmap(plot_tc_beta,
+                data = x,
+                plot_subtitle = y,
+                baseline_lab = z,
+                hide_baseline = FALSE,
+                show_connector = FALSE) %>%
+           map2(., time_rlm$plot_colors,
+                ~.x +
+                  scale_color_manual(values = .y,
+                                     labels = function(x) exchange(x,
[…]

### 2023-05-18 paper scripts/figures.R

@@ -18,7 +18,7 @@
               caption = paste('Inclusion scheme for the SIMMUN',
                               'and INCOV cohorts, and analysis strategy.'),
               w = 180,
-              h = 180 * 2832/4044)
+              h = 180 * 2946/4098)
 
 # Figure 2: linear modeling -----
 
@@ -57,123 +57,9 @@
               w = 180,
               h = 180)
 
-# Figure 3: univariate analysis for TRP/KYN ------
+# Figure 3: multi-parameter modeling of neurotransmitters in the INCOV cohort -----
 
-  insert_msg('Figure 3: Univariate analysis for TRP/KYN')
-
-  ## upper panel: correlogram
-
-  figs$uni_trp$upper <-
-    plot_grid(tst_trp$correlation$correlogram +
-                theme(plot.subtitle = element_text(size = 7.4),
-                      legend.position = 'none') +
-                labs(subtitle = paste('Pearson correlation, n =',
-                                      nrow(stigma$analysis_tbl))),
-              get_legend(tst_trp$correlation$correlogram),
-              ncol = 2,
-              rel_widths = c(0.65, 0.35),
-              align = 'hv',
-              axis = 'tblr')
-
-  ## middle panel: correlations for neopterin
-
-  figs$uni_trp$middle <-
-    tst_trp$correlation$plots[c("log_neo|trp",
-                                "log_neo|log_kyn",
-                                "log_neo|log_kyn_trp")] %>%
-    map(~.x + theme(plot.subtitle = element_text(size = 7.4))) %>%
-    plot_grid(plotlist = .,
-              ncol = 3,
-              align = 'hv',
-              axis = 'tblr')
-
-  ## bottom panel: comparisons
-
-  figs$uni_trp$bottom <- tst_trp$comparison$plots %>%
-    map(~.x + theme(plot.subtitle = element_text(size = 7))) %>%
-    plot_grid(plotlist = .,
-              ncol = 3,
-              align = 'hv',
-              axis = 'tblr')
-
-  figs$uni_trp <- plot_grid(figs$uni_trp$upper,
-                            figs$uni_trp$middle,
-                            figs$uni_trp$bottom,
-                            nrow = 3,
-                            rel_heights = c(0.75, 1, 2),
-                            labels = LETTERS,
-                            label_size = 10) %>%
-    as_figure(label = 'figure_3_univariate_analysis_trp',
-              ref_name = 'uni_trp',
-              caption = paste('Effects of age, serum inflammatory markers',
[…]

### 2023-05-18 paper scripts/supplement.R

@@ -57,55 +57,119 @@
               w = 180,
               h = 110)
 
-# Figure S3: scatter plots for TRP, KYN and KYN/TRP -----
+# Figure S3: univariate analysis for TRP/KYN ------
 
-  insert_msg('Figure S3: scatter plots for TRP, KYN and KYN/TRP')
+  insert_msg('Figure S3: Univariate analysis for TRP/KYN')
 
-  suppl$corr_trp <-
-    tst_trp$correlation$plots[c("age|trp", "age|log_kyn", "age|log_kyn_trp",
-                                "log_nlr|trp", "log_nlr|log_kyn",
-                                "log_nlr|log_kyn_trp", "pss_stress_score|trp",
-                                "pss_stress_score|log_kyn",
-                                "pss_stress_score|log_kyn_trp")] %>%
+  ## upper panel: correlogram
+
+  suppl$uni_trp$upper <-
+    plot_grid(tst_trp$correlation$correlogram +
+                theme(plot.subtitle = element_text(size = 7.4),
+                      legend.position = 'none') +
+                labs(subtitle = paste('Pearson correlation, n =',
+                                      nrow(stigma$analysis_tbl))),
+              get_legend(tst_trp$correlation$correlogram),
+              ncol = 2,
+              rel_widths = c(0.65, 0.35),
+              align = 'hv',
+              axis = 'tblr')
+
+  ## middle panel: correlations for neopterin
+
+  suppl$uni_trp$middle <-
+    tst_trp$correlation$plots[c("log_neo|trp",
+                                "log_neo|log_kyn",
+                                "log_neo|log_kyn_trp")] %>%
     map(~.x + theme(plot.subtitle = element_text(size = 7.4))) %>%
     plot_grid(plotlist = .,
               ncol = 3,
               align = 'hv',
-              axis = 'tblr',
-              labels = c('A', '', '',
-                         'B', '', '',
-                         'C', '', ''),
-              label_size = 10) %>%
-    as_figure(label = 'figure_s3_trp_system_correlation',
-              ref_name = 'corr_trp',
-              caption = paste('Correlation of tryptophan and',
-                              'its metabolites with age, serum levels',
-                              'of neopterin, neutrophil - lymphocyte ratio,',
-                              'and mental stress scoring in the SIMMUN',
-                              'cohort.'),
+              axis = 'tblr')
+
+  ## bottom panel: comparisons
+
+  suppl$uni_trp$bottom <- tst_trp$comparison$plots %>%
[…]

### 2023-05-18 paper scripts/tables.R

@@ -140,9 +140,22 @@
             caption = paste('Number of available samples and sampling',
                             'timepoints in the INCOV cohort.'))
 
-# Table S5: modeling, SIMMUN ------
+# Table S5: comparison of the cohorts -------
 
-  insert_msg('Table S5: modeling SIMMUN')
+  insert_msg('Table S5: comparison of the cohorts')
+
+  suppl_tabs$cohort_comp <- comp$result_tbl %>%
+    mdtable(label = 'table_s5_cohort_comparison',
+            ref_name = 'cohort_comp',
+            caption = paste('Comparison of the SIMMUN and INCOV cohorts.',
+                            'Numeric variables are presented as medians with',
+                            'interquartile ranges (IQR) and ranges.',
+                            'Categorical variables are presented as percentages',
+                            'and counts within the complete observation set.'))
+
+# Table S6: modeling, SIMMUN ------
+
+  insert_msg('Table S6: modeling SIMMUN')
 
   suppl_tabs$modeling_simmun <- mod_eli$inference %>%
     map_dfr(transmute,
@@ -167,15 +180,15 @@
                             'and phenylalanine - tyrosine ratios in the',
                             'SIMMUN cohort.'))
 
-# Table S6 - S7: univariate analysis ------
+# Table S7 - S8: univariate analysis ------
 
-  insert_msg('Tables S6 - S7, univariable analysis')
+  insert_msg('Tables S7 - S8, univariable analysis')
 
   suppl_tabs[c('univariate_correlation',
                'univariate_complarison')] <-
     list(x = tst_results,
-         label = c('table_s6_uni_correlation',
-                   'table_s7_uni_comparison'),
+         label = c('table_s7_uni_correlation',
+                   'table_s8_uni_comparison'),
          ref_name = c('univariate_correlation',
                       'univariate_complarison'),
          caption = c(paste('Correlation of age, stress scoring',
@@ -188,9 +201,9 @@
                            'and SARS-CoV-2 infection status.'))) %>%
     pmap(mdtable)
 
-# Table S8: modeling, INCOV ------
+# Table S9: modeling, INCOV ------
 
-  insert_msg('Table S8: robust linear modeling in the INCOV cohort')
+  insert_msg('Table S9: robust linear modeling in the INCOV cohort')
 
   suppl_tabs$modeling_incov <- incov_neuro$inference %>%
     map_dfr(transmute,
@@ -208,25 +221,38 @@
[…]

### 2023-05-23 paper/markdown/rebuttal_letter.Rmd
Manuscript and Rebuttal Letter

---
title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism"
subtitle: "Rebuttal Letter"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Editor

<span custom-style = 'reviewer'>Thank you for the timely and well written piece. 
I have received three reviews from exceptional scholars in the field. 
As you can see, there are many strengths to the paper that they note and the publication is timely. 
However, there are also several concerns raised and I am not entirely sure they can all be sufficiently addressed to warrant publication. 
However, I think it is worth a shot. 
Please be attentive to reviewer concerns for a resubmission.</span>

Thank you for appreciation of our manuscript and giving us a chance to address Reviewers' issues. 
Rewrote the paper and put efforts to incorporate the critic in the text. 
In particular, we defined clearly the analysis endpoints, and decided to present our story in a 'top - bottom' approach. 
In brief, we employed the most robust and comprehensive analytic technique, multi-parameter modeling, to identify the most relevant explanatory factors affecting systemic serotonin and dopamine metabolism. 
Subsequently, we explore such factors in more detail by correlation and network analysis, and canonical statistical tests. 
Finally, we improved quality of language and Discussion. 
Please see below for details. 
We hope that, our revised manuscript is now suitable for publication in _Psychoneuroendocrinology_.

\newpage

# Reviewer 1

## Issue 1

<span custom-style = 'reviewer'>The study assesses the relationship between psychological variables, inflammatory markers, and concentrations of aminoacids in individuals after SARS-CoV-2 infection compared to uninfected controls in two distinct cohorts (SIMMUN; INCOV). 
Moreover, the difference between the control group and different states after SARS-CoV-2 infection (acute, sub-acute, recovery) are compared to the control group within the longitudinal INCOV study cohort.
In addition, the study aims to test the hypothesis that Phenylalanine, Tyrosine, Tryptophan, Kynurenine, and their respective ratios, are predicted by psychological variables inflammation, and SARS-CoV-2 infection-related variables (infection yes vs no, Anti-Antibody titers).
First, it should be mentioned that the graphs are of very high quality and aesthetics and summarize the results very well. 
In addition, the use of a very sophisticated analysis (multiparametric linear modeling) and the detailed presentation of this analysis, and the R packages used are highly creditable. 
However, several weaknesses, including, biochemically wrong assumptions regarding the assessed parameters, lack of transparency regarding the pre-processing of the data, and measurement comparability between the assessed cohorts limit its significance and impact. 
The quality of the paper also suffers from a lack of references, scientifically inadequate language, and multiple spelling and grammatical errors. Because of the many issues that arise, I would not recommend the manuscript for publication in _Psychoneuroendocrinology_.</span>

We thank the Reviewer for the careful lecture of our manuscript and valuable feedback. 
In the revised manuscript, we did our best to address the weaknesses pointed out be the Reviewer.
In order to improve the clarity of our research aims and to facilitate interpretation of the results, we re-formatted the analysis pipeline and the manuscript in the following way:

* we streamlined the __Introduction__ section. Here, we describe briefly the serotonin and dopamine metabolism, and the competitor kynurenine pathway, functions of those neurotransmitter in the central nervous system and impact on serotonin, dopamine on mental health. We outline the role of inflammation and mutual relationship between tryptophan metabolism and mental health. Furthermore, we review literature evidence for the SARS-CoV-2/COVID-19 - neurotransmitter connex.

* In __Introduction__, we also define clearly the 'evidence gap' and our research aims in the last paragraph of the section.

* In  __Methods/Analysis endpoints__ we define unequivocally the outcomes, i.e. plasma levels of metabolites involved in serotonin and dopamine turnover, and explanatory factors, i.e. demographic, clinical, inflammatory, psychometric and SARS-CoV-2-associated variables.

* We restrict the analyzed SIMMUN and INCOV cohorts to participants with the complete set of outcome and explanatory variables (__Figure 1__, __Supplementary Table S1__ and __S3__). By this change, results of particular analyses can be more easily compared with each other. Additionally, to investigate if and how exclusion of participant affected the SIMMUN cohort characteristic and if it could be regarded as a source of bias, we compared th excluded and analyzed individuals as presented in __Supplementary Table S2__ and __Results/Characteristic of the study cohorts__.

* We excluded CRP and IL6 from analyses presented in the revised manuscript. Those markers were found to be slightly above their reference ranges (0.5 mg/dL and 7 pg/mL) solely in 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of SIMMUN individuals, respectively. They are hence unlikely to play a clinically relevant role in regulation of TRP, KYN, KYN/TRP, PHE, TYR and PHE/TYR, in spite of statistically significant correlations presented in the initial manuscript. We argument this decision in __Supplementary Methods__ of the revised manuscript.

* We provided information on all analyzed variables in the SIMMUN (__Supplementary Table S1__) and INCOV cohort (__Supplementary Table S3__) with their format, units and, if applicable, normality-stabilizing transformations.

* In order to improve clarity of our results, we decided to present our results following a 'top - bottom' scheme. In brief, both in the SIMMUN and the INCOV cohort, we employ multi-parameter modeling (linear regression or robust linear regression), to identify key explanatory factors significantly affecting the outcome. Subsequently, we investigate those significant predictions in more detail by 'univariable' correlation analysis and statistical hypothesis testing. This analysis strategy was pursued in our recent papers as well [@Hufner2022; @Hufner2023].

## Issue 2

<span custom-style = 'reviewer'>__Introduction:__ In the introduction the term "psychoneuroimmunological mechanisms" is mentioned twice. 
I would appreciate it if at least some of the mechanisms which are relevant to the paper are specified.</span>

As outlined in __Issue 1__, we re-wrote thoroughly the __Introduction__ section. 
Following discussions in the core author team, we decided no to replace the general and a bit obscure term 'psychoneuroimmunological' with more direct formulations referring to the prime role of inflammation in modulating serotonin and dopamine biosynthesis. 
For this reason, we describe now explicitly the effects of inflammatory cytokines on IDO- and PAH-mediated regulation of serotonin and dopamine biosynthesis and changed the manuscript title to "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism". 
In fact as discussed below and in the revised manuscript text, the question how systemic biosynthesis of serotonin and dopamine, involving primarily mesenteric organs and the liver, relates to the neurotransmitter availability in the brain is vigorously disputed in literature [@Goldstein1999; @Moncrieff2022; @Marx2020].

## Issue 3

<span custom-style = 'reviewer'>__Introduction:__ Some statements made in the introduction are without reference, very general, and the conclusions weakly justified. 
For instance, TRP, KYN, and KYN/TRP are listed as markers of serotonin availability. 
However, the available evidence for the relationship between increased degradation of TRP to KYN and its effects on serotonin availability is controversial as only 1-2% of tryptophan metabolism goes into the serotonin pathway (https://doi.org/10.3390/metabo12060514). 
In addition, line 60 mentions a depressiogenic and an anxiogenic effect of QUIN, and an anxiety- and depression-amplifying effect of kynurenine and catabolites is stated in line 9 (page 3). 
Should this statement refer to existing cross-sectional findings (e.g., higher QUIN in patients with MDD compared to healthy controls) it is very generalized and not sufficiently justified. 
If the statement refers to evidence that suggests a causal relationship, a reference is urgently needed. </span>

We agree with the Reviewer, that __Introduction__ needs to be more strict and requires additional references - please refer to the revised text. 
Concerning your more specific issues: we agree that the great majority majority of tryptophan is catabolized via the KYN and indole pathways and only a marginal fraction is diverted to serotonin synthesis [@Teunis2022; @Lukic2022], and for both serotonin and dopamine, the systemic metabolism taking place in liver, mesenteric organs and vasculature may not reflect the situation in the central nervous system [@Brown2021; @Marx2020; @Schwarcz2012; @Moncrieff2022; @Goldstein1999]. 
We take up this controversy in __Introduction__ and __Discussion__. 
Additionally, we addressed the question how TRP levels may affect systemic serotonin availability, we modeled circulating serotonin levels in the INCOV cohort [@Su2022] as a function of its precursors and competitor metabolites (TRP, KYN, QUIN), timepoint of SARS-CoV-2 infection, inflammatory cytokines and age. 
As presented in __Figure 3__, we could indeed find a significant positive association of plasma TRP and serotonin in multi-parameter modeling, which is expected to be independent of age, inflammation and infection time course. 
In a more detailed correlation and network analysis of the INCOV dataset (__Figure 5A__), we could observe a moderate-strength correlation of circulating serotonin and TRP levels in acute and sub-acute infection. 
This suggests that systemic serotonin synthesis may be limited by blood TRP availability to some extent during SARS-CoV-2-dependent inflammation, which was characterized by profoundly decreased TRP and elevated KYN levels (__Figure 4__). 
In the revised __Introduction__ and __Discussion__, we also critically review the evidence on brain metabolism and effects of KYN, kynurenic acid and QUIN on neuronal signaling and mental health [@Brown2021; @Schwarcz2012], which is indeed limited.

## Issue 4

<span custom-style = 'reviewer'>__Introduction:__ The introduction sentence of the last paragraph in the introduction seems not adequate for the following hypotheses. 
It highlights that low-grade inflammation (as I understand it, in the case of this study IL-6 and CRP concentrations) is associated with mental health. 
However, the presented hypotheses use IL-6 and CRP and mental stress, anxiety, and depressive symptoms as predictors in their analysis (multi-parameter linear modeling) and TRP, KYN, KYN/TRP, PHE, TRY, PHE/TRY as dependent variables.</span>

We are sorry for this unclarity. 
In the revised manuscript, we clearly formulate the aims, and define outcomes and explanatory factors. 
Please refer to the last paragraph of __Introduction__, __Methods/Analysis endpoints__, __Supplementary Tables S1__ and __S3__. 
In the revised text, we additionally avoid the misleading term 'low-grade' inflammation. 
In fact, as described in __Issue 1__ and __Supplementary Methods__, only a minute subset of the SIMMUN study participants had CRP or IL-6 concentrations above the reference levels. 
Thus, biological effect of those two inflammatory markers is doubtful and we decided to exclude them from the revised analysis pipeline.

## Issue 5

<span custom-style = 'reviewer'>__Methods:__ while the SIMMUN cohort is described in detail, basic information about the INCOV cohort is missing and should be provided.</span>

We agree with the Reviewer that the INCOV cohort deserves a more thorough description. This is provided in the __Methods/Procedures__ section of the revised manuscript and in __Supplementary Methods__ and in references to the seminal papers [@Su2020; @Su2022].

## Issue 6

<span custom-style = 'reviewer'>__Methods:__ Before the statistical analysis, several transformations (log2, ln, and square root) were applied. 
However, it is not transparent how each parameter is transformed. I would appreciate it if this could be further specified.</span>

We provide now these information in __Methods__ and __Supplementary Methods__. 
Variable format, transformations and, if appropriate, stratification schemes are presented in __Supplementary Tables S1__ and __S3__.

## Issue 7

<span custom-style = 'reviewer'>Methods: Furthermore, when measured within clinical routine laboratory values of CRP and IL-6 are in many cases lower the limits of detection (e.g., 0.3 mg/ml). 
Was this the case in this study? 
If so, it would be useful for the replicability to know how the authors dealt with this issue.</span>

As specified now in __Methods/Procedures__, values beyond or at their detection limits were substituted with the lower or upper detection limit. 
CRP and IL6 were removed from the revised analysis pipeline for the reasons specified in __Issue 1__.

## Issue 8

<span custom-style = 'reviewer'>__Methods:__ Kruskal-Wallis test with Man-Whitney U tests are applied to assess the differences be-tween uninfected individuals and infected individuals at different time points. 
While the pairwise comparisons are adequate for this research question, the overall statistic of the Kruskal-Wallis test is not useful in this context, as it tests whether there is an over-all difference between independent groups. 
I would therefore suggest using a linear model with predefined contrasts (uninfected vs. acute; uninfected vs. sub-acute; uninfected vs. recovery) inclusive control for multiple tests. 
If the assumptions of linear regression are not fulfilled, robust methods can be used (e.g., R-package: MASS; https://www.rdocumentation.org/packages/MASS/versions/7.3-58.2/topics/rlm). 
Additionally, the correlations between the assessed dependent variables seem rather high, why controlling for multiple comparisons using Benjamini & Yekutieli procedure (2001) or even robust MANOVAS (see: https://cran.r-project.org/web/packages/MANOVA.RM/vignettes/Introduction_to_MANOVA.RM.html) could be also necessary. Additionally, the terms "acute", "sub-acute and "recovery" should be described in the notes of __Figure 3__.</span>

We are grateful for the hints. 
In the revised manuscript, we applied robust linear regression with the MM algorithm and Huber's psi function [@Huber2011; @Ripley2022] for multi-parameter modeling (__Figure 3__) and time-course modeling of the INCOV dataset (__Figure 4__). 
Additionally, we specify in the figure legends of the INCOV results the median time interval between diagnosis and sampling for the acute, sub-acute and recovery timepoints. 
As suggested, we applied multiple testing adjustment throughout the analysis pipeline. 
Except for multi-parameter modeling, p values were adjusted separately for each analysis step (e.g. correlation analysis, time-course modeling) with the false discovery rate method known also as 'Benjamini-Hochberg' [@Benjamini1995].

## Issue 9

<span custom-style = 'reviewer'>__Psychometric assessment:__ In the description of the psychometric assessment, the source of the HADS is missing as well as a calculation of the reliability of the used questionnaires for the SIMMUN study cohort.</span>

We apologize for the missing references and cite the seminal HADS paper and a report dealing with the clinically relevant HADS cutoffs [@Bjelland2002; @Zigmond1983] in the revised text. 
We also checked consistency of the PSS-4 and HADS psychometric tools employed in our report. 
Since the HADS scales, and in particular, the HADS depression tool were found to be non-tau-equivalent by factors analysis (__Supplementary Figure S1A__), we decided use McDonald's $\omega$ as a consistency measure instead of the more popular Cronbach's $\alpha$ [@Revelle2022; @McDonald1999]. 
The PSS-4 tool demonstrated a good and the HADS scales an excellent consistency, as presented in __Supplementary Figure S1B__.

## Issue 10

<span custom-style = 'reviewer'>__Results:__ First, it would be useful to distinguish between cohorts in the descriptive characterization of cohorts. 
Furthermore, concrete numbers should be given instead of rough descriptions (e.g., "about half" (page 8, line 19)). 
Second, assessing the effect as "dramatic" is not appropriate in a scientific context. 
If at all, existing assessments based on effect sizes can be made using generally accepted definitions (e.g., according to Cohen).</span>

We concur with the Reviewer, that a comparison between the cohorts is useful for interpretation of the results. 
We present significant results of the comparison between the SIMMUN and INCOV cohort in __Supplementary Table S4__ and elaborate on them in __Results/Characteristic of the study cohorts__. 
The collectives differed in age and sex distribution, rates of SARS-CoV-2 infection and infection severity. 
In particular, most INCOV study participants experienced moderate-to-critical COVID-19 and required hospitalization. 
This may explain the way stronger effects of inflammation and infection timepoint on inflammatory cytokines and the metabolites of interest, which we discuss in __Discussion__ and __Limitations__.
In the revised text, Figures and Tables we also present effect size metrics for statistical hypothesis tests, correlation and modeling. 
We interpret them as proposed in the excellent handbook by Jacob Cohen [@Cohen2013] (__Supplementary Methods__).

## Issue 11

<span custom-style = 'reviewer'>__Results:__ In general, numbers (test statistics, p-values) are missing in the results part and should be added. Additionally, effect sizes should be presented.</span>

We present now p values and effect sizes or model $\beta$ with 95% confidence intervals [@Cohen2013], as appropriate, for the most important findings in Abstract and Results. 
Yet, for the sake of readability, we were not able to provide significance and effect size information for all effects. 
In particular, we do not specify p values and effect sizes for non-significant effects and state instead clearly that the effect missed statistical significance. 
We abstained from referencing test statistics such as $\chi^2$ or $t$, since they depend on the number of complete observations and hence do not deliver reliable information on the effect size.

## Issue 12

<span custom-style = 'reviewer'>__Results:__ "Tends to be higher"/ "Slightly more" is scientifically incorrect. 
Please replace the terms and add the test statistic and p-value</span>

Following your suggestion, we avoid this wording in the revised manuscript. 
Please refer to __Issue 11__ for details.

## Issue 13

<span custom-style = 'reviewer'>__Results:__ The terminology of strata (page 10, line 10) can be misleading as no stratification was performed. Probably it would be better to use the term group or subgroup.</span>

Following your suggestion, we avoid using the term 'strata' in the revised text. 
We use group, subgroup or subset instead.

## Issue 14

<span custom-style = 'reviewer'>__Introduction:__ The use of subheadings would be helpful to facilitate the understanding of the first part of the introduction.</span>

This point is no relevant anymore, since we re-wrote and consolidated __Introduction__ of the revised manuscript. 
We describe each thread in a separate paragraph.

## Issue 15

<span custom-style = 'reviewer'>__Results:__ The statement that TRP is a precursor of IDO is not correct!</span>

Thanks for careful lecture, this erroneous statement was removed.

## Issue 16

<span custom-style = 'reviewer'>__Results:__ The description of the time between the infection and the assessment of 138.5 days leads to the question of how the time between the infection and the assessment was assessed. Maybe it would be better to use complete instead of half and quarter days as units.</span>

This time interval refereed to the SARS-CoV-2 infection diagnosis via PCR test, which we specify in the revised manuscript. 
We also round them to complete days, as suggested.

## Issue 17

<span custom-style = 'reviewer'>__Results:__ h. Table 1: I would suggest adding further lines in the table as "heading" lines specifying the cohorts described in the following parts.</span>

For the sake of clarity, we present characteristic of the SIMMUN and INCOV cohorts in separate tables (__Table 1__ and __Table 2__).

## Issue 18

<span custom-style = 'reviewer'>__Results:__ Page 9 lines 24-41; This paragraph should be in the Methods instead of the Results. Furthermore, the rationale is missing: Why was INCOV used as a comparison or additional study to your own data? In addition, information on the measurement /measurement system (for INCOV) should be provided. Were the measuring probes used in SIMMUN and INCOV comparable?</span>

We concur with your suggestion, the variables of interests are now listed in the __Methods__ section instead of __Results__. 
Furthermore, we describe the rationale for analyses in the INCOV cohort in the last paragraph of __Introduction__, __Methods/Analysis endpoints__ and __Results__. 
In brief, we employed the SIMMUN collective providing more extensive demographic and clinical characteristic in an initial search for the most relevant factors impacting on the indirect readouts of serotonin and dopamine metabolism: TRP, KYN, PHE and TYR [@Brown2021; @Marx2020; @Capuron2011; @Widner1997; @Neurauter2008]. 
The INCOV dataset with measurements of more direct markers of the neurotransmitter availability, i.e. plasma serotonin and dopamine 3-O-sulfate, the most abundant dopamine metabolite present in human blood [@Meiser2013; @Goldstein1999], allowed us to validate effects of inflammation, SARS-CoV-2 infection and age, i.e. significant predictors identified in the SIMMUN study. 
Furthermore, due to longitudinal design of the INCOV study, we were able to explore not only the impact of infection recovery but also acute and sub-acute COVID-19 on serotonin and dopamine turnover. 
Still, the incompatibility of designs and datasets of these two collectives hindered us from developing consistent models in one cohort and direct validation in the outer (canonical training - test or external validation setup). 
We list this issue in __Limitations__.

## Issue 19

<span custom-style = 'reviewer'>__Results__: The term acute and sub-acute should be defined.</span>

We provide the time intervals for the acute, sub-acute and recovery timepoints in the INCOV cohort in the revised manuscript (__Methods__, __Results__, __Figures__). 
Of note, those time intervals fit well with the definition of acute, sub-acute and long-COVID/recovery timepoints in our recent population study [@Sahanic2021].

## Issue 20

<span custom-style = 'reviewer'>__Results:__ What does the term "virtually age-independent" mean? I would appreciate it if concrete numbers are stated here instead.</span>

We apologize for this imprecise term, it was removed from the revised manuscript.

## Issue 21

<span custom-style = 'reviewer'>__Results:__ Page 11 lines 35 to 54 should be part of the statistical description in the Method section.</span>

This part of the manuscript was removed during revision. 
Following your suggestion, we keep details on procedures and statistical analysis in the __Methods__ section.

## Issue 22

<span custom-style = 'reviewer'>__Results:__ Page 15 line 28: standard inflammatory cytokines should be named more specifically.</span>

We are sorry for this unclarity. 
In the revised text, we name inflammatory markers (NEO, NLR, cytokines) more explicitly.

## Issue 23

<span custom-style = 'reviewer'>__Discussion:__  The discussion of the study results is rather shallow: Results should be better integrated into existing evidence. Furthermore, there is no discussion of the results of depressive symptoms (HADS).</span>

In the revised manuscript, we restructured and improved scientific quality of __Discussion__. 
The key explanatory factors proposed by us to influence systemic serotonin and dopamine availability: age, inflammation, SARS-CoV-2 infection and, for serotonin, availability of the precursor aminoacid TRP, are discussed in a context of literature evidence. 
We also critically elaborate on the link between plasma serotonin, TRP and KYN levels and depressive disorders, which was recently challenged [@Capuron2011; @Hufner2021; @Sun2020; @Hufner2015; @Fellendorf2022; @Paul-Savoie2011; @Hunt2020; @Moncrieff2022; @Marx2020; @Brown2021].

## Issue 24

<span custom-style = 'reviewer'>__Discussion:__ The statement that "BH4 is also important for nitric oxide synthesis and therefore involved in oxidative stress" is not correct. (Page 16) BH4 is an essential Co-factor of NO-synthases that produce NO with its function as a vasodilator and additional anti-pathogenic effects. Only if BH4 is missing, there is an uncoupling of NOS with subsequently increased reactive oxygen/nitrogen species. Thus, your "therefore involved in oxidative stress" is misleading in normal physiological states.</span>

Following discussion in the study team, we decided to suppress the ROS thread of the discussion since it is likely not so important for interpretation of our data. 
Instead we focus on direct effects of inflammation on PAH, the enzyme catalyzing the PHE - TYR conversion, i.e. the initial step of dopamine synthesis and its crucial co-factor BH4 [@Geisler2013; @Luporini2021; @Fanet2021; @Capuron2011; @Neurauter2008].

## Issue 25

<span custom-style = 'reviewer'>__Discussion:__ Page 14 line 60: "pre-pandemic inflammation markers" is probably not the right term. Maybe rewrite the sentence.</span>

We apologize for this obscure term. 
We meant levels of inflammatory markers in SARS-CoV-2-uninfected individuals, which we state now explicitly.

## Issue 26

<span custom-style = 'reviewer'>__Highlights:__ In the Highlights, the results and perhaps the conclusion should be emphasized more than the theoretical background (Highlights 1 and 2) or the methodology (Highlights 3 and 4).</span>

_Katharina: please respond, I don't have this manuscript part at all in pre-revision manuscript files._

## Issue 28

<span custom-style = 'reviewer'>__General:__ Furthermore, there are several grammatical flaws and spelling mistakes.
Some of these mistakes are listed here. 
However, the manuscript would benefit from professional language correction.
</span>

We are grateful for the careful language check. 
We have paid particular attention to the style, grammar and proper scientific language in the revised manuscript. 
For particular language issues:

* _Page 3 line 24: mediate instead of mediated._ - corrected as suggested

* _Page 3 line 29: sentence incomplete_ - rewritten

* _Page 4 l. 35. Results of the SIMMUN study_ - does not apply, __Results__ were rewritten

* _Page 4: line 37: in-person study visit sounds not familiar: better use on-site investigation_ - we use the wording 'on-site study visit' in the revised manuscript

* _Page 4 line 38 is grammatically not correct: probably better […] the results of which are re-ported here_ - rewritten

* _Page 5 line 14: The term physician assessment should be replaced by a better term_ - we use 'general medical assessment' instead.

* _Page 6 line 11: The hyphen is missing at "inflammatory"_ - does not apply, the section was rewritten

* _Page 8: lines 55-56: In the participants who were tested…_ - rewritten

* _Page 8 line 58: timepoint in [the?] the course sounds strange (and the article is missing), may-be better time between infection and assessment._ - changed to 'timepoints of SARS-CoV-2 infection' and 'days after diagnosis via PCR'

* _Page 9 line 52: What are nadir TRP concentrations?_ - we do not use the term any more, changed to 'minimum'

* _Page 11 line 7: a comma is missing._ - does not apply, the section was rewritten.

* _Page 12 line 24: "on the other hand" without "on the one hand"_ - we removed this colloquial phrase and other similar wording from the revised text

* _Page 13: line 4: a comma is missing_ - rewritten

* _Page 14 line 14: […] via their bidirectional influence_ - rewritten

* _Page 14 line 34: please delete the term "larger"_ - done as suggested.

* _Page 15, line 35: of in individuals_ - rewritten

* _Page 16, line 16: incomplete_ - rewritten

## Issue 29

<span custom-style = 'reviewer'>__Supplementary Material:__ As a criterion to compare models, the Bayesian information criterion is stated with AIC as an abbreviation.</span>

We apologize for the typo, it was corrected to 'BIC'.

## Issue 30

<span custom-style = 'reviewer'>__Supplementary Figure S2:__ It is not clear why fitted generalized adaptive models are presented as graphs, as the associated statistic (Spearman's rho) assesses monotone relationships.</span>

We concur with the Reviewer's view, that visualization of trends for monotone associations is problematic. 
Due to profound changes in the manuscript structure, we do not show scatter plots to illustrate monotone correlations between non-normally distributed variables. 
Correlation analysis results in the INCOV cohort are presented now as simple bubble correlograms or visualized as force-directed graphs [@Briatte2021; @Csardi2006].

\newpage

# Reviewer 2

## Issue 31

<span custom-style = 'reviewer'>In this paper, the authors show the relationship between alterations in markers of inflammation, aminoacid precursors, and symptoms of depression, anxiety, and stress in individuals COVID-19-positive or not. 
They relate the results of these observations to the results obtained in another immune and metabolic study carried out at different time points of the COVID-19 trajectory and already published by a group of different investigators.
The methodology is excellent as can be expected from a research team specialized in the study of inflammation and its impact on the kynurenine metabolism pathway. 
The results are original and sufficiently important to be published. 
The main problem with the current manuscript is the confusion coming from the mode of presentation that mixes up in the results section data from the two studies. 
The discussion is circumvoluted and difficult to follow.</span>

We thank the Reviewer for appreciation of out text and the constructive feedback. 
Weaknesses in structure of the test were already pointed out be the Reviewer 1. 
We addressed them by employing the multi-parameter modeling results for identification of the most vital explanatory factors that affect neurotransmitter metabolism and subsequent more specific correlation analyses and statistical testing with such relevant predictors. 
To tackle with the confusion around the two investigated cohorts, we separated the results obtained with the SIMMUN and INCOV collective. 
In more details, we employ the SIMMUN cohort as a kind of 'exploratory' collective to identify factors affecting TRP, KYN, KYN/TRP, PHE and PHE/TYR among a wider palette of demographic, clinical, psychometric, inflammatory and SARS-CoV-2 associated variables. 
Subsequently, we attempted to validate the most important of them: age, inflammation and SARS-CoV-2 infection in the INCOV cohort. 
Finally, we re-wrote the __Discussion__ section with a special focus on the crosstalk between inflammation and neurotransmitter biosynthesis and, in part controversial, relevance of systemic neurotransmitter availability for the central nervous system and mental health.

## Issue 32

<span custom-style = 'reviewer'>There are a few typos in the text, e.g., othe (line 6 instead of other).</span>

Thanks for the careful lecture. 
We paid special attention to the grammar and spelling of the revised text.

## Issue 33

<span custom-style = 'reviewer'>The text would benefit from further editing by a specialized scientific English editing service</span>

_Katharina: could you provide an answer?_

## Issue 34

<span custom-style = 'reviewer'>Line 7: by collectives you probably mean cohorts or population samples.</span>

The term 'collective' has the same meaning as 'cohort' in the manuscript text. 
For practical reasons, we avoid the wording 'population sample', which may suggest a random and hence representative draw from the general population. 
In fact neither the SIMMUN cohort enriched in psychiatric patients not the INCOV cohort enriched in hospitalized COVID-19 patients pose a representative sample of the general population during the pandemic (see __Limitations__) .

## Issue 35

<span custom-style = 'reviewer'>The introduction ends with the statement that the study was conducted in an exploratory manner. Was that the objective of the present study in the absence of any specific hypothesis?</span>

As pointed out by Reviewer 1 as well, the last paragraph of __Introduction__ was quite obscure. 
In the final part of __Introduction__, we present now the 'evidence gap', which is the still incompletely characterized effect of SARS-CoV-2 infection on neurotransmitter metabolism, as well as the aim of our analysis, which identification of factors affecting the neurotransmitter turnover in a cohort of SARS-CoV-2-infected individuals. 
Still, our study does not have a fully exploratory character, since the candidate explanatory factors: age, sex, body mass class, mental and somatic conditions, smoking and alcohol, depression, anxiety, stress and inflammation have already been described to affect serotonin and dopamine metabolism.

## Issue 36

<span custom-style = 'reviewer'>Page 9, line 34 and elsewhere in the text when referring to immune factors that are measured, use the term variables instead of parameters.</span>

We agree that the term 'parameter' may be misleading, especially in context of modeling and its results. 
We changed it to 'variable' throughout the revised text.

## Issue 37

<span custom-style = 'reviewer'>Legend of Figures: the use of the term decay is inappropriate, you probably mean metabolites</span>

We avoid using the term 'decay' in the revised text and utilize 'catabolism' or 'conversion' instead.

## Issue 38

<span custom-style = 'reviewer'>The results are difficult to read as you mix up the two cohorts both in the text and in the figures. It would be preferable to first present the results of your own cohort, the SIMMUN cohort, and then test the generality of the results by assessing whether the same variations are found in the INCOV cohort with the limitation that this can only be done for the relationship between immune factors and neurotransmitter precursors.</span>

We concur with your opinion and present the analysis results as suggested. 
Please refer to __Issues 1__ and __31__ for details.

## Issue 39

<span custom-style = 'reviewer'> You utilize the terms mental disease, mental stress, mental symptoms, etc… Be more specific and replace these terms by what you measured, i.e., HADS-anxiety, HADS-depression, and PSS. 
You mention somatic symptoms in the discussion but what they refer to remains obscure. 
You also refer to somatic diseases. 
Pathologists would rather speak of physical disorders as opposed to psychiatric disorders.</span>

We are grateful for this hint. 
In the revised manuscript we provide the psychometric scale when referencing depression, anxiety and stress signs. 
We removed the sentence of __Discussion__ referring to somatic symptoms, since those were not systematically surveyed in the SIMMUN cohort. 
Following your suggestion, we term 'somatic conditions' 'physical disorders' consequently in the revised text.

## Issue 40

<span custom-style = 'reviewer'>The discussion is convoluted and needs to be simplified. 
It would be easier to follow if you would first discuss the significance of the association between inflammation and neurotransmitter precursors and then the association between these two sets of factors and what you call mental health measured by HADS and PSS, first in the immune -> HDAS symptomatology or kynurenine -> HDAS symptomatology and then in the stress -> immune direction, remembering that in the case of your own cohort, the cross-sectional nature of your study does not allow you to get further than describing associations.</span>

We agree with the Reviewer that there was a lot of room for improvement in __Discussion__.
We rewrote this part of the manuscript during the revision according to the following scheme:

* A brief generalizing summary of the results obtained with both cohorts. In this initial paragraph, we highlight the most relevant factors for systemic serotonin and dopamine metabolism: inflammation, SARS-CoV-2 infection, TRP availability, age, signs of depression and mental stress. In this paragraph, we also discuss some discrepancy in our data such as way higher effects of inflammation on neurotransmitter availability in the INCOV cohort than in the SIMMUN dataset.

* Role of inflammation and SARS-CoV-2 in the KYN pathway and PHE - TYR conversion with implication of systemic availability of serotonin and dopamine. In this paragraph we discuss clinical literature and mechanistic reports tackling with IDO and PAH activity as well as BH4, the key co-factor of the serotonin and dopamine pathway enzymes.

* Effects of mental health-related factors (psychiatric disorders, mental stress) on systemic serotonin and dopamine availability.

* Relevance of systemic metabolism of serotonin and dopamine for the central nervous system and development of psychiatric disorders. Here we also underline that evidence for effects of peripheral serotonin, systemic activity of the KYN pathway and peripheral dopamine synthesis from PHE and TYR is limited and bases in many cases on clinical association of psychiatric disorders with blood levels of the neurotransmitters, their precursors or competitor pathway products.

## Issue 41

<span custom-style = 'reviewer'>Page 15: the study by Benedetti is at best suggestive but does not demonstrate anything.</span>

We agree with the Reviewer, that the report by Benedetti et al. does not provide any solid mechanistic link between the COVID-19-related inflammation and depression. 
We removed this paper from __Discussion__ of the revised manuscript.

# Reviewer 3

## Issue 42

<span custom-style = 'reviewer'>Considering the disease SARS CoV-2 and its consequences in the form of post-covid syndrome, the presented work is highly relevant.</span>

We thank the Reviewer for appreciation of our manuscript and the valuable feedback.

## Issue 43

<span custom-style = 'reviewer'>In the abstract, the authors introduce ...markers of serotonin availability (TRP, KYN, KYN/TRP ratio)....(line 24). 
This statement is not accurate, as Serotonin is formed from TRP, but there is no KYN in its catabolic pathway. 
Only if the catabolism of TRP moves more to the kynurenine pathway, less TRP goes to the serotonin pathway, and the availability of serotonin is reduced. 
This argument is similar to the line 44.</span>

We concur with the Reviewer's opinion. 
Certainly, kynurenine and the kynurenine/tryptophan ratio are markers of the pathway which competes with the serotonin biosynthesis. 
For the sake of clarity, we we term kynurenine, quinolinate and kynurenine/tryptophan ratio 'competitor pathway products/metabolites' or products/metabolites of the competitive kynurenine pathway' in the revised text.

## Issue 44

<span custom-style = 'reviewer'>The principle point is that it would help the reader to orient himself in such a large amount of data if the authors completed the manuscript with a design scheme of the INCOV and SIMMUN study</span>

To improve readability of the revised manuscript, we present the results in a cohort-wise manner, i.e. beginning with the SIMMUN cohort as an 'exploratory' collective and proceeding with the INCOV cohort as a king of 'validation' collective used to verify and study in more detail (e.g. by time course analyses) the key findings from the SIMMUN cohort. 
Please refer to __Issues 1__ and __Issue 38__ for details. 
The modified analysis strategy is presented in __Figure 1__.

## Issue 45

<span custom-style = 'reviewer'>In line 16 on p. 6 are introducing breakdown products as KYN, TRP, KYN/TRP ratio ..... It would be more precise to introduce them in the order they are formed: TRP, KYN, KYN/TRP ratio ....pls, correct everywhere in the manuscript similar p. 9, line 29 ..... For example, correct position is given on the line 52 of p.11</span>

Thanks for the hint. 
We stick in the revised manuscript to the proposed metabolite order.

## Issue 46

<span custom-style = 'reviewer'>p. 14, line 34 : a large meta-analyses ... include more citations</span>

We re-wrote the Discussion part of the revised manuscript and provide now more evidence [@Hunt2020; @Maes2012; @Brown2021; @Marx2020;  @Fellendorf2022] on the link between the so called sickness behavior, inflammation, serotonin and dopamine metabolism in the final part of the scion tackling the relevance of our findings for mental health. 
We also discuss the controversial relationship between circulating markers of neurotransmitter availability such as tryptophan or serotonin, and the competitor pathway products (kynurenine, quinolinate) on central nervous system and mental health [@Brown2021; @Moncrieff2022].

## Issue 47

<span custom-style = 'reviewer'>p. 15, line 38 - correct as a note in the abstract</span>

_Katharina, could you please respond? I don't have the submission text with the line numbering and can't find the fragment. Thanks!_

## Issue 48

<span custom-style = 'reviewer'>p. 15, line 62 - where ???</span>

_As above, thanks!_

## Issue 49

<span custom-style = 'reviewer'>put one line above the grey text stating that it is INCOR</span>

_Same as above!_

## Issue 50

<span custom-style = 'reviewer'> the given data do not match mathematically - check or explain why</span>

We apologize for the typos concerning numbers in the study inclusion scheme. 
Please refer to the correct numbers in the revised __Figure 1__.

## Issue 51

<span custom-style = 'reviewer'>explain used abbreviations in the text below</span>

We are sorry for that. 
We explain now all abbreviations used in __Figures__ in their legends.

\newpage

# References

### 2023-05-23 paper/markdown/main_manuscript.Rmd
Manuscript and Rebuttal Letter

--- 2023-05-15 paper/markdown/main_manuscript.Rmd
@@
 ---
-title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
+title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism"
 subtitle: "Manuscript"
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
+
+author:
+  - Katharina Hüfner:
+      institute: [psych2]
+      email: katharina.huefner@tirol-kliniken.at
+      correspondence: true
+  - Sophia Vedova:
+      institite: [psych2]
+  - Piotr Tymoszuk:
+      institute: [daas]
+  - Philipp Nelles:
+      institute: [psych2]
+  - Markus Anliker: 
+      institute: [zimcl]
+  - Tobias Bruckner: 
+      institute: [psych2]
+  - Eberhard Deisenhammer: 
+      institute: [psych2, psych1]
+  - Jonas Egeter: 
+      institute: [psych2]
+  - Alexander Egger: 
+      institute: [zimcl]
+  - Matyas Galffy: 
+      institute: [psych2]
+  - Johannes Giesinger: 
+      institute: [psych2]
+  - Jens Lehmann: 
+      institute: [psych1]
+  - Maria Oberhammer: 
+      institute: [psych1]
+  - Joachim Rockenschaub: 
+      institute: [psych2]
+  - Magdalena Sacher: 
+      institute: [surg]
+  - Bernhard Holzner: 
+      institute: [psych2]
+  - Johanna M Gostner: 
+      institute: [bio]
+  - Barbara Sperner-Unterweger: 
+      institute: [psych2]
+
+institute:
+  - psych2: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Anichstrasse 35, Innsbruck, Austria
+  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria
+  - zimcl: Central Institute of Medical and Chemical Laboratory Diagnostics, University Hospital Innsbruck, Anichstrasse 35, Innsbruck, Austria
+  - psych1: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry I, Medical University of Innsbruck, Anichstrasse 35, Innsbruck, Austria
+  - surg: Department of Visceral, Transplant and Thoracic Surgery, Center of Operative Medicine, Medical University of Innsbruck, Innsbruck, Austria
+  - bio:  Institute of Medical Biochemistry, Biocenter, Medical University of Innsbruck, Innrain 80/82, Innsbruck, Austria
 
@@
 
-## Data and code availability
+# Keywords
+
+SARS-CoV-2, inflammation, serotonin, dopamine, tryptophan, kynurenine, psychiatric disorders
+
+\newpage
+
+# Abstract
+
+__Background:__ Serotonin and dopamine metabolism poses a junction of mental and physical health. 
+We aimed to identify demographic, clinical, psychometric and SARS-CoV-2 infection-related factors affecting systemic serotonin and dopamine turnover.
+
+__Methods:__ The cross-sectional SIMMUN (n = `r nrow(stigma$analysis_tbl)`, Austria) and longitudinal INCOV cohort (n = `r  length(unique(incov$complete_ids))`, Su et al. 2022) were investigated. 
+Explanatory variables encompassed age, sex, clinical and inflammatory variables, SARS-CoV-2 infection, signs of anxiety and depression (HADS), and mental stress scoring (PSS-4). 
+Their effects on serotonin biosynthesis markers (serotonin, tryptophan) and competitor kynurenine pathway products (kynurenine, quinolinate), and dopamine availability readouts (dopamine sulfate, phenylalanine, tyrosine) were assessed by linear modeling, correlation analysis and two-tailed T tests.
+
+__Results:__ In the SIMMUN cohort, the inflammatory marker neopterin, SARS-CoV-2 infection, age, depression symptoms and mental stress were positively associated with catabolism of the serotonin precursor tryptophan to kynurenine.
+Inflammation and age were also suppressed the first step of dopamine biosynthesis, phenylalanine - tyrosine conversion. 
+In the INCOV cohort, inflammation was associated with lowered serotonin (IL6: $\beta$ = `r incov_neuro$inference$serotonin %>% filter(variable == 'IL6_INF') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`) and the dopamine metabolite, dopamine sulfate (interferon-gamma: $\beta$ = `r incov_neuro$inference$dopamine %>% filter(variable == 'IFNG_INF') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`).
+Serotonin ($\beta$ = `r incov_neuro$inference$serotonin %>% filter(variable == 'timepoint', level == 'recovery') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`)
+and dopamine sulfate ($\beta$ = `r incov_neuro$inference$dopamine %>% filter(variable == 'timepoint', level == 'recovery') %>% mutate(beta_lab = paste0(signif(estimate, 2), ' [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), ']')) %>% .$beta_lab`) were significantly increased during SARS-CoV-2 infection recovery.
+
+__Conclusion:__ 
+Inflammation limits systemic serotonin and dopamine by activating the competitor kynurenine pathway and inhibiting the phenylalanine - tyrosine conversion, respectively, and may link SARS-CoV-2 infection with mental health.
+Advanced age and psychiatric disorders can additionally suppress neurotransmitter synthesis.
+
+\newpage
+
+# Introduction
+
+Physical conditions such as infections, trauma, frailty, autoimmune illness or malignancy are often accompanied by sings of depression, anxiety, metal stress and other psychiatric disorders [@Sorgdrager2019; @Dantzer2008; @Hufner2019; @Hufner2020; @Maes2012; @Geisler2013; @Haroon2012; @Hufner2015].
+Imbalance of two depressiolytic and anxiolytic neurotransmitters, serotonin and dopamine, elicited by inflammation as proposed as a connex of physical pathology and mental health deficits [@Dantzer2008; @Brown2021; @Schwarcz2012].
+Serotonin is synthesized from the essential aminoacid tryptophan (TRP) via reactions catalyzed by tryptophan hydroxylase (TPH) and aromatic L-amino acid decarboxylase. 
+Yet the great majority of TRP is by the so called kynurenine (KYN) pathway, whose first step is the breakdown of TRP to KYN catalyzed by indoleamine 2,3-dioxygenase (IDO) [@Badawy2017; @Lukic2022]. 
+Expression of an inducible and highly active form of IDO, IDO1, is strongly stimulated by a plethora of cytokines, pathogen products and other inflammatory stimuli [@Robinson2003; @Dai1990; @Litzenburger2014; @Badawy2017].
+During systemic or central nervous system-localized inflammation, TRP is depleted by high IDO1 activity, which, in turn, suppresses serotonin biosynthesis [@Brown2021; @Schwarcz2012; @Vancassel2018; @Haroon2012].
+Consequently, elevated blood levels of KYN pathway metabolites, decreased TRP and serotonin were reported for numerous physical conditions and associated with symptoms of psychiatric deficits [@Brown2021; @Marx2020; @Hunt2020; @Fellendorf2022; @Hufner2015; @Hufner2021; @Hufner2019; @Paul-Savoie2011; @Capuron2011].
+Dopamine synthesis involves the conversion of another essential amino acid phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH) [@Meiser2013].
+Both PAH and TH as well as the serotonin-synthesizing enzyme TPH rely on tetrahydrobiopterin (BH4) as an essential cofactor. 
+During inflammation, BH4 is depleted by oxidation to its inactive derivative and gets consumed in a competitor reaction of nitric oxide formation. 
+As a result of such functional BH4 deficiency, biosynthesis of dopamine and serotonin is further suppressed [@Geisler2013; @Fanet2021; @Vancassel2018]. 
+Accordingly, imbalance of PHE and TYR levels suggestive of systemic dopamine deficiency was reported in cancer, infections and inflammatory conditions and associated with depression and anxiety [@Hufner2015; @Neurauter2008; @Geisler2013; @Fanet2021; @Vancassel2018].
+
+SARS-CoV-2 virus is the causal pathogen of coronavirus disease 2019 (COVID-19). 
+Apart from sustained physical disability, psychiatric disorders in millions of SARS-CoV-2 patient amount to the persistent burden of the COVID-19 pandemic [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Staudt2022].
+Of note, high activity of the KYN pathway and signs of inflammatory BH4 deficiency manifesting by distorted PHE and TYR levels were reported for COVID-19 patients and associated with inflammation and disease severity [@Gietl2023; @Lionetto2021; @Santiago-Mujika2022; @Luporini2021; @Bizjak2022; @Dewulf2022; @Saito2022].
+Furthermore, reduced systemic availability of serotonin and dopamine was postulated to contribute to mental health deficits during SARS-CoV-2 infection recovery [@Bower2022; @Kucukkarapinar2022; @Matits2023]
+
+Still, the impact of inflammation on systemic metabolism of serotonin and dopamine availability in acute SARS-CoV-2 infection and recovery, along with its mutual interaction with psychiatric disorders is still incompletely characterized.
+To address that, we explored effects of demographic and clinical factors, inflammation, SARS-CoV-2 infection, anti-SARS-CoV-2 humoral response and psychiatric disorder symptoms on readouts of systemic serotonin and dopamine availability and kynurenine pathway activity in two cross-sectional cohorts of SARS-CoV-2 patients: the SIMMUN cohort and the published longitudinal INCOV study [@Su2022].
+
+\newpage
+
+# Materials and Methods
+
+## Ethics statement
+
+The SIMMUN study was conducted in accordance with the Declaration of Helsinki and European Data Policy. 
+All participants gave written informed consent prior to enrollment. 
+Participants' data were stored and analyzed in anonymized form.
+This study was approved by the ethics committee of the Medical University Innsbruck (Austria, approval number: 1132/2020). 
+No approval by the ethics committee was required for analysis of the published INCOV dataset [@Su2022].
+
+## Study cohorts
+
+Details on study cohorts, procedures and analyses are provided in __Supplementary Methods__.
+
+### SIMMUN study
+
+Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR were invited to participate.
+The study was conducted between June 2020 and May 2021. 
+The inclusion criteria were age of 18 - 70 years, proficiency in German, residence in the study region (Tyrol, Austria), and a SARS-CoV-2 PCR test conducted at the study site. 
+The exclusion criteria were active SARS-CoV-2 infection (< 14 days after diagnosis), pregnancy, active malignancy, organ transplantation, surgery in the past 3 months, inflammatory illness and oral corticosteroid treatment.
+The analysis inclusion criterion was the complete study variable dataset (__Figure \@ref(fig:fig-inclusion)__, __Supplementary Table S1__). 
+Significant differences between the analyzed and excluded participants are listed in __Supplementary Table S2__.
+
+### INCOV study
+
+Proteome and metabolome data and clinical information for the INCOV cohort were extracted from supplementary tables accompanying the report by Su and colleagues [@Su2022]. 
+In the current analysis, samples obtained for individuals with the complete age information and complete study variable dataset (__Figure 1__, __Table \@ref(tab:tab-incov-cohort)__, __Supplementary Table S3__).
+
+## Procedures
+
+### SIMMUN study
+
+The SIMMUN study data were gathered during a single on-site study visit including a general medical assessment, supervised completion of self-rating questionnaire and a blood sample collection (__Supplementary Table S1__).
+The median time interval between the study visit and SARS-CoV-2 PCR was 139 days (interquartile range: 119 - 157).
+
+Demographic and clinical variables: age, sex, body mass index, professionally diagnosed psychiatric disorders, self-reported physical disorders, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR test were surveyed during the study visit or extracted from electronic patient records.
+Inflammatory markers: plasma neopterin concentrations (NEO) and neutrophil - lymphocyte ratio (NLR) were determined by the certified clinical routine laboratory at the University Hospital of Innsbruck.
+Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, and the KYN/TRP and PHE/TYR ratios calculated [@Capuron2011; @Widner1997; @Neurauter2008].
+Plasma titer of immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) were quantified by ELISA [@Deisenhammer2021]. 
+Laboratory measurements at and beyond the detection limits were substituted with the lower or upper detection limit value, respectively. 
+Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
+Anxiety and depression signs were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
+Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points [@Bjelland2002; @Zigmond1983].
+
+In order to improve normality of some numeric study variables prior to linear modeling and statistical hypothesis testing with parametric tools, logarithm or square root transformations were applied (__Supplementary Table S1__)
+
+### INCOV cohort
+
+Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and ultra-high-performance liquid chromatography/tandem accurate mass spectrometry (Metabolon, USA) [@Su2022; @Su2020].
+Normalized, age- and sex-adjusted, log~2~-transformed plasma concentrations of metabolites and cytokines, and clinical information were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) of the report by Su et al.
+[@Su2022].
+Metabolites implicated in systemic serotonin availability (serotonin, TRP), KYN pathway activity (KYN, quinolinate [QUIN]) and dopamine availability (PHE, TYR and dopamine 3-O-sulfate [DA sulfate]), cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and age were analyzed (__Supplementary Table S3__).
+
+Plasma features were sampled in uninfected controls and SARS-CoV-2 individuals during 
+acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']`) and 
+sub-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']`) infection, and recovery (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR) (__Supplementary Table S3__).
+
+## Analysis endpoints
+
+Our analysis pursued two endpoints. 
+The first endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing plasma levels of serotonin and dopamine precursors and KYN pathway metabolites (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR).
+This endpoint was addressed by multi-parameter modeling in the SIMMUN cohort. 
+The second endpoint was to investigate how more direct readouts of systemic serotonin and dopamine availability, plasma serotonin [@Paul-Savoie2011; @Sun2020; @Tateishi2021; @Holck2019] and DA sulfate [@Meiser2013; @Hashizume1987; @Goldstein1999], are influenced by age, inflammatory cytokines, timepoint of the SARS-CoV-2 infection, neurotransmitter precursors (TRP, PHE, TYR) and KYN pathway products (KYN, QUIN).
+This endpoint was addressed by multi-parameter robust linear modeling, time course modeling and correlation analysis in the INCOV cohort (__Figure \@ref(fig:fig-inclusion)__).
+
+## Statistical analysis
+
+R version 4.2.3 was used for statistical analysis.
+
+Numeric variables were presented as medians with interquartile ranges and ranges. 
+Categorical variables are presented as percentages and counts within the complete observation set. 
+Distribution normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively. 
+In the SIMMUN cohort, non-normally distributed numeric variables were logarithm- or square root-transformed prior to modeling and parametric tests (__Supplementary Table S1__). 
+Since most of the INCOV study variables were non-normally distributed, robust linear modeling and non-parametric testing were employed. 
+Since the tau-equivalence assumption investigated by factor analysis for the HADS scales was violated, consistency of psychometric tools was investigated by global McDonald's $\omega$ [@McDonald1999; @Bartlett1937; @Revelle2022].
+Except for multi-parameter modeling, p values were corrected for multiple testing with the false discovery rate method [@Benjamini1995] separately for each analysis task. 
+Effects with p < 0.05 were considered significant.
+
+Correlation was assessed by Spearman's rank or Pearson test. 
+Significance of comparisons of numeric variables between two groups was determined by Mann-Whitney test with r effect size statistic or by two-tailed T test with Cohen's d effect size metric. 
+Comparisons of categorical variable distributions were evaluated by $\chi^2$ test with Cramer V effect size statistic. 
+Correlation was assessed by Pearson's or Spearman's rank test [@Cohen2013; @Kassambara2021].
+
+In the SIMMUN cohort, effects of age, sex, body mass class, physical and psychiatric disorders, body mass class, smoking and alcohol consumption, inflammation markers (NEO, NLR), SARS-CoV-2 infection, anti-RBD IgG titer, depression and anxiety signs (HADS), mental stress (PSS-4) on TRP, KYN, KYN/TRP, PHE, TYR and PHE/TYR were assessed by multi-parameter linear regression with backward elimination.
+Modeling responses and explanatory variables were subjected to normality-stabilizing transformations (__Supplementary Table S1__) and normalized. 
+Normality and homogeneity of model residuals were evaluated by Shapiro-Wilk and Levene test, respectively, and visual inspection (residuals versus fitted and quantile-quantile plots). 
+In the INCOV cohort, effects of age, cytokine markers of inflammation (IL6, IL10, TNF, IFNG), SARS-CoV-2 infection timepoint (acute, sub-acute, recovery versus uninfected control), neurotransmitter precursors (TRP, PHE, TYR) and KYN pathway products (QUIN, KYN) on plasma serotonin and DA sulfate were modeled by multi-parameter robust linear regression with the MM algorithm and Huber psi function [@Huber2011; @Ripley2022].
+Reproducibility and proper parameterization of the multi-parameter linear and robust models was investigated by RMSE and $R^2$ statistics in 10-fold cross-validation [@Kuhn2008]. 
+Significance of the model estimates was assessed by two-tailed T test.
+
+Differences in cytokines and metabolites between SARS-CoV-2 infection timepoints in the INCOV collective were investigated by robust linear modeling (MM algorithm, Huber's psi function) with uninfected subset or acute infection serving as baselines.
+
+Pairwise Spearman's correlation coefficients $\rho$ were calculated for cytokines and metabolites in the INCOV cohort. 
+The correlation matrices were subsequently scaled into the [0, 1] range and converted to undirected force directed graphs [@Csardi2006]. 
+The graphs were visualized as two-dimensional network plots with the node proximity determined by the $\rho$ value and distance-dependent repulsion, and edge color and width coding for the $\rho$ value and sign [@Briatte2021].
+
+\newpage
+
+# Results
+
+## Characteristic of the study cohorts
+
+Herein, two cohorts of uninfected and SARS-CoV-2-positive individuals were analyzed. 
+Out of `r nrow(stigma$data)` individuals enrolled in the SIMMUN study, `r nrow(stigma$analysis_tbl)` participants with a complete study variable dataset were analyzed (__Figure \@ref(fig:fig-inclusion)__, __Supplementary Table S1__).
+The excluded individuals were characterized by more frequent psychiatric disorders, depression and anxiety signs, elevated mental stress scoring and less frequent SARS-CoV-2 infections as compared with the analyzed participants (__Supplementary Table S2__).
+SARS-CoV-2-positive individuals accounted for 
+`r my_percent(cohort$analysis_tbl$stigma, 'cov')['healthy']`% of the analyzed SIMMUN cohort. 
+None of the participants had received anti-SARS-CoV-2 vaccination prior to enrollment. 
+Males represented 
+`r my_percent(cohort$analysis_tbl$stigma, 'sex')['male']`% of the cohort and the median age was 
+`r median(cohort$analysis_tbl$stigma$age)` years. 
+The gender and age structure of the SARS-CoV-2-negative and -positive subsets was comparable. 
+In the entire SIMMUN cohort, 
+`r signif(sum(table(cohort$analysis_tbl$stigma$bmi_class)[c(2, 3)])/nrow(cohort$analysis_tbl$stigma), 2) * 100`% of participants were overweight or obese, the physical disorder rate was 
+`r my_percent(cohort$analysis_tbl$stigma, 'somatic_comorb')['yes']`%; these figures were comparable between SARS-CoV-2-negative and -positive individuals. 
+Psychiatric disorders affected 
+`r my_percent(cohort$analysis_tbl$stigma, 'psych_comorb')['yes']`% of participants and were significantly more common in SARS-CoV-2-negative 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% my_percent('psych_comorb') %>% .['yes']`%) than in SARS-CoV-2-positive individuals 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('psych_comorb') %>% .['yes']`%, 
+`r cohort$clin_test$stigma %>% filter(variable == 'psych_comorb') %>% .$significance`, 
+effect size: 
+`r cohort$clin_test$stigma %>% filter(variable == 'psych_comorb') %>% .$significance`). 
+Signs of depression and anxiety, and intensity of mental stress were gauged with the HADS depression, HADS anxiety and PSS-4 tools, respectively [@Zigmond1983; @Cohen1983]. 
+The psychometric tools displayed good-to-excellent internal consistency 
+($\omega$ = `r cons$omega_stats %>% filter(factor == 'g') %>% .$total %>% range %>% signif(2) %>% paste(collapse = ' - ')`) [@McDonald1999] (__Supplementary Figure S1__). 
+Clinically relevant anxiety signs (HADS $\geq$ 8) [@Bjelland2002; @Zigmond1983] were more frequent in the uninfected 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% my_percent('hads_anx_score') %>% .[2]`%) than in the SARS-CoV-2-positive subset 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('hads_anx_score') %>% .[2]`%, 
+`r cohort$clin_test$stigma %>% filter(variable == 'hads_anx_score') %>% .$significance`, 
+effect size: 
+`r cohort$clin_test$stigma %>% filter(variable == 'hads_anx_score') %>% .$eff_size`). 
+Depression signs were more common in SARS-CoV-2-negative 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% my_percent('hads_dpr_score') %>% .[2]`%) than in SARS-CoV-2-positive participants 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('hads_dpr_score') %>% .[2]`%), yet this effect was no significant. 
+Scores of mental stress were comparable in both SARS-CoV-2 subsets 
+(negative: median `r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% .$pss_stress_score %>% median`, 
+positive: median `r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% .$pss_stress_score %>% median` points). 
+As expected, titer of antibodies against the S1/S2 SARS-CoV-2 protein (anti-RBD IgG) was significantly higher in the convalescent subset 
+(median: `r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% .$anti_rbd %>% median %>% signif(2)` AU) as compared with uninfected individuals 
+(median: `r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% .$anti_rbd %>% median %>% signif(2)` AU, 
+`r cohort$clin_test$stigma %>% filter(variable == 'anti_rbd') %>% .$significance`, 
+effect size: `r cohort$clin_test$stigma %>% filter(variable == 'anti_rbd') %>% .$eff_size`). 
+In 73% of SARS-CoV-2-infected SIMMUN study participants, the infection was mild and ambulatory (__Table \@ref(tab:tab-stigma-cohort)__).
+
+Out of 
+`r full_join(incov$proteome, incov$metabolome, by = 'patient_id') %>% filter(!duplicated(patient_id)) %>% nrow` individuals initially enrolled in the INCOV study [@Su2020; @Su2022], 
+`r length(unique(incov$complete_ids))` participants with `r length(incov$complete_ids)` samples with a complete study variable dataset were analyzed (__Figure \@ref(fig:fig-inclusion)__, __Supplementary Table S3 - S4__).
+SARS-CoV-2-positive individuals comprised 
+`r my_percent(cohort$analysis_tbl$incov, 'cov')['SARS-CoV-2']`% of the analyzed INCOV collective. 
+Males constituted 
+`r my_percent(cohort$analysis_tbl$incov, 'sex')['male']`% of participants, the median age was 
+`r signif(median(cohort$analysis_tbl$incov$age), 2)` years. 
+SARS-CoV-2-positive participants 
+(median age: `r cohort$analysis_tbl$incov %>% filter(cov == 'SARS-CoV-2') %>% .$age %>% median %>% signif(2)` years) were significantly older than the uninfected subset 
+(median age: `r cohort$analysis_tbl$incov %>% filter(cov == 'healthy') %>% .$age %>% median %>% signif(2)` years, 
+`r cohort$clin_test$incov %>% filter(variable == 'age') %>% .$significance`, 
+effect size: `r cohort$clin_test$incov %>% filter(variable == 'age') %>% .$eff_size`). 
+Shares of overweight or obese individuals were higher among SARS-COV-2-positive 
+(`r cohort$analysis_tbl$incov %>% filter(cov == 'SARS-CoV-2') %>% my_percent('bmi_class') %>% .[c(2, 3)] %>% sum`%) 
+than SARS-CoV-2-negative participants 
+(`r cohort$analysis_tbl$incov %>% filter(cov == 'healthy') %>% my_percent('bmi_class') %>% .[c(2, 3)] %>% sum`%), yet this effect missed statistical significance. 
+Nearly all 
+(`r cohort$analysis_tbl$incov %>% filter(cov == 'SARS-CoV-2') %>% my_percent('severity') %>% .[c('moderate', 'severe', 'critical')] %>% sum`%) of SARS-CoV-2-positive INCOV individuals were hospitalized due to a moderately or critically severe infection (__Table \@ref(tab:tab-incov-cohort)__).
+
+As compared with the SIMMUN collective, the INCOV cohort was characterized by a significantly higher percentage of males (`r comp$test %>% filter(variable == 'sex') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'sex') %>% .$eff_size`), 
+more advanced age (`r comp$test %>% filter(variable == 'age') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'age') %>% .$significance`), 
+elevated rates of overweight or obesity (`r comp$test %>% filter(variable == 'bmi_class') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'bmi_class') %>% .$eff_size`), 
+higher share of SARS-CoV-2 cases (`r comp$test %>% filter(variable == 'cov') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'cov') %>% .$eff_size`) and 
+higher hospitalization rates (`r comp$test %>% filter(variable == 'severity') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'severity') %>% .$eff_size`) (__Supplementary Table S5__).
+
+## Inflammation, SARS-CoV-2 infection, age, mental stress and depression influence systemic levels of neutrotransmitter precursors and kynurenine pathway activity
+
+We employed the SIMMUN dataset in an initial search for predictors of systemic neurotransmitter availability. 
+As markers of serotonin turnover, we modeled plasma levels of the serotonin precursor TRP and readouts of the competitor KYN pathway, KYN and KYN/TRP [@Capuron2011; @Lukic2022; @Badawy2017]. 
+Plasma concentrations of PHE, TYR and PHE/TYR were investigated as markers of dopamine availability [@Meiser2013; @Neurauter2008; @Geisler2013]. 
+The candidate explanatory variables were age, sex, body mass, physical and psychiatric disorders, alcohol and tobacco consumption, SARS-CoV-2 infection, anti-SARS-CoV-2 antibody titer, NEO and NLR as inflammatory markers, mental stress, depression and anxiety signs (__Supplementary Table S1__).
+All those variables were recorded at a single timepoint, median 139 days after the SARS-CoV-2 test (interquartile range: 119 - 157).
+
+Full multi-parameter linear models were optimized by backwards elimination of non-significant terms [@Ripley2022] (__Supplementary Table S6__). 
+Meaningful models could be established for TRP, KYN, KYN/TRP, TYR and PHE/TYR. 
+The PHE model had a negligible explanatory performance, proved non-significant as compared with the intercept-only null model and was hence not further analyzed. 
+The remaining models were characterized by good reproducibility and proper parameterization as indicated by comparable fit errors in the genuine dataset and cross-validation. 
+The KYN and KYN/TRP ratio models had the best explanatory performance measured by cross-validated $R^2$ of 
+`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn') %>% .$estimate %>% signif(2)` and 
+`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn_trp') %>% .$estimate %>% signif(2)`, respectively. 
+The TRP, TYR and PHE/TYR models could explain between 
+`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[1] * 100`% and 
+`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[2] * 100`% of the cross-validated response variances ($R^2$, __Supplementary Figure S2__).
+
+The inflammatory marker NEO 
+($\beta$ = `r my_beta(mod_eli$inference$trp, variable = 'log_neo')`) 
+and depression signs 
+(HADS $\geq$ 8, $\beta$ = `r my_beta(mod_eli$inference$trp, variable = 'hads_dpr_score')`) were independently associated with reduced plasma TRP. 
+Increased KYN and KYN/TRP were significantly associated with 
+SARS-CoV-2 infection (KYN: $\beta$ = `r my_beta(mod_eli$inference$kyn, variable = 'infection')`, 
+KYN/TRP: $\beta$ = `r my_beta(mod_eli$inference$kyn_trp, variable = 'infection')`), 
+elevated NEO (KYN: $\beta$ = `r my_beta(mod_eli$inference$kyn, variable = 'log_neo')`, 
+KYN/TRP: $\beta$ = `r my_beta(mod_eli$inference$kyn_trp, variable = 'log_neo')`) and 
+age (KYN: $\beta$ = `r my_beta(mod_eli$inference$kyn, variable = 'age')`, 
+KYN/TRP: $\beta$ = `r my_beta(mod_eli$inference$kyn_trp, variable = 'age')`). 
+Additionally, high mental stress measured by PSS-4 was linked to increased KYN/TRP 
+($\beta$ = `r my_beta(mod_eli$inference$kyn_trp, variable = 'pss_stress_score')`).  
+Low NLR, in turn, was related to increased KYN (`r my_beta(mod_eli$inference$kyn, variable = 'log_nlr')`) (__Figure \@ref(fig:fig-modeling)A__). 
+Plasma TYR concentrations were negatively associated with the inflammatory marker NEO 
+($\beta$ = `r my_beta(mod_eli$inference$tyr, variable = 'log_neo')`) and rose significantly with age 
+($\beta$ = `r my_beta(mod_eli$inference$tyr, variable = 'age')`). 
+SARS-CoV infection ($\beta$ = `r my_beta(mod_eli$inference$phe_tyr, variable = 'infection')`) 
+and age ($\beta$ = `r my_beta(mod_eli$inference$phe_tyr, variable = 'age')`) were linked to lower PHE/TYR (__Figure \@ref(fig:fig-modeling)B__).
+
+By univariable analysis, we could corroborate significant, positive, moderate-to-strong correlations of age and NEO with increased KYN and higher KYN/TRP in the SIMMUN dataset.
+Participant's age was paralleled by elevated circulating TYR and lowered PHE/TYR in univariable correlation analysis (__Supplementary Figure S3 - S4__, __Supplementary Tables S7 - S8__).
+
+## Inflammatory cytokines, SARS-CoV-2 infection course, age and availability of biosynthesis precursors regulate systemic serotonin and dopamine turnover
+
+Next, we sought to validate the effects of inflammation, SARS-CoV-2 infection and age on systemic serotonin and dopamine availability in the longitudinal INCOV cohort [@Su2022]. 
+The outcome variables in such analyses were plasma levels of serotonin [@Paul-Savoie2011; @Sun2020; @Tateishi2021; @Holck2019; @Lukic2022; @Moncrieff2022], and DA sulfate, a the main circulating dopamine catabolite [@Meiser2013; @Hashizume1987; @Goldstein1999].
+The explanatory variables were age, infection timepoint, major cytokines induced by inflammatory signaling (IL6, IL10, TNF, IFNG), the serotonin precursor TRP, KYN pathway metabolites (KYN, QUIN) [@Capuron2011; @Lukic2022; @Badawy2017], and dopamine precursors (PHE, TYR) [@Meiser2013; @Neurauter2008] (__Supplementary Table S3__).
+Metabolites and cytokines were sampled in SARS-CoV-2 INCOV patients at three timepoints of infection: 
+acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']`), 
+sub-acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']`) and 
+recovery (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after infection diagnosis via PCR) (__Supplementary Table S4__).
+
+Multi-parameter robust regression models [@Huber2011; @Ripley2022] could explain 
+`r (incov_neuro$cv_stats %>% filter(data_type == 'cv', response == 'serotonin', statistic == 'rsq') %>% .$estimate %>% signif(2)) * 100`% and 
+`r (incov_neuro$cv_stats %>% filter(data_type == 'cv', response == 'dopamine', statistic == 'rsq') %>% .$estimate %>% signif(2)) * 100`% of cross-validated variance ($R^2$) of serotonin and DA sulfate, respectively.
+Comparable fit errors in the training dataset and cross-validation indicated good reproducibility and proper parameterization of the models (__Figure \@ref(fig:fig-incov-neuro)A__). 
+In the INCOV collective, 
+`r incov_neuro$inference$serotonin %>% filter(variable != 'Intercept', p_value < 0.05) %>% filter(!duplicated(variable)) %>% nrow` predictors of plasma serotonin were identified. 
+Age ($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'age')`) and 
+the inflammatory cytokine 
+IL6 ($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'IL6_INF')`) were associated with significantly lower serotonin levels, whereas 
+TRP ($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'tryptophan')`) and the SARS-CoV-2 infection recovery 
+($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'timepoint', 'recovery')`) were predictors of increased plasma serotonin. 
+Infection recovery ($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'timepoint', 'recovery')`) 
+and TNF ($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'TNF_INF')`) predicted higher DA sulfate levels, and IFNG was associated with significantly decreased DA sulfate concentrations 
+($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'IFNG_INF')`) (__Figure \@ref(fig:fig-incov-neuro)__, __Supplementary Table S9__).
+
+This association of serotonin and DA sulfate levels wuth inflammation was confirmed by time course modeling. 
+The maximum plasma concentrations of inflammation markers IL6, IL10, TNF and IFNG were observed during acute infection. 
+They were paralleled by a significant decrease of the serotonin precursor 
+TRP ($\beta$ = `r my_beta(time_rlm$inference$uninfected$tryptophan, variables = 'timepoint', levels = 'acute')`) and upregulated TRP catabolism products 
+KYN ($\beta$ = `r my_beta(time_rlm$inference$uninfected$kynurenine, variables = 'timepoint', levels = 'acute')`) and 
+QUIN ($\beta$ = `r my_beta(time_rlm$inference$uninfected$quinolinate, variables = 'timepoint', levels = 'acute')`) as compared with uninfected individuals. 
+The inflammatory milieu of acute infection was associated with a significant rise in 
+PHE ($\beta$ = `r my_beta(time_rlm$inference$uninfected$phenylalanine, variables = 'timepoint', levels = 'acute')` versus uninfected), suggestive of an inhibited PHE - TYR conversion [@Neurauter2008].
+Resolution of systemic inflammation during sub-acute SARS-CoV-2 infection and 
+recovery was reflected by decreasing plasma levels of IL6, IL10, TNF and INFG as compared with acute disease. 
+The inflammation resolution was paralleled by decreased 
+KYN (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$kynurenine, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$kynurenine, 'timepoint', 'recovery')`) and 
+QUIN (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$quinolinate, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$quinolinate, 'timepoint', 'recovery')`), paralleled by elevated 
+TRP (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$tryptophan, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$tryptophan, 'timepoint', 'recovery')`) and 
+serotonin (sub-acute: `r my_beta(time_rlm$inference$acute$serotonin, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$serotonin, 'timepoint', 'recovery')`) as compared with acute SARS-CoV-2 infection. 
+Similarly, infection recovery was associated with significantly decreased 
+PHE (recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$phenylalanine, 'timepoint', 'recovery')`) and elevated 
+DA sulfate (recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$dopamine.3.O.sulfate, 'timepoint', 'recovery')`) as compared with acute infection.
+
+In correlation analysis, the serotonin precursor TRP was negatively associated with all investigated cytokines during 
+acute ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'tryptophan', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) and 
+sub-acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'tryptophan', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`). 
+Plasma serotonin levels correlated negatively with IL6 concentrations during acute infection 
+($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'serotonin', 'IL6_INF')`) and was negatively associated with IL6, IL10, TNF and IFNG in sub-acute infection 
+($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'serotonin', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`). 
+By contrast, the TRP catabolites 
+KYN (acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'kynurenine', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`; 
+sub-acute. $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'kynurenine', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) and 
+QUIN (acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'quinolinate', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`; 
+sub-acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'quinolinate', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) correlated positively with circulating IL6, IL10, TNF and IFNG.
+Serotonin levels correlated with TRP during 
+acute ($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'serotonin', 'tryptophan')`) and 
+sub-acute infection ($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'serotonin', 'tryptophan')`) with weak and moderate effect size, respectively. 
+This may indicate that systemic availability of tryptophan may be particularly crucial for serotonin synthesis at the peak of the infection-related inflammation (__Figure \@ref(fig:fig-incov-corr)A__, __Supplementary Figure S5__).
+Concerning the dopamine availability readouts, positive correlations of PHE with the cytokine markers of inflammation IL6, IL10 and TNF were observed in acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['dopamine.acute']]), 'phenylalanine', c('IL6_INF', 'IL10_INF', 'TNF_INF')) %>% range %>% paste(collapse = ' - ')`).
+PHE was positively associated with IL10 in sub-acute SARS-CoV-2 infection 
+($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['dopamine.sub-acute']]), 'phenylalanine', 'IL10_INF')`). 
+TYR plasma concentrations correlated in turn negatively with TNF and IFNG during acute 
+($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['dopamine.acute']]), 'tyrosine', c('TNF_INF', 'IFNG_INF'))[1]`) and 
+sub-acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['dopamine.sub-acute']]), 'tyrosine', c('TNF_INF', 'IFNG_INF')) %>% range() %>% paste(collapse = ' - ')`). 
+DA sulfate was negatively associated with IFNG and IL6 in acute 
+(IFNG, $\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['dopamine.acute']]), 'dopamine.3.O.sulfate', 'IFNG_INF')`) and sub-acute infection 
+(IL6, $\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['dopamine.sub-acute']]), 'dopamine.3.O.sulfate', 'IL6_INF')`), respectively (__Figure \@ref(fig:fig-incov-corr)B__, __Supplementary Figure S5__).
+
+\newpage
+
+# Discussion
+
+Herein, we explored effects of demographic, clinical, psychometric, infection-, inflammation- and SARS-CoV-2-related factors on systemic availability of serotonin and dopamine in SARS-CoV-2 infection.
+Our results underline the pivotal role of SARS-CoV-2-dependent inflammation in systemic neurotransmitter metabolism. 
+Furthermore, they put forward plasma TRP levels as a one of limiting factors for systemic serotonin biosynthesis during SARS-CoV-2 infection. 
+Age, mental stress and depression were proposed as infection-independent factors contributing to activity of the KYN pathway, lowered TRP and, consequently, reduced serotonin formation. 
+Our data suggest similar inhibitory effects of inflammation and age on PHE - TYR conversion, which is the initial step of dopamine formation.
+
+Two discrepancies of our results need a more thorough discussion. 
+First, in the SIMMUN cohort, the inflammatory marker NEO was associated with higher plasma levels of KYN [@Capuron2011]. 
+Unexpectedly, another inflammatory variable, NLR could be linked to lower KYN. 
+This may be explained e.g. by sustained suppression of neutrophil counts reported for up to 1 year after infection [@Lin2022]. 
+Second, effects of inflammatory markers on neurotransmitter-associated metabolites were stronger in the INCOV than in the SIMMUN cohort. 
+This phenomenon can be traced back to the more severe infection in INCOV study participants.
+
+Mechanistically, multiple inflammatory cytokines such as type I and type II interferon, TNF or IL6, identified here as a predictor of plasma serotonin in the INCOV cohort, signal via JAK/STAT or NK-$\kappa$B pathways, which in turn activate transcription of the _IDO1_ gene [@Robinson2003; @Dai1990; @Litzenburger2014; @Badawy2017].
+The IDO1 protein, together with inflammation-independent enzyme forms, TDO and IDO2, catalyze the first step of the KYN pathway, which catabolizes >90% TRP in the body [@Badawy2017; @Lukic2022; @Sathyasaikumar2022].
+As such, conversion of the serotonin precursor TRP to KYN and subsequent breakdown to downstream KYN pathway products such as 3-hydroxykynurenine, 3-hydroxyanthanilic acid, QUIN or kynurenic acid is believed to limit systemic serotonin production [@Badawy2017; @Lukic2022].
+Additionally, such KYN pathway intermediates were ascribed standalone activity in neuronal signaling, e.g.
+by interfering with glutamatergic NMDA receptors [@Brown2021; @Schwarcz2012; @Vancassel2018].
+Elevated KYN and KYN pathway products were reported for multiple inflammatory conditions [@Badawy2017]. 
+In COVID-19, activity of the KYN pathway was found to correlate positively with inflammatory markers [@Dewulf2022; @Saito2022; @Santiago-Mujika2022], disease severity [@Dewulf2022; @Saito2022; @Lionetto2021] and was implicated in persistent symptom presence [@Bizjak2022; @Gietl2023; @Matits2023; @Cysique2022]
+Interestingly, our INCOV data demonstrating the positive association of serotonin and TRP levels suggest the KYN pathway activity may especially efficiently compete for TRP with systemic serotonin synthesis in the highly inflammatory milieu of acute and sub-acute COVID-19.
+Increasing TRP and serotonin in infection recovery in INCOV study participants suggest re-routing of the from IDO-mediated catabolism to the serotonin biosynthesis [@Capuron2011; @Lukic2022; @Badawy2017].
+The results of multi-parameter modeling in the SIMMUN cohort demonstrate an additional, inflammation-independent effect of SARS-CoV-2 infection on KYN and KYN/TRP. 
+While its mechanism remains obscure, an analogical sustained upregulation of KYN in absence of the inflammatory marker C-reactive protein was described by Bizjak et al.
+in long-term COVID-19 recovery [@Bizjak2022].
+
+Tetrahydrobiopterin (BH4) is a co-factor of PAH, the enzyme catalyzing the PHE - TYR conversion, and of TH, catalyzing the TYR - L-DOPA (L-3,4-dihydroxyphenylalanine) conversion [@Geisler2013; @Fanet2021].
+Importantly, BH4 serves also as a co-factor of TPH, which converts TRP to 5-hydroxytryptophan, an intermediate in serotonin synthesis from TRP [@Fanet2021; @Meiser2013]. 
+BH4 poses, apart from IDO1, another junction between neurotransmitter metabolism and inflammation [@Geisler2013; @Fanet2021; @Werner1990; @Werner-Felmayer1993; @Meiser2013]. 
+Activity of GTP cyclohydrolase I, the enzyme catalyzing the rate limiting step of the BH4 synthesis is strongly stimulated by IFNG and other inflammatory stimuli [@Fanet2021; @Werner1990; @Werner-Felmayer1993].
+However, during inflammation, BH4 is oxidized to the inactive derivative dihydrobiopterin and consumed during nitric oxide formation by the BH4-dependent nitric oxide synthase, which restricts BH4 from neurotransmitter synthesis [@Geisler2013; @Fanet2021; @Vancassel2018; @Huber1984; @Hufner2019; @Neurauter2008].
+Furthermore, upon high activity of GTP cyclohydrolase I, NEO, a side product of BH4 biosynthesis exploited as an inflammation marker in the SIMMUN cohort, is secreted by myeloid leukocytes [@Geisler2013; @Fanet2021; @Vancassel2018; @Huber1984].
+In line with a previous report [@Luporini2021], we could demonstrate an increase in circulating PHE in SARS-CoV-2 infection and associate lowered TYR with NEO. 
+Additionally, we observed negative correlations of plasma levels of the dopamine derivative DA sulfate [@Meiser2013; @Goldstein1999] and TYR with the canonical GTP cyclohydrolase I and neopterin inducer, IFNG.
+Collectively, these phenomena support the model of inflammatory BH4 deficiency culminating at inefficient PHE - TYR conversion, hyperphenylalaninemia and impaired catecholamine biosynthesis during acute SARS-CoV-2 infection.
+By contrast, the rise in DA sulfate during infection recovery in the INCOV cohort and the significantly lower PHE/TYR in SARS-CoV-2 infection convalescents in the SIMMUN collective may indicate a restored BH4 homeostasis leading to an efficient PHE - TYR conversion, dopamine synthesis and subsequent dopamine sulfonylation [@Meiser2013; @Hashizume1987; @Neurauter2008; @Goldstein1999; @Geisler2013].
+
+Rising activity of the KYN pathway with individual's age was evident in both the SIMMUN and INCOV cohort data. 
+This phenomenon was attributed to 'inflammaging', a chronic low-grade inflammation observed in elderly [@Capuron2011; @Sorgdrager2019; @Valdiglesias2018]. 
+Our multi-parameter modeling results suggest, that the lowered TRP and elevated KYN and KYN/TRP in older SIMMUN study participants, and plasma serotonin levels decreasing with age of the INCOV cohort are at least partly independent of inflammatory marker levels.
+This may implicate e.g. increased KYN formation by inflammation-independent IDO2 and TDO in elderly participants. 
+This fits well to the data of Martilla and colleagues, who describe similar expression of _IDO1_ in immune cells from young and elderly individuals [@Marttila2011]. 
+Contrary to the expected age-related suppression of catecholamine synthesis [@Peters2006; @Sorgdrager2019], we observed 
+higher blood levels of TYR and lower PHE/TYR in older participants of the SIMMUN study. 
+However, reproducibility and clinical significance of this phenomenon and is questionable.
+The effect of age on PHE and TYR could not be corroborated in the INCOV collective (not shown), and no significant effects of age on circulating levels of the dopamine derivative DA sulfate were discerned.
+
+Depression [@Paul-Savoie2011; @Hufner2021; @Hufner2015; @Fellendorf2022; @Hunt2020; @Marx2020; @Capuron2011; @Brown2021], 
+depression treatment resistance [@Holck2019; @Tateishi2021; @Sun2020; @Ertugrul2007], 
+schizophrenia [@Marx2020], 
+suicidal behavior [@Tyano2006],  
+pathological mental stress [@Hufner2020] were associated with decreased circulating TRP and serotonin, high KYN pathway activity and increased blood concentrations of neuroactive KYN pathway metabolites.
+Systemic and central nervous system inflammation was put forward as a neuroimmune mechanism contributing to those mental health disorders [@Capuron2011; @Lukic2022; @Dantzer2008; @Miller2016; @Schwarcz2012; @Haroon2012; @Brown2021; @Hunt2020; @Maes2012; @Vancassel2018].
+In infections, inflammation was proposed to suppress serotonin biosynthesis and trigger KYN pathway activity that mediate the 'sickness behavior' characterized by reduced locomotor activity, social avoidance, reduced appetite, lethargy and concentration problems [@Maes2012].
+Inflammatory stimuli were also found to reduce dopamine availability measured by PHE/TYR ratio and hence contribute to depression in cancer [@Hufner2015] and trauma patients [@Hufner2019]. 
+In the SIMMUN cohort, depression signs [@Bjelland2002; @Zigmond1983] and mental stress [@Cohen1983] along with the inflammatory marker NEO could be linked to decreased TRP and higher KYN/TRP, respectively.
+In the INCOV collective, inflammation was also identified as the main diver of alterations in circulating serotonin, its precursor TRP and the competitor pathway products KYN and QUIN. 
+Others also identified lowered TRP in COVID-19 convalescents with signs of depression and anxiety [@Kucukkarapinar2022] and linked KYN pathway activity to depressive symptoms [@Matits2023] and cognitive impairment [@Cysique2022] during long-term COVID-19 recovery.
+These observations may hence support the neuroimmune model assuming inflammation as a link between physical conditions, neurotransmitter disturbance and psychiatric disorders. 
+Still, most evidence for this mechanism and relevance of the peripheral neurotransmitter availability is delivered by observational and in vitro studies, whereas in vivo experimental reports are scarce [@Brown2021; @Schwarcz2012; @Vancassel2018].
+In the periphery, liver, mesenteric organs and vasculature are the main sites of serotonin and dopamine synthesis and catabolism, and KYN pathway activity [@Badawy2017; @Lukic2022; @Meiser2013; @Goldstein1999; @Hashizume1987].
+Although the neurotransmitter precursors TRP and TYR, and neuroactive KYN pathway products were postulated to pass the blood-brain barrier [@Brown2021; @Schwarcz2012; @Vancassel2018], psychiatric disorders are not consistently paralleled by changes in dopamine, serotonin and neuroactive KYN metabolites or expression of KYN pathway enzymes in the central nervous system [@Miller2008; @Clark2016; @Brown2021].
+Of note, quality of evidence of the serotonin theory of depression was also criticized in a recent systematic review [@Moncrieff2022]. 
+Hence, the hypothesis that inflammatory stimuli impact on systemic and central nervous system serotonin and dopamine availability and, hence, contribute to psychiatric disorders, like those frequently observed during COVID-19 recovery [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Bower2022], needs validation in a robust experimental or prospective setting.
+
+# Limitations
+
+Our study has limitations. 
+First, circulating serotonin, dopamine or any abundant dopamine metabolites (e.g.
+DA sulfate) [@Meiser2013; @Hashizume1987; @Goldstein1999], i.e.
+more direct readouts of neurotransmitter turnover than TRP, KYN, PHE and TYR, were not measured in the SIMMUN cohort.
+Additionally, the SIMMUN study variables were recorded at a single timepoint with a highly variable infection - sampling interval. 
+For the INCOV collective, KYN/TRP and PHE/TYR could not be computed with normalized metabolome data, and measurements of stress, depression and anxiety were not available. 
+This incompatibility of the datasets precluded development of comprehensive multi-parameter models in one of the cohorts and subsequent direct validation in the other. 
+Second, relevance of circulating markers serotonin and dopamine availability, and KYN pathway metabolites for the central nervous system and psychiatric disorders is controversial, as discussed above.
+Third, the SIMMUN cohort suffered from a selection bias due to enrichment in psychiatric patients with a high rate of physical and psychiatric disorders. 
+Analogically, hospitalized COVID-19 patients constituted the majority of the INCOV cohort. 
+Hence, none of the collectives is representative for the entire pandemic population. 
+Fourth, both the SIMMUN and INCOV cohorts were recruited during initial phases of the pandemic and do not include any (SIMMUN) or systematically vaccinated patients (INCOV). 
+Similarly, the analyzed cohorts were exposed to wild-type-like SARS-CoV-2 variants and do not allow to assess effects of repeated, seasonal infections with highly transmissible but far less virulent omicron pathogens.
+For these reasons, studies with recent, real world post-pandemic collectives are urgently needed to validate our findings. 
+Finally, the cross-sectional SIMMUN and INCOV cohorts encompassed uninfected controls and SARS-CoV-2 infections ranging from asymptomatic to critical disease. 
+In particular, the shares of pathogen-positive individuals and the infection severity differed significantly between the analyzed collectives, which may have contributes to stronger effects of inflammation on neurotransmitter metabolism observed in the INCOV cohort.
+
+# Conclusions
+
+SARS-CoV-2-dependent inflammation can lower systemic availability of serotonin and dopamine by depletion of the tryptophan via the competitive kynurenine pathway and inhibition of the phenylalanine - tyrosine suppression, respectively.
+Those effects can be further amplified by advanced age, mental stress and depression.  
+It remains to be investigated, if and how this mechanism may contribute to neurotransmitter metabolism in the central nervous system and, consequently, to psychiatric disorders following SARS-CoV-2 infection.
+
+\newpage
+
+# Acknowledgements
+
+We thank all participants and patients for the participation in the study.
+
+# Funding
+
+The study was supported by the Science Fund of the Land Tirol (grant number GZ71134 to Katharina Hüfner).
+
+# Conflict of interest
+
+Katharina Hüfner has received research grants from Austria Wirtchaftsservice (AWS) and the State of Tyrol as well as lecturer's honoraria from Forum Medizinische Fortbildung (FOMF), the Anton Proksch Institute and the Hospital of Schwaz.
+Piotr Tymoszuk owns a data science company, Data Analytics as a Service Tirol, and receives payments from statistical data analysis, bioinformatic and scientific writing services.
+Other authors declare that no conflict of interest exists.
+
+# Data and code availability
 
@@
 
-```{r tab-incov-cohort, tab.cap = 'Characteristic of the external INCOV cohort.
+```{r tab-incov-cohort, tab.cap = 'Characteristic of the INCOV cohort.
 Numeric variables are presented as medians with interquartile ranges.
@@
 
+_Sampling timepoints in the INCOV cohort: acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after diagnosis of SARS-CoV-2 infection via PCR._
+
 _TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine - tryptophan ratio; PHE/TYR: phenylalanine - tyrosine ratio; QUIN: quinolinate; 5-HT: serotonin; DA: dopamine; Ab: antibody; IL6: interleukin-6; IL10: interleukin-10; TNF: tumor-necrosis factor alpha; IFNG: interferon gamma._
@@
 
-_Effects of systemic inflammation markers (neopterin [NEO], neutrophil-lymphocyte ratio [NLR]), SARS-CoV-2 infection status, titre of immunoglobulin gamma against the receptor binding domain of the S1 SARS-CoV-2 protein (anti-RBD IgG), scores of anxiety, depression (hospital depression and anxiety scale [HADS]) and mental stress (perceived stress scale, 4 item [PSS4]), age and sex was investigated by multi-parameter linear regression with backward elimination of non-significant terms.
-Overall model validity was assessed by likelihood-ratio test (LRT).
-Significant model coefficient estimates with 95% confidence intervals are presented in Forest plots.
+_Effects of systemic inflammation markers (neopterin [NEO], neutrophil-lymphocyte ratio [NLR]), SARS-CoV-2 infection status, titre of immunoglobulin gamma against the receptor binding domain of the S1/S2 SARS-CoV-2 protein (anti-RBD IgG), symptoms of anxiety and depression (hospital depression and anxiety scale [HADS] $\geq$ 8 points), intensity of mental stress (perceived stress scale, 4 item [PSS4]), age and sex on readouts of serotonin and dopamine availability and KYN pathway activity were investigated by multi-parameter linear regression with backward elimination of non-significant terms.
+Numeric variables were normalized prior to modeling.
+Overall model validity was assessed by likelihood-ratio test (LRT) as compared with the respective null models.
+Significant coefficient estimates ($\beta$) with 95% confidence intervals are presented in Forest plots.
 LRT results (F values, degrees of freedom and p values) and total numbers of complete observations are indicated in the plot captions.
@@
 
-```{r fig-uni-trp, fig.width = figur::convert(figs$uni_trp, to = 'in')$w, fig.height = figur::convert(figs$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}
-
-figs$uni_trp$plot
-
-```
-
-__Figure \@ref(fig:fig-uni-trp).
-Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__
-
-_Age, systemic inflammatory markers: neopterin (NEO), neutrophil - lymphocyte ratio (NLR), SARS-CoV-2 infection status, depression signs (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of systemic levels of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP).
-Their association with serum concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
-
-_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method.
-(A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
-(B) Scatter plots of serum concentrations of NEO, TRP, KYN and KYN/TRP.
-Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
-Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._
-
-_(C) Levels of the metabolite levels were compared between infected and uninfected participants, and between participants with and without depression signs by two-tailed T test with Cohen's d effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points.
-Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the strata are indicated in the X axes._
-
-\newpage
-
-```{r fig-uni-tyr, fig.width = figur::convert(figs$uni_tyr, to = 'in')$w, fig.height = figur::convert(figs$uni_tyr, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.'}
-
-figs$uni_tyr$plot
-
-```
-
-__Figure \@ref(fig:fig-uni-tyr). Effects of age, serum inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.__
-
-_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of systemic levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR).
-Their association with serum concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
-
-_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method.
-(A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
-(B) Scatter plots of patient's age, PHE, TYR and PHE/TYR.
-Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
-Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._
-
-_(C) Levels of the metabolite levels were compared between infected and uninfected participants by two-tailed T test with Cohen's d effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points.
-Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the strata are indicated in the X axes._
-
-\newpage
-
-```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of serum levels of serotonin and dopamine sulfate in the INCOV cohort.'}
+```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine 3-O-sulfate in the INCOV cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of serum levels of serotonin and dopamine sulfate in the INCOV cohort.__
+__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine 3-O-sulfate in the INCOV cohort.__
 
-_Effects of patient's age, timepoint after SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection as compared with healthy controls), serum levels of metabolites related to neurotransmitter biosynthesis and competitor pathways (tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE] and tyrosine [TYR]), and serum concentrations of inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on serum concentrations of serotonin (5-hydroxy tryptophan [5-HT]) and dopamine sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
+_Effects of patient's age, timepoint after SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR as compared with healthy controls), plasma levels of metabolites related to neurotransmitter biosynthesis and KYN pathway activity (tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE] and tyrosine [TYR]), and plasma concentrations of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin (5-hydroxy tryptamine [5-HT]) and dopamine 3-O-sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
 
@@
 
-_(B) Estimated of model coefficients with 95% confidence intervals presented in Forest plots.
+_(B) Estimates of model coefficients ($\beta$) with 95% confidence intervals presented in Forest plots.
 Numbers of complete observations are indicated in the plot captions.
@@
 
-```{r fig-time-ourse, fig.width = figur::convert(figs$time_course, to = 'in')$w, fig.height = figur::convert(figs$time_course, to = 'in')$h, fig.cap = 'Time course of inflammatory cytokines, serotonin, dopamine sulfate and the neurotransmitter biosynthesis precursors and competitor pathway products during SARS-CoV infection and recovery in the INCOV cohort.'}
+```{r fig-time-ourse, fig.width = figur::convert(figs$time_course, to = 'in')$w, fig.height = figur::convert(figs$time_course, to = 'in')$h, fig.cap = 'Time course of cytokine markers of inflammation, serotonin, dopamine 3-0-sulfate and the neurotransmitter biosynthesis precursors and kynurenine pathway products during SARS-CoV infection and recovery in the INCOV cohort.'}
 
@@
 __Figure \@ref(fig:fig-time-ourse).
-Time course of inflammatory cytokines, serotonin, dopamine sulfate and the neurotransmitter biosynthesis precursors and competitor pathway products during SARS-CoV infection and recovery in the INCOV cohort.__
+Time course of cytokine markers of inflammation, serotonin, dopamine 3-0-sulfate and the neurotransmitter biosynthesis precursors and kynurenine pathway products during SARS-CoV infection and recovery in the INCOV cohort.__
 
-_Time courses of normalized log~2~-transformed serum concentrations of inflammatory cytokines (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF] and interferon gamma [INFG]), metabolites implicated in serotonin synthesis (B, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN] and serotonin/5-hydroxy tryptophan [5-HT]) and metabolites implicated in dopamine turnover (C, phenylalanine [PHE], tyrosine [TYR] and dopamine sulfate [DA sulfate]).
-Differences between uninfected controls and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after infection, respectively) were modeled by robust linear regression with the uninfected controls as baseline.
+_Time courses of normalized log~2~-transformed plasma concentrations of cytokine makers of inflammation (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF] and interferon gamma [INFG]), metabolites implicated in serotonin synthesis (B, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN] and serotonin/5-hydroxy tryptamine [5-HT]) and metabolites implicated in dopamine turnover (C, phenylalanine [PHE], tyrosine [TYR] and dopamine 3-O-sulfate [DA sulfate]).
+Differences between uninfected controls and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after SARS-CoV-2 infection diagnosis via PCR, respectively) were modeled by robust linear regression with the uninfected controls as baseline.
 Mean normalized concentrations with 2 standard errors of the mean (2 $\times$ SEM) are depicted as solid lines with tinted ribbons (plots on the left).
-Robust linear model coefficient estimates with 95% confidence intervals are visualized in Forest plots (right).
+Robust linear model coefficient estimates ($\beta$) with 95% confidence intervals are visualized in Forest plots (right).
 Numbers of complete observations per timepoint are indicated in the X axes._
@@
 
-```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of serum levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.'}
+```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of plasma levels of cytokine markers of inflammation, serotonin and dopamine 3-O-sulfate, neurotransmitter precursors and kynurenine pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.'}
 
@@
 __Figure \@ref(fig:fig-incov-corr).
-Correlation of serum levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.__
+Correlation of plasma levels of cytokine markers of inflammation, serotonin and dopamine 3-O-sulfate, neurotransmitter precursors and kynurenine pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.__
 
-_Serum levels of metabolites implicated in serotonin synthesis (A, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin [5-hydroxy tryptophan, 5-HT]), dopamine turnover (B, phenylalanine [PHE], tyrosine [TYR], dopamine sulfate [DA sulfate]) and inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in uninfected and SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection).
-Correlation coefficient matrices for at least moderate effects ($\rho$ > 0.2) were visualized as force-directed network plots.
-Node color codes for the parameter type (gray: inflammatory cytokines, orange: serotonin, green: dopamine turnover), edge width and color codes for the value and sign of the correlation coefficient._
+_Plasma levels of metabolites implicated in serotonin synthesis (tryptophan [TRP], serotonin [5-hydroxy tryptamine, 5-HT]), kynurenine pathway activity (kynurenine [KYN], quinolinic acid [QUIN]) and dopamine metabolism (phenylalanine [PHE], tyrosine [TYR], dopamine sulfate [DA sulfate]) and cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in uninfected and SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR).
+Correlation coefficient matrices for correlation coefficients $\rho$ > 0.2 were visualized as force-directed network plots.
+Node color codes for the variable type (gray: inflammatory markers, orange: serotonin metabolism and KYN pathway, green: dopamine metabolism), edge width and color codes for the value and sign of the correlation coefficient._
 
@@
 
-# Acknowledgements
-
-The scheme of report findings was created with bioicons by Simon Duerr. 
-The SARS-CoV-2 virus image was created by Hanna Vega and licensed under the terms of a Creative Commons CC-BY SA 4.0.
-
 # References

### 2023-05-23 paper/markdown/supplementary_material.Rmd
Manuscript and Rebuttal Letter

@@
 ---
-title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic levels of aminoacid neurotransmitter precursors"
+title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism"
 subtitle: "Supplementary Material"
@@
 Figures and tables were created with _cowplot_ [@Wilke2019] and _flextable_ [@Gohel2022], respectively. 
-Parts of the manuscript and Supplementary Material were written in the _rmarkdown_ environment [@Allaire2022] and rendered with the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages. 
+The manuscript and Supplementary Material were written in the _rmarkdown_ environment [@Allaire2022] and rendered with the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages, with the `author-info-blocks.lua` and `scholarly-metadata.lua` Pandoc filters developed by Albert Krewinkel, Robert Winkler and Jörn Krenzer.
 Management of figures and tables in the rmarkdown documents was accomplished with [_figur_](https://github.com/PiotrTymoszuk/figur).
@@
 The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
-The SIMMUN study data were gathered during in-person visits at the study center conducted between 17.June 2020 and 27.May 2021. 
+The SIMMUN study data were gathered during a single on-site study visit at the study center at median 139 days after the SARS-CoV-2 test (interquartile range: 119 - 157). 
+The study visits were conducted between 17.June 2020 and 27.May 2021. 
 The SIMMUN analysis inclusion scheme is presented in __Figure 1__. 
@@
 
-Demographic and clinical history variables: age, sex, body mass index (BMI), professionally diagnosed mental illness, self-reported chronic somatic conditions, smoking and alcohol consumption history were surveyed during the study visit or extracted from electronic patient records.
+Demographic and clinical history variables: age, sex, body mass index (BMI), professionally diagnosed psychiatric disorders, self-reported chronic physical disorders, smoking and alcohol consumption history were surveyed during the study visit or extracted from electronic patient records.
 Body mass classes were defined as follows: normal for BMI < 25 kg/m^2^, overweight for BMI 25 - 30 kg/m^2^ and obesity for BMI > 30 kg/m^2^. 
@@
 `r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
-elevated IL6 > 7 pg/mL observed in 
+elevated IL6 > 7 pg/mL was observed in 
 `r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of the analyzed SIMMUN collective. 
 Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, as described elsewhere [@Widner1997; @Neurauter2008].
-Kynurenine - tryptophan and phenylalanine - tyrosine ratios (KYN/TRP and PHE/TYR) serving as readouts of systemic indoleamine 2,3-dioxygenase (IDO) and phenylalanine hydroxylase (PAH) were computed [@Capuron2011].
+Kynurenine - tryptophan and phenylalanine - tyrosine ratios (KYN/TRP and PHE/TYR) serving as readouts of systemic indoleamine 2,3-dioxygenase (IDO) and phenylalanine hydroxylase (PAH) activity were computed [@Capuron2011].
 Plasma titer of anti-RBD IgG were quantified by ELISA as described before [@Deisenhammer2021]. 
@@
 In the current analysis, `r nrow(incov$analysis_tbl)` INCOV study samples obtained for 
-`r length(unique(incov$complete_ids))` individuals with the complete dataset of metabolites of interest (serotonin, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE], tyrosine [TYR] and dopamine 3-O-sulfate [DA sulfate]), inflammatory cytokines of interest (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and complete information of age at enrollment were included.
+`r length(unique(incov$complete_ids))` individuals with the complete dataset of metabolites of interest (serotonin, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE], tyrosine [TYR] and dopamine 3-O-sulfate [DA sulfate]), inflammatory cytokine markers of interest (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and complete information of age at enrollment were included.
 The INCOV study variables are listed in __Supplementary Table S\@ref(tab:tab-incov-study-vars)__. 
@@
 Concerning the INCOV study design, plasma metabolites and proteome were determined in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
-acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['acute']` days), 
-sub-acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['sub-acute']` days) and 
-recovery (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['recovery']` days after diagnosis). 
+acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']` days), 
+sub-acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']` days) and 
+recovery (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR). 
 The measurements were only partially matched by participant (i.e. multiple longitudinal measurements for an individual) hence precluding a classical repeated measurement analysis approach. 
@@
 
-Numeric variables were presented in the text and tables as medians with interquartile ranges and ranges. 
+Numeric variables were presented in the tables as medians with interquartile ranges and ranges. 
 Categorical variables are presented as percentages and counts within the complete observation set. 
@@
 
-Effects of age, sex, body mass class, somatic and metal conditions, body mass class, smoking and alcohol consumption history, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, signs of depression and anxiety (HADS), mental stress scoring (PSS-4) on systemic levels of aminoacid neurotransmitter precursor and products of competitor pathways (TRP, KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio) was assessed by multi-parameter linear regression with backward elimination.
+Effects of age, sex, body mass class, physical and psychiatric disorders, body mass class, smoking and alcohol consumption history, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, signs of depression and anxiety (HADS), mental stress scoring (PSS-4) on systemic levels of amino acid neurotransmitter precursor and products of competitor pathways (TRP, KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio) was assessed by multi-parameter linear regression with backward elimination.
 Responses and numeric explanatory variables were transformed with log or squared root functions to improve normality as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__ and normalized prior to modeling.
@@
 Full models including the complete set of explanatory variables listed above were constructed (function `make_lm()`, package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) and optimized by Bayesian information criterion (BIC) driven backwards elimination of non-significant terms (method `step()`, [_lmqc_](https://github.com/PiotrTymoszuk/lmqc), employing internally `stepAIC()` from the _MASS_ package) [@Ripley2022].
-Normality and homogeneity of distribution of the model residuals was checked by Shapiro-Wilk and Levene test, respectively (method `summary(type = 'assumptions')`, package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) and visually inspected in standard diagnostic plots of model residuals (residuals vs fitted, quantile-quantile plots, method `plot()` called for the model objects).
+Normality and homogeneity of distribution of the model residuals were checked by Shapiro-Wilk and Levene tests, respectively (method `summary(type = 'assumptions')`, package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) and visually inspected in standard diagnostic plots of model residuals (residuals vs fitted, quantile-quantile plots, method `plot()` called for the model objects).
 Fit stats ($R^2$ and root mean squared error [RMSE]) were retrieved from the model objects by calling `summary(type = 'fit')` (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)). 
@@
 Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on systemic neurotransmitter availability were subsequently tested in the publicly available INCOV collective [@Su2022].
-To this end we modeled normalized plasma levels of serotonin as the end product of serotonin biosynthesis associated with mental disorders [@Holck2019; @Tateishi2021; @Sun2020; @Kheirouri2016; @Paul-Savoie2011; @Ertugrul2007], and the major product of systemic dopamine decay dopamine 3-O-sulfate (DA-sulfate) [@Goldstein1999; @Meiser2013].
-The explanatory variables were age, plasma levels of inflammatory cytokines (IL6, IL10, TNF, IFNG), plasma concentrations of neurotransmitter precursors and competitor pathway products (for serotonin: TRP, KYN and quinolinate [QUIN], for DA sulfate: PHE and TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control).
+To this end we modeled normalized plasma levels of serotonin as the end product of serotonin biosynthesis associated with psychiatric disorders [@Holck2019; @Tateishi2021; @Sun2020; @Kheirouri2016; @Paul-Savoie2011; @Ertugrul2007], and the major product of systemic dopamine catabolism, dopamine 3-O-sulfate (DA-sulfate) [@Goldstein1999; @Meiser2013].
+The explanatory variables were age, plasma levels of cytokines induced by inflammatory signaling (IL6, IL10, TNF, IFNG), plasma concentrations of neurotransmitter precursors and KYN pathway products (for serotonin: TRP, KYN and quinolinate [QUIN], for DA sulfate: PHE and TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control).
 Since most numeric features were non-normally distributed, MM algorithm robust linear regression with Huber's psi function was chosen [@Huber2011; @Ripley2022] as a modeling approach.
@@
 
-## Time course of inflammatory cytokines and metabolites in the INCOV cohort
+## Time course of cytokines and metabolites in the INCOV cohort
 
-Differences in normalized blood concentrations of inflammatory cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset or acute infection serving as baselines.
-Models were constructed with the `make_lm()` function (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) employing internally the MM`rlm()` algorithm with Huber's psi function (package _MASS_) [@Huber2011; @Ripley2022].
+Differences in normalized blood concentrations of cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset or acute infection serving as baselines.
+Models were constructed with the `make_lm()` function (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) employing internally the MM `rlm()` algorithm with Huber's psi function (package _MASS_) [@Huber2011; @Ripley2022].
 Inference statistics were computed as described for multi-parameter robust linear modeling. 
@@
 
-Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of inflammatory cytokines (IL6, IL10, TNF and IFNG), and metabolite involved in serotonin (serotonin, TRP, KYN and QUIN) and dopamine turnover (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
+Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of cytokines (IL6, IL10, TNF and IFNG), and metabolite involved in serotonin (serotonin, TRP, KYN and QUIN) and dopamine turnover (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
 The correlation matrices were subsequently scaled into the [0, 1] range with the $0.5\times(x + 1)$ function and converted to undirected force directed graphs with the `graph_from_adjacency_matrix()` function from the _igraph_ package [@Csardi2006].
 The graphs were visualized with tools provided by the _ggnetwork_ package [@Briatte2021] as two-dimensional network plots with the node proximity determined by the $\rho$ value and distance-dependent repulsion, and edge color and width coding for the $\rho$ value and sign (__Figure 5__).
+Additionally, correlation analysis results were presented as correlograms in __Supplementary Figure \@ref(fig:fig-corr-invcov)__.
 
@@
 
-```{r tab-excluded, tab.cap = 'Significant differences between participant of the SIMMUN study included in the analysis and SIMMUN participants excluded due to data missingness.
+```{r tab-excluded, tab.cap = 'Significant differences between participants of the SIMMUN study included in the analysis and SIMMUN participants excluded due to data missingness.
 Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.
@@
 
-```{r tab-univariate-correlation, tab.cap = "Correlation of age, stress scoring blood neopterin levels and neutrophils - leukocyte ratio with plasma metabolite levels in the SIMMUN cohort investigated by Pearson's test."}
+```{r tab-univariate-correlation, tab.cap = "Correlation of age, stress scoring, blood neopterin levels and neutrophil - leukocyte ratio with plasma metabolite levels in the SIMMUN cohort investigated by Pearson's test."}
 
@@
 
-```{r tab-modeling-incov, tab.cap = 'Results of multi-parameter robust linear modeling of plasma concentrations of serotonin, and dopamine sulfate in the INCOV cohort.'}
+```{r tab-modeling-incov, tab.cap = 'Results of multi-parameter robust linear modeling of plasma concentrations of serotonin, and dopamine 3-O-sulfate in the INCOV cohort.'}
 
@@
 
-```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokines, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.
+```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokine markers, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.
 The uninfected subset served as a baseline.'}
@@
 
-```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokines, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.
+```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokine markers, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.
 The acute SARS-CoV-2 infection subset served as a baseline.'}
@@
 
-_Multi-parameter linear regression models of plasma levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by repeated cross-validation (CV, 10 folds).
+_Multi-parameter linear regression models of plasma levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by cross-validation (CV, 10 folds).
 Values of root mean square error (RMSE, A) and $R^2$ (B) in the training data set and CV are plotted._
@@
 Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the strata are indicated in the X axes._
+Numbers of observations are indicated in the X axes._
 
@@
 Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the strata are indicated in the X axes._
+Numbers of observations are indicated in the X axes._
+
+\newpage
+
+```{r fig-corr-invcov, fig.width = figur::convert(suppl$corr_incov, to = 'in')$w, fig.height = figur::convert(suppl$corr_incov, to = 'in')$h, fig.cap = 'Correlation of metabolites with cytokine markers of inflammation in the INCOV cohort.'}
+
+suppl$corr_incov$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-corr-invcov). Correlation of metabolites with cytokine markers of inflammation in the INCOV cohort.__
+
+_Plasma levels of metabolites implicated in serotonin biosynthesis and the serotonin biosynthesis competitor kynurenine pathway (5-hydroxy tryptamine/serotonin [5-HT], tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN]) and metabolites related to dopamine metabolism (phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]) were correlated with plasma levels of inflammatory cytokine markers (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) in uninfected individuals, and during acute (median 10 days), sub-acute (median 14 days) SARS-CoV-2 infection and recovery (median 64 days after SARS-CoV-2 infection diagnosis via PCR).
+Statistical significance was determined by Spearman's rank test corrected for multiple testing with the false discovery rate method.
+Correlation coefficients $\rho$ are presented in bubble plots with point size and color corresponding to values of correlation coefficients; points are labeled with their $\rho$ values, significant effects are highlighted in bold._
 

### 2023-05-23 exploration scripts/cohort.R

@@ -18,8 +18,8 @@
       bmi = 'BMI, kg/m\u00B2',
       bmi_class = 'Body mass class',
       ethnics = 'Ethnics',
-      somatic_comorb = 'Somatic comorbidity',
-      psych_comorb = 'Psychiatric comorbidity',
+      somatic_comorb = 'Physical disorder',
+      psych_comorb = 'Psychiatric disorder',
       hads_anx_score = 'HADS anxiety score',
       hads_dpr_score = 'HADS depression score',
       hads_signs = 'Depression or anxiety signs, HADS \u2265 8',
@@ -184,9 +184,9 @@
     summarize(median = median(time_po),
               lower_q = quantile(time_po, 0.25, na.rm = TRUE),
               upper_q = quantile(time_po, 0.75, na.rm = TRUE)) %>%
-    mutate(days_pi = paste0(signif(median, 2),
-                            ' [', signif(lower_q, 2),
-                            ' - ', signif(upper_q, 2),
+    mutate(days_pi = paste0(round(median),
+                            ' [', round(lower_q),
+                            ' - ', round(upper_q),
                             ']'),
            days_pi = ifelse(timepoint == 'healthy', NA, days_pi))
 

### 2023-05-23 exploration scripts/correlation.R

@@ -1,17 +1,17 @@
-# Inter-correlations of each of the outcomes, inflammatory markers
+# Inter-exp_correlations of each of the outcomes, inflammatory markers
 # and psychometric scores
 
   insert_head()
 
 # container ------
 
-  corr <- list()
+  exp_corr <- list()
 
 # analysis globals ------
 
   insert_msg('Analysis globals')
 
-  corr$variables <-
+  exp_corr$variables <-
     list(outcomes = c('trp', 'log_kyn', 'log_kyn_trp',
                       'log_phe', 'log_tyr', 'sqrt_phe_tyr'),
          inflammation = c('log_neo', 'log_nlr'),
@@ -19,16 +19,20 @@
                          'hads_anx_score',
                          'hads_dpr_score'))
 
-  corr$pairs <- corr$variables %>%
+  exp_corr$pairs <- exp_corr$variables %>%
     map(~combn(.x, m = 2, simplify = FALSE))
+
+  exp_corr$pairs$all <- exp_corr$variables %>%
+    reduce(union) %>%
+    combn(m = 2, simplify = FALSE)
 
 # Correlation -----
 
   insert_msg('Correlation')
 
-  corr$test <-
-    list(x = corr$pairs,
-         y = c('pearson', 'pearson', 'spearman')) %>%
+  exp_corr$test <-
+    list(x = exp_corr$pairs,
+         y = c('pearson', 'pearson', 'spearman', 'spearman')) %>%
     pmap(function(x, y) x %>%
            map_dfr(~correlate_variables(stigma$analysis_tbl,
                                         variables = .x,
@@ -36,7 +40,7 @@
                                         type = y,
                                         pub_styled = FALSE))) %>%
     compress(names_to = 'task') %>%
-    mutate(task = factor(task, names(corr$variables))) %>%
+    mutate(task = factor(task, names(exp_corr$pairs))) %>%
     re_adjust %>%
     mutate(significant = ifelse(p_adjusted < 0.05,
                                 'yes', 'no'),
@@ -48,14 +52,17 @@
 
   insert_msg('Correlograms')
 
-  corr$plots <-
-    list(x = corr$test,
-         y = c('Metabolites',
-               'Inflammatory markers',
-               'Psychometry'),
-         z = corr$variables,
+  exp_corr$plots <-
+    list(x = exp_corr$test,
+         y = c('Metabolites, INCOV',
+               'Inflammatory markers, INCOV',
+               'Psychometry, INCOV',
[…]

### 2023-05-23 import.R

@@ -86,8 +86,8 @@
       'hads_anx_score' = 'Anxiety, HADS',
       'hads_dpr_score' = 'Depression, HADS',
       ## comorbidity, medical history
-      'psych_comorb' = 'Mental illness',
-      'somatic_comorb' = 'Somatic illness',
+      'psych_comorb' = 'Psychiatric disorder',
+      'somatic_comorb' = 'Physical disorder',
       'bmi_class' = 'BMI',
       'smoking' = 'Smoking',
       'alcohol' = 'Alcohol') %>%

### 2023-05-23 incov scripts/correlation.R

@@ -117,7 +117,7 @@
 
   insert_msg('Bubble plots')
 
-  cyt_corr$bubbles <-
+  corr$bubbles <-
     list(data = corr$test %>%
            map(mutate,
                variable1 = variable2,

### 2023-05-23 incov scripts/cytokine_correlation.R

@@ -21,13 +21,13 @@
   ## variable pairs
 
   cyt_corr$pairs <- globals$incov_proteins %>%
-    map(function(prot) globals$incov_metabolites[!globals$incov_metabolites %in% c('serotonin', 'dopamine 3-O-sulfate')] %>%
+    map(function(prot) globals$incov_metabolites %>%
           map(~c(prot, .x))) %>%
     unlist(recursive = FALSE)
 
-# cyt_correlation analysis ------
+# Correlation analysis ------
 
-  insert_msg('correlation')
+  insert_msg('Correlation')
 
   cyt_corr$test <- cyt_corr$analysis_tbl %>%
     map(function(dat) cyt_corr$pairs %>%
@@ -59,7 +59,21 @@
   insert_msg('Bubble plots')
 
   cyt_corr$bubbles <-
-    list(data = cyt_corr$test,
+    list(data = cyt_corr$test %>%
+           map(mutate,
+               variable2 = factor(variable2,
+                                  c('phenylalanine',
+                                    'tyrosine',
+                                    'dopamine 3-O-sulfate',
+                                    'quinolinate',
+                                    'kynurenine',
+                                    'tryptophan',
+                                    'serotonin')),
+               variable1 = factor(variable1,
+                                  c('IL6_INF',
+                                    'IL10_INF',
+                                    'TNF_INF',
+                                    'IFNG_INF'))),
          plot_title =  c('uninfected, INCOV',
                          'acute CoV, INCOV',
                          'sub-acute CoV, INCOV',
@@ -68,24 +82,27 @@
          signif_only = FALSE,
          rotate_x_labs = FALSE)
 
-  ## extra adjustment of scales
+
+  ## faceting for easier interpretation
+
+  for(i in names(cyt_corr$bubbles)) {
+
+    cyt_corr$bubbles[[i]]$data <-
+      cyt_corr$bubbles[[i]]$data %>%
+      mutate(pathway = ifelse(variable2 %in% c('tryptophan',
+                                               'kynurenine',
+                                               'quinolinate',
+                                               'serotonin'),
+                              'serotonin', 'dopamine'),
+             pathway = factor(pathway, c('serotonin', 'dopamine')))
+
+  }
 
   cyt_corr$bubbles <- cyt_corr$bubbles %>%
     map(~.x +
-          scale_y_discrete(limits = c('phenylalanine',
-                                      'tyrosine',
-                                      'tryptophan',
-                                      'kynurenine',
-                                      'quinolinate'),
[…]

### 2023-05-23 paper scripts/figures.R

@@ -18,7 +18,7 @@
               caption = paste('Inclusion scheme for the SIMMUN',
                               'and INCOV cohorts, and analysis strategy.'),
               w = 180,
-              h = 180 * 2946/4098)
+              h = 180 * 3426/4098)
 
 # Figure 2: linear modeling -----
 
@@ -119,6 +119,7 @@
 
   figs$incov_neuro$bottom <- figs$incov_neuro$bottom %>%
     map(~.x +
+          labs(title = paste0(.x$labels$title, ', INCOV'))  +
           expand_limits(x = -0.5) +
           expand_limits(x = 1.11) +
           facet_grid(subset ~ .,

### 2023-05-23 paper scripts/render.R

@@ -22,6 +22,14 @@
   insert_msg('Rendering the paper')
 
   render('./paper/markdown/main_manuscript.Rmd',
+         output_format = my_word(),
+         output_dir = './paper')
+
+# rebuttal letter -----
+
+  insert_msg('Rendering the Rebuttal Letter')
+
+  render('./paper/markdown/rebuttal_letter.Rmd',
          output_format = word_document2(number_sections = FALSE,
                                         reference_docx = 'ms_template.docx'),
          output_dir = './paper')

### 2023-05-23 paper scripts/supplement.R

@@ -11,12 +11,16 @@
   insert_msg('Figure S1: consistency of psychometric tools')
 
   suppl$consistency <- cons$fa_loading_plots %>%
-    map(~.x + theme(plot.tag = element_blank())) %>%
+    map(~.x +
+          labs(title = paste0(.x$labels$title, ', SIMMUN')) +
+          theme(plot.tag = element_blank())) %>%
     plot_grid(plotlist = .,
               ncol = 2,
               align = 'hv',
               axis = 'tblr') %>%
-    plot_grid(plot_grid(cons$omega_plot,
+    plot_grid(plot_grid(cons$omega_plot +
+                          labs(title = paste0(cons$omega_plot$labels$title,
+                                              ', SIMMUN')),
                         ncol = 2),
               nrow = 2,
               rel_heights = c(1, 0.7),
@@ -68,7 +72,9 @@
                 theme(plot.subtitle = element_text(size = 7.4),
                       legend.position = 'none') +
                 labs(subtitle = paste('Pearson correlation, n =',
-                                      nrow(stigma$analysis_tbl))),
+                                      nrow(stigma$analysis_tbl)),
+                     title = paste0(tst_trp$correlation$correlogram$labels$title,
+                                    ', SIMMUN')),
               get_legend(tst_trp$correlation$correlogram),
               ncol = 2,
               rel_widths = c(0.65, 0.35),
@@ -81,7 +87,10 @@
     tst_trp$correlation$plots[c("log_neo|trp",
                                 "log_neo|log_kyn",
                                 "log_neo|log_kyn_trp")] %>%
-    map(~.x + theme(plot.subtitle = element_text(size = 7.4))) %>%
+    map(~.x +
+          labs(title = paste0(.x$labels$title,
+                              ', SIMMUN')) +
+          theme(plot.subtitle = element_text(size = 7.4))) %>%
     plot_grid(plotlist = .,
               ncol = 3,
               align = 'hv',
@@ -90,7 +99,9 @@
   ## bottom panel: comparisons
 
   suppl$uni_trp$bottom <- tst_trp$comparison$plots %>%
-    map(~.x + theme(plot.subtitle = element_text(size = 7))) %>%
+    map(~.x +
+          labs(title = paste0(.x$labels$title, ', SIMMUN')) +
+          theme(plot.subtitle = element_text(size = 7))) %>%
     plot_grid(plotlist = .,
               ncol = 3,
               align = 'hv',
[…]

### 2023-05-23 paper scripts/tables.R

@@ -30,8 +30,8 @@
                            'Sex',
                            'Age, years',
                            'Body mass class',
-                           'Somatic comorbidity',
-                           'Psychiatric comorbidity',
+                           'Physical disorder',
+                           'Psychiatric disorder',
                            'HADS anxiety score',
                            'HADS depression score',
                            'Depression or anxiety signs, HADS ≥ 8',

### 2023-05-23 testing scripts/trp_system.R

@@ -141,7 +141,8 @@
     list(variable = tst_trp$comparison$test$variable,
          split_factor = tst_trp$comparison$test$split_factor,
          plot_title = exchange(tst_trp$comparison$test$variable,
-                               dict = globals$stigma_lexicon),
+                               dict = globals$stigma_lexicon,
+                               value = 'base_label'),
          plot_subtitle = tst_trp$comparison$test$plot_cap,
          y_lab = exchange(tst_trp$comparison$test$variable,
                           dict = globals$stigma_lexicon,

### 2023-05-23 testing scripts/tyr_system.R

@@ -141,7 +141,8 @@
     list(variable = tst_tyr$comparison$test$variable,
          split_factor = tst_tyr$comparison$test$split_factor,
          plot_title = exchange(tst_tyr$comparison$test$variable,
-                               dict = globals$stigma_lexicon),
+                               dict = globals$stigma_lexicon,
+                               value = 'base_label'),
          plot_subtitle = tst_tyr$comparison$test$plot_cap,
          y_lab = exchange(tst_tyr$comparison$test$variable,
                           dict = globals$stigma_lexicon,

### 2023-05-23 tools/globals.R

@@ -51,8 +51,8 @@
          log_age = c('log age', 'years', NA),
          sqrt_age = c('sqrt age', 'years', NA),
          sex = c('sex', NA, NA),
-         psych_comorb = c('mental illness', NA, NA),
-         somatic_comorb = c('somatic illness', NA, NA),
+         psych_comorb = c('psychiatric disorder', NA, NA),
+         somatic_comorb = c('physical disorder', NA, NA),
          infection = c('SARS-CoV2', NA, NA),
          bmi_class = c('body mass', NA, NA),
          smoking = c('smoking', NA, NA),

### 2023-05-23 tools/tools.R

@@ -8,6 +8,10 @@
   library(stringi)
   library(trafo)
   library(sciplot)
+  library(rmarkdown)
+  library(knitr)
+  library(bookdown)
+  library(clipr)
 
 # Descriptive stats -----
 
@@ -797,5 +801,98 @@
          },
          prob = NULL,
          sor = function(x) x)
+
+# markdown -----
+
+  my_word <- function(...) {
+
+    form <- word_document2(number_sections = FALSE,
+                           reference_docx = 'ms_template.docx')
+
+    form$pandoc$lua_filters <- c(form$pandoc$lua_filters,
+                                 'scholarly-metadata.lua',
+                                 'author-info-blocks.lua')
+
+    form
+
+  }
+
+  insert_issue <- function(text = NULL) {
+
+    if(!is.null(text)) {
+
+      text <- paste0("<span custom-style = 'reviewer'>", text, "</span>")
+
+    } else {
+
+      text <- paste0("<span custom-style = 'reviewer'></span>")
+
+    }
+
+    write_clip(content = text,
+               object_type = "character",
+               breaks = "\n")
+
+    return(text)
+
+  }
+
+  my_percent <- function(data,
+                         variable,
+                         return_n = FALSE,
+                         signif_digits = 2) {
+
+    ## percentages or count for categories of a qualitative variable
+
+    counts <- table(data[[variable]])
+
+    if(return_n) return(counts)
+
+    complete <- sum(table(data[[variable]]))
+
+    signif(counts/complete * 100,
+           signif_digits)
+
+  }
+
+  my_beta <- function(data,
+                      variables,
+                      levels = NULL,
+                      signif_digits = 2) {
+
+    ## extracts beta with 95% confidence interval from an inference summary
+
+    beta_dat <- data %>%
+      filter(.data[['variable']] %in% variables)
+
+    if(!is.null(levels)) {
+
+      beta_dat <- beta_dat %>%
+        filter(.data[['level']] == levels)
+
+    }
+
+    beta_dat %>%
+      mutate(txt_lab = paste0(signif(estimate, signif_digits),
+                              ' [95% CI: ', signif(lower_ci, signif_digits),
+                              ' - ', signif(upper_ci, signif_digits), ']')) %>%
+      .$txt_lab
+
+  }
+
+  my_rho <- function(data,
+                     variables_x,
+                     variables_y,
+                     signif_digits = 2) {
+
[…]

### 2023-06-07 paper/markdown/rebuttal_letter.Rmd
Analysis and manuscript files with feedback
Author's feedback included in the analyses and manuscript files. Effect of gender and BMI included in the INCOV cohort multi-parameter models. Changes and shortening of the manuscript text.

@@
 ---
-title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism"
+title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
 subtitle: "Rebuttal Letter"
@@
 Thank you for appreciation of our manuscript and giving us a chance to address Reviewers' issues. 
-Rewrote the paper and put efforts to incorporate the critic in the text. 
-In particular, we defined clearly the analysis endpoints, and decided to present our story in a 'top - bottom' approach. 
-In brief, we employed the most robust and comprehensive analytic technique, multi-parameter modeling, to identify the most relevant explanatory factors affecting systemic serotonin and dopamine metabolism.
-Subsequently, we explore such factors in more detail by correlation and network analysis, and canonical statistical tests. 
-Finally, we improved quality of language and Discussion. 
-Please see below for details. 
-We hope that, our revised manuscript is now suitable for publication in _Psychoneuroendocrinology_.
+First we would like to thank the Reviewers for their very careful, thoughtful and helpful assessment and comments. 
+We acknowledge that such detailed reviews take a lot of time and demand a lot of commitment and hope that we have done them justice in the revised manuscript.
+
+Overall, we agree that the initial manuscript version was not sufficiently coherent and clear because of the involvement of a large team of researchers in the study. 
+Following your advice and the Reviewers' suggestions, we reworked thoroughly the analysis and the text. 
+We would like to point out that this is one of very few manuscripts that characterize activity of the serotonin, kynurenine and catecholamine pathways, i.e.
+the metabolic system that controls systemic availability of indolamine and catecholamine neurotransmitters in COVID-19 [@Santiago-Mujika2022; @Saito2022; @Lionetto2021; @Bizjak2022; @Dewulf2022; @Ceballos2022].
+Importantly, their role in development of symptoms of mental health disorders in COVID-19 patients was proposed [@Kucukkarapinar2022; @Bower2022]. 
+By analysis of two cohorts, we were able to identify key factors affecting activity of those metabolic pathways and investigate their temporal activity profile in acute and sub-acute SARS-CoV-infection and recovery, which adds to novelty of our work.
+
+We have re-analysed the data to address the reviewers´ issues and have rewritten many parts of the manuscript following the reviewers' suggestions. 
+In particular, we now clearly and explicitly state the gap in the literature in __Introduction__ and the analysis endpoints in __Methods__ and __Results__. 
+To facilitate interpretation of our results, we now follow a top - bottom analysis approach. 
+In brief, we employed the most robust and comprehensive analytic technique, multi-parameter modeling, to identify the most relevant explanatory factors affecting the key metabolites and activity markers of the kynurenine, serotonin and catecholamine pathways.
+Subsequently, we explore these factors in more detail by correlation and network analysis, and canonical statistical tests. 
+Following the Reviewer's suggestion we also keep analysis results in the SIMMUN and INCOV cohort separately for the sake of clarity. 
+Concerning the manuscript text we also improved the preciseness of the language and adapted the __Discussion__ part. 
+Please see below the point-to-point response for details. 
+Please notice that due to the scale of changes in the text, it was not feasible to mark all changes introduced during revision. 
+The most relevant changes are <span custom-style = 'revision'>highlighted in red</span>. 
+We hope that our revised manuscript is now suitable for publication in _Psychoneuroendocrinology_.
 
@@
 
-We thank the Reviewer for the careful lecture of our manuscript and valuable feedback. 
-In the revised manuscript, we did our best to address the weaknesses pointed out be the Reviewer.
-In order to improve the clarity of our research aims and to facilitate interpretation of the results, we re-formatted the analysis pipeline and the manuscript in the following way:
+We thank the Reviewer for the careful assessment of our manuscript and valuable feedback which were very helpful in revising our work. 
+First we would like to apologize for imprecise wording, not including enough references and any language errors. 
+We have reworked large parts of the manuscript in order to convey our research aims, results and their interpretation in a more clear way. 
+For these reasons, we re-formatted the analysis pipeline and the manuscript in the following way:
 
-* we streamlined the __Introduction__ section.
-Here, we describe briefly the serotonin and dopamine metabolism, and the competitor kynurenine pathway, functions of those neurotransmitter in the central nervous system and impact on serotonin, dopamine on mental health.
-We outline the role of inflammation and mutual relationship between tryptophan metabolism and mental health.
-Furthermore, we review literature evidence for the SARS-CoV-2/COVID-19 - neurotransmitter connex.
+* we unified nomenclature of the metabolic pathways investigated in our manuscript and term them now 'serotonin pathway', 'kynurenine pathway' and 'catecholamine pathway' in line with prominent review articles [@Badawy2017; @Lukic2022; @Meiser2013]
 
-* In __Introduction__, we also define clearly the 'evidence gap' and our research aims in the last paragraph of the section.
+* we reorganized and rewrote many parts of __Introduction__.
 
-* In  __Methods/Analysis endpoints__ we define unequivocally the outcomes, i.e.
-plasma levels of metabolites involved in serotonin and dopamine turnover, and explanatory factors, i.e.
-demographic, clinical, inflammatory, psychometric and SARS-CoV-2-associated variables.
+* we also more explicitly and clearly state the ‘gap in the literature’ and our research aims in the last paragraph of __Introduction__.
 
-* We restrict the analyzed SIMMUN and INCOV cohorts to participants with the complete set of outcome and explanatory variables (__Figure 1__, __Supplementary Table S1__ and __S3__).
-By this change, results of particular analyses can be more easily compared with each other.
-Additionally, to investigate if and how exclusion of participant affected the SIMMUN cohort characteristic and if it could be regarded as a source of bias, we compared th excluded and analyzed individuals as presented in __Supplementary Table S2__ and __Results/Characteristic of the study cohorts__.
+* we included the __Analysis endpoints__ section of __Methods__. Here we describe our analytic approach, define outcome variables and candidate explanatory factors.
 
-* We excluded CRP and IL6 from analyses presented in the revised manuscript. Those markers were found to be slightly above their reference ranges (0.5 mg/dL and 7 pg/mL) solely in 
+* we show the analysis strategy for the SIMMUN and INCOV cohort in __Supplementary Figure S1__.
+We hope that this makes it easier to follow our analysis.
+Additionally we restrict the analyzed SIMMUN and INCOV cohorts to participants with the complete set of outcome and explanatory variables (__Supplementary Figure S1__, __Supplementary Table S1__ and __S3__).
+This makes the results of particular analyses more easily comparable to one another.
+Additionally, to investigate if and how exclusion of participant affected the SIMMUN cohort characteristic and if it could be regarded as a source of bias, we compared the excluded and analyzed individuals as presented in __Supplementary Table S2__ and __Results/Characteristic of the cohorts__.
+
+* We excluded CRP and IL6 from analyses presented in the revised manuscript. Both  markers were found to be slightly above their reference ranges (0.5 mg/dL and 7 pg/mL) solely in 
 `r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
 `r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of SIMMUN individuals, respectively.
-They are hence unlikely to play a clinically relevant role in regulation of TRP, KYN, KYN/TRP, PHE, TYR and PHE/TYR, in spite of statistically significant correlations presented in the initial manuscript.
-We argument this decision in __Supplementary Methods__ of the revised manuscript.
+They are hence unlikely to be biologically and clinically relevant predictors of the kynurenine and catecholamine pathway activity.
+Details on this are now described in __Supplementary Methods__.
 
-* We provided information on all analyzed variables in the SIMMUN (__Supplementary Table S1__) and INCOV cohort (__Supplementary Table S3__) with their format, units and, if applicable, normality-stabilizing transformations.
+* We provided information on all analyzed variables of the SIMMUN (__Supplementary Table S1__) and INCOV cohort (__Supplementary Table S3__) with their format, units and, if applicable, normality-stabilizing transformations.
 
-* In order to improve clarity of our results, we decided to present our results following a 'top - bottom' scheme.
+* In order to improve clarity of __Results__, we have now reordered and reformatted it.
+We now present the results in a ‘top - bottom’ approach.
 In brief, both in the SIMMUN and the INCOV cohort, we employ multi-parameter modeling (linear regression or robust linear regression), to identify key explanatory factors significantly affecting the outcome.
-Subsequently, we investigate those significant predictions in more detail by 'univariable' correlation analysis and statistical hypothesis testing.
-This analysis strategy was pursued in our recent papers as well [@Hufner2022; @Hufner2023].
+Subsequently, we investigate those significant predictors in more detail by ‘univariable’ correlation analysis and statistical hypothesis testing.
+This analysis strategy has recently been used in two COVID-19-related publications of our group [@Hufner2022; @Hufner2023].
 
@@
 
-As outlined in __Issue 1__, we re-wrote thoroughly the __Introduction__ section. 
-Following discussions in the core author team, we decided no to replace the general and a bit obscure term 'psychoneuroimmunological' with more direct formulations referring to the prime role of inflammation in modulating serotonin and dopamine biosynthesis.
-For this reason, we describe now explicitly the effects of inflammatory cytokines on IDO- and PAH-mediated regulation of serotonin and dopamine biosynthesis and changed the manuscript title to "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism".
-In fact as discussed below and in the revised manuscript text, the question how systemic biosynthesis of serotonin and dopamine, involving primarily mesenteric organs and the liver, relates to the neurotransmitter availability in the brain is vigorously disputed in literature [@Goldstein1999; @Moncrieff2022; @Marx2020].
+As outlined in __Issue 1__, we re-ordered and re-wrote the entire __Introduction__ section.
+Following discussions in the core author team, we decided to replace the general term ‘psychoneuroimmunological’ with more direct descriptions of the mechanisms we are referring to i.e.
+the role of inflammation on IDO activity are the effect of glucocorticoids on TDO activity.
+We have also changed the title of the manuscript.
 
@@
 Should this statement refer to existing cross-sectional findings (e.g., higher QUIN in patients with MDD compared to healthy controls) it is very generalized and not sufficiently justified. 
-If the statement refers to evidence that suggests a causal relationship, a reference is urgently needed. </span>
+If the statement refers to evidence that suggests a causal relationship, a reference is urgently needed.</span>
 
-We agree with the Reviewer, that __Introduction__ needs to be more strict and requires additional references - please refer to the revised text. 
-Concerning your more specific issues: we agree that the great majority majority of tryptophan is catabolized via the KYN and indole pathways and only a marginal fraction is diverted to serotonin synthesis [@Teunis2022; @Lukic2022], and for both serotonin and dopamine, the systemic metabolism taking place in liver, mesenteric organs and vasculature may not reflect the situation in the central nervous system [@Brown2021; @Marx2020; @Schwarcz2012; @Moncrieff2022; @Goldstein1999].
-We take up this controversy in __Introduction__ and __Discussion__. 
-Additionally, we addressed the question how TRP levels may affect systemic serotonin availability, we modeled circulating serotonin levels in the INCOV cohort [@Su2022] as a function of its precursors and competitor metabolites (TRP, KYN, QUIN), timepoint of SARS-CoV-2 infection, inflammatory cytokines and age.
-As presented in __Figure 3__, we could indeed find a significant positive association of plasma TRP and serotonin in multi-parameter modeling, which is expected to be independent of age, inflammation and infection time course.
-In a more detailed correlation and network analysis of the INCOV dataset (__Figure 5A__), we could observe a moderate-strength correlation of circulating serotonin and TRP levels in acute and sub-acute infection.
-This suggests that systemic serotonin synthesis may be limited by blood TRP availability to some extent during SARS-CoV-2-dependent inflammation, which was characterized by profoundly decreased TRP and elevated KYN levels (__Figure 4__).
-In the revised __Introduction__ and __Discussion__, we also critically review the evidence on brain metabolism and effects of KYN, kynurenic acid and QUIN on neuronal signaling and mental health [@Brown2021; @Schwarcz2012], which is indeed limited.
+We agree that the role of the kynurenine pathway metabolites as well as systemic, i.e.
+peripheral availability of tryptophan, serotonin and dopamine in mental health are controversial, not sufficiently supported by direct experimental evidence and definitely needs a critical view on the topic [@Marx2020; @Brown2021; @Moncrieff2022; @Goldstein1999].
+We find that one of the many challenges is that many mechanistic studies are done in the animal model, making the transfer to the human situation questionable [@Brown2021]. 
+On the other hand human studies are quite equivocal and, while there are influential theoretical manuscripts on the issue, surprisingly few original research has been published so far [@Brown2021]. 
+To address you issue, we decided to give a more balanced view on the serotonin, kynurenine, catecholamine pathways and their role in mental health while still keeping the text relatively compact. 
+We now give a more clear, but short introduction to there three biochemical pathways in the introduction section. 
+We provide an ample discussion of the possible role of these pathways in linking mental health and physical conditions. 
+In __Discussion__ and __Limitations__ we comment on controversy over the role of systemic serotonin, kynurenine or catecholamine pathway in depression and other psychiatric disorders [@Brown2021; @Marx2020; @Schwarcz2012; @Moncrieff2022; @Goldstein1999].
+
+Additionally, to address the question how TRP levels may affect systemic serotonin pathway activity in COVID-19 patients, we modeled circulating serotonin levels in the INCOV cohort [@Su2022] as a function of its precursors and competitor metabolites (TRP, KYN, QUIN), timepoint of SARS-CoV-2 infection, inflammatory cytokines and age.
+As presented in __Figure 2__, we could indeed find a significant positive association of plasma TRP and serotonin in multi-parameter modeling, which is expected to be independent of age, inflammation and infection time course.
+In a more detailed correlation and network analysis of the INCOV dataset (__Figure 4A__, __Supplementary Figure S6__), we could observe significant, weak-to-moderate strength correlation of circulating serotonin and TRP levels in acute and sub-acute infection.
+This suggests that systemic serotonin synthesis may be limited by blood TRP availability to some extent during SARS-CoV-2-dependent inflammation, which was characterized by lowered plasma TRP concentration and elevated KYN levels.
 
@@
 We are sorry for this unclarity. 
-In the revised manuscript, we clearly formulate the aims, and define outcomes and explanatory factors. 
-Please refer to the last paragraph of __Introduction__, __Methods/Analysis endpoints__, __Supplementary Tables S1__ and __S3__. 
-In the revised text, we additionally avoid the misleading term 'low-grade' inflammation. 
-In fact, as described in __Issue 1__ and __Supplementary Methods__, only a minute subset of the SIMMUN study participants had CRP or IL-6 concentrations above the reference levels. 
-Thus, biological effect of those two inflammatory markers is doubtful and we decided to exclude them from the revised analysis pipeline.
+In the revised manuscript, we formulate the aims, and define outcomes and explanatory factors. 
+Please refer to the last paragraph of __Introduction__ and the subsection __Analysis endpoints__ of __Methods__.
 
@@
 We agree with the Reviewer that the INCOV cohort deserves a more thorough description. 
-This is provided in the __Methods/Procedures__ section of the revised manuscript and in __Supplementary Methods__ and in references to the seminal papers [@Su2020; @Su2022].
+We included short methodological information of the INCOV cohort in __Methods/Procedures__ and __Supplementary Methods__ [@Su2020; @Su2022].
 
@@
 
-We are grateful for the hints. 
-In the revised manuscript, we applied robust linear regression with the MM algorithm and Huber's psi function [@Huber2011; @Ripley2022] for multi-parameter modeling (__Figure 3__) and time-course modeling of the INCOV dataset (__Figure 4__).
-Additionally, we specify in the figure legends of the INCOV results the median time interval between diagnosis and sampling for the acute, sub-acute and recovery timepoints. 
+We are grateful for these suggestions and included them in our analyses. 
+In the revised manuscript, we applied robust linear regression with the MM algorithm and Huber's psi function [@Huber2011; @Ripley2022] for multi-parameter modeling (__Figure 2__) and time-course modeling of the INCOV dataset (__Figure 3__).
+Additionally, we specify in legends of figures with INCOV cohort data the median time interval between diagnosis and sampling for the acute, sub-acute and recovery timepoints. 
 As suggested, we applied multiple testing adjustment throughout the analysis pipeline. 
@@
 We also checked consistency of the PSS-4 and HADS psychometric tools employed in our report. 
-Since the HADS scales, and in particular, the HADS depression tool were found to be non-tau-equivalent by factors analysis (__Supplementary Figure S1A__), we decided use McDonald's $\omega$ as a consistency measure instead of the more popular Cronbach's $\alpha$ [@Revelle2022; @McDonald1999].
-The PSS-4 tool demonstrated a good and the HADS scales an excellent consistency, as presented in __Supplementary Figure S1B__.
+Since the HADS scales, and in particular, the HADS depression tool were found to be non-tau-equivalent by factors analysis (__Supplementary Figure S2A__), we decided use McDonald's $\omega$ as a consistency measure instead of the more popular Cronbach's $\alpha$ [@Revelle2022; @McDonald1999].
+The PSS-4 tool demonstrated a good and the HADS scales an excellent consistency, as presented in __Supplementary Figure S2B__.
 
@@
 We concur with the Reviewer, that a comparison between the cohorts is useful for interpretation of the results. 
-We present significant results of the comparison between the SIMMUN and INCOV cohort in __Supplementary Table S4__ and elaborate on them in __Results/Characteristic of the study cohorts__. 
-The collectives differed in age and sex distribution, rates of SARS-CoV-2 infection and infection severity. 
+We present significant results of the comparison between the SIMMUN and INCOV cohort in __Supplementary Table S5__ and elaborate on them in __Results/Characteristic of the study cohorts__. 
+The cohorts differed in age and sex distribution, rates of SARS-CoV-2 infection and infection severity. 
 In particular, most INCOV study participants experienced moderate-to-critical COVID-19 and required hospitalization. 
-This may explain the way stronger effects of inflammation and infection timepoint on inflammatory cytokines and the metabolites of interest, which we discuss in __Discussion__ and __Limitations__.
+This may explain the stronger effects of inflammation and infection timepoint on inflammatory cytokines and the metabolites of interest, which we discuss in __Discussion__. 
 In the revised text, Figures and Tables we also present effect size metrics for statistical hypothesis tests, correlation and modeling. 
-We interpret them as proposed in the excellent handbook by Jacob Cohen [@Cohen2013] (__Supplementary Methods__).
+We interpret them as proposed in the handbook by Cohen [@Cohen2013] (__Supplementary Methods__).
 
@@
 
-We present now p values and effect sizes or model $\beta$ with 95% confidence intervals [@Cohen2013], as appropriate, for the most important findings in Abstract and Results. 
+We present now p values and effect sizes or model $\beta$ with 95% confidence intervals, as appropriate, for the most important findings in Abstract and Results. 
 Yet, for the sake of readability, we were not able to provide significance and effect size information for all effects. 
@@
 
-Following your suggestion, we avoid using the term 'strata' in the revised text. 
-We use group, subgroup or subset instead.
+Following your suggestion, we have replaced the term ‘strata’ in the revised text by 'subgroup' or 'subset'.
 
@@
 
-This point is no relevant anymore, since we re-wrote and consolidated __Introduction__ of the revised manuscript. 
-We describe each thread in a separate paragraph.
+Since we re-wrote and shortened the __Introduction__ section, we abstained from using subheadings; such format is also not common in original research manuscripts in _Psychoneuroendocrinology_. 
+To facilitate interpretation, each topic of __Introduction__ is described in a separate paragraph. 
+In turn, we employ subheadings in __Discussion__.
 
@@
 
-Thanks for careful lecture, this erroneous statement was removed.
+Thanks for the attentive assessment, this misleading wording was removed. 
+Please refer to the metabolic pathway nomenclature in __Issue 1__.
 
@@
 This time interval refereed to the SARS-CoV-2 infection diagnosis via PCR test, which we specify in the revised manuscript. 
-We also round them to complete days, as suggested.
+We now also round them to complete days, as suggested.
 
@@
 
-We concur with your suggestion, the variables of interests are now listed in the __Methods__ section instead of __Results__. 
-Furthermore, we describe the rationale for analyses in the INCOV cohort in the last paragraph of __Introduction__, __Methods/Analysis endpoints__ and __Results__. 
-In brief, we employed the SIMMUN collective providing more extensive demographic and clinical characteristic in an initial search for the most relevant factors impacting on the indirect readouts of serotonin and dopamine metabolism: TRP, KYN, PHE and TYR [@Brown2021; @Marx2020; @Capuron2011; @Widner1997; @Neurauter2008].
-The INCOV dataset with measurements of more direct markers of the neurotransmitter availability, i.e.
-plasma serotonin and dopamine 3-O-sulfate, the most abundant dopamine metabolite present in human blood [@Meiser2013; @Goldstein1999], allowed us to validate effects of inflammation, SARS-CoV-2 infection and age, i.e.
-significant predictors identified in the SIMMUN study.
-Furthermore, due to longitudinal design of the INCOV study, we were able to explore not only the impact of infection recovery but also acute and sub-acute COVID-19 on serotonin and dopamine turnover. 
-Still, the incompatibility of designs and datasets of these two collectives hindered us from developing consistent models in one cohort and direct validation in the outer (canonical training - test or external validation setup).
+As suggested, we now list the variables of interest in __Methods__ instead of __Results__. 
+Furthermore, we describe the rationale for using the INCOV dataset in the last paragraph of __Introduction__ and the __Analysis endpoints__ subsection of __Methods__. 
+Information on measurement of proteins and metabolites is provided in __Methods/Procedures__ and __Supplementary Methods__.
+Unfortunately, the incompatibility of designs and datasets of these two cohorts hindered us from developing a comprehensive model in one cohort and direct validation in the outer (canonical training - test setup).
 We list this issue in __Limitations__.
@@
 We provide the time intervals for the acute, sub-acute and recovery timepoints in the INCOV cohort in the revised manuscript (__Methods__, __Results__, __Figures__).
-Of note, those time intervals fit well with the definition of acute, sub-acute and long-COVID/recovery timepoints in our recent population study [@Sahanic2021].
 
@@
 
-We apologize for this imprecise term, it was removed from the revised manuscript.
+We have removed all imprecise wording while rewriting the __Results__ section. 
+Thank you for pointing this out.
 
@@
 We are sorry for this unclarity. 
-In the revised text, we name inflammatory markers (NEO, NLR, cytokines) more explicitly.
+In the revised text, we name inflammatory markers (SIMMUN: NEO, NLR, INCOV: IL6, IL10, TNF and IFNG) more explicitly.
 
@@
 
-In the revised manuscript, we restructured and improved scientific quality of __Discussion__. 
-The key explanatory factors proposed by us to influence systemic serotonin and dopamine availability: age, inflammation, SARS-CoV-2 infection and, for serotonin, availability of the precursor aminoacid TRP, are discussed in a context of literature evidence.
-We also critically elaborate on the link between plasma serotonin, TRP and KYN levels and depressive disorders, which was recently challenged [@Capuron2011; @Hufner2021; @Sun2020; @Hufner2015; @Fellendorf2022; @Paul-Savoie2011; @Hunt2020; @Moncrieff2022; @Marx2020; @Brown2021].
+We restructured and improved __Discussion__. 
+Large parts of __Discussion__ were reorganized and rewritten as suggested by Reviewer 2 (__Issue 40__). 
+We hope that you agree with the structure proposed by the other Reviewer which we have now followed. 
+We have also included new references to back out results. 
+The __Limitations__ section was also expanded and improved.
 
@@
 <span custom-style = 'reviewer'>__Discussion:__ The statement that "BH4 is also important for nitric oxide synthesis and therefore involved in oxidative stress" is not correct.
-(Page 16) BH4 is an essential Co-factor of NO-synthases that produce NO with its function as a vasodilator and additional anti-pathogenic effects.
+(Page 16) BH4 is an essential co-factor of NO-synthases that produce NO with its function as a vasodilator and additional anti-pathogenic effects.
 Only if BH4 is missing, there is an uncoupling of NOS with subsequently increased reactive oxygen/nitrogen species.
@@
 
-Following discussion in the study team, we decided to suppress the ROS thread of the discussion since it is likely not so important for interpretation of our data. 
-Instead we focus on direct effects of inflammation on PAH, the enzyme catalyzing the PHE - TYR conversion, i.e.
-the initial step of dopamine synthesis and its crucial co-factor BH4 [@Geisler2013; @Luporini2021; @Fanet2021; @Capuron2011; @Neurauter2008].
+We have removed most of the discussion on oxidative stress from the revised text since it is likely not so important for interpretation of our data. 
+Instead we focus more on BH4 as the co-factor for synthesis of indolamine and catecholamine neurotransmitters [@Geisler2013; @Capuron2011; @Neurauter2008; @Vancassel2018].
 
@@
 
-We apologize for this obscure term. 
+We apologize, this is really misleading and was reworded. 
 We meant levels of inflammatory markers in SARS-CoV-2-uninfected individuals, which we state now explicitly.
@@
 
-_Katharina: please respond, I don't have this manuscript part at all in pre-revision manuscript files._
+The __Highlights__ were adapted according to your suggestions.
 
@@
 
-We are grateful for the careful language check. 
+Thank you for your careful assessment.  
 We have paid particular attention to the style, grammar and proper scientific language in the revised manuscript. 
@@
 
-* _Page 9 line 52: What are nadir TRP concentrations?_ - we do not use the term any more, changed to 'minimum'
+* _Page 9 line 52: What are nadir TRP concentrations?_ - this was changed to 'minimum'
 
@@
 
-* _Page 14 line 34: please delete the term "larger"_ - done as suggested.
+* _Page 14 line 34: please delete the term "larger"_ - done as suggested
 
@@
 We concur with the Reviewer's view, that visualization of trends for monotone associations is problematic. 
-Due to profound changes in the manuscript structure, we do not show scatter plots to illustrate monotone correlations between non-normally distributed variables. 
-Correlation analysis results in the INCOV cohort are presented now as simple bubble correlograms or visualized as force-directed graphs [@Briatte2021; @Csardi2006].
+Due to profound changes in the structure of the revised manuscript, we do not use scatter plots to illustrate monotone correlations between non-normally distributed variables any more. 
+Correlation analysis results in the INCOV cohort are now presented as simple bubble correlograms or visualized as force-directed graphs [@Briatte2021; @Csardi2006].
 
@@
 We thank the Reviewer for appreciation of out text and the constructive feedback. 
-Weaknesses in structure of the test were already pointed out be the Reviewer 1. 
-We addressed them by employing the multi-parameter modeling results for identification of the most vital explanatory factors that affect neurotransmitter metabolism and subsequent more specific correlation analyses and statistical testing with such relevant predictors.
-To tackle with the confusion around the two investigated cohorts, we separated the results obtained with the SIMMUN and INCOV collective. 
-In more details, we employ the SIMMUN cohort as a kind of 'exploratory' collective to identify factors affecting TRP, KYN, KYN/TRP, PHE and PHE/TYR among a wider palette of demographic, clinical, psychometric, inflammatory and SARS-CoV-2 associated variables.
-Subsequently, we attempted to validate the most important of them: age, inflammation and SARS-CoV-2 infection in the INCOV cohort. 
-Finally, we re-wrote the __Discussion__ section with a special focus on the crosstalk between inflammation and neurotransmitter biosynthesis and, in part controversial, relevance of systemic neurotransmitter availability for the central nervous system and mental health.
+Weaknesses in structure of the analyses were already pointed out by the Reviewer 1. 
+We addressed them by employing the multi-parameter modeling results for identification of the most vital explanatory factors that affect neurotransmitter metabolism and subsequent more specific correlation analyses and statistical testing with those relevant predictors.
+To make the analysis of the two investigated cohorts easier to follow, we separated the results obtained with the SIMMUN and INCOV cohort. 
+In more detail, we employ the SIMMUN cohort as an ‘exploratory’ cohort to identify factors affecting TRP, KYN, KYN/TRP, PHE and PHE/TYR from a wider palette of demographic, clinical, psychometric, inflammatory and SARS-CoV-2 associated variables.
+Subsequently, we attempted to validate the most important ones, age, inflammation and SARS-CoV-2 infection, in the INCOV cohort.
+Finally, we re-wrote __Discussion__ with a special focus on the crosstalk between inflammation and neurotransmitter biosynthesis and, in part controversial, relevance of systemic neurotransmitter availability for the central nervous system and mental health.
 
@@
 
-Thanks for the careful lecture. 
+Thanks for the careful assessment. 
 We paid special attention to the grammar and spelling of the revised text.
@@
 
-_Katharina: could you provide an answer?_
+The first author of the manuscript has extensive experience in scientific English. 
+We are assuming that some inconsistencies in the first version of the manuscript arose from working on the manuscript in a large study team. 
+The first author has now carefully reworked the text and we hope that this has now lead to a more consistent style and better quality.
 
@@
 
-The term 'collective' has the same meaning as 'cohort' in the manuscript text. 
-For practical reasons, we avoid the wording 'population sample', which may suggest a random and hence representative draw from the general population. 
-In fact neither the SIMMUN cohort enriched in psychiatric patients not the INCOV cohort enriched in hospitalized COVID-19 patients pose a representative sample of the general population during the pandemic (see __Limitations__) .
+Yes, in the manuscript  text the term ‘collective’ had the same meaning as ‘cohort’. 
+We removed the word 'collective' since this obviously was not easy to understand and use consistently 'cohort' instead. 
+For practical reasons, we avoid the wording ‘population sample’, which may suggest a random and hence representative draw from the general population. 
+In fact neither the SIMMUN cohort enriched in psychiatric patients nor the INCOV cohort enriched in hospitalized COVID-19 patients pose a representative sample of the general population during the pandemic (see __Limitations__).
 
@@
 
-As pointed out by Reviewer 1 as well, the last paragraph of __Introduction__ was quite obscure. 
-In the final part of __Introduction__, we present now the 'evidence gap', which is the still incompletely characterized effect of SARS-CoV-2 infection on neurotransmitter metabolism, as well as the aim of our analysis, which identification of factors affecting the neurotransmitter turnover in a cohort of SARS-CoV-2-infected individuals.
-Still, our study does not have a fully exploratory character, since the candidate explanatory factors: age, sex, body mass class, mental and somatic conditions, smoking and alcohol, depression, anxiety, stress and inflammation have already been described to affect serotonin and dopamine metabolism.
+As also pointed out by Reviewer 1, the last paragraph of __Introduction__ was not written clearly enough in the initial manuscript version. 
+In the final paragraph of __Introduction__, we present now the ‘gap in the literature’, which is the still incompletely characterized effect of SARS-CoV-2 infection on metabolic pathways determining systemic availability of serotonin and dopamine.
+We also state the aim of our analysis, which was characterization of factors affecting the serotonin, kynurenine and catecholamine pathway activity during SARS-CoV-2 infection and recovery. 
+Our study did  not have a fully exploratory character, since the candidate explanatory factors: age, sex, body mass index, mental and somatic conditions, smoking and alcohol, depression, anxiety, stress and inflammation have already been described to affect serotonin and dopamine metabolism in previous research, also from our own research consortium [@Hufner2019; @Hufner2020; @Hufner2021; @Hufner2015].
 
@@
 
-We avoid using the term 'decay' in the revised text and utilize 'catabolism' or 'conversion' instead.
+This was reworded in the revised version, we utilize 'catabolism', 'breakdown' or 'conversion' instead.
 
@@
 
-We concur with your opinion and present the analysis results as suggested. 
+We agree and rearranged the complete __Results__ section. 
 Please refer to __Issues 1__ and __31__ for details.
@@
 
-<span custom-style = 'reviewer'> You utilize the terms mental disease, mental stress, mental symptoms, etc… Be more specific and replace these terms by what you measured, i.e., HADS-anxiety, HADS-depression, and PSS.
+<span custom-style = 'reviewer'> You utilize the terms mental disease, mental stress, mental symptoms, etc.
+Be more specific and replace these terms by what you measured, i.e., HADS-anxiety, HADS-depression, and PSS.
 You mention somatic symptoms in the discussion but what they refer to remains obscure. 
@@
 
-We are grateful for this hint. 
-In the revised manuscript we provide the psychometric scale when referencing depression, anxiety and stress signs. 
-We removed the sentence of __Discussion__ referring to somatic symptoms, since those were not systematically surveyed in the SIMMUN cohort. 
-Following your suggestion, we term 'somatic conditions' 'physical disorders' consequently in the revised text.
+Thank you for your suggestions. 
+We have specified the terms according to your suggestions throughout the revised manuscript. 
+We removed the sentence from __Discussion__ referring to somatic symptoms, since those were not systematically surveyed in the SIMMUN cohort. 
+Following your suggestion, we term ‘somatic conditions’ ‘physical disorders’ consequently in the revised text.
 
@@
 <span custom-style = 'reviewer'>The discussion is convoluted and needs to be simplified. 
-It would be easier to follow if you would first discuss the significance of the association between inflammation and neurotransmitter precursors and then the association between these two sets of factors and what you call mental health measured by HADS and PSS, first in the immune -> HDAS symptomatology or kynurenine -> HDAS symptomatology and then in the stress -> immune direction, remembering that in the case of your own cohort, the cross-sectional nature of your study does not allow you to get further than describing associations.</span>
+It would be easier to follow if you would first discuss the significance of the association between inflammation and neurotransmitter precursors and then the association between these two sets of factors and what you call mental health measured by HADS and PSS, first in the immune -> HDAS symptomatology or kynurenine -> HADS symptomatology and then in the stress -> immune direction, remembering that in the case of your own cohort, the cross-sectional nature of your study does not allow you to get further than describing associations.</span>
 
-We agree with the Reviewer that there was a lot of room for improvement in __Discussion__.
-We rewrote this part of the manuscript during the revision according to the following scheme:
+We thank the Reviewer for this specific suggestion on the discussion section which we have gratefully taken up and rewritten large parts of __Discussion__ according to the following scheme:
 
-* A brief generalizing summary of the results obtained with both cohorts.
-In this initial paragraph, we highlight the most relevant factors for systemic serotonin and dopamine metabolism: inflammation, SARS-CoV-2 infection, TRP availability, age, signs of depression and mental stress.
-In this paragraph, we also discuss some discrepancy in our data such as way higher effects of inflammation on neurotransmitter availability in the INCOV cohort than in the SIMMUN dataset.
+* A brief generalizing summary of the results obtained with both cohorts and discussion of discrepant findings
 
-* Role of inflammation and SARS-CoV-2 in the KYN pathway and PHE - TYR conversion with implication of systemic availability of serotonin and dopamine.
-In this paragraph we discuss clinical literature and mechanistic reports tackling with IDO and PAH activity as well as BH4, the key co-factor of the serotonin and dopamine pathway enzymes.
+* Effects of inflammation and infection, including SARS-CoV-2, on the serotonin and kynurenine pathways, and systemic levels of serotonin
 
-* Effects of mental health-related factors (psychiatric disorders, mental stress) on systemic serotonin and dopamine availability.
+* Effects of inflammation and infection on the catecholamine pathway and systemic availability of dopamine
 
-* Relevance of systemic metabolism of serotonin and dopamine for the central nervous system and development of psychiatric disorders.
-Here we also underline that evidence for effects of peripheral serotonin, systemic activity of the KYN pathway and peripheral dopamine synthesis from PHE and TYR is limited and bases in many cases on clinical association of psychiatric disorders with blood levels of the neurotransmitters, their precursors or competitor pathway products.
+* Mutual effects of mental health-related factors (e.g.
+depression, anxiety and mental stress) on activity of the serotonin, kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability.
+Here we also underline that evidence for effects of peripheral serotonin, kynurenine and catecholamine pathways is limited and based in many cases on clinical association of psychiatric disorders with blood levels of the neurotransmitters, their precursors or competitor pathway products.
 
@@
 
-We agree with the Reviewer, that the report by Benedetti et al. does not provide any solid mechanistic link between the COVID-19-related inflammation and depression. 
-We removed this paper from __Discussion__ of the revised manuscript.
+We removed this reference from __Discussion__ of the revised manuscript.
 
@@
 
-We concur with the Reviewer's opinion. 
-Certainly, kynurenine and the kynurenine/tryptophan ratio are markers of the pathway which competes with the serotonin biosynthesis. 
-For the sake of clarity, we we term kynurenine, quinolinate and kynurenine/tryptophan ratio 'competitor pathway products/metabolites' or products/metabolites of the competitive kynurenine pathway' in the revised text.
+We absolutely agree with the Reviewer. 
+To clarify that, we use consistently the terms 'serotonin pathway', 'kynurenine pathway' and 'catecholamine pathway' in the revised manuscript.
 
@@
 
-To improve readability of the revised manuscript, we present the results in a cohort-wise manner, i.e.
-beginning with the SIMMUN cohort as an 'exploratory' collective and proceeding with the INCOV cohort as a king of 'validation' collective used to verify and study in more detail (e.g.
-by time course analyses) the key findings from the SIMMUN cohort.
-Please refer to __Issues 1__ and __Issue 38__ for details. 
-The modified analysis strategy is presented in __Figure 1__.
+Thank you for the suggestions. 
+As described in __Issues 1__ and __38__, we present the results in cohort-wise manner, for the sake of better readability. 
+We also included the analysis strategy scheme in __Supplementary Figure S1__.
 
@@
 
-Thanks for the hint. 
-We stick in the revised manuscript to the proposed metabolite order.
+Thank you, we made the changes you suggest throughout the revised manuscript.
 
@@
 
-We re-wrote the Discussion part of the revised manuscript and provide now more evidence [@Hunt2020; @Maes2012; @Brown2021; @Marx2020;  @Fellendorf2022] on the link between the so called sickness behavior, inflammation, serotonin and dopamine metabolism in the final part of the scion tackling the relevance of our findings for mental health.
-We also discuss the controversial relationship between circulating markers of neurotransmitter availability such as tryptophan or serotonin, and the competitor pathway products (kynurenine, quinolinate) on central nervous system and mental health [@Brown2021; @Moncrieff2022].
+We have improved this part of __Discussion__ and now report citations [@Hunt2020; @Maes2012; @Brown2021; @Marx2020;  @Fellendorf2022; @Moncrieff2022]] on the pros and cons of KYN pathway involvement in depression in the text and the __Limitations__ section.
 
@@
 
-_Katharina, could you please respond? I don't have the submission text with the line numbering and can't find the fragment. Thanks!_
+This was re-written as suggested.
 
@@
 
-_As above, thanks!_
+The '???' was likely a typo in a figure reference in the manuscript markdown document. 
+We have paid special attention to the proper syntax in the revised text.
 
@@
 
-_Same as above!_
+This was re-written as suggested.
 
@@
 We apologize for the typos concerning numbers in the study inclusion scheme. 
-Please refer to the correct numbers in the revised __Figure 1__.
+Please refer to the correct numbers in the revised __Supplementary Figure S1__.
 

### 2023-06-07 paper/markdown/issues.Rmd
Analysis and manuscript files with feedback
Author's feedback included in the analyses and manuscript files. Effect of gender and BMI included in the INCOV cohort multi-parameter models. Changes and shortening of the manuscript text.

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
subtitle: "Issues"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# General

* Title page of the main manuscript: you will have to add the address and phone number per hand for re-submission. The header of the markdown manuscript document with this info gets lost during rendering to docx.

* I suggest sticking to three commonly used terms 'serotonin pathway' (TRP -> 5HTRP -> serotonin -> 5HIAA and other breakdown products), 'kynurenine pathway' (TRP -> KYN -> ... -> QUIN and other breakdown products) and 'catecholamine pathway' (PHE -> TYR -> DOPA -> DA -> ... -> DA sulfate and other DA breakdown products) when referring to metabolic processes, metabolites and activity markers in our manuscript. Terms like 'neurotransmitter precursor' or 'pathway leading to reduced TRP availability' are misleading. This nomenclature is used in prominent reviews [@Badawy2017; @Lukic2022; @Meiser2013].

* Similarly, I suggest sticking to the terms 'systemic serotonin availability' or 'systemic serotonin levels' and 'systemic dopamine availability' or 'systemic dopamine levels' when referring to neurotransmitters of interest.

* I'm spelling out 'kynurenine', when referencing the 'kynurenine pathway'. The abbreviated KYN form is used to reference the metabolite.

* I think (and my Open AI as well:) that the term 'status post SARS-CoV-2 infection' may pertain to a kind of illness, persistent symptoms or disability (details below). In addition, it suggests that there a difference between SARS-CoV-2-positive participants of the SIMMUN ('status post SARS-CoV-2 infection') and INCOV study ('SARS-CoV-2-positive' or 'infected'). Hence I'm using consistently the terms 'SARS-CoV-2-positive', 'SARS-CoV-2-infected' and 'SARS-CoV-2' infection in the text.

* Title changed to "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity". I think event if this title does not sound too attractive, is describes quite precisely the matter of our research without extrapolating it to neuro- or immunology.

* Modeling in the INCOV cohort. I've included age, sex and BMI class in the models as well, as suggested, details below.

* As [required by the journal](https://www.elsevier.com/wps/find/journaldescription.cws_home/473?generatepdf=true), the manuscript must have < 6000 words (Introduction, Materials and Methods, Results and Discussion/Limitations/Conclusion as I suppose), the Introduction part must be max. 1000 words and Discussion max. 2000 words long. I've tried to observe these limits quite strictly and shortened the text. I reduced also the number of subheadings in Discussion (details below).

* We are allowed to have a total of 6 Figures and Tables (together!) in the main manuscript text. For this reason, I moved to the study inclusion diagram to Supplementary Material. I'm not sure, we need a scheme of the serotonin, kynurenine and catecholamine pathways in the manuscript. There are plenty of them in review papers we cite anyway. I don't think, our hypothesis is particularly complex or particularly novel, so that need to help the reader understand it.

* They recommend as well up to 50 references, at the moment we have approximately 70, which should be OK.

* Marked changes in the text. As I wrote in Rebuttal Letter, I'd simply highlight parts of the text, which we he most profoundly altered. These were Abstract, Results (without characteristic of the study cohorts) and Discussion. Methods are more or less the the same.

# Highlights

* Please refer to the file in the docx folder of the repository. You will find there also Word files with your feedback.

* Per point 85 characters are allowed!

# Rebuttal Letter

* Please note the updated Figure and Supplementary Figure order.

* Reviewer 1 and the controversial role of peripheral neurotransmitters for mental health. Like Reviewer 1 I am not so convinced that biochemical changes in periphery are really key players of mental health :) To me, the clinical evidence is simply correlative and most studies include biased or 'impure' patient collectives (i.e. patients with and without medication with SSRI, comorbidities that affect serotonin levels like stroke, vascular diseases, CKD or cancers that employ the kynurenine pathway to trick T cell immunity, different mental diseases thrown into one category etc.), don't correct for demographic confounders (I was not able to find any consistent info on serotonin concentrations in the genders!), social confounders, diet, expected disease outcome (e.g. palliative vs curative). We can write a critical methodological review on that (seriously!). I'm not sure if we have place to discuss that in the manuscript, so I'd suppress that a bit and cite balanced [@Brown2021] and critical papers on that [@Moncrieff2022; @Goldstein1999]. As a machine learning fan, I'd say, there is also another factor that makes me skeptical about systemic biochemical mechanisms of mental disorders - complexity. Beginning from say 100 neurons or lets say 1000 random trees in silico and a nice training dataset, the algorithm is not deterministic any more and less and less sensitive to changes in its parameters like learning rate (neurons) or splitting rule (trees). I.e. it starts to adapt, like a human brain. If you translate that to network of few billions of neurons: you may have less serotonin at the beginning, but the brain will cope anyway with tasks and challenges:)

# Supplementary Material

* Please note the updated Supplementary Figure order.

# Manuscript

## Abstract

* I've tried to include your Abstract issues but I'd recommend sticking to the 250 Word limit. I'd avoid abbreviations. Importantly, the comparison of dopamine 3-O-sulfate and serotonin for COVID-19 time course were done with uninfected or acute SARS-CoV-2 infection as baselines but not versus sub-acute infection - the effects were significant primarily for the acute - recovery comparison. Yes the n number for the ICOV cohort is different - now I refer to the number of patients instead of samples; note, that most of the patients were measured more than once. I also stick to the results of linear modeling for the main readouts: KYN/TRP, PHE/TYR, serotonin and dopamine 3-O-sulfate and omit correlation results in Abstract to spare words. For the same reason I would not show R^2^ but just beta coefficients.

## Keywords

* I'd replace the wording 'neurotransmitter precursor amino acids', since we work also with other metabolites. I'd put 'mental health' instead of 'mental stress', I suppose our text will be easier to find. The Keyword section need to be placed below Abstract.

## Introduction

* This section must be < 1000 words according to the journal's guidelines. A above I've tried to include your comments in succinct form, we have now approx. 780 words.

* 'interact': means that something happens in a two-way manner, i.e. two things influence each other - the first sentence of Introduction was shortened.

* 'interferon treatment' is not a physical condition. This setting was used in the past to treat some types of cancer and MS and was linked to kynurenine and depression. I would keep it under malignancy and autoimmunity.

* Figures: we can have [up to 6 combined figures and tables](https://www.elsevier.com/wps/find/journaldescription.cws_home/473?generatepdf=true) - so there's definitively no room for a scheme. The pathways of interest are excellently investigated and we cite really nice reviews on them. I think the reader can follow our story without a scheme. Because of the figure/table number limit, I've moved Figure 1 with the inclusion and analysis scheme to Supplementary Material (Supplementary Figure S1).

* The journal wants approx. 50 references and, possibly, more after revision. Because of the citation style, they take 2 - 4 words per reference and cause hence problems with the word limit. I've tried to stay at 70 references in the current version and removed some less important ones, which we do not discuss later on.

* 'Alternatively TRP can be metabolized along the kynurenine pathway': suggests erroneously that the kynurenine is a minor metabolic pathway of TRP breakdown: in fact it catabolizes 99% of TRP in the body. I've used another wording.

* Knox 1951 is actually the sole experimental evidence for glucocorticoids and TDO activity. I'd remove the thread if we have problems with the word limit.

* I've tried to organize Introduction as follows: (A) physical - mental health, (B) serotonin and kynurenine pathway, (C) catecholamine pathway, (D) the serotonin, kynurenine, catecholamine pathways and COVID-19, (E) literature gap and research aims.

* I would not introduce the ROS/microglia story in too much details in Introduction. We are aiming anyway at suppressing the PHE/TYR story a bit. I feel short description of the catecholamine pathway and the role of BH4 is important here.

* I'm starting the COVID-19 paragraph explaining that there's indeed a huge psychiatric burden of COVID-19. In turn, I'd remove the sentence on blood brain permeability and the reference with the cytokine storm (which affects virtually only the hospitalized patients). In mild cases of COVID-19 I can't believe that there's a BBB breach or the virus in the brain.

* The last paragraph of Introduction. I'm describing what our explanatory variables and responses were without listing them in detail. Such details are described anyway in Methods and Results. Frankly speaking, markers of systemic serotonin, kynurenine and catecholamine pathways are exactly the our readouts. The rest i.e. what happens to serotonin, dopamine and mental health is a matter of discussion.

## Materials and Methods

* Definition of complete cases: explanatory and response variables presented in Supplementary Figure S1, and Supplementary Tables S1 and S3.

* NEO is routinely determined in ZIMCL by [ELISA](https://zimcl.tirol-kliniken.at/page.cfm?vpath=parameterdetails&genericpageid=648)

* Instead of Swan 1976, who does not tackle DF sulfate at all, I'm referencing a newer paper of [Suominen at al.](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0068007)

* Alcohol and smoking: you have to talk to Sophia. I've extracted the information from two variables 'Nikotin' and 'Alkohol' in the SPSS file.

* Materials and Methods were shortened considerably to stay within the word limit.

## Results

* As far as I can tell, there are no 'old' manuscript parts in Results.

* I'd stick to the wording 'infected' or 'SARS-CoV-2-positive' instead of 'status post-SARS-CoV-2 infection'. I'm not sure if the reader gets it correctly. GPT-4 says that it 'refers to the condition of a person who has recovered from COVID-19' but then elaborates on PASC. Google returns only pages on 'post-COVID-19 condition' and something like 'health/functional status post-SARS-CoV-2 infection'. This may suggest a kind of illness, persistent disability or persistent symptoms. Another question which may come to Reviewer's mind: why are SARS-CoV-2-positive participants in the SIMMUN study are called 'with status post-SARS-CoV-2 infection' and why are the INCOV study participants are called 'SARS-CoV-2-positive'? What is the difference? To my opinion it's enough to provide the median time interval between the PCR and survey and stress it again in Discussion. In fact, we used the "SARS-CoV-2-positive" and "-negative" wording in our Frontiers 2022 paper in cohorts with even longer observation times and the reviewers were happy:)

* Percentage of SARS-CoV-2+ SIMMUN participants corrected, thanks!

* Information on hospitalization was added to Table 2 (footnote).

* There was a typo in the R code returning the initial number of participants and samples in the INCOV study in the manuscript text. Now the figures are correct (text and Supplementary Figure S1).

* Please note the new numbering of Figures and Supplementary Figures, since the study inclusion and analysis schemes were moved to Supplementary Material.

* I would leave the network analysis figure, i.e. Figure 4 as it is. Event though this is a quite shallow network analysis (we can consider e.g. checking which of the metabolites/cytokines is the most important for the process/connectivity or try to define so called cliques), I would let it stay in the manuscript for re-submission and await the Reviewers' response. In my recent cancer paper the reviewers really liked the networks:)

* INCOV analyses: I've modified the initial paragraph, since it may sound that serotonin is a readout the kynurenine pathway activity, which was criticized by one of the reviewers.

* Generally, I've get rid of most references in Results, in particular, purely technical ones (e.g. Ripley 2022 referring to an R package).

* Referring to the difference between the SIMMUN and INCOV models. In the SIMMUN cohort, I use ordinal least square regression, which allows to 'optimize' a multi-parameter model by dropping non-significant terms if the model with the term removed is not substantially worse than the initial model (measured by BIC) - this is the so called 'backward elimination'. So, in an optimal case, I'll end up with a model with all terms significant (Figure 2). In the INCOV cohort, I'm constructing robust multi-parameter models which, instead of squares, minimize a special psi function (e.g. Huber's psi) which assigns weights for each observation (outliers have lower weights, nicely fitting observations are weighted higher). Those weights vary from model to model, hence if I skip a non-significant variable in a robust model, there's no way to check if this worsened the model fit. The full model and the model with the variable skipped have namely different weights assigned to the same observations so I can't compare them:) For this reason, I have to leave all variables, also non-significant ones in the robust model (Figure 3).

* Back to the INCOV models. Following your suggestion, I've included gender and BMI in the models as well. Pertaining to the previous issue, because we leave all variables in the final robust model, it's prone to overfitting. So my initial idea was to include in the robust models only those explanatory factors, which proved significant in the SIMMUN analysis (age and BMI was not significant). The robust models with gender and BMI are still acceptable, but there are already sings of overfitting, if you compare R^2^ between the training and CV datasets (Figure 3)- R^2 is namely more sensitive to overfitting than the model error/RMSE because it takes a square error to calculate it: $R^2 \approx 1 - \frac{RMSE^2}{Variance}$. I'd leave the model as they are for re-submission, but I'm quite sure, that Reviewer 1, who seems to be a psych-statistic aficionado will spot it.

* Correlation analysis in the INCOV cohort: all correlation coefficients with their significance are presented in Supplementary Figure S6, so I'm not including a table. I prefer the bubble plot representation to a table, which needs to huge with 112 cells.

* Correlations in the SIMMUN cohort. I see, this may be a bit uncomfortable to have less figures with the 'own' dataset than with an external one. I have personally no problem with that, and it is actually a standard in most oncology papers nowadays, where you end up with lets say 2 - 20 published validation cohorts for a single 'own' collective (see: [Pichler et al 2023](https://www.frontiersin.org/articles/10.3389/fimmu.2023.1095195/full) for an extreme example - one single panel of a figure with 6 own patients; 800 external cases). Unfortunately, almost all effects investigated by univariable analysis (Figures S4 - S5) in the SIMMUN cohort were not significant and I'm not inclined to expose them and discuss too much.

* Effect size descriptions were added to the correlation analysis results in the INCOV cohort. I'm also elaborating on effect sizes of the R^2^ of our models.

* The Result section has now approx. 1800 Words.

# Discussion

* As above I've tried to accommodate all issues, but we are limited by the hard 2000 word limit according to the [journal's guidelines](https://www.elsevier.com/wps/find/journaldescription.cws_home/473?generatepdf=true). Please note that references cost 2 - 4 words per citation, I shorted chained citations as well.

* For this reason I'm not so sure if subheadings in Discussion are a good idea. I'm providing provisional subheadings (yellow), which I'd remove prior to re-submission.

* The opening sentence: "We investigated the bidirectional relationship between the mental and physical health in individuals infected with SARS-CoV-2 and uninfected controls in two separate cohorts" - I fear the Reviewers won't be happy to see that, we actually had a look at activity of metabolic pathways which regulate systemic neurotransmitter availability and MAY have an effect on mental health. It suggest that psychometric measures were analyzed as independent variables as well, which was not the case. Our model was: demography + clinical history + inflammation + infection + SARS-CoV-2 immunity + psychometry -> metabolic pathway activity.

* My idea for the first section was to provide a short summary which unites the results from the SIMMUN and INCOV cohort, i.e. to provide a bigger picture without referencing single effects which are anyway presented in Results. I feel this may convince Reviewer 2, who claimed our data is complex. I would point out only the most relevant explanatory factors and mechanisms there and go into details later on in Discussion. I would also keep two-three sentences in this initial paragraph which discuss the major discrepancies of out data.

* The following sections discuss roles of inflammation, infection and infection-independent factors in the serotonin, kynurenine and catecholamine pathway activity.

* The final section tackles the effects and controversies over serotonin, kynurenine and catecholamine metabolism and mental health. I decided not to go too much into the mental stress thread of Discussion because of the word limit. In our SIMMUN cohort, likely due to a lot of psychiatric patients there's a huge overlap between stress, depression and anxiety (rho > 0.65). I have somehow a feeling that this purely random, which of them associate with KYN or KYN/TRP.

* I'm keeping the sentence stating that > 90% TRP is catabolized by IDO/TDO, this was a thread proposed by Reviewer 1.

* I'd omit the study of Hamilton 2021, with pre-pandemic inflammation. It has actually nothing to do with COVID-19, but says that people with inflammation for whatever reasons were mentally hit by the pandemic, irrespectively of the infection. It's obvious since people with chronic inflammation were canonical COVID-19 risk groups (multi-morbidity, frailty) and hence subject to particularly harsh containment measures like quarantine, social distancing. They do not control for that at all!

* I'm not referencing Alzahrani 2021. This paper tackles severe COVID-19, which, like sepsis (I used to work with sepsis for years in mice and men), is characterized by a dramatic corticosteroid release (you may see it as a way to harness cytokine storm during COVID-19) or, in some patients, by no release at all (you may see it as a multi-organ failure). If somebody experiences one of those is happy if survived and mental stress is probably just a minor problem.

# References

### 2023-06-07 paper/markdown/main_manuscript.Rmd
Analysis and manuscript files with feedback
Author's feedback included in the analyses and manuscript files. Effect of gender and BMI included in the INCOV cohort multi-parameter models. Changes and shortening of the manuscript text.

--- 2023-05-16 paper/markdown/main_manuscript.Rmd
@@
 ---
-title: "Persistent low-grade inflammation, SARS-CoV-2 infection and mental health deterioration signs impact on systemic levels of aminoacid neurotransmitter precursors"
+title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
 subtitle: "Manuscript"
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
+
+author:
+  - Katharina Hüfner:
+      institute: [psych2]
+      email: katharina.huefner@tirol-kliniken.at
+      correspondence: true
+  - Sophia Vedova:
+      institite: [psych2]
+  - Piotr Tymoszuk:
+      institute: [daas]
+  - Philipp Nelles:
+      institute: [psych2]
+  - Tobias Bruckner: 
+      institute: [psych2]    
+  - Eberhard Deisenhammer: 
+      institute: [psych2, psych1]
+  - Jonas Egeter: 
+      institute: [psych2]
+  - Alexander Egger: 
+      institute: [zimcl]
+  - Matyas Galffy: 
+      institute: [psych2]
+  - Johannes Giesinger: 
+      institute: [psych2]
+  - Jens Lehmann: 
+      institute: [psych1]
+  - Maria Oberhammer: 
+      institute: [psych1]
+  - Joachim Rockenschaub: 
+      institute: [psych2]
+  - Magdalena Sacher: 
+      institute: [surg]
+  - Bernhard Holzner: 
+      institute: [psych2]
+  - Johanna M Gostner: 
+      institute: [bio]
+  - Barbara Sperner-Unterweger: 
+      institute: [psych2]
+
+institute:
+  - psych2: University Hospital of Psychiatry II, Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, Medical University Innsbruck, Innsbruck, Austria.
+  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria. 
+  - psych1: University Hospital of Psychiatry I, Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, Medical University Innsbruck, Innsbruck, Austria.
+  - surg: Department of Visceral, Transplant and Thoracic Surgery (VTT), Medical University Innsbruck, Innsbruck, Austria. 
+  - zimcl: Central Institute of Clinical and Chemical Laboratory Diagnostics, University Hospital of Innsbruck, Innsbruck, Austria.
+  - bio:  Insitute of Medical Biochemistry, Medical University Innsbruck, Innsbruck, Austria
 
@@
 
-csl: frontiers-in-immunology.csl
+csl: psychoneuroendocrinology.csl
 
@@
 
+Running title: Activity of the kynurenine, serotonin and dopamine metabolic pathways and COVID-19
+
+\newpage
+
+# Abstract
+
+__Background:__ COVID-19 patients frequently develop symptoms of depression, anxiety and mental stress. 
+The serotonin, kynurenine and catecholamine metabolic pathways determine systemic levels of serotonin and dopamine, whose insufficiency was linked to mental health disorders. 
+We investigated factors that affect their activity in SARS-CoV-2 infection and recovery.
+
+__Methods:__ The cross-sectional SIMMUN (n = `r nrow(stigma$analysis_tbl)`) and longitudinal INCOV cohort (n = `r  length(unique(incov$complete_ids))`, Su et al. 2022) were analyzed. 
+Demographic and clinical characteristic, inflammatory markers, SARS-CoV-2 infection, signs of depression and anxiety (HADS), and mental stress (PSS-4) served as explanatory variables. 
+Plasma markers of the serotonin and kynurenine (serotonin, kynurenine/tryptophan ratio), and catecholamine pathway activity (dopamine 3-O-sulfate, phenylalanine/tyrosine ratio) were modeled by multi-parameter linear regression.
+
+__Results:__ In the SIMMUN cohort, the inflammatory marker 
+neopterin ($\beta$ = `r my_beta(mod_eli$inference$kyn_trp, 'log_neo')`), 
+SARS-CoV-2-positivity (`r my_beta(mod_eli$inference$kyn_trp, 'infection', show_ci_txt = FALSE)`), 
+mental stress (`r my_beta(mod_eli$inference$kyn_trp, 'pss_stress_score', show_ci_txt = FALSE)`) and age were positively associated with elevated kynurenine/tryptophan ratio. 
+Phenylalanine/tyrosine ratio was lower in 
+SARS-CoV-2-positive participants (`r my_beta(mod_eli$inference$phe_tyr, 'infection', show_ci_txt = FALSE)`). 
+In the INCOV cohort, inflammation markers were associated with decreased serotonin 
+(IL6: `r my_beta(incov_neuro$inference$serotonin, 'IL6_INF', show_ci_txt = FALSE)`) and dopamine 3-O-sulfate 
+(interferon-gamma: `r my_beta(incov_neuro$inference$dopamine, 'IFNG_INF', show_ci_txt = FALSE)`). 
+Serotonin (`r my_beta(incov_neuro$inference$serotonin, 'timepoint', 'recovery', show_ci_txt = FALSE)`) and 
+dopamine 3-O-sulfate levels (`r my_beta(incov_neuro$inference$dopamine, 'timepoint', 'recovery', show_ci_txt = FALSE)`) were higher during recovery than in acute SARS-CoV-2 infection.
+
+__Conclusion:__ SARS-CoV-2 infection, inflammation, mental stress and age are the key independent predictors of high kynurenine pathway activity, which decreases systemic serotonin levels by depleting its precursor tryptophan.
+These factors are also associated with suppressed catecholamine pathway activity. 
+This lowered serotonin and dopamine availability may contribute to mental health problems in COVID-19 patients.
+
+# Keywords
+
+serotonin and dopamine availability, COVID-19, kynurenine, tryptophan, inflammation, mental health
+
 \newpage
@@
 
+The immune system and the brain interact at multiple levels [@Dantzer2008]. 
+Physical conditions such as infections, cardiovascular disease, frailty, autoimmune illness or malignancy are often accompanied by symptoms of depression, anxiety, mental stress and other psychiatric disorders [@Mitchell2011; @Dickens2001; @Mayerl2020; @Rajan2020; @Renault1987; @Hufner2015; @Hufner2022].
+Inflammation interferes with key metabolic processes controlling serotonin and dopamine availability, the serotonin, kynurenine and catecholamine pathways, and was proposed as an important link between mental and physical health [@Bonaccorso2002; @Gostner2020; @Valdiglesias2018; @Dantzer2008; @Hufner2015; @Brown2021].
+
+Serotonin is synthesized from the essential aminoacid tryptophan (TRP) via serotonin pathway reactions catalyzed by tryptophan hydroxylase (TPH) and aromatic L-amino acid decarboxylase. 
+Most systemic TRP is catabolized by the kynurenine pathway, which limits serotonin pathway activity [@Lukic2022]. 
+Its first rate-limiting step, the TRP - kynurenine (KYN) breakdown catalyzed by tryptophan 2,3-dioxygenase (TDO) or indoleamine 2,3-dioxygenases (IDO1, IDO2), controls TRP levels available for serotonin synthesis [@Lukic2022; @Badawy2017].
+Importantly, IDO1 is stimulated by inflammatory signals [@Badawy2017; @Robinson2003]
+and TDO activity is increased by glucocorticoids released upon depression and mental stress [@Knox1951; @Maes1990]. 
+Kynurenine pathway metabolites such as KYN or quinolinic acid [QUIN] can additionally pass the blood-brain barrier, interfere e.g.
+glutamatergic signaling and exert neurotoxicity [@Brown2021; @Schwarcz2012].
+Hence, the kynurenine pathway, whose activity can be assessed by the ratio of blood KYN to TRP (KYN/TRP), poses a connex of inflammatory and hypothalamic–pituitary–adrenal axis signals, systemic serotonin availability and neural signal transmission [@Badawy2017; @Brown2021; @Schwarcz2012].
+Consequently, elevated levels of circulating kynurenine pathway metabolites, high KYN/TRP, decreased TRP or lowered serotonin were reported for numerous physical conditions and associated with symptoms of mental disorders [@Cervenka2017; @Marx2020; @Hunt2020; @Fellendorf2022; @Hufner2015; @Hufner2021; @Hufner2019; @Capuron2011].
+
+Dopamine synthesis via the catecholamine pathway involves the conversion of phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH).
+PAH, TH and the serotonin pathway enzyme TPH rely on tetrahydrobiopterin (BH4) as an essential cofactor [@Meiser2013; @Neurauter2008]. 
+During inflammation, BH4 is oxidized to its inactive derivative and consumed in a competitor reaction of nitric oxide formation [@Rahimian2022a; @Neurauter2008]. 
+As a result of such functional BH4 deficiency, reflected e.g. by an elevated blood PHE to TYR ratio, biosynthesis of dopamine and serotonin is suppressed [@Geisler2013; @Vancassel2018; @Capuron2011]. 
+Accordingly, changes in PHE and TYR levels have been reported in different physical disorders such as cancer, infections and inflammatory conditions and associated with depression and anxiety [@Vancassel2018; @Geisler2013; @Hufner2015; @Capuron2011].
+
+SARS-CoV-2 virus is the causal pathogen of coronavirus disease 2019 (COVID-19). 
+Apart from sustained physical disability, psychiatric disorders in millions of patients amount to the persistent burden of the COVID-19 pandemic [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Staudt2022].
+Increased kynurenine pathway activity suggestive of lowered systemic serotonin availability has recently been identified in acute COVID-19 and during COVID-19 recovery and associated with infection severity [@Santiago-Mujika2022; @Lionetto2021; @Bizjak2022; @Dewulf2022].
+Theoretical considerations [@Bower2022] and preliminary experimental data suggest that highly active kynurenine pathway along with suppression of serotonin and catecholamine biosynthesis may contribute to symptoms of mental stress, anxiety and depression in COVID-19 patients [@Matits2023; @Kucukkarapinar2022].
+
+Still, clinical and experimental evidence for interaction of inflammation, mental health and systemic availability of serotonin and dopamine in COVID-19 is scarce. 
+To address this question, we explored effects of demographic, clinical, psychometric (anxiety, depression, mental stress), inflammation- and SARS-CoV-2-related factors on markers of systemic activity of the serotonin, kynurenine and catecholamine pathways in two cross-sectional cohorts of SARS-CoV-2 patients: the SIMMUN cohort and the published longitudinal INCOV study [@Su2022].
+
 \newpage
@@
 # Materials and Methods
+
+Details on study cohorts, procedures and analysis are provided in __Supplementary Methods__.
 
@@
 Participants' data were stored and analyzed in anonymized form.
-This study was approved by the ethics committee of the Medical University Innsbruck, Austria (approval number: 1132/2020). 
-No approval by the ethics committee was required for analysis of the published anonymized INCOV data set [@Su2022].
+This study was approved by the ethics committee of the Medical University Innsbruck (Austria, approval number: 1132/2020). 
+No approval by the ethics committee was required for the published INCOV dataset [@Su2022].
 
@@
 
-Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck Medical University (Innsbruck, Austria) and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR screening were invited to participate in the SIMMUN study.
-The study enrollment was conducted between 10. June 2020 (first patient enrolled) and 27.May 2021 (last study visit). 
-The inclusion criteria were age of 18 - 70 years, proficiency in German language, residence in the study region (Tyrol, Austria), results of a SARS-CoV-2 PCR test conducted at the University Hospital of Innsbruck.
-The exclusion criteria were active SARS-CoV-2 infection (< 14 days following a positive test), pregnancy, active malignancies, organ transplantation, prior surgery in the past 3 months, or acute or chronic inflammatory illness and treatment with oral corticosteroids.
-The analysis inclusion criterion was the complete study variable dataset (__Supplementary Table S1__). 
-A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were included in the analysis (__Figure \@ref(fig:fig-inclusion)__, __Table \@ref(tab:tab-stigma-cohort)__). 
-Significant differences between the analyzed participants and participants excluded due to data missingness are listed in __Supplementary Table S2__.
+Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR were invited to participate.
+The study was conducted between June 2020 and May 2021. 
+The inclusion criteria were age of 18 - 70 years, proficiency in German, residence in the study region (Tyrol, Austria), and a SARS-CoV-2 PCR conducted at the study site. 
+The exclusion criteria were active SARS-CoV-2 infection (< 14 days after diagnosis), pregnancy, active malignancy, organ transplantation, surgery in the past 3 months, inflammatory illness and oral corticosteroid treatment.
+Only individuals with complete explanatory and response variables presented in __Supplementary Table S1__ and __Supplementary Figure S1__ were analyzed. 
+Significant differences between the analyzed and excluded participants are listed in __Supplementary Table S2__.
 
@@
 
-Proteome and metabolome data and clinical information for the INCOV cohort were extracted from supplementary tables accompanying the report by Su and colleagues [@Su2022]. 
-In the current analysis, `r nrow(incov$analysis_tbl)` INCOV study participants with the complete age information and dataset of metabolites related to serotonin and dopamine turnover and major inflammatory cytokines were included (__Figure 1__, __Table \@ref(tab:tab-incov-cohort)__, __Supplementary Table S3__).
+Proteome, metabolome and clinical data are [publicly available](https://data.mendeley.com/datasets/96v329bg7g/1) [@Su2022]. 
+Samples with complete explanatory and response variables presented in __Supplementary Figure S1__ and __Supplementary Table S3__ were analyzed.
 
@@
 
-Details on study procedures are provided in __Supplementary Methods__.
-
 ### SIMMUN study
 
-The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
-The SIMMUN study data were gathered during in-person visits including a physician assessment, supervised completion of self-rating questionnaire and a blood sample collection.
-
-Demographic and clinical history variables: age, sex, body mass index (BMI), professionally diagnosed mental illness, self-reported chronic somatic conditions, smoking and alcohol consumption history, result and date of the SARS-CoV-2 PCR test were surveyed during the study visit or extracted from electronic patient records.
-
-Inflammatory markers: plasma neopterin concentrations (NEO) and neutrophil - lymphocyte ratio (NLR) were determined by the certified clinical routine laboratory at the University Hospital of Innsbruck.
-Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, and the KYN/TRP and PHE/TYR ratios calculated [@Capuron2011; @Widner1997; @Neurauter2008].
-Plasma titer of immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) were quantified by ELISA as described before [@Deisenhammer2021]. 
-Laboratory measurements at and beyond the detection limits were substituted with the lower or upper detection limit value, respectively.
-
+The SIMMUN data were collected during a single on-site visit at median 139 days after the SARS-CoV-2 PCR (interquartile range: 119 - 157) or extracted from electronic patient records (__Supplementary Table S1__).
+Demographic and clinical characteristic included age, sex, body mass index (BMI), mental disorders diagnosed by a physician, self-reported physical disorders, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR.
+Inflammatory markers, plasma neopterin concentrations (NEO) measured by ELISA [@Werner1990] and neutrophil/lymphocyte ratio (NLR) were determined by a certified laboratory at the University Hospital of Innsbruck.
+Plasma TRP, KYN, PHE and TYR were determined by high-performance liquid chromatography [@Widner1997; @Neurauter2008]. 
+Immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) was quantified by ELISA [@Deisenhammer2021]. 
+Laboratory measurements beyond the detection limits were substituted with the respective detection limit value. 
 Mental stress was scored with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
-Anxiety and depression signs were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
-The total possible score range for each subscale is 0 to 21, with higher scores indicating more intense symptoms of anxiety/depression. 
-Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points [@Bjelland2002; @Zigmond1983].
-
-In order to improve normality of some numeric study variables prior to linear modeling and statistical hypothesis testing with parametric tools, logarithm or square root transformations were applied (__Supplementary Table S1__)
+Clinically relevant symptoms of anxiety and depression were defined as $\geq$ 8 points in the hospital anxiety and depression scale (HADS) [@Bjelland2002; @Zigmond1983].
 
@@
 
-Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and ultra-high-performance liquid chromatography/tandem accurate mass spectrometry (Metabolon, USA) [@Su2022; @Su2020].
-The proteome, metabolome data in form of normalized, age- and sex-adjusted, log~2~-transformed plasma concentrations as well as clinical information were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) of the report by Su et al.
-[@Su2022].
-In the current report plasma levels of metabolites implicated in biosynthesis and turnover, and competitor pathway products for serotonin (serotonin [5-HT], TRP, KYN, quinolinate [QUIN]) and dopamine (PHE, TYR and dopamine 3-O-sulfate [DA sulfate]), as well as major inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) were analyzed (__Supplementary Table S3__).
+Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and liquid chromatography/tandem mass spectrometry (Metabolon, USA), respectively, and expressed as age- and sex-adjusted, log~2~-transformed concentrations [@Su2022; @Su2020].
+Markers of the serotonin (serotonin, TRP) [@Lukic2022], kynurenine (KYN, kynurenine/tryptophan ratio [KYN/TRP], QUIN) [@Badawy2017] and catecholamine pathway activity (PHE, TYR, phenylalanine/tyrosine ratio [PHE/TYR] and dopamine 3-O-sulfate [DA sulfate]) [@Meiser2013; @Goldstein1999; @Suominen2013], cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]), age, sex and BMI were analyzed (__Supplementary Table S3__).
 
-Plasma metabolites and proteome were measured in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['acute']` days), 
-sub-acute (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['sub-acute']`) and 
-recovery (median: `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~median(.x$time_po)) %>% signif(2))['recovery']` after diagnosis) (__Supplementary Table S3__).
+Plasma samples were collected in during in acute and sub-acute infection, and recovery at median 
+`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']`, 
+`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']`), 
+`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR, respectively (__Supplementary Table S4__).
 
-## Study endpoints
+## Analysis endpoints
 
-The primary analysis endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing plasma levels of serotonin and dopamine precursors and products of competitor metabolic pathways (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR) in a real world collective of medical service users during the SARS-CoV-2 pandemics.
-This endpoint was addressed by multi-parameter modeling and univariable statistical hypothesis testing in the SIMMUN cohort. 
-The secondary analysis endpoint was to investigate how the readouts of systemic serotonin and dopamine availability, plasma 5-HT and DA sulfate, are influenced by age, inflammatory cytokines, timepoint of the SARS-CoV-2 infection and serum levels of their precursors (TRP, PHE, TYR) and competitor pathway products (KYN, QUIN).
-This endpoint was addressed by multi-parameter robust linear modeling, univariate time course modeling and correlation analysis in the INCOV cohort (__Figure \@ref(fig:fig-inclusion)__).
+The first endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing plasma markers of kynurenine and catecholamine pathway activity (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR) by multi-parameter modeling in the SIMMUN cohort.
+The second endpoint was to investigate how markers of systemic serotonin and dopamine availability (DA sulfate), are influenced by age, sex, body mass, inflammatory cytokines, timepoint of SARS-CoV-2 infection, neurotransmitter precursor amino acids (TRP, PHE, TYR) and kynurenine pathway metabolites (KYN, QUIN).
+This endpoint was addressed by multi-parameter robust linear modeling, time course modeling and correlation analysis in the INCOV cohort (__Supplementary Figure S1__).
 
@@
 
-Details of data transformation and analysis are provided in __Supplementary Methods__.
-
 R version 4.2.3 was used for statistical analysis.
 
-Numeric variables were presented in the text and tables as medians with interquartile ranges and ranges. 
-Categorical variables are presented as percentages and counts within the complete observation set. 
-Distribution normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively. 
-In the SIMMUN cohort, non-normally distributed numeric variables were transformed with the logarithm or square root functions prior to modeling and analyses with parametric tests (__Supplementary Table S1__).
-Since most of the INCOV study variables were non-normally distributed, robust linear modeling and non-parametric testing were employed in the analysis of those data. 
-Uniformity and consistency of psychometric tools was investigated by factor analysis and global McDonald's $\omega$ statistic [@McDonald1999; @Bartlett1937; @Revelle2022]. 
-Except for results of multi-parameter modeling, p values were corrected for multiple testing with the false discovery rate method [@Benjamini1995] separately for each analysis task. 
+Numeric variables were presented as medians with interquartile ranges and ranges. 
+Categorical variables are presented as percentages and counts. 
+Normality and variance homogeneity was assessed by Shapiro-Wilk and Levene test, respectively. 
+In the SIMMUN cohort, non-normally distributed numeric variables were logarithm- or square root-transformed prior to modeling and parametric tests (__Supplementary Table S1__). 
+Since most of the INCOV study variables were non-normally distributed, robust linear modeling and non-parametric testing were employed. 
+Consistency of psychometric tools was investigated by global McDonald's $\omega$ [@McDonald1999]. 
+Except for multi-parameter modeling, p values were corrected for multiple testing with the false discovery rate method [@Benjamini1995] separately for each analysis task. 
 Effects with p < 0.05 were considered significant.
 
-Correlation was assessed by Spearman's rank or Pearson test. 
-Significance of comparisons of numeric variables between two groups was determined by Mann-Whitney test with r effect size statistic or by two-tailed T test with Cohen's d effect size metric. 
-Comparisons of categorical variable distributions were evaluated by $\chi^2$ test with Cramer V effect size statistic. 
-Correlation was assessed by Pearson's or Spearman's rank test [@Cohen2013; @Kassambara2021].
+Significance was determined by Mann-Whitney test with r effect size, two-tailed T test with Cohen's d effect size statistic or $\chi^2$ test with Cramer's V effect size statistic. 
+Correlation was assessed by Pearson's or Spearman's rank test and visualized as bubble or force-directed network plots [@Csardi2006].
 
-In the SIMMUN cohort, effects of age, sex, body mass class, somatic and metal conditions, body mass class, smoking and alcohol consumption history, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, signs of depression and anxiety (HADS), mental stress scoring (PSS-4) on systemic levels of aminoacid neurotransmitter precursor and competitor pathway products (TRP, KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio) was assessed by multi-parameter linear regression with backward elimination.
-Modeling responses and explanatory variables were subjected to normality-stabilizing transformations (__Supplementary Table S1__) and normalized prior to modeling in the SIMMUN cohort. 
-The normality and homogeneity of linear model residuals were evaluated by Shapiro-Wilk and Levene test, respectively, and visual inspection (residuals versus fitted and quantile-quantile plots). 
-In the INCOV cohort, effects of age, plasma levels of inflammatory cytokines (IL6, IL10, TNF, IFNG), timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control) and plasma levels of neurotransmitter precursors and competitor pathway products (TRP, KYN, QUIN, PHE, TYR) on plasma levels of 5-HT and DA sulfate were modeled by multi-parameter robust linear regression with the MM algorithm and Huber psi function [@Huber2011; @Ripley2022].
-Reproducibility and proper parameterization of the the linear and robust multi-parameter models was investigated by comparison of the RMSE and $R^2$ statistics in with the training dataset and in 10-fold cross-validation [@Kuhn2008].
-Significance of the model estimates was assessed by two-tailed T test.
-
-Differences in normalized blood concentrations of inflammatory cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery in the INCOV collective were investigated with robust linear modeling (MM algorithm, Huber's psi function) with uninfected subset serving as baseline.
-
-Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of inflammatory cytokines and metabolites in the INCOV cohort. 
-The correlation matrices were subsequently scaled into the [0, 1] range and converted to undirected force directed graphs [@Csardi2006]. 
-The graphs were visualized as two-dimensional network plots with the node proximity determined by the $\rho$ value and random repulsion, and edge color and width coding for the $\rho$ value and sign [@Briatte2021].
+Responses and explanatory variables were normalized prior to modeling. 
+In the SIMMUN cohort, effects of age, sex, body mass class, physical and psychiatric disorders, BMI, smoking and alcohol consumption, inflammation markers (NEO, NLR), SARS-CoV-2 infection, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress (PSS-4) on pathway activity markers were assessed by multi-parameter linear regression with backward elimination.
+In the INCOV cohort, effects of age, sex, BMI, cytokine markers of inflammation (IL6, IL10, TNF, IFNG), SARS-CoV-2 infection timepoint (uninfected, acute, sub-acute, recovery), serotonin and catecholamine pathway metabolites (TRP, PHE, TYR) and kynurenine pathway metabolites (QUIN, KYN) on plasma serotonin and DA sulfate were modeled by multi-parameter robust linear regression [@Huber2011].
+Explanatory performance, reproducibility and proper parameterization of the multi-parameter models was investigated by R^2^ and root mean square error (RMSE) in 10-fold cross-validation. 
+Differences in cytokines and metabolites between SARS-CoV-2 infection timepoints in the INCOV collective were investigated by robust linear modeling with uninfected subset or acute infection serving as baselines.
 
@@
 
+## Characteristic of the cohorts
+
+Two cohorts of uninfected and SARS-CoV-2-infected individuals were analyzed (SIMMUN and INCOV). 
+Out of `r nrow(stigma$data)` individuals enrolled in the SIMMUN study, 
+`r nrow(stigma$analysis_tbl)` participants with complete study variables were analyzed (__Supplementary Figure S1__, __Supplementary Table S1__). 
+The excluded individuals were characterized by more frequent mental disorders and clinically relevant symptoms of depression and anxiety (HADS), higher scores of mental stress (PSS-4) and less frequent SARS-CoV-2 infections than the analyzed participants (__Supplementary Table S2__).
+In the analyzed SIMMUN cohort, males represented 
+`r my_percent(cohort$analysis_tbl$stigma, 'sex')['male']`% and the median age was 
+`r median(cohort$analysis_tbl$stigma$age)` years. 
+The SIMMUN study data were recorded at median 139 days after the SARS-CoV-2 PCR. 
+None of the participants had received anti-SARS-CoV-2 vaccination prior to enrollment. 
+SARS-CoV-2-positive individuals accounted for 
+`r my_percent(cohort$analysis_tbl$stigma, 'cov')['SARS-CoV-2']`% of the analyzed participants. 
+The gender and age structure of the SARS-CoV-2-negative and -positive subsets was comparable. 
+In the entire SIMMUN cohort, 
+`r signif(sum(table(cohort$analysis_tbl$stigma$bmi_class)[c(2, 3)])/nrow(cohort$analysis_tbl$stigma), 2) * 100`% of participants were overweight or obese, and `r my_percent(cohort$analysis_tbl$stigma, 'somatic_comorb')['yes']`% reported a physical disorder; these figures were comparable between SARS-CoV-2-negative and -positive individuals.
+Mental disorders diagnosed by a physician affected 
+`r my_percent(cohort$analysis_tbl$stigma, 'psych_comorb')['yes']`% of participants and were significantly more common in SARS-CoV-2-negative 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% my_percent('psych_comorb') %>% .['yes']`%) than in SARS-CoV-2-positive individuals 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('psych_comorb') %>% .['yes']`%, 
+`r cohort$clin_test$stigma %>% filter(variable == 'psych_comorb') %>% .$significance`, 
+effect size: 
+`r cohort$clin_test$stigma %>% filter(variable == 'psych_comorb') %>% .$eff_size`) (__Table \@ref(tab:tab-stigma-cohort)__). 
+The psychometric tools for assessment of perceived mental stress (PSS-4), depression and anxiety (HADS) displayed good-to-excellent internal consistency 
+($\omega$ = `r cons$omega_stats %>% filter(factor == 'g') %>% .$total %>% range %>% signif(2) %>% paste(collapse = ' - ')`) [@McDonald1999] (__Supplementary Figure S2__). 
+Clinically relevant symptoms of anxiety (HADS $\geq$ 8) were more frequent in SARS-CoV-2-negative
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% my_percent('hads_anx_score') %>% .[2]`%) than in SARS-CoV-2-positive participants 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('hads_anx_score') %>% .[2]`%, 
+`r cohort$clin_test$stigma %>% filter(variable == 'hads_anx_score') %>% .$significance`, 
+effect size: 
+`r cohort$clin_test$stigma %>% filter(variable == 'hads_anx_score') %>% .$eff_size`). 
+Clinically relevant symptoms of depression (HADS $\geq$ 8) were more common in SARS-CoV-2-negative 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% my_percent('hads_dpr_score') %>% .[2]`%) than in SARS-CoV-2-positive participants 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('hads_dpr_score') %>% .[2]`%), yet this effect was no statistically significant. 
+Scores of mental stress were comparable in both SARS-CoV-2 subsets 
+(negative: median `r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% .$pss_stress_score %>% median`, 
+positive: median `r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% .$pss_stress_score %>% median` points). 
+As expected, titer of antibodies against the S1/S2 SARS-CoV-2 protein (anti-RBD IgG) was significantly higher in the SARS-CoV-2-positive subset 
+(median: `r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% .$anti_rbd %>% median %>% signif(2)` AU) as compared with uninfected individuals 
+(median: `r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% .$anti_rbd %>% median %>% signif(2)` AU, 
+`r cohort$clin_test$stigma %>% filter(variable == 'anti_rbd') %>% .$significance`, 
+effect size: `r cohort$clin_test$stigma %>% filter(variable == 'anti_rbd') %>% .$eff_size`). 
+In 73% of SARS-CoV-2-infected SIMMUN study participants, the infection was mild and treated on an outpatient basis (__Table \@ref(tab:tab-stigma-cohort)__).
+
+Out of 
+`r intersect(incov$proteome$patient_id, incov$metabolome$patient_id) %>% length` individuals initially enrolled in the INCOV study [@Su2022], 
+`r length(unique(incov$complete_ids))` participants with complete study variables providing 
+`r length(incov$complete_ids)` metabolome and proteome samples were analyzed (__Supplementary Figure S1__, __Supplementary Tables S3 - S4__). 
+In the analyzed INCOV cohort, the median age was 
+`r signif(median(cohort$analysis_tbl$incov$age), 2)` years and 
+`r my_percent(cohort$analysis_tbl$incov, 'sex')['male']`% of participants were male. 
+SARS-CoV-2-positive individuals comprised 
+`r my_percent(cohort$analysis_tbl$incov, 'cov')['SARS-CoV-2']`% of the cohort. 
+SARS-CoV-2-positive individuals 
+(median age: `r cohort$analysis_tbl$incov %>% filter(cov == 'SARS-CoV-2') %>% .$age %>% median %>% signif(2)` years) were significantly older than uninfected participants 
+(median age: `r cohort$analysis_tbl$incov %>% filter(cov == 'uninfected') %>% .$age %>% median %>% signif(2)` years, 
+`r cohort$clin_test$incov %>% filter(variable == 'age') %>% .$significance`, 
+effect size: `r cohort$clin_test$incov %>% filter(variable == 'age') %>% .$eff_size`). 
+Percentages of overweight or obese individuals were higher among SARS-COV-2-positive 
+(`r cohort$analysis_tbl$incov %>% filter(cov == 'SARS-CoV-2') %>% my_percent('bmi_class') %>% .[c(2, 3)] %>% sum`%) 
+than SARS-CoV-2-negative participants 
+(`r cohort$analysis_tbl$incov %>% filter(cov == 'uninfected') %>% my_percent('bmi_class') %>% .[c(2, 3)] %>% sum`%), this effect was not statistically significant. 
+Nearly all 
+(`r cohort$analysis_tbl$incov %>% filter(cov == 'SARS-CoV-2') %>% my_percent('severity') %>% .[c('moderate', 'severe', 'critical')] %>% sum`%) of SARS-CoV-2-positive participants of the INCOV study experienced moderate-to-critically severe infections and were hospitalized during COVID-19 (__Table \@ref(tab:tab-incov-cohort)__).
+
+As compared with the SIMMUN cohort, the INCOV cohort had a significantly higher percentage of males (`r comp$test %>% filter(variable == 'sex') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'sex') %>% .$eff_size`), 
+higher age (`r comp$test %>% filter(variable == 'age') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'age') %>% .$significance`), 
+higher rates of overweight or obese participants (`r comp$test %>% filter(variable == 'bmi_class') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'bmi_class') %>% .$eff_size`), 
+more frequent SARS-CoV-2 infections (`r comp$test %>% filter(variable == 'cov') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'cov') %>% .$eff_size`), and 
+higher hospitalization rates (`r comp$test %>% filter(variable == 'severity') %>% .$significance`, 
+effect size: `r comp$test %>% filter(variable == 'severity') %>% .$eff_size`) (__Supplementary Table S5__).
+
+## Inflammation, SARS-CoV-2 infection, age, mental stress and depression influence systemic activity of the kynurenine pathway
+
+In the SIMMUN dataset, we initially searched for predictors of activity of the kynurenine and catecholamine pathways. 
+TRP, KYN and KYN/TRP served as markers of the kynurenine pathway [@Badawy2017], and PHE, TYR and PHE/TYR were investigated as markers of the catecholamine pathway activity [@Meiser2013]. 
+The candidate explanatory variables were age, sex, BMI, the presence of physical and mental disorders, smoking, alcohol consumption, SARS-CoV-2 infection, anti-SARS-CoV-2 antibody titer, NEO and NLR as inflammatory markers, perceived mental stress, and symptoms of depression and anxiety (__Supplementary Table S1__).
+
+Meaningful multi-parameter models optimized by backward elimination could be established for TRP, KYN, KYN/TRP, TYR and PHE/TYR.
+The PHE model had a negligible explanatory performance, proved non-significant as compared with the intercept-only null model and was hence not further analyzed  (__Supplementary Table S6__).
+The remaining models were characterized by good reproducibility and proper parameterization as indicated by comparable fit errors (RMSE) in the training dataset and cross-validation. 
+The KYN and KYN/TRP ratio models had the best, moderate-to-substantial explanatory performance measured by cross-validated R^2^ of 
+`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn') %>% .$estimate %>% signif(2)` and 
+`r mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response == 'kyn_trp') %>% .$estimate %>% signif(2)`, respectively. 
+The cross-validated R^2^ metric values for the TRP, TYR and PHE/TYR models ranged from 
+`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[1]`% to 
+`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[2]` indicative of weak-to-moderate explanatory value (__Supplementary Figure S3__).
+
+Regarding the kynurenine pathway, the inflammatory marker NEO 
+($\beta$ = `r my_beta(mod_eli$inference$trp, variable = 'log_neo')`) 
+and clinically relevant depression symptoms 
+(HADS $\geq$ 8, $\beta$ = `r my_beta(mod_eli$inference$trp, variable = 'hads_dpr_score')`) were independently associated with reduced plasma TRP. 
+KYN and KYN/TRP were significantly associated with 
+SARS-CoV-2 infection (KYN: $\beta$ = `r my_beta(mod_eli$inference$kyn, variable = 'infection')`, 
+KYN/TRP: $\beta$ = `r my_beta(mod_eli$inference$kyn_trp, variable = 'infection')`), 
+higher NEO (KYN: $\beta$ = `r my_beta(mod_eli$inference$kyn, variable = 'log_neo')`, 
+KYN/TRP: $\beta$ = `r my_beta(mod_eli$inference$kyn_trp, variable = 'log_neo')`) and 
+age (KYN: $\beta$ = `r my_beta(mod_eli$inference$kyn, variable = 'age')`, 
+KYN/TRP: $\beta$ = `r my_beta(mod_eli$inference$kyn_trp, variable = 'age')`). 
+Additionally, high perceived mental stress (PSS-4) was identified as a significant predictor of increased KYN/TRP 
+($\beta$ = `r my_beta(mod_eli$inference$kyn_trp, variable = 'pss_stress_score')`).  
+High KYN concentrations were related to low NLR 
+($\beta$ = `r my_beta(mod_eli$inference$kyn, variable = 'log_nlr')`) (__Figure \@ref(fig:fig-modeling)A__).
+
+Concerning the catecholamine pathway, plasma TYR concentrations were negatively associated with NEO 
+($\beta$ = `r my_beta(mod_eli$inference$tyr, variable = 'log_neo')`) and positively with age 
+($\beta$ = `r my_beta(mod_eli$inference$tyr, variable = 'age')`). 
+SARS-CoV infection ($\beta$ = `r my_beta(mod_eli$inference$phe_tyr, variable = 'infection')`) 
+and age ($\beta$ = `r my_beta(mod_eli$inference$phe_tyr, variable = 'age')`) were significant predictors of lower PHE/TYR (__Figure \@ref(fig:fig-modeling)B__).
+
+By analysis of single significant predictors identified by modeling, we could corroborate significant, positive, moderate-to-strong correlations of age and NEO with KYN and KYN/TRP in the SIMMUN cohort.
+This analysis revealed also a significant positive correlation of age with TYR and a significant negative correlation of age with PHE/TYR (__Supplementary Figure S4 - S5__, __Supplementary Tables S7 - S8__).
+
+## Inflammatory cytokines, SARS-CoV-2 infection course, age and availability of biosynthesis precursors regulate systemic serotonin and dopamine levels
+
+Next, we investigated factors affecting plasma levels of serotonin [@Lukic2022] and a major circulating catabolite of dopamine, DA sulfate [@Meiser2013; @Goldstein1999], as direct markers of the serotonin and catecholamine pathway activity, respectively.
+This analysis was done in the INCOV cohort [@Su2022] with age, sex, BMI, cytokines markers of inflammation (IL6, IL10, TNF, IFNG), timepoint of SARS-CoV-2 infection, concentrations of serotonin and catecholamine pathway metabolites (TRP, PHE, TYR), and kynurenine pathway metabolite levels (KYN, QUIN) as candidate explanatory variables (__Supplementary Table S3__).
+Metabolites and cytokines were sampled in SARS-CoV-2-positive participants during acute and sub-acute infection, and recovery, at median 
+`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']`, 
+`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']` and 
+`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after infection diagnosis via PCR, respectively (__Supplementary Table S4__).
+
+The multi-parameter robust regression model [@Huber2011] for plasma serotonin was characterized by moderate explanatory performance 
+(cross-validated R^2^ = `r incov_neuro$cv_stats %>% filter(data_type == 'cv', response == 'serotonin', statistic == 'rsq') %>% .$estimate %>% signif(2)`). 
+The explanatory performance of the DA sulfate model was weak 
+(cross-validated R^2^ = `r incov_neuro$cv_stats %>% filter(data_type == 'cv', response == 'dopamine', statistic == 'rsq') %>% .$estimate %>% signif(2)`). 
+Comparable fit errors (RMSE) in the training dataset and cross-validation indicated good reproducibility and proper parameterization of the models (__Figure \@ref(fig:fig-incov-neuro)A__).
+
+In the INCOV collective, 
+`r incov_neuro$inference$serotonin %>% filter(variable != 'Intercept', p_value < 0.05) %>% filter(!duplicated(variable)) %>% nrow %>% int_to_text` significant predictors of plasma serotonin were identified.
+Age ($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'age')`), 
+male sex ($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'sex', 'male')`) and 
+the inflammatory cytokine 
+IL6 ($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'IL6_INF')`) were associated with significantly lower serotonin levels, whereas 
+TRP ($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'tryptophan')`) and recovery from SARS-CoV-2 infection 
+($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'timepoint', 'recovery')` versus uninfected controls) were predictors of higher plasma serotonin. 
+Significantly higher DA sulfate levels were identified for recovery from SARS-CoV-2 infection 
+($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'timepoint', 'recovery')` versus uninfected control) 
+and high TNF concentrations ($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'TNF_INF')`). 
+In turn, obesity ($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'bmi_class', 'obesity')` versus normal weight) and IFNG ($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'IFNG_INF')`) were independently associated with significantly decreased DA sulfate levels (__Figure \@ref(fig:fig-incov-neuro)B__, __Supplementary Table S9__).
+
+The association of serotonin and DA sulfate levels with inflammation was confirmed by time course modeling (__Figure \@ref(fig:fig-time-ourse)__, __Supplementary Tables S10 - S11__). 
+The maximum plasma concentrations of inflammation markers IL6, IL10, TNF and IFNG were observed during acute infection. 
+They were paralleled by changes in metabolites of the serotonin and kynurenine pathways with a significantly decreased 
+TRP ($\beta$ = `r my_beta(time_rlm$inference$uninfected$tryptophan, variables = 'timepoint', levels = 'acute')`), increased 
+KYN ($\beta$ = `r my_beta(time_rlm$inference$uninfected$kynurenine, variables = 'timepoint', levels = 'acute')`) and 
+elevated QUIN ($\beta$ = `r my_beta(time_rlm$inference$uninfected$quinolinate, variables = 'timepoint', levels = 'acute')`) during acute infection as compared with uninfected individuals.
+The inflammatory milieu of acute infection was also associated with significantly increased levels of the catecholamine pathway substrate PHE ($\beta$ = `r my_beta(time_rlm$inference$uninfected$phenylalanine, variables = 'timepoint', levels = 'acute')` versus uninfected controls).
+Resolution of systemic inflammation during sub-acute SARS-CoV-2 infection and 
+recovery was reflected by decreasing plasma levels of IL6, IL10, TNF and INFG as compared with acute disease. 
+This was paralleled by decreasing levels of 
+KYN (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$kynurenine, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$kynurenine, 'timepoint', 'recovery')`) and 
+QUIN (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$quinolinate, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$quinolinate, 'timepoint', 'recovery')`), and increasing 
+TRP (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$tryptophan, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$tryptophan, 'timepoint', 'recovery')`) and 
+serotonin levels (sub-acute: `r my_beta(time_rlm$inference$acute$serotonin, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$serotonin, 'timepoint', 'recovery')`) as compared with acute SARS-CoV-2 infection. 
+Similarly, infection recovery was associated with significantly decreasing 
+PHE (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$phenylalanine, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$phenylalanine, 'timepoint', 'recovery')`) and elevated 
+DA sulfate (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$dopamine.3.O.sulfate, 'timepoint', 'sub-acute')`, 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$dopamine.3.O.sulfate, 'timepoint', 'recovery')`) as compared with acute infection.
+
+In correlation analysis of inflammation-associated cytokines and metabolites (__Figure \@ref(fig:fig-incov-corr)A__, __Supplementary Figure S6__), 
+TRP  was negatively associated with all investigated cytokines during 
+acute ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'tryptophan', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) and 
+sub-acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'tryptophan', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) with weak-to-moderate and moderate-to-large effect size.
+Plasma serotonin levels correlated negatively with IL6 concentrations during acute infection 
+($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'serotonin', 'IL6_INF')`, effect size: weak) and correlated negatively with IL6, IL10, TNF and IFNG in sub-acute infection 
+($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'serotonin', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`, effect size: weak-to-moderate).
+By contrast, KYN (acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'kynurenine', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`; 
+sub-acute. $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'kynurenine', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) and 
+QUIN (acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'quinolinate', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`; 
+sub-acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'quinolinate', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) correlated positively with circulating IL6, IL10, TNF and IFNG with a moderate-to-large strength.
+Serotonin levels correlated with TRP during 
+acute ($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'serotonin', 'tryptophan')`) and 
+sub-acute infection ($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'serotonin', 'tryptophan')`) with weak and moderate effect size, respectively.
+
+Concerning the catecholamine pathway metabolites (__Figure \@ref(fig:fig-incov-corr)B__, __Supplementary Figure S6__), positive weak-to-moderate correlations of PHE with IL6, IL10 and TNF were observed in acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['dopamine.acute']]), 'phenylalanine', c('IL6_INF', 'IL10_INF', 'TNF_INF')) %>% range %>% paste(collapse = ' - ')`).
+PHE was weakly positively associated with IL10 in sub-acute SARS-CoV-2 infection 
+($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['dopamine.sub-acute']]), 'phenylalanine', 'IL10_INF')`). 
+In turn, TYR plasma concentrations correlated negatively with TNF and IFNG during acute 
+($\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['dopamine.acute']]), 'tyrosine', c('TNF_INF', 'IFNG_INF'))[1]`) and 
+sub-acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['dopamine.sub-acute']]), 'tyrosine', c('TNF_INF', 'IFNG_INF')) %>% range() %>% paste(collapse = ' - ')`) with weak effect size. 
+DA sulfate correlated negatively with IFNG and IL6 in acute 
+(IFNG, $\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['dopamine.acute']]), 'dopamine.3.O.sulfate', 'IFNG_INF')`) and sub-acute infection 
+(IL6, $\rho$ = `r my_rho(t(incov_mds$analysis_tbl[['dopamine.sub-acute']]), 'dopamine.3.O.sulfate', 'IL6_INF')`), respectively, with weak effect size.
+
 \newpage
@@
 
+## Result summary
+
+In two separate cohorts, we investigated factors affecting the key metabolic pathways determining systemic availability of serotonin and dopamine in SARS-CoV-2 infection. 
+Our analysis underlines the pivotal role of SARS-CoV-2-dependent and -independent inflammation on the serotonin and kynurenine pathway activity. 
+In particular, levels of the kynurenine pathway markers, KYN, KYN/TRP and QUIN, were higher, whereas concentration of the serotonin precursor TRP was lower, in the presence of inflammation reflected by elevated NEO or IL6.
+This observation and the significant association of serotonin with TRP levels in the INCOV cohort suggest collectively that depletion of TRP via highly active kynurenine pathway can limit systemic serotonin availability during SARS-CoV-2 infection [@Badawy2017].
+Age, male sex, mental stress and depression were proposed as infection-independent predictors of high kynurenine pathway activity, lowered TRP and, consequently, reduced serotonin formation. 
+Our data suggest similar inhibitory effects of inflammation on the first step of the catecholamine pathway, PHE - TYR conversion, and a reduced systemic dopamine availability reflected by lowered circulating DA sulfate [@Meiser2013].
+
+Few discrepancies need a more thorough discussion. 
+First, explanatory performance of analyses of the serotonin and kynurenine pathway activity measured by R^2^ and correlation coefficients were generally higher than explanatory values of the catecholamine pathway analyses.
+This indicates that the predictors identified by our analysis, inflammation, SARS-CoV-2-positivity and infection timepoint impact the TRP metabolic pathways to a larger degree than catecholamine metabolism, in line with a metabolomic study in COVID-19 individuals [@Thomas2020].
+Second, effects of inflammation on markers of the kynurenine and catecholamine pathway activity were stronger in the INCOV than in the SIMMUN cohort, which may reflect more severe infection in INCOV study participants.
+Third, in the SIMMUN cohort, the inflammatory marker NLR could be unexpectedly linked to lower KYN, which  may be explained by a sustained low-grade neutropenia reported for up to 1 year after COVID-19 [@Lin2022].
+
+## Effects of inflammation and infection on the serotonin and kynurenine pathways, and systemic levels of serotonin
+
+IDO1, together with TDO and IDO2, catalyzes the first reaction of the kynurenine pathway, the TRP - KYN conversion, which catabolizes >90% of TRP and may limit serotonin availability [@Badawy2017; @Lukic2022].
+Expression and activity of IDO1 are induced by inflammatory stimuli such as IL6, TNF and IFNG [@Robinson2003; @Badawy2017].
+Consequently, elevated metabolite levels and activity readouts of the kynurenine pathway were reported for multiple inflammatory conditions [@Badawy2017].
+In COVID-19, levels of metabolites of the kynurenine pathway in blood and urine were found to correlate positively with inflammatory markers [@Dewulf2022; @Santiago-Mujika2022], disease severity [@Dewulf2022; @Lionetto2021; @Ceballos2022] and was implicated in persistent COVID-19 symptom [@Bizjak2022; @Matits2023; @Cysique2022].
+These findings were summarized in a recent metaanalysis demonstrating increased KYN/TRP in COVID-19 and particularly in its severe manifestations [@Almulla2022]. 
+Using high-throughput metabolomics, tryptophan metabolism via serotonin and kynurenine pathways was identified as the most prominently affected metabolic system in acute COVID-19. 
+In this study, TRP and serotonin were demonstrated to decrease and KYN was shown to increase in an IL6-dependent way [@Thomas2020]. 
+Of note, IL6 was also identified by us as a predictor of lower plasma serotonin in the INCOV study. 
+We demonstrated significant, positive associations of serotonin and TRP levels in acute and sub-acute infection in the INCOV cohort. 
+This suggests that kynurenine pathway may especially efficiently compete for TRP with systemic serotonin synthesis in the highly inflammatory milieu of early COVID-19. 
+In turn, increasing TRP and serotonin during infection recovery in the INCOV cohort may reflect decreasing IDO1 activity and re-routing of TRP to the serotonin biosynthesis [@Lukic2022; @Badawy2017]. 
+The multi-parameter modeling in the SIMMUN cohort revealed an additional, inflammation-independent effect of SARS-CoV-2-positivity at median 139 days after infection on KYN and KYN/TRP. 
+While its mechanism remains unclear, an analogical sustained increase in KYN in absence of the inflammatory marker C-reactive protein was described in long-term COVID-19 recovery [@Bizjak2022]. 
+Age was identified by us as another predictor of high kynurenine pathway activity and low serotonin availability, which may be a result of a chronic low-grade inflammation observed in elderly [@Capuron2011; @Sorgdrager2019].
+Furthermore, significantly lower levels of serotonin were observed in male participants of the INCOV study. 
+This may reflect a generally stronger inflammatory response to SARS-CoV-2 in males [@Ceballos2022], which was also evident in our data (INCOV, IL6: $\beta$ = `r my_beta(gender_incov$inference$IL6_INF, 'sex', 'male')`, male versus female, robust linear modeling).
+
+## Effects of inflammation and infection on the serotonin and kynurenine pathways, and systemic levels of serotonin
+
+BH4 poses, another junction between neurotransmitter metabolism and inflammation. 
+Activity of GTP cyclohydrolase I, the enzyme catalyzing the rate-limiting step of BH4 synthesis, is strongly stimulated by IFNG and other inflammatory stimuli. 
+However, during inflammation, BH4 availability for catecholamine and serotonin pathway enzymes is limited due to oxidation and depletion by nitric oxide synthesis [@Geisler2013; @Meiser2013; @Neurauter2008].
+As previously reported [@Luporini2021], we could demonstrate an increased plasma PHE during acute and sub-acute infection in the INCOV cohort. 
+We could also correlate readouts of suppressed catecholamine pathway activity such as increased PHE and PHE/TYR, and low TYR and DA sulfate levels with inflammatory markers NEO, IL6, TNF and IFNG. 
+Collectively, these observations are consistent with the inflammatory BH4 deficiency leading to suppressed PHE - TYR conversion, hyperphenylalaninemia and impaired catecholamine biosynthesis during acute and sub-acute SARS-CoV-2 infection.
+By contrast, the increased DA sulfate during infection recovery in the INCOV cohort and the significantly lower PHE/TYR at median 139 days after SARS-CoV-2 infection in the SIMMUN cohort may indicate a restored BH4 homeostasis and efficient dopamine synthesis [@Meiser2013; @Goldstein1999; @Geisler2013; @Neurauter2008].
+Contrary to the reported suppression of catecholamine synthesis with age [@Peters2006], we observed 
+higher blood levels of TYR and lower PHE/TYR in older participants of the SIMMUN study. 
+However, since we could not corroborate this effect in the INCOV cohort, its reproducibility remains questionable.
+Analogically, although low DA sulfate levels were associated with obesity in the INCOV cohort, we could no observe any significant effects of obesity on markers of the catecholamine pathway activity in the SIMMUN cohort.
+
+## Interaction of mental health with activity of the serotonin, kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability
+
+Peripheral and central nervous system inflammation was put forward as a neuroimmune mechanism contributing to mental health disorder symptoms such as depression, anxiety and mental stress [@Brown2021; @Schwarcz2012; @Vancassel2018].
+Sickness behavior accompanying e.g. infections and inflammatory conditions comprises symptoms of depression and anxiety, social disconnection, fatigue, cognitive disturbance, and psychomotor slowing. 
+Mechanistically, sickness behavior was postulated to involve kynurenine pathway along with suppression of serotonin and catecholamine metabolism mediated by acute or protracted low grade inflammation [@Dantzer2008; @Maes2012; @Vancassel2018].
+Additionally, kynurenine pathway metabolites were ascribed standalone activity in neuronal signaling, e.g.
+by interfering with glutamatergic receptors [@Brown2021; @Schwarcz2012].
+Accordingly, markers of the kynurenine pathway activity and suppressed serotonin pathway such as serotonin, TRP, KYN or KYN/TRP were associated with symptoms of depression in recent metaanalyses [@Hunt2020; @Marx2020; @Fellendorf2022].
+Inflammatory markers were also associated with elevated PHE/TYR indicative of reduced dopamine availability and proposed to contribute to depression in cancer [@Hufner2015] and trauma [@Hufner2019]. 
+The potential role of protracted inflammation and TRP depletion in persistent symptoms and mental disorders in COVID-19 convalescents was proposed in recent hypothesis papers [@Eroglu2021; @Bower2022].
+There is also clinical evidence for concomitantly elevated markers of inflammation, kynurenine pathway activity and lowered TRP availability in COVID-19 patients suffering from persistent symptoms [@Bizjak2022], cognitive impairment [@Cysique2022], or symptoms of anxiety, depression and mental stress [@Kucukkarapinar2022; @Matits2023].
+In the SIMMUN cohort, clinically relevant depression symptoms (HADS) and mental stress (PSS-4) along with the inflammatory marker NEO could be linked to decreased TRP and higher KYN/TRP, respectively.
+In the INCOV collective, inflammation was also identified as a major diver of alterations in circulating serotonin, TRP, as well as the kynurenine pathway metabolites KYN and QUIN. 
+Hence, the additive effects of inflammation, infection, mental stress and age may disrupt the homeostasis of the serotonin, kynurenine and catecholamine pathways and predispose to depressive or anxious disorders in COVID-19 patients [@Dantzer2008; @Bower2022].
+
+Still, most evidence for relevance of the peripheral neurotransmitter levels in mental disorders is delivered by observational studies, whereas in vivo experimental reports are scarce [@Brown2021]. 
+In the periphery, liver, mesenteric organs and vasculature are the main sites of tryptophan, serotonin and dopamine metabolism [@Badawy2017; @Lukic2022; @Meiser2013; @Goldstein1999]. 
+Although the neurotransmitter precursors TRP and TYR, and kynurenin pathway metabolites were postulated to pass the blood-brain barrier [@Brown2021; @Schwarcz2012; @Vancassel2018], psychiatric disorders are not consistently paralleled by alterations in dopamine, serotonin, and kynurenine pathway metabolites in the central nervous system [@Miller2008; @Clark2016; @Brown2021].
+Of note, clinical evidence for the serotonin theory of depression was also questioned in a recent systematic review [@Moncrieff2022]. 
+Whether inflammatory and infection-related factors impact on peripheral and brain availability of serotonin and dopamine availability and, hence, contribute to psychiatric disorders frequently observed in COVID-19 patients [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023], needs validation in a robust experimental and prospective setting.
+
+## Limitations
+
+Our study has limitations. 
+First, incompatibility of study designs and variable sets of the SIMMUN and INCOV studies precluded direct comparison and validation of our findings in the classical training - test cohort analysis setting.
+In particular, we were not able to validate effects of early infection on the metabolic pathways of interest in the SIMMUN cohort. 
+Conversely, we could not validate the effects of stress, depression and anxiety in the INCOV cohort due to unavailability of psychometric data. 
+Second, relevance of the systemic serotonin, kynurenine and catecholamine pathway activity for the central nervous system and mental health is controversial, as discussed above. 
+Third, the SIMMUN cohort had a limited sample size and a highly variable diagnosis - sampling interval, which likely diminished effect sizes of the SARS-CoV-2 infection. 
+Fourth, the SIMMUN cohort suffered from a selection bias due to enrichment in psychiatric patients with a high rate of physical and mental disorders. 
+Analogically, hospitalized COVID-19 patients constituted the majority of the INCOV cohort. 
+Hence, none of the cohorts is representative for the entire population. 
+Fifth, both cohorts were recruited during initial phases of the pandemic and do not include vaccinated patients. 
+Similarly, the analyzed cohorts were exposed to wild-type-like SARS-CoV-2 variants and do not allow to assess effects of infection with highly transmissible but less virulent omicron pathogens. 
+For these reasons, studies with recent, post-pandemic collectives are urgently needed to validate our findings. 
+Finally, the cross-sectional SIMMUN and INCOV cohorts encompassed uninfected controls and SARS-CoV-2 infections ranging from asymptomatic to critical disease.
+
+## Conclusions
+
+SARS-CoV-2-dependent and -independent inflammation alters activity of the serotonin, kynurenine and catecholamine pathway which may limit systemic availability of serotonin and dopamine. 
+Those effects can be further amplified by advanced age, mental stress and depression. 
+Further research is needed to explore the mechanistic interplay of SARS-CoV-2 infection, inflammation and mental health parameters. 
+It remains to be investigated, if and how this mechanism contributes to psychiatric disorders following SARS-CoV-2 infection.
+
 \newpage
+
+# Acknowledgements
+
+We thank all participants and patients for the participation in the study.
+
+# Funding
+
+The study was supported by the Science Fund of the Land Tirol (grant number GZ71134 to Katharina Hüfner).
+
+# Conflict of interest
+
+Katharina Hüfner has received research grants from Austria Wirtchaftsservice (AWS) and the State of Tyrol as well as lecturer's honoraria from Forum Medizinische Fortbildung (FOMF), the Anton Proksch Institute and the Hospital of Schwaz.
+Piotr Tymoszuk owns a data science company, Data Analytics as a Service Tirol, and receives payments from statistical data analysis, bioinformatic and scientific writing services.
+Other authors declare that no conflict of interest exists.
 
@@
 
-```{r tab-stigma-cohort, tab.cap = 'Characteristic of the local SIMMUN cohort.
+```{r tab-stigma-cohort, tab.cap = 'Characteristic of the SIMMUN cohort.
 Numeric variables are presented as medians with interquartile ranges.
@@
   footnote(1, 5, 
-           value = as_paragraph('corrected for multiple testing with the false discovery rate method.'), 
+           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
            part = 'header', 
@@
   width(5:6, width = 2.7, unit = 'cm') %>% 
-  footnote(3, 1, 
+  footnote(1, 5, 
+           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(4, 1, 
            value = as_paragraph('overweight: body mass index (BMI) 25 - 30 kg/mm\u00B2, obesity: BMI > 30 kg/mm\u00B2'), 
            part = 'body', 
-           ref_symbols = 'a') %>% 
-  footnote(5, 1, 
-           value = as_paragraph('WHO ordinal scale for clinical improvement; mild: 1 - 2, moderate: 3 - 4, severe: 5 - 6, critical: 7'), 
+           ref_symbols = 'b') %>% 
+  footnote(6, 1, 
+           value = as_paragraph('WHO ordinal scale for clinical improvement; mild: 1 - 2, moderate: 3 - 4, severe: 5 - 6, critical: 7; moderate, severe and critical infection patients were hospitalized during COVID-19.'),
            part = 'body', 
-           ref_symbols = 'b') %>% 
+           ref_symbols = 'c') %>% 
   theme_vanilla
@@
 # Figures
-
-```{r fig-inclusion, fig.width = figur::convert(figs$inclusion, to = 'in')$w, fig.height = figur::convert(figs$inclusion, to = 'in')$h, fig.cap = 'Inclusion scheme for the SIMMUN and INCOV cohorts, and analysis strategy.'}
-
-figs$inclusion$plot
-
-```
-
-__Figure \@ref(fig:fig-inclusion). Inclusion scheme for the SIMMUN and INCOV cohorts, and analysis strategy.__
-
-_TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine - tryptophan ratio; PHE/TYR: phenylalanine - tyrosine ratio; QUIN: quinolinate; 5-HT: serotonin; DA: dopamine; Ab: antibody; IL6: interleukin-6; IL10: interleukin-10; TNF: tumor-necrosis factor alpha; IFNG: interferon gamma._
-
-\newpage
 
@@
 
-_Effects of systemic inflammation markers (neopterin [NEO], neutrophil-lymphocyte ratio [NLR]), SARS-CoV-2 infection status, titre of immunoglobulin gamma against the receptor binding domain of the S1 SARS-CoV-2 protein (anti-RBD IgG), scores of anxiety, depression (hospital depression and anxiety scale [HADS]) and mental stress (perceived stress scale, 4 item [PSS4]), age and sex was investigated by multi-parameter linear regression with backward elimination of non-significant terms.
-Overall model validity was assessed by likelihood-ratio test (LRT).
-Significant model coefficient estimates with 95% confidence intervals are presented in Forest plots.
+_Effects of markers of systemic inflammation (neopterin [NEO], neutrophil/lymphocyte ratio [NLR]), SARS-CoV-2 infection status, titre of immunoglobulin gamma against the receptor binding domain of the S1/S2 SARS-CoV-2 protein (anti-RBD IgG), clinically relevant symptoms of anxiety and depression (hospital depression and anxiety scale [HADS] $\geq$ 8 points), intensity of mental stress (perceived stress scale, 4 item [PSS-4]), age and sex on readouts of the kynurenine pathway (A, tryptophan [TRP], kynurenine [KYN], kynurenine/tryptophan ratio [KYN/TRP]) and the catecholamine pathway activity (B, phenylalanine [PHE], tyrosine [TYR], phenylalanine/tyrosine ratio [PHE/TYR]) were investigated by multi-parameter linear regression with backward elimination of non-significant terms.
+Numeric variables were normalized prior to modeling.
+Overall model validity was assessed by likelihood-ratio test (LRT) as compared with the respective null models.
+Coefficient estimates ($\beta$) with 95% confidence intervals are presented in Forest plots.
 LRT results (F values, degrees of freedom and p values) and total numbers of complete observations are indicated in the plot captions.
@@
 
-_TRP: tryptophan; KYN: kynurenin; KYN/TRP: kynurenine - tryptophan ratio; PHE: phenylalanine; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio._
-
 \newpage
 
-```{r fig-uni-trp, fig.width = figur::convert(figs$uni_trp, to = 'in')$w, fig.height = figur::convert(figs$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, plasma inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}
-
-figs$uni_trp$plot
-
-```
-
-__Figure \@ref(fig:fig-uni-trp).
-Effects of age, plasma inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__
-
-_Age, systemic inflammatory markers: neopterin (NEO), neutrophil - lymphocyte ratio (NLR), SARS-CoV-2 infection status, depression signs (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of systemic levels of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP).
-Their association with plasma concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
-
-_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method.
-(A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
-(B) Scatter plots of plasma concentrations of NEO, TRP, KYN and KYN/TRP.
-Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
-Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._
-
-_(C) Levels of the metabolite levels were compared between infected and uninfected participants, and between participants with and without depression signs by two-tailed T test with Cohen's d effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points.
-Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the strata are indicated in the X axes._
-
-\newpage
-
-```{r fig-uni-tyr, fig.width = figur::convert(figs$uni_tyr, to = 'in')$w, fig.height = figur::convert(figs$uni_tyr, to = 'in')$h, fig.cap = 'Effects of age, plasma inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.'}
-
-figs$uni_tyr$plot
-
-```
-
-__Figure \@ref(fig:fig-uni-tyr). Effects of age, plasma inflammatory marker neopterin and SARS-CoV-2 infection on phenylalanine, tyrosine and phenylalanine/tyrosine ratio in the SIMMUN cohort.__
-
-_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of systemic levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR).
-Their association with plasma concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
-
-_(A, B) Correlation of the metabolite levels and the numeric explanatory factors of interest was assessed by Pearson's test corrected for multiple testing by the false discovery rate method.
-(A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
-(B) Scatter plots of patient's age, PHE, TYR and PHE/TYR.
-Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
-Correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._
-
-_(C) Levels of the metabolite levels were compared between infected and uninfected participants by two-tailed T test with Cohen's d effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Medians with interquartile ranges (IQR) are represented by boxes, whiskers span over 150% IQR, single observations are depicted as points.
-Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the strata are indicated in the X axes._
-
-\newpage
-
-```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine sulfate in the INCOV cohort.'}
+```{r fig-incov-neuro, fig.width = figur::convert(figs$incov_neuro, to = 'in')$w, fig.height = figur::convert(figs$incov_neuro, to = 'in')$h, fig.cap = 'Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine 3-O-sulfate in the INCOV cohort.'}
 
@@
 
-__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine sulfate in the INCOV cohort.__
+__Figure \@ref(fig:fig-incov-neuro). Results of multi-parameter robust linear modeling of plasma levels of serotonin and dopamine 3-O-sulfate in the INCOV cohort.__
 
-_Effects of patient's age, timepoint after SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection as compared with healthy controls), plasma levels of metabolites related to neurotransmitter biosynthesis and competitor pathways (tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE] and tyrosine [TYR]), and plasma concentrations of inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin (5-hydroxy tryptophan [5-HT]) and dopamine sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
+_Effects of age, timepoint of SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR), plasma levels of metabolites of the serotonin and kynurenine pathways (tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]), plasma levels of metabolites of the catecholamine pathway (phenylalanine [PHE], tyrosine [TYR]), and plasma concentrations of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin (5-hydroxy tryptamine [5-HT]) and dopamine 3-O-sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
 
-_(A) Model fit error (root mean square error [RMSE]) and fraction of explained variances (R-squared) of the robust linear models assessed in the genuine training dataset and 10-fold cross-validation (CV)._
+_(A) Model fit error (root mean square error [RMSE]) and fraction of explained variance (R^2^) of the robust linear models assessed in the training dataset and infection timepoint-stratified 10-fold cross-validation (CV)._
 
-_(B) Estimated of model coefficients with 95% confidence intervals presented in Forest plots.
+_(B) Estimates of model coefficients ($\beta$) with 95% confidence intervals presented in Forest plots.
 Numbers of complete observations are indicated in the plot captions.
@@
 
-```{r fig-time-ourse, fig.width = figur::convert(figs$time_course, to = 'in')$w, fig.height = figur::convert(figs$time_course, to = 'in')$h, fig.cap = 'Time course of inflammatory cytokines, serotonin, dopamine sulfate and the neurotransmitter biosynthesis precursors and competitor pathway products during SARS-CoV infection and recovery in the INCOV cohort.'}
+```{r fig-time-ourse, fig.width = figur::convert(figs$time_course, to = 'in')$w, fig.height = figur::convert(figs$time_course, to = 'in')$h, fig.cap = 'TTime course of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.'}
 
@@
 __Figure \@ref(fig:fig-time-ourse).
-Time course of inflammatory cytokines, serotonin, dopamine sulfate and the neurotransmitter biosynthesis precursors and competitor pathway products during SARS-CoV infection and recovery in the INCOV cohort.__
+Time course of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.__
 
-_Time courses of normalized log~2~-transformed plasma concentrations of inflammatory cytokines (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF] and interferon gamma [INFG]), metabolites implicated in serotonin synthesis (B, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN] and serotonin/5-hydroxy tryptophan [5-HT]) and metabolites implicated in dopamine turnover (C, phenylalanine [PHE], tyrosine [TYR] and dopamine sulfate [DA sulfate]).
-Differences between uninfected controls and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after infection, respectively) were modeled by robust linear regression with the uninfected controls as baseline.
+_Time courses of normalized log~2~-transformed plasma concentrations of cytokine makers of inflammation (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF], interferon gamma [INFG]), metabolites of the serotonin and kynurenine pathways (B, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin/5-hydroxy tryptamine [5-HT]), and metabolites of the catecholamine pathway (C, phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]).
+Differences between uninfected controls and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after SARS-CoV-2 infection diagnosis via PCR, respectively) were modeled by robust linear regression with the uninfected controls as baseline.
 Mean normalized concentrations with 2 standard errors of the mean (2 $\times$ SEM) are depicted as solid lines with tinted ribbons (plots on the left).
-Robust linear model coefficient estimates with 95% confidence intervals are visualized in Forest plots (right).
+Robust linear model coefficient estimates ($\beta$) with 95% confidence intervals are visualized in Forest plots (right).
 Numbers of complete observations per timepoint are indicated in the X axes._
@@
 
-```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of plasma levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.'}
+```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of plasma levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways during SARS-CoV-2 infection in the INCOV cohort.'}
 
@@
 __Figure \@ref(fig:fig-incov-corr).
-Correlation of plasma levels of inflammatory cytokines, serotonin and dopamine sulfate, their precursors and competitor pathway products in healthy controls and SARS-CoV-2 patients of the INCOV cohort.__
+Correlation of plasma levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.__
 
-_plasma levels of metabolites implicated in serotonin synthesis (A, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin [5-hydroxy tryptophan, 5-HT]), dopamine turnover (B, phenylalanine [PHE], tyrosine [TYR], dopamine sulfate [DA sulfate]) and inflammatory cytokines (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in uninfected and SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after infection).
-Correlation coefficient matrices for at least moderate effects ($\rho$ > 0.2) were visualized as force-directed network plots.
-Node color codes for the parameter type (gray: inflammatory cytokines, orange: serotonin, green: dopamine turnover), edge width and color codes for the value and sign of the correlation coefficient._
+_Plasma levels of metabolites of the kynurenine and serotonin pathways (tryptophan [TRP], serotonin [5-hydroxy tryptamine, 5-HT], kynurenine [KYN], quinolinic acid [QUIN]) and metabolites of the catecholamine pathway (phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]) and cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR).
+Correlation coefficient matrices for correlation coefficients $\rho$ > 0.2 were visualized as force-directed network plots.
+Node color codes for the variable type (gray: inflammatory markers, orange: serotonin and kynurenine pathway, green: catecholamine pathway), edge width and color codes for the value and sign of the correlation coefficient._
 

### 2023-06-07 paper/markdown/supplementary_material.Rmd
Analysis and manuscript files with feedback
Author's feedback included in the analyses and manuscript files. Effect of gender and BMI included in the INCOV cohort multi-parameter models. Changes and shortening of the manuscript text.

@@
 ---
-title: "Inflammation, SARS-CoV-2 infection and mental health disorders impact on systemic serotonin and dopamine metabolism"
+title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
 subtitle: "Supplementary Material"
@@
 
-Descriptive statistics, statistical hypothesis testing and correlation analysis was done with the packages _rstatix_ [@Kassambara2021] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA). 
+Descriptive statistics, statistical hypothesis testing and correlation analysis were done with the packages _rstatix_ [@Kassambara2021] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA). 
 For linear modeling and robust linear modeling, base R functions, the packages _stats_, _MASS_ [@Ripley2022], _caret_ [@Kuhn2008], [_lmqc_](https://github.com/PiotrTymoszuk/lmqc) and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) were employed.
@@
 
-Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck Medical University (Innsbruck, Austria), as well as inpatients and outpatients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR screening were invited to participate in the SIMMUN study.
+Individuals tested for SARS-CoV-2 via PCR at the University Hospital of Innsbruck (Innsbruck, Austria) and patients of the University Clinic for Psychiatry I and II (Innsbruck, Austria) undergoing routine SARS-CoV-2 PCR screening were invited to participate in the SIMMUN study.
 The study enrollment was initiated on 10. June 2020. 
-The inclusion criteria for the SIMMUN study were age of 18 - 70 years, proficiency in German language, residence in the study region (Tyrol, Austria), results of a SARS-CoV-2 PCR test conducted at the University Hospital of Innsbruck.
-The exclusion criteria were active SARS-CoV-2 infection (< 14 days following a positive test), pregnancy, active malignancies, organ transplantation, prior surgery in the past 3 months, or acute or chronic inflammatory illness and treatment with oral corticosteroids.
+The inclusion criteria for the SIMMUN study were age of 18 - 70 years, proficiency in German, residence in the study region (Tyrol, Austria), and a SARS-CoV-2 PCR test conducted at the University Hospital of Innsbruck.
+The exclusion criteria were active SARS-CoV-2 infection (< 14 days following a positive test), pregnancy, active malignancies, organ transplantation, prior surgery in the past 3 months, acute or chronic inflammatory illness and treatment with oral corticosteroids.
 The analysis inclusion criterion was the complete study variable dataset consisting of 
 basic demographic and medical history variables, 
-SARS-CoV-2 infection status and titer of anti-receptor-binding domain S12/S2 protein immunoglobulin gamma (anti-RBD IgG), 
-psychometric scoring of depression, anxiety and stress as well as blood levels of inflammatory markers and metabolites related to serotonin and dopamine biosynthesis. 
+SARS-CoV-2 PCR result, titer of anti-receptor-binding domain S12/S2 protein immunoglobulin gamma (anti-RBD IgG), 
+psychometric scoring of depression, anxiety and mental stress as well as blood levels of inflammatory markers and metabolites of the kynurenine and catecholamine pathway. 
 A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were analyzed here. 
 The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
-The SIMMUN study data were gathered during a single on-site study visit at the study center at median 139 days after the SARS-CoV-2 test (interquartile range: 119 - 157). 
-The study visits were conducted between 17.June 2020 and 27.May 2021. 
-The SIMMUN analysis inclusion scheme is presented in __Figure 1__. 
+The SIMMUN study data were gathered during a single on-site study visit at the study center at median 139 days after the SARS-CoV-2 PCR test (interquartile range: 119 - 157). 
+The study visits were conducted between 17. June 2020 and 27. May 2021. 
+The SIMMUN inclusion scheme is presented in __Supplementary Figure S\@ref(fig:fig-inclusion)__. 
 Significant differences between participants included in the analysis and excluded due to data missingness are shown in __Supplementary Table S\@ref(tab:tab-excluded)__. 
-Characteristic of the analyzed SIMMUN collective split by the SARS-CoV-2 infection status is presented in __Table 1__.
+Characteristic of the analyzed SIMMUN cohort split by the SARS-CoV-2 infection status is presented in __Table 1__.
 
-Demographic and clinical history variables: age, sex, body mass index (BMI), professionally diagnosed psychiatric disorders, self-reported chronic physical disorders, smoking and alcohol consumption history were surveyed during the study visit or extracted from electronic patient records.
+Demographic and clinical history variables, age, sex, body mass index (BMI), professionally diagnosed psychiatric disorders, self-reported chronic physical disorders, smoking and alcohol consumption, and result and date of the SARS-CoV-2 PCR were surveyed during the study visit or extracted from electronic patient records.
 Body mass classes were defined as follows: normal for BMI < 25 kg/m^2^, overweight for BMI 25 - 30 kg/m^2^ and obesity for BMI > 30 kg/m^2^.
-Result and date of the SARS-CoV-2 PCR test were retrieved from electronic patient records.
 
-The inflammatory markers included in the analysis: plasma concentration of neopterin (NEO) and neutrophil - lymphocyte ratio (NLR) were determined by the certified clinical routine laboratory at the University Hospital of Innsbruck.
-C-reactive protein (CRP) and interleukin-6 (IL6) were determined for the SIMMUN cohort as well but excluded from the analysis, since their levels were within the reference range in the great majority of participants.
+The inflammatory markers included in the analysis, plasma concentration of neopterin (NEO) measured by ELISA and neutrophil/lymphocyte ratio (NLR) were determined by a certified clinical routine laboratory at the University Hospital of Innsbruck.
+C-reactive protein (CRP) and interleukin-6 (IL6) were determined in the SIMMUN cohort as well but excluded from the analysis, since their levels were within the reference range in the great majority of participants.
 Elevated CRP > 0.5 mg/dL was detected in 
@@
 elevated IL6 > 7 pg/mL was observed in 
-`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of the analyzed SIMMUN collective. 
+`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of the analyzed SIMMUN cohort. 
 Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, as described elsewhere [@Widner1997; @Neurauter2008].
-Kynurenine - tryptophan and phenylalanine - tyrosine ratios (KYN/TRP and PHE/TYR) serving as readouts of systemic indoleamine 2,3-dioxygenase (IDO) and phenylalanine hydroxylase (PAH) activity were computed [@Capuron2011].
+Kynurenine - tryptophan and phenylalanine - tyrosine ratios (KYN/TRP and PHE/TYR) were used as readouts of systemic indoleamine 2,3-dioxygenase (IDO, kynurenine pathway) and phenylalanine hydroxylase (PAH, catecholamine pathway) activity, respectively [@Capuron2011].
 Plasma titer of anti-RBD IgG were quantified by ELISA as described before [@Deisenhammer2021]. 
-Strongly non-normally distributed anti-RBD IgG levels were expressed as arbitrary units (AU) were stratified with the $\leq$ 1, 1 - 16.3 and > 16.3 AU cutpoints defined by the maximal titer in the uninfected participants (1 AU) and median titer in the SARS-CoV-2 subset (16.3 AU).
-Laboratory measurements at and below the lower detection limit were substituted with the lower detection limit. 
-Laboratory measurements at and above the upper detection limit were substituted with the upper detection limit.
+Strongly non-normally distributed anti-RBD IgG levels were expressed as arbitrary units (AU) and stratified with the $\leq$ 1, 1 - 16.3 and > 16.3 AU cutpoints defined by the maximal titer in the uninfected participants (1 AU) and median titer in the SARS-CoV-2 subset (16.3 AU).
+Laboratory measurements at and beyond detection limits were substituted with the respective detection limit value.
 
-Mental stress was gauged with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
-Anxiety and depression signs were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
-The total possible score range for each subscale is 0 to 21, with higher scores indicating more intense symptoms of anxiety/depression. 
-Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points at the respective subscale as proposed in the seminal HADS paper and a recent metaanalysis [@Bjelland2002; @Zigmond1983].
+Mental stress was assessed with the 4-item perceived stress scale (PSS-4) [@Cohen1983]. 
+Clinically relevant symptoms of anxiety and depression were scored with the hospital anxiety and depression scale (HADS) including 7 items for anxiety and 7 items for depression [@Zigmond1983]. 
+The total possible score range for each subscale is 0 to 21, with higher scores indicating more severe symptoms of anxiety/depression. 
+Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points at the respective subscale as proposed [@Bjelland2002; @Zigmond1983]. 
 PSS-4, HADS anxiety and depression scales displayed good-to-excellent internal consistency as measured by the McDonald $\omega$ statistic [@McDonald1999] (__Supplementary Figure S\@ref(fig:fig-consistency)__).
@@
 
-SIMMUN cohort data were stored as SPSS files and imported in R with the `read.spss()` function (package _foreign_). 
-Transformation, stratification of numeric study variables and coding of categorical features was done with an in-hose developed script.
+The SIMMUN cohort data were stored as a SPSS file and imported in R with the `read.spss()` function (package _foreign_). 
+Transformation, stratification of numeric study variables and coding of categorical features was done with an in-house developed script.
 
@@
 Plasma proteome measurements in the INCOV cohort were obtained by proximity extension assay (PEA, Olink, Sweden). 
-Plasma metabolome was quantified by ultra-high-performance liquid chromatography/tandem accurate mass spectrometry (Metabolon, USA) [@Su2022; @Su2020]. 
-Proteome and metabolome data in form of normalized, age- and sex-adjusted, log~2~-transformed plasma levels as well as clinical information (sex, SARS-CoV-2 infection status, COVID-19 severity, timepoint) for the INCOV cohort were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) accompanying the report by Su and colleagues [@Su2022] (function `read_xlsx`, package _readxl_) [@Wickham2022].
+Plasma metabolome was quantified by high-performance liquid chromatography/tandem mass spectrometry (Metabolon, USA) [@Su2022; @Su2020]. 
+Normalized, age- and sex-adjusted, log~2~-transformed plasma levels of metabolites and cytokines as well as clinical information (sex, SARS-CoV-2 infection status, COVID-19 severity, timepoint) for the INCOV cohort were extracted from [supplementary tables](https://data.mendeley.com/datasets/96v329bg7g/1) accompanying the report by Su and colleagues [@Su2022] (function `read_xlsx`, package _readxl_) [@Wickham2022].
 In the current analysis, `r nrow(incov$analysis_tbl)` INCOV study samples obtained for 
-`r length(unique(incov$complete_ids))` individuals with the complete dataset of metabolites of interest (serotonin, tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN], phenylalanine [PHE], tyrosine [TYR] and dopamine 3-O-sulfate [DA sulfate]), inflammatory cytokine markers of interest (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and complete information of age at enrollment were included.
+`r length(unique(incov$complete_ids))` individuals with the complete dataset of metabolites of interest (serotonin, TRP, KYN, quinolinic acid [QUIN], PHE, TYR and dopamine 3-O-sulfate [DA sulfate]), inflammatory cytokine markers of interest (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) and complete information of age, sex and BMI at enrollment were included.
+BMI classes were defined as in the SIMMUN cohort. 
 The INCOV study variables are listed in __Supplementary Table S\@ref(tab:tab-incov-study-vars)__. 
-The analysis inclusion scheme for the INCOV cohort is presented in __Figure 1__. 
-Characteristic of the INCOV collective is shown in __Table 2__.
+The analysis inclusion scheme for the INCOV cohort is presented in __Supplementary Figure S\@ref(fig:fig-inclusion)__. 
+Characteristics of the INCOV cohort are shown in __Table 2__.
+
 Concerning the INCOV study design, plasma metabolites and proteome were determined in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
@@
 
-The tau-equivalence assumption was clearly violated for both the HADS depression and HADS anxiety scale in the SIMMUN collective as investigated by factor analysis (function `reduce_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), employing internally `factanal()` from the _stats_ package).
+The tau-equivalence assumption was clearly violated for both the HADS depression and HADS anxiety scale in the SIMMUN cohort as investigated by factor analysis (function `reduce_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), employing internally `factanal()` from the _stats_ package).
 For this reason, global McDonald's $\omega$ statistic was used as a consistency measure for all psychometric tools in the SIMMUN cohort (function `omega()`, package _psych_) [@McDonald1999; @Bartlett1937; @Revelle2022] (__Supplementary Figure S\@ref(fig:fig-consistency)__).
@@
 Effects with p < 0.05 were considered significant.
+
 The following effect size measures were used [@Cohen2013]:
-Cramer's V for comparison of distribution of categorical variables (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5), 
-Cohen's d for two group comparisons of normally distributed variables (weak: < 0.5, moderate: 0.5 - 0.8, large effect: $\geq$ 0.8), 
-r for two group comparisons of non-normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5), 
-$\rho$ for correlation of non-normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, strong correlation: $\geq$ 0.5), 
-r for correlation of normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, strong correlation: $\geq$ 0.5), 
-$R^2$ as a measure of explained variance in modeling (weak: < 0.13, moderate: 0.13 - 0.26, substantial: $\geq$ 0.26).
+
+* Cramer's V for comparison of distribution of categorical variables (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5)
+
+* Cohen's d for two group comparisons of normally distributed variables (weak: < 0.5, moderate: 0.5 - 0.8, large effect: $\geq$ 0.8)
+
+* r for two group comparisons of non-normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5)
+
+* $\rho$ for correlation of non-normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, strong correlation: $\geq$ 0.5)
+
+* r for correlation of normally distributed variables (weak: < 0.3, moderate: 0.3 - 0.5, strong correlation: $\geq$ 0.5)
+
+* $R^2$ as a measure of explained variance in modeling (weak: < 0.13, moderate: 0.13 - 0.26, substantial: $\geq$ 0.26).
 
@@
 Differences in characteristic of SIMMUN participants included in the analysis and excluded due to data missingness were assessed by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Supplementary Table S\@ref(tab:tab-excluded)__).
-Differences in characteristic of SIMMUN participants or INCOV study participants stratified by the SARS-CoV-2 infection status were investigated by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Tables 1__ and __2__).
-Correlation of metabolite concentrations with age, stress scoring (PSS-4) and inflammation measures (NEO and NLR) in the SIMMUN cohort was done by Pearson's test (__Supplementary Table S\@ref(tab:tab-univariate-correlation)__).
-Differences in metabolite concentrations and metabolite ratios in the SIMMUN cohort stratified by the SARS-CoV-2 infection status or presence of depression symptoms (HADS) were assessed by two-tailed T test with Cohen's d effect size metric (__Supplementary Table S\@ref(tab:tab-univariate-complarison)__).
+Differences in characteristic of SIMMUN participants or INCOV study participants divided by the SARS-CoV-2 infection status were investigated by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Tables 1__ and __2__).
+Correlation of metabolite concentrations with age, scoring of mental stress (PSS-4) and markers of inflammation (NEO and NLR) in the SIMMUN cohort was done by Pearson's test (__Supplementary Table S\@ref(tab:tab-univariate-correlation)__).
+Pairwise correlations of cytokine and metabolite levels in the INCOV cohort were assessed by Spearman's rank test. 
+Differences in metabolite concentrations and metabolite ratios in the SIMMUN cohort split by the SARS-CoV-2 infection status or presence of clinically relevant depression symptoms (HADS) were assessed by two-tailed T test with Cohen's d effect size metric (__Supplementary Table S\@ref(tab:tab-univariate-complarison)__).
 
@@
 
-Effects of age, sex, body mass class, physical and psychiatric disorders, body mass class, smoking and alcohol consumption history, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, signs of depression and anxiety (HADS), mental stress scoring (PSS-4) on systemic levels of amino acid neurotransmitter precursor and products of competitor pathways (TRP, KYN, KYN/TRP ratio, PHE, TYR and PHE/TYR ratio) was assessed by multi-parameter linear regression with backward elimination.
+Effects of age, sex, body mass class, physical and mental disorders, BMI class, smoking and alcohol consumption, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress scoring (PSS-4) on readouts of the kynurenine (TRP, KYN, KYN/TRP) and catecholamine pathway activity (PHE, TYR, PHE/TYR) were assessed by multi-parameter linear regression with backward elimination.
 Responses and numeric explanatory variables were transformed with log or squared root functions to improve normality as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__ and normalized prior to modeling.
@@
 Validity of the optimized models was determined by likelihood-ratio test (LRT) versus the respective null models (method `anova()`). 
-Reproducibility and proper parameterization of the optimized multi-parameter models was investigated by repeated cross-validation (10 folds, function `train(method = 'lm')`, package _caret_)[@Kuhn2008] and by comparison of the RMSE and $R^2$ statistics obtained with in the training dataset and in cross-validation (method `summary()`, package _caretExtra_).
-As presented in __Supplementary Figure S\@ref(fig:fig-fit-stats)A__, similar values of error fit in the training and cross-validation data suggest good reproducibility of the optimized models and lack of over-parameterization.
+Explanatory performance, reproducibility and proper parameterization of the optimized multi-parameter models was investigated by cross-validation (10 folds, function `train(method = 'lm')`, package _caret_)[@Kuhn2008] and by comparison of the RMSE (root mean square  error) and $R^2$ statistics obtained in the training dataset and in cross-validation (method `summary()`, package _caretExtra_).
+As presented in __Supplementary Figure S\@ref(fig:fig-fit-stats)A__, similar values of model error (RMSE) in the training and cross-validation data suggest good reproducibility of the optimized models and lack of over-parameterization.
 
-Inference statistics for the model estimates ($\beta$, expected values and 95% confidence intervals) were extracted with the `summary()` method. 
+Inference statistics for the model estimates ($\beta$, expected values and 95% confidence intervals) were extracted from the model objects with the `summary()` method. 
 Significance of the estimates ($p(\beta\neq0)$) was determined by two-tailed T test. 
@@
 
-## Multi-paramater robust linear modeling in the INCOV cohort
+## Multi-parameter robust linear modeling in the INCOV cohort
 
-Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on systemic neurotransmitter availability were subsequently tested in the publicly available INCOV collective [@Su2022].
-To this end we modeled normalized plasma levels of serotonin as the end product of serotonin biosynthesis associated with psychiatric disorders [@Holck2019; @Tateishi2021; @Sun2020; @Kheirouri2016; @Paul-Savoie2011; @Ertugrul2007], and the major product of systemic dopamine catabolism, dopamine 3-O-sulfate (DA-sulfate) [@Goldstein1999; @Meiser2013].
-The explanatory variables were age, plasma levels of cytokines induced by inflammatory signaling (IL6, IL10, TNF, IFNG), plasma concentrations of neurotransmitter precursors and KYN pathway products (for serotonin: TRP, KYN and quinolinate [QUIN], for DA sulfate: PHE and TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control).
+Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on systemic neurotransmitter availability were subsequently tested in the publicly available INCOV cohort [@Su2022].
+We modeled normalized plasma levels of serotonin as the product of the serotonin pathway [@Lukic2022], and the major product of systemic dopamine catabolism, dopamine 3-O-sulfate (DA sulfate) [@Goldstein1999; @Meiser2013].
+The explanatory variables were age, sex and BMI class, plasma levels of cytokines markers of inflammation (IL6, IL10, TNF, IFNG), plasma concentrations of metabolites of the serotonin and kynurenine pathway (TRP, KYN, QUIN), plasma concentrations of metabolites of the catecholamine pathway (PHE, TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control).
 Since most numeric features were non-normally distributed, MM algorithm robust linear regression with Huber's psi function was chosen [@Huber2011; @Ripley2022] as a modeling approach.
@@
 The robust linear models were built with the `make_lm()` wrapper ([_lmqc_](https://github.com/PiotrTymoszuk/lmqc) package) around the MASS package `rlm()` function [@Ripley2022]. 
-The parameterization and reproducibility of the models were tested by 10-fold cross-validation with the _caret_ package [@Kuhn2008] as described for multi-parameter linear regression. 
-As presented in __Figure 5A__, similar values of error fit in the training and cross-validation data suggest good reproducibility of the robust linear models and proper model parameterization. 
+The parameterization and reproducibility of the models were tested by infection timepoint-stratified 10-fold cross-validation with the _caret_ package [@Kuhn2008] as described for multi-parameter linear regression.
+As presented in __Figure 5A__, similar values of error fit (RMSE) in the training and cross-validation data suggest good reproducibility of the robust linear models and proper model parameterization. 
 Inference statistics were extracted from the models with `summary()` method. 
@@
 
-Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of cytokines (IL6, IL10, TNF and IFNG), and metabolite involved in serotonin (serotonin, TRP, KYN and QUIN) and dopamine turnover (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
+Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of cytokines (IL6, IL10, TNF and IFNG), metabolites of the serotonin and kynurenine pathways (serotonin, TRP, KYN and QUIN), and metabolites of the catecholamine pathway (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
 The correlation matrices were subsequently scaled into the [0, 1] range with the $0.5\times(x + 1)$ function and converted to undirected force directed graphs with the `graph_from_adjacency_matrix()` function from the _igraph_ package [@Csardi2006].
@@
 flextable::flextable(suppl_tabs$study_vars) %>% 
-  width(1, width = 3, unit = 'cm') %>% 
-  width(2, width = 3.7, unit = 'cm') %>% 
+  width(1, width = 2.7, unit = 'cm') %>% 
+  width(2, width = 4, unit = 'cm') %>% 
   width(3, width = 2.7, unit = 'cm') %>% 
-  width(4, width = 2.5, unit = 'cm') %>% 
-  width(5, width = 2.5, unit = 'cm') %>% 
-  width(6, width = 3.7, unit = 'cm') %>% 
+  width(4, width = 2.2, unit = 'cm') %>% 
+  width(5, width = 2.7, unit = 'cm') %>% 
+  width(6, width = 3.9, unit = 'cm') %>% 
   merge_v(1) %>% 
   footnote(1, 2, 
-           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine - tryptophan ratio; PHE/TYR: phenylalanine/tyrosine ratio; HADS: hospital anxiety and depression score; PSS-4: perceived stress scale, 4 item; NEO: neopterin; NLR: neutrophil - lymphocyte ratio; anti-RBD IgG: anti-receptor-binding domain S1/S2 immunoglobulin.'),
+           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine/tryptophan ratio; PHE/TYR: phenylalanine/tyrosine ratio; HADS: hospital anxiety and depression score; PSS-4: perceived stress scale, 4 item; NEO: neopterin; NLR: neutrophil - lymphocyte ratio; anti-RBD IgG: anti-receptor-binding domain S1/S2 immunoglobulin.'),
            part = 'header', 
@@
   width(2, width = 3, unit = 'cm') %>% 
-  width(3, width = 6, unit = 'cm') %>% 
+  width(3, width = 3, unit = 'cm') %>%
+  width(4, width = 6, unit = 'cm') %>% 
   merge_v(1) %>% 
   footnote(1, 2, 
-           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinate; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'),
+           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinic acid; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'),
            part = 'header', 
            ref_symbols = 'a') %>% 
+   footnote(14, 2, 
+           value = as_paragraph('normal: body mass index (BMI) < 25 kg/m\u00B2, overweight: BMI 25 - 30 kg/m\u00B2, obesity: BMI > 30 kg/m\u00B2.'), 
+           part = 'body', 
+           ref_symbols = 'b') %>% 
   theme_vanilla
@@
 ```{r tab-cohort-comp, tab.cap = 'Comparison of the SIMMUN and INCOV cohorts.
+Significant differences are shown.
 Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.
@@
   width(6, width = 2.5, unit = 'cm') %>% 
+  footnote(1, 4:5, 
+           value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
+           part = 'header', 
+           ref_symbols = 'b') %>% 
   footnote(4, 1, 
@@
   footnote(1, 1, 
-           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine - tryptophan ratio; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio.'), 
+           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine/tryptophan ratio; TYR: tyrosine; PHE/TYR: phenylalanine/tyrosine ratio.'), 
            part = 'header', 
@@
 
-```{r tab-univariate-correlation, tab.cap = "Correlation of age, stress scoring, blood neopterin levels and neutrophil - leukocyte ratio with plasma metabolite levels in the SIMMUN cohort investigated by Pearson's test."}
+```{r tab-univariate-correlation, tab.cap = "Correlation of age, perceived mental stress scoring, plasma neopterin levels and neutrophil/leukocyte ratio with plasma metabolite levels in the SIMMUN cohort investigated by Pearson's test."}
 
@@
   width(1, width = 3.7, unit = 'cm') %>% 
-  width(2, width = 3.7, unit = 'cm') %>% 
+  width(2, width = 4, unit = 'cm') %>% 
   width(3, width = 2, unit = 'cm') %>% 
@@
   footnote(1, 1, 
-           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine - tryptophan ratio; PHE: phenylalanine; TYR: tyrosine; PHE/TYR: phenylalanine - tyrosine ratio.'), 
+           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; KYN/TRP: kynurenine/tryptophan ratio; PHE: phenylalanine; TYR: tyrosine; PHE/TYR: phenylalanine/tyrosine ratio.'), 
            part = 'header', 
@@
 
-```{r tab-univariate-complarison, tab.cap = "Comparison of plasma metabolite levels in SIMMUN study participants stratified by depression signs and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}
+```{r tab-univariate-complarison, tab.cap = "Comparison of plasma metabolite concentrations in SIMMUN study participants split by clinically relevant depression signs and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}
 
@@
   width(2, width = 4.2, unit = 'cm') %>% 
-  width(3, width = 2.7, unit = 'cm') %>% 
-  width(4, width = 2.5, unit = 'cm') %>% 
+  width(3, width = 2.4, unit = 'cm') %>% 
+  width(4, width = 2.2, unit = 'cm') %>% 
   width(5, width = 3.7, unit = 'cm') %>% 
@@
   width(7, width = 2.4, unit = 'cm') %>% 
+  merge_v(1) %>% 
   footnote(1, 1, 
@@
            ref_symbols = 'c') %>% 
-  footnote(1, 5, 
+  footnote(1, 6, 
            value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
@@
 flextable::flextable(suppl_tabs$modeling_incov) %>% 
-  width(1, width = 3, unit = 'cm') %>% 
-  width(2, width = 3.3, unit = 'cm') %>% 
+  width(1, width = 2.5, unit = 'cm') %>% 
+  width(2, width = 3.5, unit = 'cm') %>% 
   width(3, width = 3, unit = 'cm') %>% 
-  width(4, width = 2.7, unit = 'cm') %>% 
-  width(5, width = 3.7, unit = 'cm') %>% 
+  width(4, width = 2.3, unit = 'cm') %>% 
+  width(5, width = 3.9, unit = 'cm') %>% 
   width(6, width = 2.7, unit = 'cm') %>% 
-  merge_v(1) %>% 
+  merge_v(1:2) %>% 
   footnote(1, 1, 
@@
   footnote(1, 2, 
-           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; QUIN: quinolinate; IL6: interleukin-6, IL10: interleukin-10; TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; PHE: phenylalanine; TYR: tyrosine.'),
+           value = as_paragraph('TRP: tryptophan; KYN: kynurenine; QUIN: quinolinic acid; IL6: interleukin-6, IL10: interleukin-10; TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; PHE: phenylalanine; TYR: tyrosine; BMI: body mass index, normal: BMI < 25 kg/m\u00B2, overweight: BMI 25 - 30 kg/m\u00B2, obesity: BMI > 30 kg/m\u00B2.'),
            part = 'header', 
@@
 
-```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokine markers, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.
+```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of serum levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort.
 The uninfected subset served as a baseline.'}
@@
   width(5, width = 2.9, unit = 'cm') %>% 
-  merge_v(1) %>% 
+  merge_v(1:2) %>% 
   footnote(1, 2, 
-           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinate; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'),
+           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinic acid; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'),
            part = 'header', 
            ref_symbols = 'a') %>% 
-  footnote(1, 4, 
+  footnote(1, 5, 
            value = as_paragraph('Corrected for multiple testing with the false discovery rate method.'), 
@@
 
-```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of serum levels of inflammatory cytokine markers, tryptophan, tyrosine and their metabolites as a function of SARS-CoV-2 infection status in the INCOV cohort.
+```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of serum levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort.
 The acute SARS-CoV-2 infection subset served as a baseline.'}
@@
 
-```{r fig-consistency, fig.width = figur::convert(suppl$consistency, to = 'in')$w, fig.height = figur::convert(suppl$consistency, to = 'in')$h, fig.cap = 'Consistency of the PSS4 stress HADS depression and HADS anxiety psychometric tools in the SIMMUN cohort.'}
+```{r fig-inclusion, fig.width = figur::convert(suppl$inclusion, to = 'in')$w, fig.height = figur::convert(suppl$inclusion, to = 'in')$h, fig.cap = 'Inclusion scheme for the SIMMUN and INCOV cohorts, and analysis strategy.'}
+
+suppl$inclusion$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-inclusion). Inclusion scheme for the SIMMUN and INCOV cohorts, and analysis strategy.__
+
+_Sampling timepoints in the INCOV cohort: acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after diagnosis of SARS-CoV-2 infection via PCR._
+
+_TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine/tryptophan ratio; PHE/TYR: phenylalanine/tyrosine ratio; NLR: neutrophil - leukocyte ratio; QUIN: quinolinic acid; 5-HT: serotonin; DA: dopamine; Ab: antibody; IL6: interleukin-6; IL10: interleukin-10; TNF: tumor-necrosis factor alpha; IFNG: interferon gamma._
+
+\newpage
+
+```{r fig-consistency, fig.width = figur::convert(suppl$consistency, to = 'in')$w, fig.height = figur::convert(suppl$consistency, to = 'in')$h, fig.cap = 'Consistency of the PSS-4 mental stress, HADS depression and HADS anxiety psychometric tools in the SIMMUN cohort.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-consistency). Consistency of the PSS4 stress HADS depression and HADS anxiety psychometric tools in the SIMMUN cohort.__
+__Supplementary Figure S\@ref(fig:fig-consistency). Consistency of the PSS-4 mental stress, HADS depression and HADS anxiety psychometric tools in the SIMMUN cohort.__
 
@@
 
-```{r fig-fit-stats, fig.width = figur::convert(suppl$fit_stats, to = 'in')$w, fig.height = figur::convert(suppl$fit_stats, to = 'in')$h, fig.cap = 'Root mean square error and R-squared statistics for multi-parameter linear models of tryptophan, tyrosine and their metabolites in the SIMMUN cohort.'}
+```{r fig-fit-stats, fig.width = figur::convert(suppl$fit_stats, to = 'in')$w, fig.height = figur::convert(suppl$fit_stats, to = 'in')$h, fig.cap = 'Root mean square error and R^2^ statistics for multi-parameter linear models of readouts of the kynurenine and catecholamine pathway activity in the SIMMUN cohort.'}
 
@@
 __Supplementary Figure S\@ref(fig:fig-fit-stats).
-Root mean square error and R-squared statistics for multi-parameter linear models of tryptophan, tyrosine and their metabolites in the SIMMUN cohort.__
+Root mean square error and R^2^ statistics for multi-parameter linear models of readouts of the kynurenine and catecholamine pathway activity in the SIMMUN cohort.__
 
 _Multi-parameter linear regression models of plasma levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by cross-validation (CV, 10 folds).
-Values of root mean square error (RMSE, A) and $R^2$ (B) in the training data set and CV are plotted._
+Values of root mean square error (RMSE, A) and R^2^ (B) in the training data set and CV are plotted._
 
@@
 
-```{r fig-uni-trp, fig.width = figur::convert(suppl$uni_trp, to = 'in')$w, fig.height = figur::convert(suppl$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}
+```{r fig-uni-trp, fig.width = figur::convert(suppl$uni_trp, to = 'in')$w, fig.height = figur::convert(suppl$uni_trp, to = 'in')$h, fig.cap = 'Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, mental stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.'}
 
@@
 __Supplementary Figure S\@ref(fig:fig-uni-trp).
-Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__
+Effects of age, serum inflammatory markers neopterin and neutrophil-lymphocyte ratio, mental stress, depression and SARS-CoV-2 infection on tryptophan, kynurenine and kynurenine/tryptophan ratio in the SIMMUN cohort.__
 
-_Age, systemic inflammatory markers: neopterin (NEO), neutrophil - lymphocyte ratio (NLR), SARS-CoV-2 infection status, depression signs (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of systemic levels of tryptophan (TRP), kynurenine (KYN) and kynurenine - tryptophan ratio (KYN/TRP).
-Their association with plasma concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
+_Age, systemic inflammatory markers: neopterin (NEO), neutrophil/lymphocyte ratio (NLR), SARS-CoV-2 infection status, clinically relevant symptoms of depression (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of plasma levels of tryptophan (TRP), kynurenine (KYN) and kynurenine/tryptophan ratio (KYN/TRP).
+Their association with plasma concentrations of these metabolites was investigated by correlation analysis and statistical hypothesis testing._
 
@@
 
-_(C) Levels of the metabolite levels were compared between infected and uninfected participants, and between participants with and without depression signs by two-tailed T test with Cohen's d effect size statistic.
+_(C) Levels of the metabolite levels were compared between infected and uninfected participants, and between participants with and without clinically relevant symptoms of depression by two-tailed T test with Cohen's d effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
@@
 
-_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of systemic levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine - tyrosine ratio (PHE/TYR).
-Their association with plasma concentrations of these metabolites was investigated by univariable correlation analysis and statistical hypothesis testing._
+_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of plasma levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TYR).
+Their association with plasma concentrations of these metabolites was investigated by correlation analysis and statistical hypothesis testing._
 
@@
 
-_Plasma levels of metabolites implicated in serotonin biosynthesis and the serotonin biosynthesis competitor kynurenine pathway (5-hydroxy tryptamine/serotonin [5-HT], tryptophan [TRP], kynurenine [KYN], quinolinate [QUIN]) and metabolites related to dopamine metabolism (phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]) were correlated with plasma levels of inflammatory cytokine markers (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) in uninfected individuals, and during acute (median 10 days), sub-acute (median 14 days) SARS-CoV-2 infection and recovery (median 64 days after SARS-CoV-2 infection diagnosis via PCR).
+_Plasma levels of metabolites of the serotonin and kynurenine pathways (5-hydroxy tryptamine/serotonin [5-HT], tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]) and metabolites of the catecholamine pathway (phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]) were correlated with plasma levels of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) in uninfected individuals, and during acute (median 10 days), sub-acute (median 14 days) SARS-CoV-2 infection and recovery (median 64 days after SARS-CoV-2 infection diagnosis via PCR).
 Statistical significance was determined by Spearman's rank test corrected for multiple testing with the false discovery rate method.

### 2023-06-07 exploration scripts/age_incov.R

@@ -1,0 +1,66 @@
+# Age,metabolites and cytokines for particular infection timepoints
+
+  insert_head()
+
+# container ------
+
+  age_incov <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+
+  # analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variable lexicon
+
+  age_incov$lexicon <- globals$incov_lexicon %>%
+    mutate(label_long = variable,
+           variable = make.names(variable))
+
+  ## analysis tables
+
+  age_incov$analysis_tbl <- incov$analysis_tbl %>%
+    select(patient_id,
+           timepoint,
+           age,
+           any_of(set_names(age_incov$lexicon$label_long,
+                            age_incov$lexicon$variable))) %>%
+    mutate(timepoint = car::recode(timepoint,
+                                   "'healthy' = 'uninfected'"),
+           timepoint = factor(timepoint,
+                              c('uninfected', 'acute', 'sub-acute', 'recovery'))) %>%
+    blast(timepoint) %>%
+    map(~filter(.x, complete.cases(.x)))
+
+# Correralation analysis -----
+
+  insert_msg('Correlation analysis')
+
+  age_incov$test <- age_incov$analysis_tbl %>%
+    map(function(data) map(age_incov$lexicon$variable, ~c('age', .x)) %>%
+          map_dfr(~correlate_variables(data,
+                                       variables = .x,
+                                       what = 'correlation',
+                                       type = 'spearman',
+                                       ci = TRUE,
+                                       pub_styled = FALSE))) %>%
+    compress(names_to = 'timepoint') %>%
+    re_adjust
+
+# Bubble plot -----
+
+  insert_msg('Bubble plot')
+
+  age_incov$plot <- age_incov$test %>%
+    mutate(variable1 = timepoint) %>%
+    corr_buble(plot_title = 'Metabolites, cytokines and age',
+               signif_only = FALSE) +
+    labs(plot_subtitle = 'Spearman correlation with age') +
+    scale_y_continuous(labels = function(x) exchange(x, dict = age_incov$lexicon))
+
+# END ------
+
+  insert_tail()

### 2023-06-07 exploration scripts/age_simmun.R

@@ -1,0 +1,61 @@
+# Effects of age on metabolites and markers of inflammation
+# correlation analysis (Pearson)
+
+  insert_head()
+
+# container ------
+
+  age_simmun <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+
+  ## variables
+
+  age_simmun$variables <- c(neo = 'log_neo',
+                               nlr = 'log_nlr',
+                               stigma$responses)
+
+  ## analysis table
+
+  age_simmun$analysis_tbl <-
+    stigma$data[c('patient_id', 'age', age_simmun$variables)] %>%
+    filter(complete.cases(.))
+
+# Correlation analysis ------
+
+  insert_msg('Correlation analysis')
+
+  age_simmun$test <- age_simmun$variables %>%
+    map_dfr(~correlate_variables(age_simmun$analysis_tbl,
+                                 variables = c('age', .x),
+                                 what = 'correlation',
+                                 type = 'pearson',
+                                 ci = TRUE,
+                                 pub_styled = TRUE)) %>%
+    re_adjust %>%
+    mutate(plot_cap = paste(eff_size, significance, sep = ', '),
+           plot_cap = paste(plot_cap, n, sep = ', n = '))
+
+# Plotting -------
+
+  insert_msg('Plotting, scatter plots')
+
+  age_simmun$plots <-
+    list(variables = map(age_simmun$variables, ~c('age', .x)),
+         plot_title = exchange(names(age_simmun$variables),
+                               globals$stigma_lexicon),
+         plot_subtitle = age_simmun$test$plot_cap,
+         y_lab = exchange(age_simmun$variables,
+                          globals$stigma_lexicon,
+                          value = 'axis_label')) %>%
+    pmap(plot_correlation,
+         age_simmun$analysis_tbl,
+         type = 'correlation',
+         cust_theme = globals$common_theme,
+         x_lab = 'age, years')
+
+# END ------
+
+  insert_tail()

### 2023-06-07 exploration scripts/bmi_incov.R

@@ -1,0 +1,161 @@
+# Testing of effects of BMI on the time course of metabolites and cytokines
+# Analyzed with robust linear modeling, I'm skipping the interaction effect
+# since it was not significant for any of the responses
+
+  insert_head()
+
+# container ----
+
+  bmi_incov <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variable lexicon
+
+  bmi_incov$lexicon <- globals$incov_lexicon %>%
+    mutate(label_long = variable,
+           variable = make.names(variable))
+
+  ## analysis tables
+
+  bmi_incov$analysis_tbl <- incov$analysis_tbl %>%
+    select(patient_id,
+           timepoint,
+           bmi_class,
+           any_of(set_names(bmi_incov$lexicon$label_long,
+                            bmi_incov$lexicon$variable))) %>%
+    mutate(timepoint = car::recode(timepoint,
+                                   "'healthy' = 'uninfected'"),
+           timepoint = factor(timepoint,
+                              c('uninfected', 'acute', 'sub-acute', 'recovery')))
+
+  ## colors for plots
+
+  bmi_incov$plot_colors <-
+    list(inflammatory = c('IL6_INF' = 'orangered2',
+                          'IL10_INF' = 'orangered4',
+                          'TNF_INF' = 'gray60',
+                          'IFNG_INF' = 'black'),
+         trp = c('tryptophan' = 'coral',
+                 'kynurenine' = 'indianred2',
+                 'quinolinate' = 'indianred4',
+                 'serotonin' = 'steelblue'),
+         tyr = c('phenylalanine' = 'darkolivegreen3',
+                 'tyrosine' = 'darkolivegreen4',
+                 'dopamine.3.O.sulfate' = 'steelblue'))
+
+  ## plot titles
+
+  bmi_incov$plot_titles <-
+    paste0(c('Inflammatory cytokines',
+             'Serotonin and kynurenine pathway',
+             'Catecholamine pathway'),
+           ', INCOV') %>%
+    set_names(c('inflammatory', 'trp', 'tyr'))
+
+  ## model formulas
+
+  bmi_incov$formulas <- bmi_incov$lexicon$variable %>%
+    map(~paste(.x, '~ bmi_class + timepoint')) %>%
+    map(as.formula) %>%
+    set_names(bmi_incov$lexicon$variable)
+
+# Construction of the models ------
+
+  insert_msg('Construction of the models')
+
+  bmi_incov$models <- bmi_incov$formulas %>%
[…]

### 2023-06-07 exploration scripts/bmi_simmun.R

@@ -1,0 +1,127 @@
+# Comparison of metabolite and inflammation markers in the BMI classes
+# and CoV infection. Done with two-way ANOVA with the interaction term.
+
+  insert_head()
+
+# container ------
+
+  bmi_simmun <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+
+  ## variables
+
+  bmi_simmun$variables <- c(neo = 'log_neo',
+                               nlr = 'log_nlr',
+                               stigma$responses)
+
+  ## analysis table
+
+  bmi_simmun$analysis_tbl <-
+    stigma$data[c('patient_id', 'bmi_class', 'cov', bmi_simmun$variables)] %>%
+    mutate(cov = car::recode(cov, "'healthy' = 'uninfected'"),
+           cov = factor(cov, c('uninfected', 'SARS-CoV-2'))) %>%
+    filter(!is.na(bmi_class))
+
+  ## model formulas
+
+  bmi_simmun$formulas <- bmi_simmun$variables %>%
+    map(~paste(.x, '~ bmi_class * cov')) %>%
+    map(as.formula)
+
+# Models -----
+
+  insert_msg('Models')
+
+  bmi_simmun$models <- bmi_simmun$formulas %>%
+    map(~make_lm(data = bmi_simmun$analysis_tbl,
+                 formula = .x,
+                 mod_fun = lm,
+                 family = NULL))
+
+# Assumptions -----
+
+  insert_msg('Assumptions')
+
+  ## there moderate violations of the normality assumption for the
+  ## model residuals measured by Shapiro-Wilk test and QQ plots of the residuals
+  ## (neopterin, kyn/trp, tyr, phe/tyr)
+
+  bmi_simmun$assumptions <- bmi_simmun$models %>%
+    map(summary, 'assumptions', type.predict = 'response')
+
+  bmi_simmun$resid_plots <- bmi_simmun$models %>%
+    map(plot, cust_theme = globals$common_theme, type.predict = 'response')
+
+# Fit stats ------
+
+  insert_msg('Fit stats')
+
+  bmi_simmun$fit_stats <- bmi_simmun$models %>%
+    map(summary, 'fit') %>%
+    map(mutate, transformation = 'response') %>%
+    compress(names_to = 'response')
+
+# ANOVA ------
+
+  insert_msg('Two-way ANOVA')
+
+  bmi_simmun$test <- bmi_simmun$models %>%
+    map(anova) %>%
[…]

### 2023-06-07 exploration scripts/cohort.R

@@ -16,14 +16,14 @@
     c(sex = 'Sex',
       age = 'Age, years',
       bmi = 'BMI, kg/m\u00B2',
-      bmi_class = 'Body mass class',
+      bmi_class = 'Body mass index',
       ethnics = 'Ethnics',
       somatic_comorb = 'Physical disorder',
-      psych_comorb = 'Psychiatric disorder',
+      psych_comorb = 'Mental disorder',
       hads_anx_score = 'HADS anxiety score',
       hads_dpr_score = 'HADS depression score',
-      hads_signs = 'Depression or anxiety signs, HADS \u2265 8',
-      pss_stress_score = 'PSS-4 stress score',
+      hads_signs = 'Clinically relevant signs of depression or anxiety, HADS \u2265 8',
+      pss_stress_score = 'PSS-4 mental stress score',
       cov = 'Infection',
       anti_rbd = 'anti-RBD SARS-CoV-2, IgG, AU',
       severity = 'COVID-19 severity') %>%
@@ -43,7 +43,12 @@
   cohort$analysis_tbl$incov <- incov$clinic %>%
     filter(patient_id %in% incov$complete_ids) %>%
     filter(!duplicated(patient_id)) %>%
-    mutate(long_cov = ifelse(long_cov == 'healthy',
+    mutate(cov = car::recode(cov, "'healthy' = 'uninfected'"),
+           cov = factor(cov, c('uninfected', 'SARS-CoV-2')),
+           timepoint = car::recode(timepoint, "'healthy' = 'uninfected'"),
+           timepoint = factor(timepoint,
+                              c('uninfected', 'acute', 'sub-acute', 'recovery')),
+           long_cov = ifelse(long_cov == 'healthy',
                              'no', as.character(long_cov)),
            long_cov = factor(long_cov),
            psy_long_cov = ifelse(psy_long_cov %in% c('healthy',
@@ -100,7 +105,7 @@
                   what = 'table',
                   pub_styled = TRUE)) %>%
     map(reduce, left_join, by = 'variable') %>%
-    map(set_names, c('variable', 'Healthy', 'SARS-CoV-2'))
+    map(set_names, c('variable', 'Uninfected', 'SARS-CoV-2 infection'))
 
 # Testing for differences between the healthy and SARS-CoV-2 ----
 
@@ -130,18 +135,18 @@
              ~.x[c('variable', 'test', 'significance', 'eff_size')]),
          left_join, by = 'variable') %>%
     map(mutate,
[…]

### 2023-06-07 exploration scripts/cohort_comparison.R

@@ -35,15 +35,18 @@
   comp$analysis_tbl$INCOV <-
     comp$analysis_tbl$INCOV %>%
     mutate(severity = ifelse(cov == 'healthy',
-                             'healthy',
+                             'uninfected',
                              ifelse(stri_detect(severity,
                                                 regex = '(1|2)$'),
-                                    'ambulatory', 'hospitalized')),
-           severity = factor(severity, c('healthy', 'ambulatory', 'hospitalized')))
+                                    'ambulatory', 'hospitalized')))
 
   comp$analysis_tbl <- comp$analysis_tbl %>%
     compress(names_to = 'cohort') %>%
-    mutate(cohort = factor(cohort, c('SIMMUN', 'INCOV')))
+    mutate(cohort = factor(cohort, c('SIMMUN', 'INCOV')),
+           cov = car::recode(cov, "'healthy' = 'uninfected'"),
+           cov = factor(cov, c('uninfected', 'SARS-CoV-2')),
+           severity = car::recode(severity, "'healthy' = 'uninfected'"),
+           severity = factor(severity, c('uninfected', 'ambulatory', 'hospitalized')))
 
   ## test type
 

### 2023-06-07 exploration scripts/consistency.R

@@ -87,7 +87,7 @@
               hjust = 1.5,
               vjust = 0.5,
               color = 'white') +
-    scale_y_discrete(labels = c(pss4 = 'Stress, PSS4',
+    scale_y_discrete(labels = c(pss4 = 'PSS-4 mental stress',
                                 hads_anxiety = 'HADS anxiety',
                                 hads_depression = 'HADS depression')) +
     globals$common_theme +

### 2023-06-07 exploration scripts/gender_incov.R

@@ -1,0 +1,163 @@
+# Testing of effects of gender on the time course of metabolites and cytokines
+# Analyzed with robust linear modeling, I'm skipping the interaction effect
+# since it was not significant for any of the responses
+
+
+  insert_head()
+
+# container ----
+
+  gender_incov <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variable lexicon
+
+  gender_incov$lexicon <- globals$incov_lexicon %>%
+    mutate(label_long = variable,
+           variable = make.names(variable))
+
+  ## analysis tables
+
+  gender_incov$analysis_tbl <- incov$analysis_tbl %>%
+    select(patient_id,
+           timepoint,
+           sex,
+           any_of(set_names(gender_incov$lexicon$label_long,
+                            gender_incov$lexicon$variable))) %>%
+    mutate(timepoint = car::recode(timepoint,
+                                   "'healthy' = 'uninfected'"),
+           timepoint = factor(timepoint,
+                              c('uninfected', 'acute', 'sub-acute', 'recovery')))
+
+  ## colors for plots
+
+  gender_incov$plot_colors <-
+    list(inflammatory = c('IL6_INF' = 'orangered2',
+                          'IL10_INF' = 'orangered4',
+                          'TNF_INF' = 'gray60',
+                          'IFNG_INF' = 'black'),
+         trp = c('tryptophan' = 'coral',
+                 'kynurenine' = 'indianred2',
+                 'quinolinate' = 'indianred4',
+                 'serotonin' = 'steelblue'),
+         tyr = c('phenylalanine' = 'darkolivegreen3',
+                 'tyrosine' = 'darkolivegreen4',
+                 'dopamine.3.O.sulfate' = 'steelblue'))
+
+  ## plot titles
+
+  gender_incov$plot_titles <-
+    paste0(c('Inflammatory cytokines',
+             'Serotonin and kynurenine pathway',
+             'Catecholamine pathway'),
+           ', INCOV') %>%
+    set_names(c('inflammatory', 'trp', 'tyr'))
+
+  ## model formulas
+
+  gender_incov$formulas <- gender_incov$lexicon$variable %>%
+    map(~paste(.x, '~ sex + timepoint')) %>%
+    map(as.formula) %>%
+    set_names(gender_incov$lexicon$variable)
+
+# Construction of the models ------
+
+  insert_msg('Construction of the models')
+
[…]

### 2023-06-07 exploration scripts/gender_simmun.R

@@ -1,0 +1,122 @@
+# Comparison of metabolite and inflammation markers in the genders
+# and CoV infection. Done with two-way ANOVA with the interaction term.
+
+  insert_head()
+
+# container ------
+
+  gender_simmun <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+
+  ## variables
+
+  gender_simmun$variables <- c(neo = 'log_neo',
+                               nlr = 'log_nlr',
+                               stigma$responses)
+
+  ## analysis table
+
+  gender_simmun$analysis_tbl <-
+    stigma$data[c('patient_id', 'sex', 'cov', gender_simmun$variables)] %>%
+    mutate(cov = car::recode(cov, "'healthy' = 'uninfected'"),
+           cov = factor(cov, c('uninfected', 'SARS-CoV-2')))
+
+  ## model formulas
+
+  gender_simmun$formulas <- gender_simmun$variables %>%
+    map(~paste(.x, '~ sex * cov')) %>%
+    map(as.formula)
+
+# Models -----
+
+  insert_msg('Models')
+
+  gender_simmun$models <- gender_simmun$formulas %>%
+    map(~make_lm(data = gender_simmun$analysis_tbl,
+                 formula = .x,
+                 mod_fun = lm,
+                 family = NULL))
+
+# Assumptions -----
+
+  insert_msg('Assumptions')
+
+  ## there moderate violations of the normality assumption for the
+  ## model residuals measured by Shapiro-Wilk test and QQ plots of the residuals
+  ## (neopterin, tyrosine and phe/tyr ratio)
+
+  gender_simmun$assumptions <- gender_simmun$models %>%
+    map(summary, 'assumptions', type.predict = 'response')
+
+  gender_simmun$resid_plots <- gender_simmun$models %>%
+    map(plot, cust_theme = globals$common_theme, type.predict = 'response')
+
+# Fit stats ------
+
+  insert_msg('Fit stats')
+
+  gender_simmun$fit_stats <- gender_simmun$models %>%
+    map(summary, 'fit') %>%
+    map(mutate, transformation = 'response') %>%
+    compress(names_to = 'response')
+
+# ANOVA ------
+
+  insert_msg('Two-way ANOVA')
+
+  gender_simmun$test <- gender_simmun$models %>%
+    map(anova) %>%
+    map(re_adjust, p_variable = 'Pr(>F)', method = 'none') %>%
+    map(~mutate(.x,
+                variable = stri_replace(variable,
[…]

### 2023-06-07 exploration scripts/severity_incov.R

@@ -1,0 +1,106 @@
+# Effects of age an gender on bi_severity of COVID-19
+
+  insert_head()
+
+# container -----
+
+  sev_incov <- list()
+
+# analysis globals -----
+
+  insert_msg('Analysis globals')
+
+  sev_incov$analysis_tbl <- incov$clinic %>%
+    filter(patient_id %in% unique(incov$complete_ids)) %>%
+    filter(!duplicated(patient_id), !is.na(bi_severity)) %>%
+    mutate(bi_severity = droplevels(bi_severity)) %>%
+    select(patient_id, age, sex, bi_severity)
+
+  sev_incov$sev_colors <-
+    c(healthy = 'steelblue',
+      mild = 'darkolivegreen4',
+      moderate = 'coral3',
+      severe = 'coral4',
+      critical = 'gray40',
+      `mild/moderate` = 'steelblue',
+      `severe/critical` = 'coral3')
+
+# Differences in age ------
+
+  insert_msg('Age')
+
+  ## stats
+
+  sev_incov$age$stats <- sev_incov$analysis_tbl %>%
+    explore(split_factor = 'bi_severity',
+            variables = 'age',
+            what = 'table',
+            pub_styled = TRUE)
+
+  ## Kruskal-Wallis test
+
+  sev_incov$age$test <- sev_incov$analysis_tbl %>%
+    compare_variables(variables = 'age',
+                      split_factor = 'bi_severity',
+                      what = 'eff_size',
+                      types = 'kruskal_etasq',
+                      exact = FALSE,
+                      ci = FALSE,
+                      pub_styled = TRUE) %>%
+    mutate(plot_cap = paste(eff_size, significance, sep = ', '))
+
+  ## plot
+
+  sev_incov$age$plot <- sev_incov$analysis_tbl %>%
+    plot_variable(variable = 'age',
+                  split_factor = 'bi_severity',
+                  type = 'box',
+                  point_hjitter = 0,
+                  cust_theme = globals$common_theme,
+                  plot_title = 'Age and SARS-CoV-2 infection bi_severity',
+                  plot_subtitle = sev_incov$age$test$plot_cap,
+                  x_n_labs = TRUE) +
+    scale_fill_manual(values = sev_incov$sev_colors)
+
+# Gender and bi_severity ------
+
+  insert_msg('Gender and bi_severity')
+
+  ## descriptive stats
+
+  sev_incov$sex$stats <- sev_incov$analysis_tbl %>%
+    explore(split_factor = 'sex',
[…]

### 2023-06-07 exploration scripts/severity_simmun.R

@@ -1,0 +1,97 @@
+# Differences in severity of COVID-19 attributed to age and gender
+
+  insert_head()
+
+# container ----
+
+  sev_simmun <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  sev_simmun$sev_colors <- c(healthy = 'steelblue',
+                             ambulatory = 'coral3',
+                             hospitalized = 'coral4')
+
+# age ------
+
+  insert_msg('Age nd severity')
+
+  ## stats
+
+  sev_simmun$age$stats <- stigma$data %>%
+    explore(split_factor = 'severity',
+            variables = 'age',
+            what = 'table',
+            pub_styled = TRUE) %>%
+    reduce(left_join, by = 'variable') %>%
+    set_names(c('variable', levels(stigma$data$severity)))
+
+  ## testing for differences in age, Kruskal-Wallis test
+
+  sev_simmun$age$test <- stigma$data %>%
+    compare_variables(variables = 'age',
+                      split_factor = 'severity',
+                      what = 'eff_size',
+                      types = 'kruskal_etasq',
+                      exact = FALSE,
+                      ci = FALSE,
+                      pub_styled = TRUE) %>%
+    mutate(plot_cap = paste(eff_size, significance, sep = ', '))
+
+  ## plot
+
+  sev_simmun$age$plot <- stigma$data %>%
+    plot_variable(variable = 'age',
+                  split_factor = 'severity',
+                  type = 'box',
+                  point_hjitter = 0,
+                  cust_theme = globals$common_theme,
+                  plot_title = 'Age and SARS-CoV-2 infection severity, SIMMUN',
+                  plot_subtitle = sev_simmun$age$test$plot_cap,
+                  x_n_labs = TRUE) +
+    scale_fill_manual(values = sev_simmun$sev_colors)
+
+# gender -----
+
+  insert_msg('Severity and gender')
+
+  ## stats
+
+  sev_simmun$sex$stats <-  stigma$data %>%
+    explore(split_factor = 'sex',
+            variables = 'severity',
+            what = 'table',
+            pub_styled = TRUE) %>%
+    reduce(left_join, by = 'variable') %>%
+    set_names(c('variable', levels(stigma$data$sex)))
+
+  ## testing: chi-squared test with Cramer's V effect size
+
+  sev_simmun$sex$test <- stigma$data %>%
+    compare_variables(variables = 'severity',
+                      split_factor = 'sex',
[…]

### 2023-06-07 exploration.R

@@ -5,6 +5,9 @@
 # Testing for the normality and EOV for the study variables (STIGMA and INCOV)
 #
 # Comparing included and excluded participants of the SIMMUN study
+#
+# Testing effects of SARS-CoV-2 and gender on metabolites
+# and inflammatory markers (needed for the paper Discussion)
 
 # tools --------
 
@@ -16,6 +19,8 @@
   library(soucer)
   library(psych)
   library(clustTools)
+  library(lmqc)
+  library(MASS)
 
   select <- dplyr::select
   reduce <- purrr::reduce
@@ -34,7 +39,15 @@
     './exploration scripts/distribution.R',
     './exploration scripts/excluded.R',
     './exploration scripts/consistency.R',
-    './exploration scripts/correlation.R') %>%
+    './exploration scripts/correlation.R',
+    './exploration scripts/gender_simmun.R',
+    './exploration scripts/gender_incov.R',
+    './exploration scripts/severity_simmun.R',
+    './exploration scripts/severity_incov.R',
+    './exploration scripts/bmi_simmun.R',
+    './exploration scripts/bmi_incov.R',
+    './exploration scripts/age_simmun.R',
+    './exploration scripts/age_incov.R') %>%
     source_all(message = TRUE, crash = TRUE)
 
 # END -----

### 2023-06-07 import.R

@@ -82,11 +82,11 @@
       'age' = 'Age, decades',
       'sex' = 'Sex',
       ## psychometrics
-      'pss_stress_score' = 'Stress, PSS-4',
+      'pss_stress_score' = 'Mental stress, PSS-4',
       'hads_anx_score' = 'Anxiety, HADS',
       'hads_dpr_score' = 'Depression, HADS',
       ## comorbidity, medical history
-      'psych_comorb' = 'Psychiatric disorder',
+      'psych_comorb' = 'Mental disorder',
       'somatic_comorb' = 'Physical disorder',
       'bmi_class' = 'BMI',
       'smoking' = 'Smoking',

### 2023-06-07 incov scripts/cytokine_correlation.R

@@ -75,9 +75,9 @@
                                     'TNF_INF',
                                     'IFNG_INF'))),
          plot_title =  c('uninfected, INCOV',
-                         'acute CoV, INCOV',
-                         'sub-acute CoV, INCOV',
-                         'CoV recovery, INCOV')) %>%
+                         'acute SARS-CoV-2 infection, INCOV',
+                         'sub-acute SARS-CoV-2 infection, INCOV',
+                         'SARS-CoV-2 infection recovery, INCOV')) %>%
     pmap(corr_buble,
          signif_only = FALSE,
          rotate_x_labs = FALSE)
@@ -93,8 +93,8 @@
                                                'kynurenine',
                                                'quinolinate',
                                                'serotonin'),
-                              'serotonin', 'dopamine'),
-             pathway = factor(pathway, c('serotonin', 'dopamine')))
+                              'serotonin/kynurenine', 'catecholamine'),
+             pathway = factor(pathway, c('serotonin/kynurenine', 'catecholamine')))
 
   }
 

### 2023-06-07 incov scripts/mds.R

@@ -103,7 +103,10 @@
 
   incov_mds$plots <-
     list(x = incov_mds$plots,
-         y =  paste0(c('Uninfected', 'Acute Cov', 'Sub-acute CoV', 'Recovery'),
+         y =  paste0(c('uninfected',
+                       'acute SARS-CoV-2 infection',
+                       'sub-acute SARS-CoV-2 infection',
+                       'SARS-CoV-2 infection recovery'),
                      ', INCOV') %>%
            rep(2),
          z = paste('n =', map_dbl(incov_mds$analysis_tbl, ncol))) %>%

### 2023-06-07 incov scripts/metabolite_correlation.R

@@ -76,9 +76,9 @@
   met_corr$bubbles <-
     list(data = met_corr$test,
          plot_title =  c('uninfected, INCOV',
-                         'acute CoV, INCOV',
-                         'sub-acute CoV, INCOV',
-                         'CoV recovery, INCOV')) %>%
+                         'acute SARS-CoV-2 infection, INCOV',
+                         'sub-acute SARS-CoV-2 infection, INCOV',
+                         'SARS-CoV-2 infection recovery, INCOV')) %>%
     pmap(corr_buble,
          signif_only = FALSE,
          rotate_x_labs = FALSE)

### 2023-06-07 incov scripts/modeling.R

@@ -1,5 +1,10 @@
 # Robust linear modeling of metabolites as a function of infection status
 # and levels of inflammatory cytokines
+#
+# A short note on cross-validation: the model are validated by 10-fold CV
+# where the folds are constructed in a timepoint-balanced wise
+# (or more precisely: timepoint-stratified CV),
+# see: tools/tools.R for code details and rationale
 
   insert_head()
 
@@ -31,6 +36,8 @@
   incov_mod$variable_lexicon <-
     c('timepoint' = 'SARS-CoV-2',
       'age' = 'Age',
+      'sex' = 'Sex',
+      'bmi_class' = 'BMI',
       'IL6_INF' = 'IL6',
       'IL10_INF' = 'IL10',
       'TNF_INF' = 'TNF',
@@ -61,6 +68,15 @@
     map(as.formula) %>%
     set_names(incov_mod$response_lexicon$variable)
 
+  ## train control object
+
+  set.seed(123456)
+
+  incov_mod$trainControl <-
+    time_balanced_folds(data = incov_mod$analysis_tbl,
+                        time_variable = 'timepoint',
+                        number = 10)
+
 # Construction of models -----
 
   insert_msg('Construction of models')
@@ -72,16 +88,6 @@
                  family = NULL,
                  method = 'MM',
                  psi = psi.huber))
-
-# Custom model list for caret -------
-
-  insert_msg('A custom list for caret')
-
-  ## merely a copy-paste from the output of getModelInfo('rlm')
-  ## but specifying the method "MM" explicitly
-
-  incov_mod$caret_list <-
-
 
 # cross-validation via caret ------
 
@@ -98,11 +104,7 @@
                                                 "'sub-acute' = 'sub'")),
                method = mm_rlm,
                metric = 'RMSE',
-               trControl = trainControl(method = 'repeatedcv',
-                                        number = 10,
-                                        savePredictions = 'final',
-                                        returnData = TRUE,
-                                        returnResamp = 'final'),
+               trControl =incov_mod$trainControl,
                tuneGrid = data.frame(intercept = TRUE,
                                      psi = 'psi.huber'))) %>%
     map(as_caretx)

### 2023-06-07 incov scripts/neuro_modeling.R

@@ -1,5 +1,10 @@
 # Modeling of neurotransmitter levels as a function of treir precursors,
 # competitive decay products, infection status and inflammatory cytokines
+#
+# A short note on cross-validation: the model are validated by 10-fold CV
+# where the folds are constructed in a timepoint-balanced wise,
+# or more precisely speaking timepoint-stratified CV,
+# see: tools/tools.R for code details and rationale
 
   insert_head()
 
@@ -32,6 +37,8 @@
     map(~c(.x,
            c('timepoint' = 'SARS-CoV-2',
              'age' = 'Age',
+             'sex' = 'Sex',
+             'bmi_class' = 'BMI',
              'IL6_INF' = 'IL6',
              'IL10_INF' = 'IL10',
              'TNF_INF' = 'TNF',
@@ -65,6 +72,15 @@
     map(as.formula) %>%
     set_names(names(incov_neuro$variable_lexicons))
 
+  ## trainControl object
+
+  set.seed(123456)
+
+  incov_neuro$trainControl <-
+    time_balanced_folds(data = incov_neuro$analysis_tbl,
+                        time_variable = 'timepoint',
+                        number = 10)
+
 # Construction of models -----
 
   insert_msg('Construction of models')
@@ -92,11 +108,7 @@
                                                 "'sub-acute' = 'sub'")),
                method = mm_rlm,
                metric = 'RMSE',
-               trControl = trainControl(method = 'repeatedcv',
-                                        number = 10,
-                                        savePredictions = 'final',
-                                        returnData = TRUE,
-                                        returnResamp = 'final'),
+               trControl = incov_neuro$trainControl,
                tuneGrid = data.frame(intercept = TRUE,
                                      psi = 'psi.huber'))) %>%
     map(as_caretx)

### 2023-06-07 incov scripts/time_modeling.R

@@ -61,6 +61,15 @@
                  'tyrosine' = 'darkolivegreen4',
                  'dopamine.3.O.sulfate' = 'steelblue'))
 
+  ## plot titles
+
+  time_rlm$plot_titles <-
+    paste0(c('Inflammatory cytokines',
+             'Serotonin and kynurenine pathway',
+             'Catecholamine pathway'),
+           ', INCOV') %>%
+    set_names(c('inflammatory', 'trp', 'tyr'))
+
 # Construction of the models -----
 
   insert_msg('Model construction')
@@ -139,10 +148,7 @@
     list(vars = list(inflammatory = c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF'),
                      trp = c('tryptophan', 'kynurenine', 'quinolinate', 'serotonin'),
                      tyr = c('phenylalanine', 'tyrosine', 'dopamine.3.O.sulfate')),
-         plot_title = paste0(c('Inflammatory cytokines',
-                               'TRP decay and 5-HT',
-                               'PHE/TYR and DA'),
-                             ', INCOV')) %>%
+         plot_title = time_rlm$plot_titles) %>%
     pmap(plot_tc_ribbon,
          data = time_rlm$analysis_tbl$uninfected,
          plot_subtitle = 'Mean, 2 \u00D7 SEM',
@@ -176,9 +182,7 @@
                                             tyr = c('phenylalanine',
                                                     'tyrosine',
                                                     'dopamine.3.O.sulfate')),
-                                plot_title = paste(c('Inflammatory cytokines',
-                                                     'TRP decay and 5-HT',
-                                                     'PHE/TYR and DA'))) %>%
+                                plot_title = time_rlm$plot_titles) %>%
            pmap(plot_tc_beta,
                 data = x,
                 plot_subtitle = y,

### 2023-06-07 modeling.R

@@ -63,10 +63,11 @@
 
   insert_msg('Analysis scripts')
 
-  c('./modeling scripts/elimination.R',
-    './modeling scripts/glmnet.R',
-    './modeling scripts/ranger.R',
-    './modeling scripts/cforest.R') %>%
+  c('./modeling scripts/elimination.R'#,
+    #'./modeling scripts/glmnet.R',
+    #'./modeling scripts/ranger.R',
+    #'./modeling scripts/cforest.R'
+    ) %>%
     source_all(message = TRUE, crash = TRUE)
 
 # END ------

### 2023-06-07 paper scripts/figures.R

@@ -6,23 +6,9 @@
 
   figs <- list()
 
-# Figure 1: CONSORT -----
+# Figure 1: linear modeling -----
 
-  insert_msg('Figure 1: study inclusion diagram')
-
-  figs$inclusion <-
-    plot_grid(ggdraw() +
-                draw_image('./schemes/inclusion.png')) %>%
-    as_figure(label = 'figure_1_study_inclusion_scheme',
-              ref_name = 'inclusion',
-              caption = paste('Inclusion scheme for the SIMMUN',
-                              'and INCOV cohorts, and analysis strategy.'),
-              w = 180,
-              h = 180 * 3426/4098)
-
-# Figure 2: linear modeling -----
-
-  insert_msg('Figure 2: linear modeling')
+  insert_msg('Figure 1: linear modeling')
 
   figs$modeling <- mod_eli$forest_plots %>%
     map(~.x + theme(legend.position = 'none'))
@@ -47,7 +33,7 @@
                          '', '',
                          'B', ''),
               label_size = 10) %>%
-    as_figure(label = 'figure_2_linear_modeling',
+    as_figure(label = 'figure_1_linear_modeling',
               ref_name = 'modeling',
               caption = paste('Results of multi-parameter linear modeling',
                               'of tryptophan,',
@@ -57,9 +43,9 @@
               w = 180,
               h = 180)
 
-# Figure 3: multi-parameter modeling of neurotransmitters in the INCOV cohort -----
+# Figure 2: multi-parameter modeling of neurotransmitters in the INCOV cohort -----
 
-  insert_msg('Figure 3: INCOV neurotransmitters')
+  insert_msg('Figure 2: INCOV neurotransmitters')
 
   ## upper panel: fit stats
 
@@ -80,7 +66,10 @@
   ## hacking for re-ordering the Y axis
 
   figs$incov_neuro$plot_order <-
-    c('age' = 'age',
+    c('age' = 'background',
+      'sexmale' = 'background',
+      'bmi_classobesity' = 'background',
+      'bmi_classoverweight' = 'background',
       'timepointacute' = 'SARS-CoV-2',
       'timepointsub-acute' = 'SARS-CoV-2',
       'timepointrecovery' = 'SARS-CoV-2',
@@ -100,7 +89,7 @@
                            rev(c('inflammation',
                                  'metabolites',
                                  'SARS-CoV-2',
-                                 'age'))))
+                                 'background'))))
 
   figs$incov_neuro$bottom <- incov_neuro$forest_plots
 
[…]

### 2023-06-07 paper scripts/render.R

@@ -8,6 +8,15 @@
 
   cov_biblio <- read_bib('./paper/markdown/infl_biblio.bib')
 
+# feedback and issues ------
+
+  insert_msg('Rendering the issue file')
+
+  render('./paper/markdown/issues.Rmd',
+         output_format = word_document2(number_sections = FALSE,
+                                        reference_docx = 'ms_template.docx'),
+         output_dir = './paper')
+
 # supplementary material ------
 
   insert_msg('Rendering the supplements')

### 2023-06-07 paper scripts/supplement.R

@@ -6,9 +6,23 @@
 
   suppl <- list()
 
-# Figure S1: consistency of psychometric tools ------
+# Figure S1: CONSORT -----
 
-  insert_msg('Figure S1: consistency of psychometric tools')
+  insert_msg('Figure S1: study inclusion diagram')
+
+  suppl$inclusion <-
+    plot_grid(ggdraw() +
+                draw_image('./schemes/inclusion.png')) %>%
+    as_figure(label = 'figure_s1_study_inclusion_scheme',
+              ref_name = 'inclusion',
+              caption = paste('Inclusion scheme for the SIMMUN',
+                              'and INCOV cohorts, and analysis strategy.'),
+              w = 180,
+              h = 180 * 3486/4098)
+
+# Figure S2: consistency of psychometric tools ------
+
+  insert_msg('Figure S2: consistency of psychometric tools')
 
   suppl$consistency <- cons$fa_loading_plots %>%
     map(~.x +
@@ -26,7 +40,7 @@
               rel_heights = c(1, 0.7),
               labels = LETTERS,
               label_size = 10) %>%
-    as_figure(label = 'figure_s1_psycho_consistency',
+    as_figure(label = 'figure_s2_psycho_consistency',
               ref_name = 'consistency',
               caption = paste('Consistency of the PSS4 stress',
                               'HADS depression and HADS anxiety psychometric',
@@ -34,9 +48,9 @@
               w = 180,
               h = 140)
 
-# Figure S2: error and R-squared, multiple linear regression ------
+# Figure S3: error and R-squared, multiple linear regression ------
 
-  insert_msg('Figure S2: Fits stats, multiple regression')
+  insert_msg('Figure S3: Fits stats, multiple regression')
 
   suppl$fit_stats <- mod_eli$fit_stat_plots %>%
     map2(c('pseudo_log', 'identity'),
@@ -52,7 +66,7 @@
                            theme(legend.position = 'bottom')),
               nrow = 2,
               rel_heights = c(0.9, 0.1)) %>%
-    as_figure(label = 'figure_s2_fit_stats',
+    as_figure(label = 'figure_s3_fit_stats',
               ref_name = 'fit_stats',
               caption = paste('Root mean square error and R-squared',
                               'statistics for multi-parameter linear models',
@@ -61,9 +75,9 @@
               w = 180,
               h = 110)
 
-# Figure S3: univariate analysis for TRP/KYN ------
[…]

### 2023-06-07 paper scripts/tables.R

@@ -20,8 +20,8 @@
                  regex = '^HADS.*\\nHADS\\+:\\s{1}',
                  replacement = '')) %>%
     map(set_names,
-        c('Variable', 'Uninfected',
-          'SARS-CoV-2', 'Test', 'Significance', 'Effect size'))
+        c('Variable', 'Uninfected', 'SARS-CoV-2 infection',
+          'Test', 'Significance', 'Effect size'))
 
   ## skipping the somatic symptoms
 
@@ -29,13 +29,13 @@
     filter(Variable %in% c('Participants, n',
                            'Sex',
                            'Age, years',
-                           'Body mass class',
+                           'Body mass index',
                            'Physical disorder',
-                           'Psychiatric disorder',
+                           'Mental disorder',
                            'HADS anxiety score',
                            'HADS depression score',
-                           'Depression or anxiety signs, HADS ≥ 8',
-                           'PSS-4 stress score',
+                           'Clinically relevant signs of depression or anxiety, HADS ≥ 8',
+                           'PSS-4 mental stress score',
                            'anti-RBD SARS-CoV-2, IgG, AU',
                            'COVID-19 severity'))
 
@@ -117,15 +117,18 @@
 
   suppl_tabs$incov_study_vars <-
     globals$incov_lexicon[c('variable', 'label', 'unit')] %>%
-    rbind(tibble(variable = 'age',
-                 label = 'age',
-                 unit = 'years')) %>%
+    rbind(tibble(variable = c('age', 'sex', 'bmi_class'),
+                 label = c('age', 'sex', 'body mass index'),
+                 unit = c('years', NA, NA))) %>%
     mutate(var_type = ifelse(variable %in% c('serotonin', 'dopamine 3-O-sulfate'),
-                             'response', 'explanatory')) %>%
+                             'response', 'explanatory'),
+           format = ifelse(is.na(unit), 'categorical', 'numeric')) %>%
+    arrange(desc(var_type)) %>%
     select(var_type,
            label,
+           format,
            unit) %>%
-    set_names(c('Variable type', 'Variable label', 'Unit')) %>%
+    set_names(c('Variable type', 'Variable label', 'Format', 'Unit')) %>%
     mdtable(label = 'table_s3_incov_study_vars',
             ref_name = 'incov_study_vars',
             caption = 'Study variables in the INCOV cohort.')
@@ -164,8 +167,8 @@
                                             variable,
                                             exchange(variable,
[…]

### 2023-06-07 testing scripts/result_tables.R

@@ -51,9 +51,9 @@
            eff_size) %>%
     set_names(c('Metabolite',
                 'Explanatory variable',
-                'Stratum',
-                'Observations/stratum',
-                'Statistic',
+                'Category',
+                'N',
+                'Median, IQR, range',
                 'Significance',
                 'Effect size'))
 

### 2023-06-07 testing scripts/trp_system.R

@@ -22,9 +22,9 @@
   tst_trp$analysis_tbl <- stigma$data %>%
     mutate(infection = car::recode(cov,
                                    "'healthy' = 'uninfected';
-                                    'SARS-CoV-2' = 'recovery'"),
+                                    'SARS-CoV-2' = 'SARS-CoV-2'"),
            infection = factor(infection,
-                              c('uninfected', 'recovery'))) %>%
+                              c('uninfected', 'SARS-CoV-2'))) %>%
     select(patient_id,
            all_of(unname(tst_trp$responses)),
            all_of(tst_trp$lexicon$variable)) %>%

### 2023-06-07 testing scripts/tyr_system.R

@@ -22,9 +22,9 @@
   tst_tyr$analysis_tbl <- stigma$data %>%
     mutate(infection = car::recode(cov,
                                    "'healthy' = 'uninfected';
-                                    'SARS-CoV-2' = 'recovery'"),
+                                    'SARS-CoV-2' = 'SARS-CoV-2'"),
            infection = factor(infection,
-                              c('uninfected', 'recovery'))) %>%
+                              c('uninfected', 'SARS-CoV-2'))) %>%
     select(patient_id,
            all_of(unname(tst_tyr$responses)),
            all_of(tst_tyr$lexicon$variable)) %>%

### 2023-06-07 tools/globals.R

@@ -51,18 +51,18 @@
          log_age = c('log age', 'years', NA),
          sqrt_age = c('sqrt age', 'years', NA),
          sex = c('sex', NA, NA),
-         psych_comorb = c('psychiatric disorder', NA, NA),
+         psych_comorb = c('mental disorder', NA, NA),
          somatic_comorb = c('physical disorder', NA, NA),
          infection = c('SARS-CoV2', NA, NA),
-         bmi_class = c('body mass', NA, NA),
+         bmi_class = c('body mass index', NA, NA),
          smoking = c('smoking', NA, NA),
          alcohol = c('alcohol', NA, NA),
          hads_anx_score = c('HADS, anxiety score', NA, NA),
          hads_dpr_score = c('HADS, depression score', NA, NA),
          hads_signs = c('anxiety/depression signs', NA, NA),
-         pss_stress_score = c('PSS-4, stress score', NA, NA),
-         log_pss_stress_score = c('log PSS-4, stress score', NA, NA),
-         sqrt_pss_stress_score = c('sqrt PSS-4, stress score', NA, NA),
+         pss_stress_score = c('PSS-4, mental stress score', NA, NA),
+         log_pss_stress_score = c('log PSS-4, mental stress score', NA, NA),
+         sqrt_pss_stress_score = c('sqrt PSS-4, mental stress score', NA, NA),
          phq_dpr_score = c('PHQ-9, depression score', NA, NA),
          log_phq_dpr_score = c('log PHQ-9, depression score', NA, NA),
          sqrt_phq_dpr_score = c('sqrt PHQ-9, depression score', NA, NA),

### 2023-06-07 tools/tools.R

@@ -12,6 +12,7 @@
   library(knitr)
   library(bookdown)
   library(clipr)
+  library(caret)
 
 # Descriptive stats -----
 
@@ -106,25 +107,25 @@
 
 # multiple testing -----
 
-  re_adjust <- function(data, method = 'BH') {
+  re_adjust <- function(data, p_variable = 'p_value', method = 'BH') {
 
     ## adjusts for multiple testing e.g. with the Benjamini-Hochberg method
 
     if(method != 'none') {
 
       data <- data %>%
-        mutate(p_adjusted = p.adjust(p_value, method = method))
+        mutate(p_adjusted = p.adjust(.data[[p_variable]], method = method))
 
     }
 
     data %>%
-      mutate(significance = ifelse(p_adjusted < 0.001,
+      mutate(significance = ifelse(.data[[p_variable]] < 0.001,
                                    'p < 0.001',
-                                   ifelse(p_adjusted >= 0.05,
+                                   ifelse(.data[[p_variable]] >= 0.05,
                                           paste0('ns (p = ',
-                                                 signif(p_adjusted, 2), ')'),
+                                                 signif(.data[[p_variable]], 2), ')'),
                                           paste('p =',
-                                                signif(p_adjusted, 2)))))
+                                                signif(.data[[p_variable]], 2)))))
 
   }
 
@@ -494,10 +495,11 @@
                              vars = c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF'),
                              average_stat = c('mean', 'median'),
                              ribbon_stat = c('SEM', '2SEM', 'IQR'),
+                             split_factor = 'timepoint',
                              plot_title = NULL,
                              plot_subtitle = NULL,
                              y_lab = expression('normalized log'[2] * ' concentration'),
-                             x_lab = 'SARS-CoV-2',
+                             x_lab = 'SARS-CoV-2 infection',
                              ribbon_alpha = 0.45) {
 
     ## plots average stats with errors as lines and ribbons
@@ -522,21 +524,23 @@
 
     ## computing stats per timepoint -------
 
-    data <- data[c('timepoint', vars)]
+    data <- data[c(split_factor, vars)]
+
+    levs <- levels(data[[split_factor]])
 
     summ_data <- data %>%
-      blast(timepoint) %>%
-      map(select, -timepoint) %>%
[…]

### 2023-06-11 paper/markdown/rebuttal_letter.Rmd
Feedback from 08 - 09.06.2023
Changes in headings and labels in Figures, Tables, Supplementary Figures and Supplementary Tables. Changes in the manuscript and supplement text.

@@
 ---
-title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
+title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
 subtitle: "Rebuttal Letter"
@@
 bibliography: infl_biblio.bib
-csl: frontiers-in-immunology.csl
+csl: psychoneuroendocrinology.csl
 header-includes:
@@
 
-Thank you for appreciation of our manuscript and giving us a chance to address Reviewers' issues. 
-First we would like to thank the Reviewers for their very careful, thoughtful and helpful assessment and comments. 
+Thank you for appreciation of our manuscript and giving us a chance to address reviewers' concerns. 
+First we would like to thank the reviewers for their very careful, thoughtful and helpful assessment and comments. 
 We acknowledge that such detailed reviews take a lot of time and demand a lot of commitment and hope that we have done them justice in the revised manuscript.
 
-Overall, we agree that the initial manuscript version was not sufficiently coherent and clear because of the involvement of a large team of researchers in the study. 
-Following your advice and the Reviewers' suggestions, we reworked thoroughly the analysis and the text. 
-We would like to point out that this is one of very few manuscripts that characterize activity of the serotonin, kynurenine and catecholamine pathways, i.e.
-the metabolic system that controls systemic availability of indolamine and catecholamine neurotransmitters in COVID-19 [@Santiago-Mujika2022; @Saito2022; @Lionetto2021; @Bizjak2022; @Dewulf2022; @Ceballos2022].
-Importantly, their role in development of symptoms of mental health disorders in COVID-19 patients was proposed [@Kucukkarapinar2022; @Bower2022]. 
-By analysis of two cohorts, we were able to identify key factors affecting activity of those metabolic pathways and investigate their temporal activity profile in acute and sub-acute SARS-CoV-infection and recovery, which adds to novelty of our work.
+Overall, we agree that the initial manuscript version was not sufficiently coherent and clear likely because of the involvement of a large team of researchers in the study. 
+Following your advice and the reviewers' suggestions, we reworked thoroughly the analysis and the text. 
+We would like to point out that this is one of very few manuscripts characterizing activity of the kynurenine and catecholamine pathways in COVID-19 [@Santiago-Mujika2022; @Saito2022; @Lionetto2021; @Bizjak2022; @Dewulf2022; @Ceballos2022].
+Importantly, their role in symptoms of mental health disorders in COVID-19 patients has been proposed [@Kucukkarapinar2022; @Bower2022]. 
+By analyzing two cohorts, we were able to identify key factors affecting the activity of those metabolic pathways and investigate their temporal activity profile in acute and sub-acute SARS-CoV-infection and recovery, which adds to novelty of our work.
 
-We have re-analysed the data to address the reviewers´ issues and have rewritten many parts of the manuscript following the reviewers' suggestions. 
-In particular, we now clearly and explicitly state the gap in the literature in __Introduction__ and the analysis endpoints in __Methods__ and __Results__. 
+We have re-analysed the data to address the reviewers´ concerns and have rewritten many parts of the manuscript following the reviewers' suggestions while keeping within the word limits of the journal.
+Despite extensive changes on the manuscript text, the main results remain unaltered.
+In particular, we now clearly and explicitly state the gap in the literature in __Introduction__ and the analysis endpoints in the __Methods__ and __Results__ sections. 
 To facilitate interpretation of our results, we now follow a top - bottom analysis approach. 
-In brief, we employed the most robust and comprehensive analytic technique, multi-parameter modeling, to identify the most relevant explanatory factors affecting the key metabolites and activity markers of the kynurenine, serotonin and catecholamine pathways.
+In brief, we employed the most robust and comprehensive analytic technique, multi-parameter modeling, to identify the most relevant explanatory factors affecting the key metabolites and activity markers of the kynurenine pathway, serotonin availability and the catecholamine pathway.
 Subsequently, we explore these factors in more detail by correlation and network analysis, and canonical statistical tests. 
-Following the Reviewer's suggestion we also keep analysis results in the SIMMUN and INCOV cohort separately for the sake of clarity. 
-Concerning the manuscript text we also improved the preciseness of the language and adapted the __Discussion__ part. 
-Please see below the point-to-point response for details. 
-Please notice that due to the scale of changes in the text, it was not feasible to mark all changes introduced during revision. 
-The most relevant changes are <span custom-style = 'revision'>highlighted in red</span>. 
+Following the reviewer's suggestion we also keep analysis results in the SIMMUN and INCOV cohort separately for the sake of clarity. 
+We have re-arrange and rewritten the Discussion section taking into account the reviewer´s suggestions.
+
+Please see below the point by point response for details. 
+Please notice that due to the scale of changes in the text, it was not feasible to mark the changes introduced during revision. 
+To meet the journals' requirements concerning the maximal number of tables and figures, we had to eliminate the original figure 1 (model of investigated pathways) and move the scheme of analysis inclusion and strategy flowchart to __Supplementary Material__ (__Supplementary Figure S1__).
+Since this improved flowchart is very helpful for the understanding of our analysis, we would move it back to the main text if we are granted to exceed the maximum figure number upon acceptance.
 We hope that our revised manuscript is now suitable for publication in _Psychoneuroendocrinology_.
@@
 
-We thank the Reviewer for the careful assessment of our manuscript and valuable feedback which were very helpful in revising our work. 
+We thank the reviewer for the careful assessment of our manuscript and valuable feedback which were very helpful in revising our work. 
 First we would like to apologize for imprecise wording, not including enough references and any language errors. 
@@
 
-* we unified nomenclature of the metabolic pathways investigated in our manuscript and term them now 'serotonin pathway', 'kynurenine pathway' and 'catecholamine pathway' in line with prominent review articles [@Badawy2017; @Lukic2022; @Meiser2013]
+* We unified nomenclature of the metabolic pathways investigated in our manuscript and term them now 'kynurenine pathway' and 'catecholamine pathway' in line with prominent review articles [@Badawy2017; @Lukic2022; @Meiser2013]
 
-* we reorganized and rewrote many parts of __Introduction__.
+* We reorganized and rewrote many parts of __Introduction__.
 
-* we also more explicitly and clearly state the ‘gap in the literature’ and our research aims in the last paragraph of __Introduction__.
+* We also more explicitly and clearly state the ‘gap in the literature’ and our research aims in the last paragraph of __Introduction__.
 
-* we included the __Analysis endpoints__ section of __Methods__. Here we describe our analytic approach, define outcome variables and candidate explanatory factors.
+* We included an __Analysis endpoints__ section in __Methods__. Here we describe our analytic approach, define outcome variables and candidate explanatory factors.
 
-* we show the analysis strategy for the SIMMUN and INCOV cohort in __Supplementary Figure S1__.
+* We show the analysis strategy for the SIMMUN and INCOV cohort in __Supplementary Figure S1__.
 We hope that this makes it easier to follow our analysis.
@@
 This makes the results of particular analyses more easily comparable to one another.
-Additionally, to investigate if and how exclusion of participant affected the SIMMUN cohort characteristic and if it could be regarded as a source of bias, we compared the excluded and analyzed individuals as presented in __Supplementary Table S2__ and __Results/Characteristic of the cohorts__.
+Additionally, we compared the excluded and analyzed individuals as presented in __Supplementary Table S2__ and __Results/Characteristic of the cohorts__.
 
@@
 
-* In order to improve clarity of __Results__, we have now reordered and reformatted it.
+* In order to improve clarity of __Results__, we have now reordered and reformatted this section.
 We now present the results in a ‘top - bottom’ approach.
@@
 
-We agree that the role of the kynurenine pathway metabolites as well as systemic, i.e.
-peripheral availability of tryptophan, serotonin and dopamine in mental health are controversial, not sufficiently supported by direct experimental evidence and definitely needs a critical view on the topic [@Marx2020; @Brown2021; @Moncrieff2022; @Goldstein1999].
-We find that one of the many challenges is that many mechanistic studies are done in the animal model, making the transfer to the human situation questionable [@Brown2021]. 
-On the other hand human studies are quite equivocal and, while there are influential theoretical manuscripts on the issue, surprisingly few original research has been published so far [@Brown2021]. 
-To address you issue, we decided to give a more balanced view on the serotonin, kynurenine, catecholamine pathways and their role in mental health while still keeping the text relatively compact. 
-We now give a more clear, but short introduction to there three biochemical pathways in the introduction section. 
-We provide an ample discussion of the possible role of these pathways in linking mental health and physical conditions. 
-In __Discussion__ and __Limitations__ we comment on controversy over the role of systemic serotonin, kynurenine or catecholamine pathway in depression and other psychiatric disorders [@Brown2021; @Marx2020; @Schwarcz2012; @Moncrieff2022; @Goldstein1999].
+We agree that the role of systemic, i.e.
+peripheral availability of serotonin as well as its association with kynurenine pathway activity in mental health are controversial, and definitely warrants a critical discussion [@Marx2020; @Brown2021; @Moncrieff2022; @Goldstein1999].
+We have tried to give a more balanced view on the topic in the revised Introduction and Discussion sections, taking the reviewer´s concerns and suggestions into account. 
+On one hand, we find that one of the many challenges is that many mechanistic studies are done in animal models and their significance for humans is not well documented [@Brown2021]. 
+On the other hand, human studies are quite equivocal and, while there are influential theoretical manuscripts on the issue, surprisingly few original research has been published so far [@Brown2021].
 
-Additionally, to address the question how TRP levels may affect systemic serotonin pathway activity in COVID-19 patients, we modeled circulating serotonin levels in the INCOV cohort [@Su2022] as a function of its precursors and competitor metabolites (TRP, KYN, QUIN), timepoint of SARS-CoV-2 infection, inflammatory cytokines and age.
-As presented in __Figure 2__, we could indeed find a significant positive association of plasma TRP and serotonin in multi-parameter modeling, which is expected to be independent of age, inflammation and infection time course.
-In a more detailed correlation and network analysis of the INCOV dataset (__Figure 4A__, __Supplementary Figure S6__), we could observe significant, weak-to-moderate strength correlation of circulating serotonin and TRP levels in acute and sub-acute infection.
-This suggests that systemic serotonin synthesis may be limited by blood TRP availability to some extent during SARS-CoV-2-dependent inflammation, which was characterized by lowered plasma TRP concentration and elevated KYN levels.
+We now give a more clear, but short introduction to the biochemical pathways in __Introduction__. 
+We provide critical discussion of the possible role of these pathways in linking mental and physical health. 
+In __Discussion__ and __Limitations__ we comment on the controversy over systemic serotonin availability as well as kynurenine and catecholamine pathway in depression and other mental disorders [@Brown2021; @Marx2020; @Schwarcz2012; @Moncrieff2022; @Goldstein1999].
+
+Additionally, to address the question whether TRP levels are associated with systemic serotonin availability in SARS-CoV-2 infection, we modeled circulating serotonin levels in the INCOV cohort [@Su2022] as a function of its precursor and competitor metabolites (TRP, KYN, QUIN), timepoint of SARS-CoV-2 infection, inflammatory cytokines and age.
+As presented in Figure 2, we found a significant positive association of plasma TRP and serotonin in multi-parameter modeling, which is expected to be independent of age, inflammation and infection timepoints.
+In a more detailed correlation and network analysis of the INCOV dataset (Figure 4A, Supplementary Figure S6), we observed significant, weak-to-moderate strength correlation of circulating serotonin and TRP levels in acute and sub-acute infection.
 
@@
 We are sorry for this unclarity. 
-In the revised manuscript, we formulate the aims, and define outcomes and explanatory factors. 
+In the revised manuscript, we rephrased the aims, and define outcomes and explanatory factors. 
 Please refer to the last paragraph of __Introduction__ and the subsection __Analysis endpoints__ of __Methods__.
@@
 
-We agree with the Reviewer that the INCOV cohort deserves a more thorough description. 
-We included short methodological information of the INCOV cohort in __Methods/Procedures__ and __Supplementary Methods__ [@Su2020; @Su2022].
+We included short methodological information of the INCOV cohort in __Methods__ (subsection __Procedures__) and in __Supplementary Methods__ [@Su2020; @Su2022].
 
@@
 
-We provide now these information in __Methods__ and __Supplementary Methods__. 
+We provide this information in __Methods__ and __Supplementary Methods__. 
 Variable format, transformations and, if appropriate, stratification schemes are presented in __Supplementary Tables S1__ and __S3__.
@@
 
-As specified now in __Methods/Procedures__, values beyond or at their detection limits were substituted with the lower or upper detection limit. 
+As specified now in __Methods__ (subsection __Procedures__), values beyond or at their detection limits were substituted with the lower or upper detection limit. 
 CRP and IL6 were removed from the revised analysis pipeline for the reasons specified in __Issue 1__.
@@
 
-We concur with the Reviewer, that a comparison between the cohorts is useful for interpretation of the results. 
+We concur with the reviewer, that a comparison between the cohorts is useful for interpretation of the results. 
 We present significant results of the comparison between the SIMMUN and INCOV cohort in __Supplementary Table S5__ and elaborate on them in __Results/Characteristic of the study cohorts__. 
@@
 We abstained from referencing test statistics such as $\chi^2$ or $t$, since they depend on the number of complete observations and hence do not deliver reliable information on the effect size. 
+Instead we present p values and effect size statistics in plots and tables.
 
@@
 
-Since we re-wrote and shortened the __Introduction__ section, we abstained from using subheadings; such format is also not common in original research manuscripts in _Psychoneuroendocrinology_. 
+Since we re-wrote and shortened the __Introduction__ section, we did not included subheadings. 
 To facilitate interpretation, each topic of __Introduction__ is described in a separate paragraph. 
-In turn, we employ subheadings in __Discussion__.
+We now included subheadings in __Discussion__.
 
@@
 Thanks for the attentive assessment, this misleading wording was removed.
-Please refer to the metabolic pathway nomenclature in __Issue 1__.
 
@@
 Furthermore, we describe the rationale for using the INCOV dataset in the last paragraph of __Introduction__ and the __Analysis endpoints__ subsection of __Methods__. 
-Information on measurement of proteins and metabolites is provided in __Methods/Procedures__ and __Supplementary Methods__.
-Unfortunately, the incompatibility of designs and datasets of these two cohorts hindered us from developing a comprehensive model in one cohort and direct validation in the outer (canonical training - test setup).
+Information on measurement of proteins and metabolites is provided in __Methods__ (subsection Procedures) and __Supplementary Methods__. 
+Unfortunately, data incompatibilities of the two cohorts hindered us from analyzing them in one comprehensive model with direct validation in the other (canonical training - test setup). 
 We list this issue in __Limitations__.
@@
 We restructured and improved __Discussion__. 
-Large parts of __Discussion__ were reorganized and rewritten as suggested by Reviewer 2 (__Issue 40__). 
-We hope that you agree with the structure proposed by the other Reviewer which we have now followed. 
-We have also included new references to back out results. 
+Large parts of __Discussion__ were reorganized and rewritten as suggested by reviewer 2 (__Issue 40__). 
+We hope that you agree with the structure proposed by the other reviewer which we have now followed. 
+We have also included new references to back and critically interpret our results.
 The __Limitations__ section was also expanded and improved.
@@
 
-We apologize, this is really misleading and was reworded. 
-We meant levels of inflammatory markers in SARS-CoV-2-uninfected individuals, which we state now explicitly.
+We apologize, this phrase was removed.
 
@@
 
-We concur with the Reviewer's view, that visualization of trends for monotone associations is problematic. 
+We concur with the reviewer's view, that visualization of trends for monotone associations is problematic. 
 Due to profound changes in the structure of the revised manuscript, we do not use scatter plots to illustrate monotone correlations between non-normally distributed variables any more. 
@@
 
-We thank the Reviewer for appreciation of out text and the constructive feedback. 
-Weaknesses in structure of the analyses were already pointed out by the Reviewer 1. 
+We thank the reviewer for appreciation of out text and the constructive feedback. 
+Weaknesses in structure of the analyses were already pointed out by the reviewer 1. 
 We addressed them by employing the multi-parameter modeling results for identification of the most vital explanatory factors that affect neurotransmitter metabolism and subsequent more specific correlation analyses and statistical testing with those relevant predictors.
@@
 
-As also pointed out by Reviewer 1, the last paragraph of __Introduction__ was not written clearly enough in the initial manuscript version. 
-In the final paragraph of __Introduction__, we present now the ‘gap in the literature’, which is the still incompletely characterized effect of SARS-CoV-2 infection on metabolic pathways determining systemic availability of serotonin and dopamine.
-We also state the aim of our analysis, which was characterization of factors affecting the serotonin, kynurenine and catecholamine pathway activity during SARS-CoV-2 infection and recovery. 
-Our study did  not have a fully exploratory character, since the candidate explanatory factors: age, sex, body mass index, mental and somatic conditions, smoking and alcohol, depression, anxiety, stress and inflammation have already been described to affect serotonin and dopamine metabolism in previous research, also from our own research consortium [@Hufner2019; @Hufner2020; @Hufner2021; @Hufner2015].
+As also pointed out by reviewer 1, the last paragraph of __Introduction__ was not written clearly enough in the initial manuscript version. 
+In the final paragraph of __Introduction__, we now present the ‘gap in the literature’ and the specific aim of our analysis. 
+Our study did  not have a fully exploratory character, since the candidate explanatory factors: age, sex, body mass index, mental and physical disorders, smoking and alcohol consumption, depression, anxiety, mental stress and inflammation have already been described to be associated with serotonin availability, and kynurenine and catecholamine pathway activity in previous research, also from our own research consortium [@Hufner2019; @Hufner2020; @Hufner2021; @Hufner2015].
 
@@
 
-This was reworded in the revised version, we utilize 'catabolism', 'breakdown' or 'conversion' instead.
+This was reworded in the revised manuscript.
 
@@
 We removed the sentence from __Discussion__ referring to somatic symptoms, since those were not systematically surveyed in the SIMMUN cohort. 
-Following your suggestion, we term ‘somatic conditions’ ‘physical disorders’ consequently in the revised text.
+Following your suggestion, we use the terms 'mental disorders' and 'physical disorders' consequently in the revised text.
 
@@
 
-We thank the Reviewer for this specific suggestion on the discussion section which we have gratefully taken up and rewritten large parts of __Discussion__ according to the following scheme:
+We thank the reviewer for this specific suggestion on the discussion section which we have gratefully taken up and rewritten large parts of __Discussion__ according to the following scheme:
 
@@
 
-* Effects of inflammation and infection, including SARS-CoV-2, on the serotonin and kynurenine pathways, and systemic levels of serotonin
+* Effects of inflammation and infection, including SARS-CoV-2, on the kynurenine pathway as well as systemic levels of serotonin
 
@@
 * Mutual effects of mental health-related factors (e.g.
-depression, anxiety and mental stress) on activity of the serotonin, kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability.
-Here we also underline that evidence for effects of peripheral serotonin, kynurenine and catecholamine pathways is limited and based in many cases on clinical association of psychiatric disorders with blood levels of the neurotransmitters, their precursors or competitor pathway products.
+depression, anxiety and mental stress) on activity of the kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability.
+We provided a balanced and critical discussion of the evidence here.
 
@@
 
-We thank the Reviewer for appreciation of our manuscript and the valuable feedback.
+We thank the reviewer for appreciation of our manuscript and the valuable feedback.
 
@@
 
-We absolutely agree with the Reviewer. 
-To clarify that, we use consistently the terms 'serotonin pathway', 'kynurenine pathway' and 'catecholamine pathway' in the revised manuscript.
+We absolutely agree with the reviewer. 
+We now consistently use the terms ‘serotonin’, ‘kynurenine pathway’ (TRP, KYN, QUIN) and ‘catecholamine pathway’ (PHE, TYR, dopamine sulfate) in the revised manuscript.
 
@@
 Thank you for the suggestions. 
-As described in __Issues 1__ and __38__, we present the results in cohort-wise manner, for the sake of better readability. 
-We also included the analysis strategy scheme in __Supplementary Figure S1__.
+As described in __Issues 1__ and __38__, we present the results in cohort-wise manner, for the sake of clarity. 
+We also included the analysis strategy scheme as __Supplementary Figure S1__. 
+Unfortunately we had already reached the maximum figure number allowed so that this could not be included into the main text.
 

### 2023-06-11 paper/markdown/issues_07062023.Rmd
Feedback from 08 - 09.06.2023
Changes in headings and labels in Figures, Tables, Supplementary Figures and Supplementary Tables. Changes in the manuscript and supplement text.

--- 2023-06-07 paper/markdown/issues.Rmd
@@
 ---
-title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
-subtitle: "Issues"
+title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
+subtitle: "Issues 07.06.2023"
 author: "PSY Team"

### 2023-06-11 paper/markdown/issues_10062023.Rmd
Feedback from 08 - 09.06.2023
Changes in headings and labels in Figures, Tables, Supplementary Figures and Supplementary Tables. Changes in the manuscript and supplement text.

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
subtitle: "Issues 10.06.2023"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# General

* I really like the title, thanks!

* Following your suggestion, I've replaced '5-HT' with 'serotonin' in Figures, Tables and Supplementary Material.

# Rebuttal Letter

* A minor language issue: I'd propose to use 'Discussion', 'Introduction' or 'Results' instead of 'the Discussion', 'the Introduction' and 'the Results', because we're referring to parts of the text, which are named so. I used to work together with US companies during my lab time at the University, this was a kind of rule in papers and reports for them.

* Issue 11. I'm not a big fan of presenting test stats in tables or figures. An example: we have two variables: one with mean of 0, the other with mean of 0.1, both have SD = 1. This difference is not significant if each variable has n = 100 observations (t = 1.4) but highly significant for n = 10000 observations (t = 6.8). The effect size of this difference is weak as determined by Cohen's d:

$$d = \frac{Mean(x1) - Mean(x2)}{SD} = \frac{0.1 - 0}{1} = 0.1$$ 
So the test stat like t or $\chi^2$ tells us only, if the effect is significant (i.e. obtained not by chance), but does not allow us to interpret its relevance! Yes, there are test stats like correlation coefficients, which provide information on effect sizes as well, but these are only special cases.

# Supplementary Material

* I prefer keeping details on methods in Supplementary Material event if they are provided in the main manuscript as well. The reason for that is pragmatic: I don't have to look up for technical details in the main manuscript, while I'm reading Supplementary Material and vice versa. If you wish, you can delete them, at the moment I'm leaving them as they are.

* Section __Supplementary Methods/INCOV cohort dataset__: I'm fine with the rewritten sentence on analysis timepoints.

* References are now in the right format.

* Variable 'Clinically relevant symptoms of depression or anxiety (HADS ≥ 8)' codes for depression or anxiety symptoms, I think we used it in previous analyses. I've removed it from the final version of Supplementary Material and Manuscript, because we don't discuss it anymore.

* Issues with Supplementary Table S2 were cleared ('healthy' -> 'uninfected')

* Supplementary Table S4, the heading is fine for me, thanks!

* Supplementary Table S5: 'SARS-CoV-2 infection', as suggested.

* Supplementary Figure S1: corrected as suggested.

# Main manuscript

## Abstract

* The Background paragraph: I've made it shorter and tried to point out, why the kynurenine and catecholamine pathways are postulated to be important for mental health. I feel this is quite important for somebody from outside the field.

* I've re-phrased slightly the conclusion paragraph, to underline direction of changes in kynurenine and catecholamine pathway activity. I think wording 'which may reduce serotonin availability' is anyway quite weak. 'associated' would suggest that we could show only a coincidence. Yes we did that, but combining our findings with literature, we can speculate that - its also about marketing:). 
I also avoided  repeating the information from the Background paragraph ('link between physical disorders and mental health').

## Introduction

* Thanks for edits in Introduction, I accepted almost all of them. And for staying far below 1000 words:) Just few issues below.

* I know I'm stubborn, but the opening sentence 'The immune system and the brain interact at multiple levels with influences on one system having consequences on the other' can be simply reduced to 'The immune system and the brain interact at multiple levels'. 'Interact' means that one thing influences another and vice versa, please find the [definition here](https://www.merriam-webster.com/dictionary/interaction). You may of course keep it as it is, but there are 10 words extra without additional information.

* 'Hence, the kynurenine pathway, whose activity can be assessed by the ratio of blood KYN to TRP (KYN/TRP) is an activation marker of IDO and TDO activity and thereby reflects inflammatory and hypothalamic–pituitary–adrenal axis signals' - there was too many 'activation' and 'activity' in one sentence, I had to re-write it. I would no say that KYN pathway activity or IDO activity reflects inflammatory signals - some say, it pertains to immunosuppression, e.g. in tumors. Anyway, it may sound that KYN pathway activity is a kind of inflammatory marker.

* 'The synthesis of __catecholamine__ neurotransmitters such has dopamine via the catecholamine pathway involves the conversion of phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH)'. I've added 'catecholamine' to stress that is concerns to one particular family of neurotransmitters.

## Methods

* INCOV cohort: of course we don't have the ratios, thanks for pointing out! This was a copy-paste error.

* Analysis endpoints. Thanks for the edits. I've reworded 'inflammatory cytokines' to 'cytokine markers or inflammation', because IL10 is actually an anti-inflammatory cytokine (but strongly induced by NF-kappa-B signals and hence an inflammatory marker).

* Statistical analysis. The sentence 'Responses and explanatory variables were normalized prior to modeling' does not refer to representation of variables. It means, that I transform numeric variables to Z scores (normalization):

$$Zscore = \frac{Mean(x)}{SD(x)}$$

This is actually a vary important statement, because it tells me, how to interpret the models. When I model normalized explanatory variables, I can compare $\beta$ coefficients for e.g. age ($\beta$ = `my_beta(mod_eli$inference$kyn, 'age')`) and NEO ($\beta$ = `r my_beta(mod_eli$inference$kyn, 'log_neo')`) and say that NEO is a stronger predictor of KYN levels. I'm allowed to state tat because Z scores of age and NEO are in the same scale:) Since the responses are also expressed as Z-scores, I can also say that e.g. difference of 1 $\times$ SD in age causes 0.25 $\times$ difference in KYN.

## Results

* Characteristic of the study cohort. I find your proposal interesting - I tend to provide extremely short characteristics of the cohorts e.g. in onco-papers. Generally, I'd keep the information on physical and mental disorders, since its important for discussion of bias of our cohorts (enrichment of mental disorders in the SIMMUN). We can remove the sentence on anti-RBD IgG (already done as suggested).

* I agree on references in Results and removed almost all of them, you're right, have anyway an extensive Introduction and Discussion.

* R^2^: thanks I've forgotten to delete the % mark. Values of 0.1 and 0.3 are correct.

* Some remarks to the sentence (subsection: 'Inflammatory cytokines, SARS-CoV-2 infection course, age and availability of biosynthesis precursors regulate systemic serotonin and dopamine levels'): 'Next, we investigated factors affecting plasma levels of serotonin and a major circulating catabolite of dopamine, DA sulfate,  as direct markers of systemic serotonin and dopamine availability, respectively.'. Because we don't employ the wording 'serotonin pathway' anymore, plasma serotonin is simply a marker of systemic serotonin availability and not a marker of 'associated kynurenine pathway'. Markers of the kynurenine pathway are explanatory variables, we have to be really strict here and follow the Analysis endpoints', because the mixing up explanatory and dependent variables was an issue of a reviewer. The same refers to the catecholamine pathway: we're using its metabolites PHE and TYR as explanatory variables.

* The sentence: 'This analysis was done in the INCOV cohort [@Su2022] with age, sex, BMI, markers of inflammation (IL6, IL10, TNF, IFNG), status and timepoint of SARS-CoV-2 infection, concentrations of kynurenine (TRP KYN, QUIN) and catecholamine pathway metabolites (TRP, PHE, TYR) as candidate explanatory variables (__Supplementary Table S3__)'. I'd propose skipping 'affecting also serotonin', since we introduce the pathway and its role in serotonin availability in Introduction and elaborate on it in Discussion. Somehow, we also assume that the kynurenine pathway affects serotonin, because we include its metabolites among explanatory variables.

## Discussion

* Thanks for the edits and feedback and staying well within the word limit:)

### Result summary

* 'neurotransmitter metabolism': I'd propose to write more precisely 'serotonin and catecholamine metabolism'.

* The sentence: 'This observation and the significant association of serotonin with TRP levels in the INCOV cohort suggest collectively that depletion of TRP via highly active kynurenine pathway can limit systemic serotonin availability during SARS-CoV-2 infection [@Badawy2017].' You may remove it if you wish. I put it here for marketing reasons. Indeed, there are many papers that show that inflammation increases KYN pathway metabolite levels, but very few that show that, that this phenomenon has something to do with serotonin.

* 'male sex' removed, I agree!

* I'd remove the sub-heading 'General discussion', we're still elaborating on our data without a broader literature context, so for me, we're still in 'Result summary'. The edits to the paragraph are really fine! I agree, 'discrepancies' is not a fortunate word:)

### Effects of inflammation and infection on the kynurenine pathways and systemic levels of serotonin

* The wording 'metabolite levels of the kynurenine pathway' is not precise (do metabolites or levels refer to the kynurenine pathway?), I've replaced it by 'levels of kynurenine pathway metabolites'

### Effects of inflammation and infection on the catecholamine pathway and systemic dopamine availability

* The sentence 'However, during inflammation, BH4 availability for catecholamine and serotonin synthesis enzymes is limited due to oxidation and depleted by nitric oxide synthesis'. It implies that availability of something is depleted. What about: 'However, during inflammation, BH4 availability for catecholamine and serotonin synthesis enzymes is reduced by oxidation and nitric oxide synthesis'?

* The sentence 'We also found an inverse association of parameters of catecholamine pathway activity such as PHE and PHE/TYR (increased), and TYR and DA sulfate levels  (decreased) with inflammatory markers NEO, IL6, TNF and IFNG'. The wording 'parameter' was criticized by a reviewer (he/she's right, 'parameter' refers to a model term), changing to 'readouts'.

### Interaction of mental health with activity of the kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability

* I've added Kim 2017 and your Brain Sci as evidence for the link of mental stress and inflammation.

* The sentence 'Additionally, kynurenine pathway metabolites can contribute to the sickness behavior and depression via interference with neuronal signaling, e.g. with glutamatergic receptors': I'm trying to be more precise here and spell out the sickness behavior and depression.

* 'Inflammatory markers were also associated with elevated PHE/TYR indicative of reduced dopamine availability and proposed to contribute to depression in cancer (Hüfner et al., 2015) and trauma (Hüfner et al., 2019).' I think this indeed fits to our data: NEO -> low TYR (SIMMUN), IFNG -> low DA sulfate (INCOV, modeling and correlation) and IL6, TNF, IL10 -> high PHE (INCOV correlation), does it? I reworded it a bit for the sake of clarity.

* 'The potential role of __persistent__ inflammation and TRP depletion in persistent somatic symptoms and mental disorders in COVID-19 convalescents was proposed in recent hypothesis papers'. I've removed the first 'persistent', the word was repeated in twice in the sentence.

* The sentence 'In the SIMMUN cohort, clinically relevant symptoms of depression (HADS) and mental stress (PSS-4) along with the inflammatory marker NEO could were associated with lower TRP values and a higher KYN/TRP ratio, respectively' is not precise enough: depression symptoms were formally associated with TRP and mental stress was a predictor of KYN/TRP. I'd say 'formally', because this two psychometric variables were extremely inter-correlated and somehow 'competed' to be a significant predictor in the models. I've split the sentence in two ones.

### Limitations

* Thanks!

### Conclusions

* As in Abstract, I've added information on direction of changes in activity of the kynurenine and catecholamine pathway, and availability of serotonin and dopamine.

* 'parameters' removed, sorry:)

## Tables

* Tables 1 and 2 versus Table S2: there's indeed a difference in format. In Tables 1 and 2 I don't have to provide extra number of complete cases for each analyzed variable, because the analyzed cohorts contain only complete cases. In Supplementary Table S2, I'm comparing the analyzed SIMMUN cohort with excluded participants. The excluded ones have per definition missing data, so I need to provide the number for each variable here.

* Tables 1 and 2: I'd stay with the 'Uninfected' and 'SARS-CoV-2 infection' header, if we introduce 'SARS-CoV-2-negative' and 'SARS-CoV-2-positive', we have to change it in all figures and tables.

* As discussed per telephone, I'd keep for the moment Table 1 and Table 2, because one of the reviewers criticized that we mix results from the cohorts. I think merging the tables is worth considering, if the editor tells us to move Supplementary Figure S1 to the main text.

## Figures

* Changes in the figures and figure legends following your suggestions and classification of TRP as a kynurenine pathway metabolite

# References

### 2023-06-11 paper/markdown/main_manuscript.Rmd
Feedback from 08 - 09.06.2023
Changes in headings and labels in Figures, Tables, Supplementary Figures and Supplementary Tables. Changes in the manuscript and supplement text.

@@
 ---
-title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
+title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
 subtitle: "Manuscript"
@@
       institute: [psych2]    
-  - Eberhard Deisenhammer: 
+  - Eberhard A. Deisenhammer: 
       institute: [psych2, psych1]
@@
       institute: [psych2]
-  - Alexander Egger: 
-      institute: [zimcl]
   - Matyas Galffy: 
@@
   - Jens Lehmann: 
-      institute: [psych1]
+      institute: [psych2]
   - Maria Oberhammer: 
@@
   - surg: Department of Visceral, Transplant and Thoracic Surgery (VTT), Medical University Innsbruck, Innsbruck, Austria. 
-  - zimcl: Central Institute of Clinical and Chemical Laboratory Diagnostics, University Hospital of Innsbruck, Innsbruck, Austria.
   - bio:  Insitute of Medical Biochemistry, Medical University Innsbruck, Innsbruck, Austria
@@
 
-Running title: Activity of the kynurenine, serotonin and dopamine metabolic pathways and COVID-19
+Running title: Activity of the serotonin, kynurenine, and catecholamine metabolic pathways and COVID-19
 
@@
 
-__Background:__ COVID-19 patients frequently develop symptoms of depression, anxiety and mental stress. 
-The serotonin, kynurenine and catecholamine metabolic pathways determine systemic levels of serotonin and dopamine, whose insufficiency was linked to mental health disorders. 
-We investigated factors that affect their activity in SARS-CoV-2 infection and recovery.
+__Background:__ Increased prevalence of mental disorders following COVID-19 has been described. 
+The kynurenine and catecholamine metabolic pathways are modulated by inflammation and can affect systemic levels of serotonin and dopamine. 
+Their activity may hence link physical disorders with mental health. 
+We investigated factors that affect kynurenine and catecholamine pathway activity in SARS-CoV-2 infection and recovery.
 
 __Methods:__ The cross-sectional SIMMUN (n = `r nrow(stigma$analysis_tbl)`) and longitudinal INCOV cohort (n = `r  length(unique(incov$complete_ids))`, Su et al. 2022) were analyzed. 
-Demographic and clinical characteristic, inflammatory markers, SARS-CoV-2 infection, signs of depression and anxiety (HADS), and mental stress (PSS-4) served as explanatory variables. 
-Plasma markers of the serotonin and kynurenine (serotonin, kynurenine/tryptophan ratio), and catecholamine pathway activity (dopamine 3-O-sulfate, phenylalanine/tyrosine ratio) were modeled by multi-parameter linear regression.
+Demographic and clinical characteristic, inflammatory markers, SARS-CoV-2 infection, symptoms of depression and anxiety (HADS), and mental stress (PSS-4) served as explanatory variables. 
+Blood serotonin and markers of the kynurenine (kynurenine/tryptophan ratio), and catecholamine pathway activity (dopamine 3-O-sulfate, phenylalanine/tyrosine ratio) were modeled by multi-parameter linear regression.
 
@@
 Phenylalanine/tyrosine ratio was lower in 
-SARS-CoV-2-positive participants (`r my_beta(mod_eli$inference$phe_tyr, 'infection', show_ci_txt = FALSE)`). 
+SARS-CoV-2-positive than SARS-CoV-2-negative participants (`r my_beta(mod_eli$inference$phe_tyr, 'infection', show_ci_txt = FALSE)`). 
 In the INCOV cohort, inflammation markers were associated with decreased serotonin 
@@
 
-__Conclusion:__ SARS-CoV-2 infection, inflammation, mental stress and age are the key independent predictors of high kynurenine pathway activity, which decreases systemic serotonin levels by depleting its precursor tryptophan.
-These factors are also associated with suppressed catecholamine pathway activity. 
-This lowered serotonin and dopamine availability may contribute to mental health problems in COVID-19 patients.
+__Conclusion:__ 
+SARS-CoV-2 infection, inflammation, age and mental stress are the key independent predictors of high kynurenine pathway activity, which may reduce systemic serotonin availability. 
+The catecholamine pathway was suppressed in SARS-CoV-2 infection. 
+Altered activity of these pathways may contribute to mental disorder symptoms following COVID-19.
 
@@
 The immune system and the brain interact at multiple levels [@Dantzer2008]. 
-Physical conditions such as infections, cardiovascular disease, frailty, autoimmune illness or malignancy are often accompanied by symptoms of depression, anxiety, mental stress and other psychiatric disorders [@Mitchell2011; @Dickens2001; @Mayerl2020; @Rajan2020; @Renault1987; @Hufner2015; @Hufner2022].
-Inflammation interferes with key metabolic processes controlling serotonin and dopamine availability, the serotonin, kynurenine and catecholamine pathways, and was proposed as an important link between mental and physical health [@Bonaccorso2002; @Gostner2020; @Valdiglesias2018; @Dantzer2008; @Hufner2015; @Brown2021].
+Physical disorders such as infections, cardiovascular disease, frailty, autoimmune illness or malignancy are often accompanied by symptoms of depression, anxiety and mental stress [@Mitchell2011; @Dickens2001; @Mayerl2020; @Rajan2020; @Renault1987; @Hufner2015; @Hufner2022].
+Inflammation interferes with key metabolic processes such as the kynurenine and catecholamine pathway, which can control serotonin and dopamine availability, respectively. 
+Inflammation was thus proposed as an important link between mental and physical health [@Bonaccorso2002; @Gostner2020; @Valdiglesias2018; @Dantzer2008; @Hufner2015; @Brown2021].
 
-Serotonin is synthesized from the essential aminoacid tryptophan (TRP) via serotonin pathway reactions catalyzed by tryptophan hydroxylase (TPH) and aromatic L-amino acid decarboxylase. 
-Most systemic TRP is catabolized by the kynurenine pathway, which limits serotonin pathway activity [@Lukic2022]. 
-Its first rate-limiting step, the TRP - kynurenine (KYN) breakdown catalyzed by tryptophan 2,3-dioxygenase (TDO) or indoleamine 2,3-dioxygenases (IDO1, IDO2), controls TRP levels available for serotonin synthesis [@Lukic2022; @Badawy2017].
+Serotonin is synthesized from the essential aminoacid tryptophan (TRP) in reactions catalyzed by tryptophan hydroxylase (TPH) and aromatic L-amino acid decarboxylase. 
+Most systemic TRP is catabolized along the kynurenine pathway. 
+The conversion of TRP to kynurenine (KYN) by tryptophan 2,3-dioxygenase (TDO) or indoleamine 2,3-dioxygenases (IDO1, IDO2) is the first, rate-limiting step of the kynurenine pathway and may thereby control TRP levels subsequently available for serotonin synthesis [@Lukic2022; @Badawy2017].
 Importantly, IDO1 is stimulated by inflammatory signals [@Badawy2017; @Robinson2003]
-and TDO activity is increased by glucocorticoids released upon depression and mental stress [@Knox1951; @Maes1990]. 
-Kynurenine pathway metabolites such as KYN or quinolinic acid [QUIN] can additionally pass the blood-brain barrier, interfere e.g.
-glutamatergic signaling and exert neurotoxicity [@Brown2021; @Schwarcz2012].
-Hence, the kynurenine pathway, whose activity can be assessed by the ratio of blood KYN to TRP (KYN/TRP), poses a connex of inflammatory and hypothalamic–pituitary–adrenal axis signals, systemic serotonin availability and neural signal transmission [@Badawy2017; @Brown2021; @Schwarcz2012].
-Consequently, elevated levels of circulating kynurenine pathway metabolites, high KYN/TRP, decreased TRP or lowered serotonin were reported for numerous physical conditions and associated with symptoms of mental disorders [@Cervenka2017; @Marx2020; @Hunt2020; @Fellendorf2022; @Hufner2015; @Hufner2021; @Hufner2019; @Capuron2011].
+and TDO activity is increased by glucocorticoids which are also released as a reaction to mental stress [@Knox1951; @Maes1990]. 
+Hence, activity of IDO and TDO, which can be assessed by the ratio of blood KYN to TRP (KYN/TRP), is modulated by inflammatory and hypothalamic–pituitary–adrenal axis signals.
+Additionally, kynurenine pathway metabolites such as KYN or quinolinic acid [QUIN] can additionally pass the blood-brain barrier, interfere e.g.
+glutamatergic signaling and exert neurotoxicity [@Badawy2017; @Brown2021; @Schwarcz2012].
+Consequently, elevated concentrations of circulating kynurenine pathway metabolites, high KYN/TRP, decreased TRP as well as lowered levels of associated neurotransmitter serotonin were reported for numerous physical disorders and associated with symptoms of mental disorders [@Cervenka2017; @Marx2020; @Hunt2020; @Fellendorf2022; @Hufner2015; @Hufner2021; @Hufner2019; @Capuron2011].
 
-Dopamine synthesis via the catecholamine pathway involves the conversion of phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH).
-PAH, TH and the serotonin pathway enzyme TPH rely on tetrahydrobiopterin (BH4) as an essential cofactor [@Meiser2013; @Neurauter2008]. 
+The synthesis of catecholamine neurotransmitters such has dopamine via the catecholamine pathway involves the conversion of phenylalanine (PHE) to tyrosine (TYR) catalyzed by phenylalanine hydroxylase (PAH) followed by hydroxylation of TYR by tyrosine 5-hydroxylase (TH).
+PAH, TH and the serotonin-synthesizing TPH enzyme rely on tetrahydrobiopterin (BH4) as an essential cofactor [@Meiser2013; @Neurauter2008]. 
 During inflammation, BH4 is oxidized to its inactive derivative and consumed in a competitor reaction of nitric oxide formation [@Rahimian2022a; @Neurauter2008]. 
-As a result of such functional BH4 deficiency, reflected e.g. by an elevated blood PHE to TYR ratio, biosynthesis of dopamine and serotonin is suppressed [@Geisler2013; @Vancassel2018; @Capuron2011]. 
+As a result of such functional BH4 deficiency, reflected e.g. by an elevated blood PHE to TYR ratio, biosynthesis of dopamine and serotonin is reduced [@Geisler2013; @Vancassel2018; @Capuron2011]. 
 Accordingly, changes in PHE and TYR levels have been reported in different physical disorders such as cancer, infections and inflammatory conditions and associated with depression and anxiety [@Vancassel2018; @Geisler2013; @Hufner2015; @Capuron2011].
@@
 SARS-CoV-2 virus is the causal pathogen of coronavirus disease 2019 (COVID-19). 
-Apart from sustained physical disability, psychiatric disorders in millions of patients amount to the persistent burden of the COVID-19 pandemic [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Staudt2022].
-Increased kynurenine pathway activity suggestive of lowered systemic serotonin availability has recently been identified in acute COVID-19 and during COVID-19 recovery and associated with infection severity [@Santiago-Mujika2022; @Lionetto2021; @Bizjak2022; @Dewulf2022].
-Theoretical considerations [@Bower2022] and preliminary experimental data suggest that highly active kynurenine pathway along with suppression of serotonin and catecholamine biosynthesis may contribute to symptoms of mental stress, anxiety and depression in COVID-19 patients [@Matits2023; @Kucukkarapinar2022].
+Apart from sustained physical disability, mental disorders in millions of patients amount to the persistent burden of the COVID-19 pandemic [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023; @Staudt2022].
+Increased kynurenine pathway activity has recently been identified in acute COVID-19 and during COVID-19 recovery and associated with infection severity [@Santiago-Mujika2022; @Lionetto2021; @Bizjak2022; @Dewulf2022].
+Theoretical considerations [@Bower2022] and preliminary experimental data suggest that a highly active kynurenine pathway as well as reduced biosynthesis of serotonin and catecholamine neurotransmitters may contribute to symptoms of mental stress, anxiety and depression in COVID-19 patients [@Matits2023; @Kucukkarapinar2022].
 
-Still, clinical and experimental evidence for interaction of inflammation, mental health and systemic availability of serotonin and dopamine in COVID-19 is scarce. 
-To address this question, we explored effects of demographic, clinical, psychometric (anxiety, depression, mental stress), inflammation- and SARS-CoV-2-related factors on markers of systemic activity of the serotonin, kynurenine and catecholamine pathways in two cross-sectional cohorts of SARS-CoV-2 patients: the SIMMUN cohort and the published longitudinal INCOV study [@Su2022].
+Clinical and experimental evidence for an interaction of mental health with inflammation, kynurenine and catecholamine pathway activity and systemic availability of serotonin and dopamine in COVID-19 is scarce.
+To address this question, we analyzed the effects of demographic, clinical, psychometric (anxiety, depression, mental stress), inflammation- and SARS-CoV-2-related factors on the kynurenine pathway activity, serotonin levels as well as the catecholamine pathway in two cohorts of SARS-CoV-2 patients: the cross-sectional SIMMUN cohort and the published longitudinal INCOV cohort [@Su2022].
 
@@
 
-The SIMMUN data were collected during a single on-site visit at median 139 days after the SARS-CoV-2 PCR (interquartile range: 119 - 157) or extracted from electronic patient records (__Supplementary Table S1__).
-Demographic and clinical characteristic included age, sex, body mass index (BMI), mental disorders diagnosed by a physician, self-reported physical disorders, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR.
-Inflammatory markers, plasma neopterin concentrations (NEO) measured by ELISA [@Werner1990] and neutrophil/lymphocyte ratio (NLR) were determined by a certified laboratory at the University Hospital of Innsbruck.
-Plasma TRP, KYN, PHE and TYR were determined by high-performance liquid chromatography [@Widner1997; @Neurauter2008]. 
+The SIMMUN data were collected during a single on-site visit at median 139 days after the SARS-CoV-2 PCR (interquartile range: 119 - 157) and extracted from electronic patient records (__Supplementary Table S1__).
+Demographic and clinical characteristics included age, sex, body mass index (BMI), mental disorders diagnosed by a physician, self-reported physical disorders, smoking and alcohol consumption, result and date of the SARS-CoV-2 PCR.
+Neutrophil/lymphocyte ratio (NLR) and serum concentration of neopterin (NEO) served as inflammatory markers. 
+NLR was determined by a certified laboratory at the University Hospital of Innsbruck. 
+NEO was measured measured by enzyme-linked immunosorbent assay (BRAHMS Diagnostics, Berlin, Germany). 
+Serum TRP, KYN, PHE and TYR were determined by high-performance liquid chromatography [@Widner1997; @Neurauter2008]. 
 Immunoglobulin gamma against receptor binding domain S1/S2 protein (anti-RBD IgG) was quantified by ELISA [@Deisenhammer2021]. 
@@
 Plasma proteomes and metabolomes in the INCOV cohort were measured by proximity extension assay (Olink, Sweden) and liquid chromatography/tandem mass spectrometry (Metabolon, USA), respectively, and expressed as age- and sex-adjusted, log~2~-transformed concentrations [@Su2022; @Su2020].
-Markers of the serotonin (serotonin, TRP) [@Lukic2022], kynurenine (KYN, kynurenine/tryptophan ratio [KYN/TRP], QUIN) [@Badawy2017] and catecholamine pathway activity (PHE, TYR, phenylalanine/tyrosine ratio [PHE/TYR] and dopamine 3-O-sulfate [DA sulfate]) [@Meiser2013; @Goldstein1999; @Suominen2013], cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]), age, sex and BMI were analyzed (__Supplementary Table S3__).
+Serotonin, metabolites of the kynurenine (TRP, KYN, QUIN) [@Lukic2022; @Badawy2017] and catecholamine pathway (PHE, TYR and dopamine 3-O-sulfate [DA sulfate]) [@Meiser2013; @Goldstein1999; @Suominen2013], cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]), age, sex and BMI were analyzed (__Supplementary Table S3__).
 
-Plasma samples were collected in during in acute and sub-acute infection, and recovery at median 
+Plasma samples were collected during acute and sub-acute infection, and recovery at median 
 `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']`, 
-`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']`), 
+`r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['sub-acute']`, 
 `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['recovery']` days after diagnosis of SARS-CoV-2 infection via PCR, respectively (__Supplementary Table S4__).
@@
 
-The first endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing plasma markers of kynurenine and catecholamine pathway activity (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR) by multi-parameter modeling in the SIMMUN cohort.
-The second endpoint was to investigate how markers of systemic serotonin and dopamine availability (DA sulfate), are influenced by age, sex, body mass, inflammatory cytokines, timepoint of SARS-CoV-2 infection, neurotransmitter precursor amino acids (TRP, PHE, TYR) and kynurenine pathway metabolites (KYN, QUIN).
+The first endpoint was to determine demographic, clinical, psychometric, inflammation- and SARS-CoV-2-related factors influencing serum markers of kynurenine and catecholamine pathway activity (TRP, KYN, KYN/TRP, PHE, TYR, PHE/TYR) by multi-parameter modeling in the SIMMUN cohort.
+The second endpoint was to investigate how systemic serotonin and DA dopamine availability (DA sulfate), are influenced by age, sex, BMI, cytokine markers of inflammation, timepoint of SARS-CoV-2 infection, kynurenine pathway (TRP, KYN, QUIN) and catecholamine metabolite levels (PHE, TYR).
 This endpoint was addressed by multi-parameter robust linear modeling, time course modeling and correlation analysis in the INCOV cohort (__Supplementary Figure S1__).
@@
 
-Responses and explanatory variables were normalized prior to modeling. 
-In the SIMMUN cohort, effects of age, sex, body mass class, physical and psychiatric disorders, BMI, smoking and alcohol consumption, inflammation markers (NEO, NLR), SARS-CoV-2 infection, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress (PSS-4) on pathway activity markers were assessed by multi-parameter linear regression with backward elimination.
-In the INCOV cohort, effects of age, sex, BMI, cytokine markers of inflammation (IL6, IL10, TNF, IFNG), SARS-CoV-2 infection timepoint (uninfected, acute, sub-acute, recovery), serotonin and catecholamine pathway metabolites (TRP, PHE, TYR) and kynurenine pathway metabolites (QUIN, KYN) on plasma serotonin and DA sulfate were modeled by multi-parameter robust linear regression [@Huber2011].
+Responses and numeric explanatory variables were normalized prior to modeling. 
+In the SIMMUN cohort, effects of age, sex, body mass class, physical and psychiatric disorders, BMI, smoking and alcohol consumption, inflammation markers (NEO, NLR), SARS-CoV-2 infection, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress (PSS-4) on kynurenine and catecholamine pathway activity markers were assessed by multi-parameter linear regression with backward elimination.
+In the INCOV cohort, effects of age, sex, BMI, cytokine markers of inflammation (IL6, IL10, TNF, IFNG), SARS-CoV-2 infection status and timepoint (uninfected, acute, sub-acute, recovery), kynurenine (TRP, KYN, QUIN) and catecholamine pathway metabolites (PHE, TYR) on plasma serotonin and DA sulfate were modeled by multi-parameter robust linear regression [@Huber2011].
 Explanatory performance, reproducibility and proper parameterization of the multi-parameter models was investigated by R^2^ and root mean square error (RMSE) in 10-fold cross-validation. 
@@
 `r nrow(stigma$analysis_tbl)` participants with complete study variables were analyzed (__Supplementary Figure S1__, __Supplementary Table S1__). 
-The excluded individuals were characterized by more frequent mental disorders and clinically relevant symptoms of depression and anxiety (HADS), higher scores of mental stress (PSS-4) and less frequent SARS-CoV-2 infections than the analyzed participants (__Supplementary Table S2__).
+The excluded individuals were characterized by more frequent mental disorders and more frequent clinically relevant symptoms of depression and anxiety (HADS), higher scores of mental stress (PSS-4) and less frequent SARS-CoV-2 infections than the analyzed participants (__Supplementary Table S2__).
 In the analyzed SIMMUN cohort, males represented 
@@
 `r cohort$clin_test$stigma %>% filter(variable == 'psych_comorb') %>% .$eff_size`) (__Table \@ref(tab:tab-stigma-cohort)__). 
-The psychometric tools for assessment of perceived mental stress (PSS-4), depression and anxiety (HADS) displayed good-to-excellent internal consistency 
+The psychometric tools for assessment of perceived mental stress (PSS-4), and depression and anxiety (HADS) displayed good-to-excellent internal consistency 
 ($\omega$ = `r cons$omega_stats %>% filter(factor == 'g') %>% .$total %>% range %>% signif(2) %>% paste(collapse = ' - ')`) [@McDonald1999] (__Supplementary Figure S2__). 
@@
 (`r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% my_percent('hads_dpr_score') %>% .[2]`%) than in SARS-CoV-2-positive participants 
-(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('hads_dpr_score') %>% .[2]`%), yet this effect was no statistically significant. 
+(`r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% my_percent('hads_dpr_score') %>% .[2]`%), yet this effect was not statistically significant. 
 Scores of mental stress were comparable in both SARS-CoV-2 subsets 
@@
 positive: median `r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% .$pss_stress_score %>% median` points). 
-As expected, titer of antibodies against the S1/S2 SARS-CoV-2 protein (anti-RBD IgG) was significantly higher in the SARS-CoV-2-positive subset 
-(median: `r cohort$analysis_tbl$stigma %>% filter(cov == 'SARS-CoV-2') %>% .$anti_rbd %>% median %>% signif(2)` AU) as compared with uninfected individuals 
-(median: `r cohort$analysis_tbl$stigma %>% filter(cov == 'healthy') %>% .$anti_rbd %>% median %>% signif(2)` AU, 
-`r cohort$clin_test$stigma %>% filter(variable == 'anti_rbd') %>% .$significance`, 
-effect size: `r cohort$clin_test$stigma %>% filter(variable == 'anti_rbd') %>% .$eff_size`). 
 In 73% of SARS-CoV-2-infected SIMMUN study participants, the infection was mild and treated on an outpatient basis (__Table \@ref(tab:tab-stigma-cohort)__).
@@
 In the SIMMUN dataset, we initially searched for predictors of activity of the kynurenine and catecholamine pathways. 
-TRP, KYN and KYN/TRP served as markers of the kynurenine pathway [@Badawy2017], and PHE, TYR and PHE/TYR were investigated as markers of the catecholamine pathway activity [@Meiser2013]. 
+TRP, KYN and KYN/TRP served as markers of the kynurenine pathway, and PHE, TYR and PHE/TYR were investigated as markers of the catecholamine pathway activity. 
 The candidate explanatory variables were age, sex, BMI, the presence of physical and mental disorders, smoking, alcohol consumption, SARS-CoV-2 infection, anti-SARS-CoV-2 antibody titer, NEO and NLR as inflammatory markers, perceived mental stress, and symptoms of depression and anxiety (__Supplementary Table S1__).
@@
 The cross-validated R^2^ metric values for the TRP, TYR and PHE/TYR models ranged from 
-`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[1]`% to 
+`r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[1]` to 
 `r (mod_eli$cv_stats %>% filter(data_type == 'cv', statistic == 'rsq', response %in% c('trp', 'tyr', 'phe_tyr')) %>% .$estimate %>% range %>% signif(2))[2]` indicative of weak-to-moderate explanatory value (__Supplementary Figure S3__).
@@
 ($\beta$ = `r my_beta(mod_eli$inference$trp, variable = 'log_neo')`) 
-and clinically relevant depression symptoms 
-(HADS $\geq$ 8, $\beta$ = `r my_beta(mod_eli$inference$trp, variable = 'hads_dpr_score')`) were independently associated with reduced plasma TRP. 
-KYN and KYN/TRP were significantly associated with 
+and clinically relevant symptoms of depression 
+(HADS depression $\geq$ 8, $\beta$ = `r my_beta(mod_eli$inference$trp, variable = 'hads_dpr_score')`) were independently associated with reduced serum TRP. 
+KYN and KYN/TRP were significantly and positively associated with 
 SARS-CoV-2 infection (KYN: $\beta$ = `r my_beta(mod_eli$inference$kyn, variable = 'infection')`, 
@@
 
-Concerning the catecholamine pathway, plasma TYR concentrations were negatively associated with NEO 
+Regarding the catecholamine pathway, serum TYR concentrations were negatively associated with NEO 
 ($\beta$ = `r my_beta(mod_eli$inference$tyr, variable = 'log_neo')`) and positively with age 
 ($\beta$ = `r my_beta(mod_eli$inference$tyr, variable = 'age')`). 
-SARS-CoV infection ($\beta$ = `r my_beta(mod_eli$inference$phe_tyr, variable = 'infection')`) 
+SARS-CoV-2 infection ($\beta$ = `r my_beta(mod_eli$inference$phe_tyr, variable = 'infection')`) 
 and age ($\beta$ = `r my_beta(mod_eli$inference$phe_tyr, variable = 'age')`) were significant predictors of lower PHE/TYR (__Figure \@ref(fig:fig-modeling)B__).
@@
 
-Next, we investigated factors affecting plasma levels of serotonin [@Lukic2022] and a major circulating catabolite of dopamine, DA sulfate [@Meiser2013; @Goldstein1999], as direct markers of the serotonin and catecholamine pathway activity, respectively.
-This analysis was done in the INCOV cohort [@Su2022] with age, sex, BMI, cytokines markers of inflammation (IL6, IL10, TNF, IFNG), timepoint of SARS-CoV-2 infection, concentrations of serotonin and catecholamine pathway metabolites (TRP, PHE, TYR), and kynurenine pathway metabolite levels (KYN, QUIN) as candidate explanatory variables (__Supplementary Table S3__).
+Next, we investigated factors affecting plasma levels of serotonin and a major circulating catabolite of dopamine, DA sulfate,  as direct markers of systemic serotonin and dopamine availability, respectively.
+This analysis was done in the INCOV cohort with age, sex, BMI, markers of inflammation (IL6, IL10, TNF, IFNG), status and timepoint of SARS-CoV-2 infection, concentrations of kynurenine (KYN, QUIN) and catecholamine pathway metabolites (TRP, PHE, TYR) as candidate explanatory variables (__Supplementary Table S3__).
 Metabolites and cytokines were sampled in SARS-CoV-2-positive participants during acute and sub-acute infection, and recovery, at median 
@@
 
-The multi-parameter robust regression model [@Huber2011] for plasma serotonin was characterized by moderate explanatory performance 
+The multi-parameter robust regression model for plasma serotonin was characterized by moderate explanatory performance 
 (cross-validated R^2^ = `r incov_neuro$cv_stats %>% filter(data_type == 'cv', response == 'serotonin', statistic == 'rsq') %>% .$estimate %>% signif(2)`). 
@@
 
-In the INCOV collective, 
+In the INCOV cohort, 
 `r incov_neuro$inference$serotonin %>% filter(variable != 'Intercept', p_value < 0.05) %>% filter(!duplicated(variable)) %>% nrow %>% int_to_text` significant predictors of plasma serotonin were identified.
@@
 TRP ($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'tryptophan')`) and recovery from SARS-CoV-2 infection 
-($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'timepoint', 'recovery')` versus uninfected controls) were predictors of higher plasma serotonin. 
+($\beta$ = `r my_beta(incov_neuro$inference$serotonin, 'timepoint', 'recovery')` versus uninfected) were predictors of higher plasma serotonin. 
 Significantly higher DA sulfate levels were identified for recovery from SARS-CoV-2 infection 
-($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'timepoint', 'recovery')` versus uninfected control) 
+($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'timepoint', 'recovery')` versus uninfected) 
 and high TNF concentrations ($\beta$ = `r my_beta(incov_neuro$inference$dopamine, 'TNF_INF')`). 
@@
 The maximum plasma concentrations of inflammation markers IL6, IL10, TNF and IFNG were observed during acute infection. 
-They were paralleled by changes in metabolites of the serotonin and kynurenine pathways with a significantly decreased 
+They were paralleled by changes in serotonin and metabolites of the kynurenine pathways with a significantly decreased 
 TRP ($\beta$ = `r my_beta(time_rlm$inference$uninfected$tryptophan, variables = 'timepoint', levels = 'acute')`), increased 
@@
 elevated QUIN ($\beta$ = `r my_beta(time_rlm$inference$uninfected$quinolinate, variables = 'timepoint', levels = 'acute')`) during acute infection as compared with uninfected individuals.
-The inflammatory milieu of acute infection was also associated with significantly increased levels of the catecholamine pathway substrate PHE ($\beta$ = `r my_beta(time_rlm$inference$uninfected$phenylalanine, variables = 'timepoint', levels = 'acute')` versus uninfected controls).
+The inflammatory milieu of acute infection was also associated with significantly increased levels of the catecholamine pathway substrate PHE ($\beta$ = `r my_beta(time_rlm$inference$uninfected$phenylalanine, variables = 'timepoint', levels = 'acute')` versus uninfected).
 Resolution of systemic inflammation during sub-acute SARS-CoV-2 infection and 
@@
 PHE (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$phenylalanine, 'timepoint', 'sub-acute')`, 
-recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$phenylalanine, 'timepoint', 'recovery')`) and elevated 
+recovery: $\beta$ = `r my_beta(time_rlm$inference$acute$phenylalanine, 'timepoint', 'recovery')`) and increasing 
 DA sulfate (sub-acute: $\beta$ = `r my_beta(time_rlm$inference$acute$dopamine.3.O.sulfate, 'timepoint', 'sub-acute')`, 
@@
 acute ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'tryptophan', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) and 
-sub-acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'tryptophan', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) with weak-to-moderate and moderate-to-large effect size.
+sub-acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'tryptophan', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) with weak-to-moderate and moderate-to-large effect sizes.
 Plasma serotonin levels correlated negatively with IL6 concentrations during acute infection 
@@
 QUIN (acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.acute']]), 'quinolinate', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`; 
-sub-acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'quinolinate', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) correlated positively with circulating IL6, IL10, TNF and IFNG with a moderate-to-large strength.
+sub-acute, $\rho$: `r my_rho(t(incov_mds$analysis_tbl[['serotonin.sub-acute']]), 'quinolinate', c('IL6_INF', 'IL10_INF', 'TNF_INF', 'IFNG_INF')) %>% range %>% paste(collapse = ' - ')`) correlated positively with circulating IL6, IL10, TNF and IFNG with a moderate-to-large effect size.
 Serotonin levels correlated with TRP during 
@@
 
-Concerning the catecholamine pathway metabolites (__Figure \@ref(fig:fig-incov-corr)B__, __Supplementary Figure S6__), positive weak-to-moderate correlations of PHE with IL6, IL10 and TNF were observed in acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['dopamine.acute']]), 'phenylalanine', c('IL6_INF', 'IL10_INF', 'TNF_INF')) %>% range %>% paste(collapse = ' - ')`).
+Concerning the catecholamine pathway metabolites (__Figure \@ref(fig:fig-incov-corr)B__, __Supplementary Figure S6__), positive correlations of PHE with IL6, IL10 and TNF were observed in acute infection ($\rho$: `r my_rho(t(incov_mds$analysis_tbl[['dopamine.acute']]), 'phenylalanine', c('IL6_INF', 'IL10_INF', 'TNF_INF')) %>% range %>% paste(collapse = ' - ')`, effect size: weak-to-moderate).
 PHE was weakly positively associated with IL10 in sub-acute SARS-CoV-2 infection 
@@
 
-In two separate cohorts, we investigated factors affecting the key metabolic pathways determining systemic availability of serotonin and dopamine in SARS-CoV-2 infection. 
-Our analysis underlines the pivotal role of SARS-CoV-2-dependent and -independent inflammation on the serotonin and kynurenine pathway activity. 
-In particular, levels of the kynurenine pathway markers, KYN, KYN/TRP and QUIN, were higher, whereas concentration of the serotonin precursor TRP was lower, in the presence of inflammation reflected by elevated NEO or IL6.
-This observation and the significant association of serotonin with TRP levels in the INCOV cohort suggest collectively that depletion of TRP via highly active kynurenine pathway can limit systemic serotonin availability during SARS-CoV-2 infection [@Badawy2017].
-Age, male sex, mental stress and depression were proposed as infection-independent predictors of high kynurenine pathway activity, lowered TRP and, consequently, reduced serotonin formation. 
-Our data suggest similar inhibitory effects of inflammation on the first step of the catecholamine pathway, PHE - TYR conversion, and a reduced systemic dopamine availability reflected by lowered circulating DA sulfate [@Meiser2013].
+We investigated factors affecting the kynurenine pathway activity, serotonin availability as well as catecholamine pathway activity in SARS-CoV-2 infection in two separate cohorts. 
+Our analysis underlines the pivotal role of SARS-CoV-2-dependent and independent inflammation in systemic serotonin and catecholamine metabolism. 
+Combining the results from both cohorts we report higher levels of the kynurenine pathway markers, KYN, KYN/TRP and QUIN, and lower concentration of the serotonin precursor TRP, in the presence of inflammation reflected by NEO or IL6.
+This observation and the significant association of serotonin with TRP levels in the INCOV cohort suggest collectively that depletion of TRP via highly active kynurenine pathway could limit systemic serotonin availability during SARS-CoV-2 infection [@Badawy2017].
+Age, mental stress and depression were proposed as infection-independent predictors of high kynurenine pathway activity, lowered TRP and, consequently, reduced serotonin synthesis. 
+Our data suggest inhibitory effects of inflammation on the first step of the catecholamine pathway, PHE - TYR conversion, and a reduced systemic dopamine availability reflected by lowered circulating DA sulfate [@Meiser2013].
 
-Few discrepancies need a more thorough discussion. 
-First, explanatory performance of analyses of the serotonin and kynurenine pathway activity measured by R^2^ and correlation coefficients were generally higher than explanatory values of the catecholamine pathway analyses.
-This indicates that the predictors identified by our analysis, inflammation, SARS-CoV-2-positivity and infection timepoint impact the TRP metabolic pathways to a larger degree than catecholamine metabolism, in line with a metabolomic study in COVID-19 individuals [@Thomas2020].
-Second, effects of inflammation on markers of the kynurenine and catecholamine pathway activity were stronger in the INCOV than in the SIMMUN cohort, which may reflect more severe infection in INCOV study participants.
-Third, in the SIMMUN cohort, the inflammatory marker NLR could be unexpectedly linked to lower KYN, which  may be explained by a sustained low-grade neutropenia reported for up to 1 year after COVID-19 [@Lin2022].
+Overall we note that explanatory performance of the analyses of serotonin availability and kynurenine pathway activity measured by R^2^ and correlation coefficients were generally higher than explanatory values of the catecholamine pathway analyses.
+This indicates that the predictors identified by our analysis, inflammation, SARS-CoV-2-positivity, and infection timepoint affect kynurenine pathway activity to a larger degree than catecholamine catecholamine metabolism, in line with a recent metabolomic study in COVID-19 individuals [@Thomas2020].
+Effects of inflammation on markers of the kynurenine and catecholamine pathway activity were stronger in the INCOV than in the SIMMUN cohort, which may reflect more severe infection in INCOV study participants.
+In the SIMMUN cohort, the inflammatory marker NLR was unexpectedly linked to lower KYN, which  may be explained by a sustained low-grade neutropenia reported for up to 1 year after COVID-19 [@Lin2022].
 
-## Effects of inflammation and infection on the serotonin and kynurenine pathways, and systemic levels of serotonin
+## Effects of inflammation and infection on the kynurenine pathways and systemic levels of serotonin
 
-IDO1, together with TDO and IDO2, catalyzes the first reaction of the kynurenine pathway, the TRP - KYN conversion, which catabolizes >90% of TRP and may limit serotonin availability [@Badawy2017; @Lukic2022].
-Expression and activity of IDO1 are induced by inflammatory stimuli such as IL6, TNF and IFNG [@Robinson2003; @Badawy2017].
-Consequently, elevated metabolite levels and activity readouts of the kynurenine pathway were reported for multiple inflammatory conditions [@Badawy2017].
-In COVID-19, levels of metabolites of the kynurenine pathway in blood and urine were found to correlate positively with inflammatory markers [@Dewulf2022; @Santiago-Mujika2022], disease severity [@Dewulf2022; @Lionetto2021; @Ceballos2022] and was implicated in persistent COVID-19 symptom [@Bizjak2022; @Matits2023; @Cysique2022].
-These findings were summarized in a recent metaanalysis demonstrating increased KYN/TRP in COVID-19 and particularly in its severe manifestations [@Almulla2022]. 
+IDO1, together with TDO and IDO2, catalyzes the first reaction of the kynurenine pathway, the TRP - KYN conversion, which catabolizes >90% of TRP and might thereby limit serotonin availability [@Badawy2017; @Lukic2022].
+Expression and activity of IDO1 is induced by inflammatory stimuli such as IL6, TNF and IFNG [@Robinson2003; @Badawy2017].
+Consequently, increased kynurenine pathway activity was reported for multiple inflammatory conditions [@Badawy2017].
+In COVID-19, levels of kynurenine pathway metabolites in blood and urine were found to correlate positively with inflammatory markers [@Dewulf2022; @Santiago-Mujika2022], disease severity [@Dewulf2022; @Lionetto2021; @Ceballos2022] and were implicated in persistent COVID-19 symptoms [@Bizjak2022; @Matits2023; @Cysique2022].
+These findings were summarized in a recent metaanalysis demonstrating an increased KYN/TRP ratio in COVID-19, particularly in severe manifestations [@Almulla2022]. 
 Using high-throughput metabolomics, tryptophan metabolism via serotonin and kynurenine pathways was identified as the most prominently affected metabolic system in acute COVID-19. 
 In this study, TRP and serotonin were demonstrated to decrease and KYN was shown to increase in an IL6-dependent way [@Thomas2020]. 
-Of note, IL6 was also identified by us as a predictor of lower plasma serotonin in the INCOV study. 
+Of note, IL6 was also identified in the present analysis by us as a predictor of lower plasma serotonin in the INCOV cohort. 
 We demonstrated significant, positive associations of serotonin and TRP levels in acute and sub-acute infection in the INCOV cohort. 
@@
 In turn, increasing TRP and serotonin during infection recovery in the INCOV cohort may reflect decreasing IDO1 activity and re-routing of TRP to the serotonin biosynthesis [@Lukic2022; @Badawy2017]. 
-The multi-parameter modeling in the SIMMUN cohort revealed an additional, inflammation-independent effect of SARS-CoV-2-positivity at median 139 days after infection on KYN and KYN/TRP. 
-While its mechanism remains unclear, an analogical sustained increase in KYN in absence of the inflammatory marker C-reactive protein was described in long-term COVID-19 recovery [@Bizjak2022]. 
-Age was identified by us as another predictor of high kynurenine pathway activity and low serotonin availability, which may be a result of a chronic low-grade inflammation observed in elderly [@Capuron2011; @Sorgdrager2019].
-Furthermore, significantly lower levels of serotonin were observed in male participants of the INCOV study. 
+The multi-parameter modeling in the SIMMUN cohort revealed an additional, inflammation-independent effect of SARS-CoV-2-positivity at a median of 139 days after infection on KYN and KYN/TRP. 
+While its mechanism remains unclear, an analogical sustained increase in KYN in the absence of the inflammatory marker C-reactive protein was described in long-term COVID-19 recovery [@Bizjak2022]. 
+Age was identified in the present analysis as another predictor of high kynurenine pathway activity and also low serotonin availability, which may be a result of a chronic low-grade inflammation observed in elderly [@Capuron2011; @Sorgdrager2019].
+Furthermore, significantly lower levels of serotonin were observed in male participants of the INCOV cohort. 
 This may reflect a generally stronger inflammatory response to SARS-CoV-2 in males [@Ceballos2022], which was also evident in our data (INCOV, IL6: $\beta$ = `r my_beta(gender_incov$inference$IL6_INF, 'sex', 'male')`, male versus female, robust linear modeling).
 
-## Effects of inflammation and infection on the serotonin and kynurenine pathways, and systemic levels of serotonin
+## Effects of inflammation and infection on the catecholamine pathway and systemic dopamine availability
 
-BH4 poses, another junction between neurotransmitter metabolism and inflammation. 
+BH4 poses another junction between neurotransmitter metabolism and inflammation. 
 Activity of GTP cyclohydrolase I, the enzyme catalyzing the rate-limiting step of BH4 synthesis, is strongly stimulated by IFNG and other inflammatory stimuli. 
-However, during inflammation, BH4 availability for catecholamine and serotonin pathway enzymes is limited due to oxidation and depletion by nitric oxide synthesis [@Geisler2013; @Meiser2013; @Neurauter2008].
-As previously reported [@Luporini2021], we could demonstrate an increased plasma PHE during acute and sub-acute infection in the INCOV cohort. 
-We could also correlate readouts of suppressed catecholamine pathway activity such as increased PHE and PHE/TYR, and low TYR and DA sulfate levels with inflammatory markers NEO, IL6, TNF and IFNG. 
-Collectively, these observations are consistent with the inflammatory BH4 deficiency leading to suppressed PHE - TYR conversion, hyperphenylalaninemia and impaired catecholamine biosynthesis during acute and sub-acute SARS-CoV-2 infection.
+However, during inflammation, BH4 availability for catecholamine and serotonin synthesis enzymes is reduced by oxidation and nitric oxide synthesis [@Geisler2013; @Meiser2013; @Neurauter2008]. 
+As previously reported [@Luporini2021], we we observed a higher plasma PHE during acute and sub-acute infection in the INCOV cohort. 
+We also found an inverse association of readouts of catecholamine pathway activity such as PHE and PHE/TYR (increased), and TYR and DA sulfate levels (decreased) with inflammatory markers NEO, IL6, TNF and IFNG.
+Taken together, these observations are consistent with the inflammatory BH4 deficiency leading to a reduced PHE - TYR conversion and impaired catecholamine biosynthesis during acute and sub-acute SARS-CoV-2 infection.
 By contrast, the increased DA sulfate during infection recovery in the INCOV cohort and the significantly lower PHE/TYR at median 139 days after SARS-CoV-2 infection in the SIMMUN cohort may indicate a restored BH4 homeostasis and efficient dopamine synthesis [@Meiser2013; @Goldstein1999; @Geisler2013; @Neurauter2008].
@@
 higher blood levels of TYR and lower PHE/TYR in older participants of the SIMMUN study. 
-However, since we could not corroborate this effect in the INCOV cohort, its reproducibility remains questionable.
-Analogically, although low DA sulfate levels were associated with obesity in the INCOV cohort, we could no observe any significant effects of obesity on markers of the catecholamine pathway activity in the SIMMUN cohort.
+However, since we could not replicate this effect in the INCOV cohort, its relevance remains questionable.
+Analogically, although low DA sulfate levels were associated with obesity in the INCOV cohort, we could not observe any significant effects of obesity on markers of catecholamine pathway activity in the SIMMUN cohort.
 
-## Interaction of mental health with activity of the serotonin, kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability
+## Interaction of mental health with activity of the kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability
 
-Peripheral and central nervous system inflammation was put forward as a neuroimmune mechanism contributing to mental health disorder symptoms such as depression, anxiety and mental stress [@Brown2021; @Schwarcz2012; @Vancassel2018].
-Sickness behavior accompanying e.g. infections and inflammatory conditions comprises symptoms of depression and anxiety, social disconnection, fatigue, cognitive disturbance, and psychomotor slowing. 
-Mechanistically, sickness behavior was postulated to involve kynurenine pathway along with suppression of serotonin and catecholamine metabolism mediated by acute or protracted low grade inflammation [@Dantzer2008; @Maes2012; @Vancassel2018].
-Additionally, kynurenine pathway metabolites were ascribed standalone activity in neuronal signaling, e.g.
-by interfering with glutamatergic receptors [@Brown2021; @Schwarcz2012].
-Accordingly, markers of the kynurenine pathway activity and suppressed serotonin pathway such as serotonin, TRP, KYN or KYN/TRP were associated with symptoms of depression in recent metaanalyses [@Hunt2020; @Marx2020; @Fellendorf2022].
-Inflammatory markers were also associated with elevated PHE/TYR indicative of reduced dopamine availability and proposed to contribute to depression in cancer [@Hufner2015] and trauma [@Hufner2019]. 
-The potential role of protracted inflammation and TRP depletion in persistent symptoms and mental disorders in COVID-19 convalescents was proposed in recent hypothesis papers [@Eroglu2021; @Bower2022].
-There is also clinical evidence for concomitantly elevated markers of inflammation, kynurenine pathway activity and lowered TRP availability in COVID-19 patients suffering from persistent symptoms [@Bizjak2022], cognitive impairment [@Cysique2022], or symptoms of anxiety, depression and mental stress [@Kucukkarapinar2022; @Matits2023].
-In the SIMMUN cohort, clinically relevant depression symptoms (HADS) and mental stress (PSS-4) along with the inflammatory marker NEO could be linked to decreased TRP and higher KYN/TRP, respectively.
-In the INCOV collective, inflammation was also identified as a major diver of alterations in circulating serotonin, TRP, as well as the kynurenine pathway metabolites KYN and QUIN. 
-Hence, the additive effects of inflammation, infection, mental stress and age may disrupt the homeostasis of the serotonin, kynurenine and catecholamine pathways and predispose to depressive or anxious disorders in COVID-19 patients [@Dantzer2008; @Bower2022].
+Peripheral and central nervous system inflammation has been identified as a neuroimmune mechanism contributing to mental stress [@Kim2017; @Hufner2020] and mental health disorders such as depression or anxiety [@Brown2021; @Schwarcz2012; @Vancassel2018].
+Mechanistically, sickness behavior which accompanies acute inflammation and shares many similarities to depression, was postulated to involve the kynurenine and catecholamine pathway [@Dantzer2008; @Maes2012; @Vancassel2018].
+Additionally, kynurenine pathway metabolites can contribute to the sickness behavior and depression via interference with neuronal signaling, e.g.
+with glutamatergic receptors [@Brown2021; @Schwarcz2012].
+Markers of kynurenine pathway activity such as TRP, KYN or KYN/TRP were associated with symptoms of depression in recent metaanalyses [@Hunt2020; @Marx2020; @Fellendorf2022]. 
+Inflammatory markers were also associated with elevated PHE/TYR, which may indicate a reduced dopamine availability, and proposed to contribute to depression in cancer [@Hufner2015] and trauma [@Hufner2019].
+The potential role of inflammation and TRP depletion in persistent somatic symptoms and mental disorders in COVID-19 convalescents was proposed in recent hypothesis papers [@Eroglu2021; @Bower2022]. 
+There is first clinical evidence for concomitantly elevated markers of inflammation, kynurenine pathway activity and lowered TRP availability in COVID-19 patients suffering from persistent symptoms [@Bizjak2022], cognitive impairment [@Cysique2022], or symptoms of anxiety, depression and mental stress [@Kucukkarapinar2022; @Matits2023].
+In the SIMMUN cohort, clinically relevant symptoms of depression (HADS) along with the inflammatory marker NEO were associated with decreased TRP. 
+Additionally, mental stress levels (PSS-4) together with NEO were identified as predictors of higher KYN/TRP. 
+In the INCOV cohort, inflammation was also found to affect levels of circulating serotonin, TRP, and kynurenine pathway metabolites KYN and QUIN. 
+Hence, the additive effects of inflammation, infection, mental stress and age may disrupt the homeostasis of the kynurenine and catecholamine pathways and predispose to depressive or anxious disorders in COVID-19 patients [@Dantzer2008; @Bower2022].
 
-Still, most evidence for relevance of the peripheral neurotransmitter levels in mental disorders is delivered by observational studies, whereas in vivo experimental reports are scarce [@Brown2021]. 
-In the periphery, liver, mesenteric organs and vasculature are the main sites of tryptophan, serotonin and dopamine metabolism [@Badawy2017; @Lukic2022; @Meiser2013; @Goldstein1999]. 
-Although the neurotransmitter precursors TRP and TYR, and kynurenin pathway metabolites were postulated to pass the blood-brain barrier [@Brown2021; @Schwarcz2012; @Vancassel2018], psychiatric disorders are not consistently paralleled by alterations in dopamine, serotonin, and kynurenine pathway metabolites in the central nervous system [@Miller2008; @Clark2016; @Brown2021].
-Of note, clinical evidence for the serotonin theory of depression was also questioned in a recent systematic review [@Moncrieff2022]. 
-Whether inflammatory and infection-related factors impact on peripheral and brain availability of serotonin and dopamine availability and, hence, contribute to psychiatric disorders frequently observed in COVID-19 patients [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023], needs validation in a robust experimental and prospective setting.
+Most evidence on the relevance of the peripheral neurotransmitter levels in mental disorders is delivered by observational studies, whereas in vivo experimental reports are scarce [@Brown2021]. 
+In the periphery, liver, mesenteric organs and vasculature are the main sites of TRP, serotonin and dopamine metabolism [@Badawy2017; @Lukic2022; @Meiser2013; @Goldstein1999]. 
+Although kynurenin pathway metabolites were postulated to pass the blood-brain barrier [@Brown2021; @Schwarcz2012; @Vancassel2018], mental disorders are not consistently paralleled by uniform and reproducible changes in dopamine, serotonin, and kynurenine pathway metabolites in the central nervous system [@Miller2008; @Clark2016; @Brown2021].
+Clinical evidence for the serotonin theory of depression was also questioned in a recent systematic review [@Moncrieff2022]. 
+Whether there is a causal association of inflammatory and infection-related factors with mental disorders frequently observed in COVID-19 patients [@Huang2021; @Al-Aly2021; @Hufner2022; @Sahanic2023], needs further validation.
 
@@
 Conversely, we could not validate the effects of stress, depression and anxiety in the INCOV cohort due to unavailability of psychometric data. 
-Second, relevance of the systemic serotonin, kynurenine and catecholamine pathway activity for the central nervous system and mental health is controversial, as discussed above. 
-Third, the SIMMUN cohort had a limited sample size and a highly variable diagnosis - sampling interval, which likely diminished effect sizes of the SARS-CoV-2 infection. 
-Fourth, the SIMMUN cohort suffered from a selection bias due to enrichment in psychiatric patients with a high rate of physical and mental disorders. 
-Analogically, hospitalized COVID-19 patients constituted the majority of the INCOV cohort. 
-Hence, none of the cohorts is representative for the entire population. 
+Second, relevance of the systemic kynurenine and catecholamine pathway activity for the central nervous system and mental health is controversial. 
+We give a balanced discussion of this controversy to account for this fact. 
+Third, the SIMMUN cohort had a limited sample size and a highly variable SARS-CoV-2 PCR test - sampling interval, which likely diminished effect sizes of the SARS-CoV-2 infection. 
+Fourth, a selection bias occurred in the SIMMUN cohort due to enrichment in patients with mental disorders. 
+In the INCOV cohort there was a selection bias towards hospitalized COVID-19 patients. Hence, none of the cohorts is representative for the general population. 
 Fifth, both cohorts were recruited during initial phases of the pandemic and do not include vaccinated patients.
 Similarly, the analyzed cohorts were exposed to wild-type-like SARS-CoV-2 variants and do not allow to assess effects of infection with highly transmissible but less virulent omicron pathogens.
-For these reasons, studies with recent, post-pandemic collectives are urgently needed to validate our findings. 
-Finally, the cross-sectional SIMMUN and INCOV cohorts encompassed uninfected controls and SARS-CoV-2 infections ranging from asymptomatic to critical disease.
+For these reasons, validation studies with further cohorts are urgently needed.
 
@@
 
-SARS-CoV-2-dependent and -independent inflammation alters activity of the serotonin, kynurenine and catecholamine pathway which may limit systemic availability of serotonin and dopamine. 
+SARS-CoV-2-dependent and -independent inflammation is associated high activity of the kynurenine pathway and reduced activity of the catecholamine pathway, which may lead to a reduced systemic availability of serotonin and dopamine.
 Those effects can be further amplified by advanced age, mental stress and depression. 
-Further research is needed to explore the mechanistic interplay of SARS-CoV-2 infection, inflammation and mental health parameters. 
-It remains to be investigated, if and how this mechanism contributes to psychiatric disorders following SARS-CoV-2 infection.
+Further research is needed to explore the mechanistic interplay of SARS-CoV-2 infection, inflammation and mental health. 
+It remains to be investigated, if and how this mechanism may contribute to mental disorders following SARS-CoV-2 infection.
 
@@
 
-_Effects of age, timepoint of SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR), plasma levels of metabolites of the serotonin and kynurenine pathways (tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]), plasma levels of metabolites of the catecholamine pathway (phenylalanine [PHE], tyrosine [TYR]), and plasma concentrations of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin (5-hydroxy tryptamine [5-HT]) and dopamine 3-O-sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
+_Effects of age, timepoint of SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR), plasma levels of metabolites of the kynurenine (tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]) and catecholamine pathways (phenylalanine [PHE], tyrosine [TYR]), and plasma concentrations of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin and dopamine 3-O-sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
 
@@
 
-```{r fig-time-ourse, fig.width = figur::convert(figs$time_course, to = 'in')$w, fig.height = figur::convert(figs$time_course, to = 'in')$h, fig.cap = 'TTime course of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.'}
+```{r fig-time-ourse, fig.width = figur::convert(figs$time_course, to = 'in')$w, fig.height = figur::convert(figs$time_course, to = 'in')$h, fig.cap = 'Time course of cytokine markers of inflammation, serotonin, and metabolites of the kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.'}
 
@@
 __Figure \@ref(fig:fig-time-ourse).
-Time course of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.__
+Time course of cytokine markers of inflammation, serotonin, and metabolites of the kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.__
 
-_Time courses of normalized log~2~-transformed plasma concentrations of cytokine makers of inflammation (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF], interferon gamma [INFG]), metabolites of the serotonin and kynurenine pathways (B, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN], serotonin/5-hydroxy tryptamine [5-HT]), and metabolites of the catecholamine pathway (C, phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]).
-Differences between uninfected controls and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after SARS-CoV-2 infection diagnosis via PCR, respectively) were modeled by robust linear regression with the uninfected controls as baseline.
+_Time courses of normalized log~2~-transformed plasma concentrations of cytokine makers of inflammation (A, interleukin-6 [IL6], interleukin-10 [IL-10], tumor necrosis factor-alpha [TNF], interferon gamma [INFG]), serotonin and metabolites of the kynurenine pathway (B, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]), and metabolites of the catecholamine pathway (C, phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]).
+Differences between uninfected individuals and acute, sub-acute SARS-CoV-2 infection and recovery (median 10, 14 and 64 days after SARS-CoV-2 infection diagnosis via PCR, respectively) were modeled by robust linear regression with the uninfected individuals as baseline.
 Mean normalized concentrations with 2 standard errors of the mean (2 $\times$ SEM) are depicted as solid lines with tinted ribbons (plots on the left).
@@
 
-```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of plasma levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways during SARS-CoV-2 infection in the INCOV cohort.'}
+```{r fig-incov-corr, fig.width = figur::convert(figs$incov_corr, to = 'in')$w, fig.height = figur::convert(figs$incov_corr, to = 'in')$h, fig.cap = 'Correlation of plasma levels of cytokine markers of inflammation, serotonin, and metabolites of the kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.'}
 
@@
 __Figure \@ref(fig:fig-incov-corr).
-Correlation of plasma levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.__
+Correlation of plasma levels of cytokine markers of inflammation, serotonin, and metabolites of the kynurenine and catecholamine pathways during SARS-CoV-2 infection and recovery in the INCOV cohort.__
 
-_Plasma levels of metabolites of the kynurenine and serotonin pathways (tryptophan [TRP], serotonin [5-hydroxy tryptamine, 5-HT], kynurenine [KYN], quinolinic acid [QUIN]) and metabolites of the catecholamine pathway (phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]) and cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR).
+_Plasma levels of serotonin and metabolites of the kynurenine pathway (A, tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]) and metabolites of the catecholamine pathway (B, phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]), and cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF], interferon-gamma [INFG]) were subjected to pairwise correlation analysis in SARS-CoV-2-infected individuals (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR).
 Correlation coefficient matrices for correlation coefficients $\rho$ > 0.2 were visualized as force-directed network plots.

### 2023-06-11 paper/markdown/supplementary_material.Rmd
Feedback from 08 - 09.06.2023
Changes in headings and labels in Figures, Tables, Supplementary Figures and Supplementary Tables. Changes in the manuscript and supplement text.

@@
 ---
-title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
+title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
 subtitle: "Supplementary Material"
@@
 psychometric scoring of depression, anxiety and mental stress as well as blood levels of inflammatory markers and metabolites of the kynurenine and catecholamine pathway. 
-A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were analyzed here. 
+A total of `r nrow(stigma$analysis_tbl)` SIMMUN study participants were analyzed. 
 The SIMMUN study variables with their transformation and stratification schemes are listed in __Supplementary Table S\@ref(tab:tab-study-vars)__.
@@
 
-The inflammatory markers included in the analysis, plasma concentration of neopterin (NEO) measured by ELISA and neutrophil/lymphocyte ratio (NLR) were determined by a certified clinical routine laboratory at the University Hospital of Innsbruck.
-C-reactive protein (CRP) and interleukin-6 (IL6) were determined in the SIMMUN cohort as well but excluded from the analysis, since their levels were within the reference range in the great majority of participants.
+Neutrophil/lymphocyte ratio (NLR) and serum neopterin concentration were used as markers of inflammation. 
+NLR was determined by a certified clinical routine laboratory at the University Hospital of Innsbruck. 
+NEO levels were measured by enzyme-linked immunosorbent assay (BRAHMS Diagnostics, Berlin, Germany). 
+In addition to NLR and NEO, C-reactive protein (CRP) and interleukin-6 (IL6) were determined in the SIMMUN cohort as well but excluded from the analysis, since their levels were within the reference range in the great majority of participants.
 Elevated CRP > 0.5 mg/dL was detected in 
@@
 elevated IL6 > 7 pg/mL was observed in 
-`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of the analyzed SIMMUN cohort. 
-Plasma concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, as described elsewhere [@Widner1997; @Neurauter2008].
+`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of the analyzed SIMMUN participants. 
+Serum concentrations of tryptophan (TRP), kynurenine (KYN), phenylalanine (PHE) and tyrosine (TYR) were determined by high-performance liquid chromatography, as described elsewhere [@Widner1997; @Neurauter2008].
 Kynurenine - tryptophan and phenylalanine - tyrosine ratios (KYN/TRP and PHE/TYR) were used as readouts of systemic indoleamine 2,3-dioxygenase (IDO, kynurenine pathway) and phenylalanine hydroxylase (PAH, catecholamine pathway) activity, respectively [@Capuron2011].
-Plasma titer of anti-RBD IgG were quantified by ELISA as described before [@Deisenhammer2021]. 
+Serum titer of anti-RBD IgG were quantified by ELISA as described before [@Deisenhammer2021]. 
 Strongly non-normally distributed anti-RBD IgG levels were expressed as arbitrary units (AU) and stratified with the $\leq$ 1, 1 - 16.3 and > 16.3 AU cutpoints defined by the maximal titer in the uninfected participants (1 AU) and median titer in the SARS-CoV-2 subset (16.3 AU).
@@
 The total possible score range for each subscale is 0 to 21, with higher scores indicating more severe symptoms of anxiety/depression. 
-Clinically relevant signs of anxiety or depression were identified with the cutoff of $\geq$ 8 points at the respective subscale as proposed [@Bjelland2002; @Zigmond1983]. 
+Clinically relevant symptoms of anxiety or depression were identified with the cutoff of $\geq$ 8 points at the respective subscale as proposed [@Bjelland2002; @Zigmond1983]. 
 PSS-4, HADS anxiety and depression scales displayed good-to-excellent internal consistency as measured by the McDonald $\omega$ statistic [@McDonald1999] (__Supplementary Figure S\@ref(fig:fig-consistency)__).
@@
 (`r suppl_tabs$study_vars %>% filter(Transformation == 'square root') %>% .[['Variable label']] %>% paste(collapse = ', ')`) were applied. 
-Strongly non-normally distributed HADS scores and anti-RBD IgG were stratified with the cutoffs described above. 
+The strongly non-normally distributed HADS scores and anti-RBD IgG titers were stratified with the cutoffs described above. 
 Transformation and stratification schemes of study variable are provided in __Supplementary Table S\@ref(tab:tab-study-vars)__.
@@
 
-Concerning the INCOV study design, plasma metabolites and proteome were determined in uninfected controls and SARS-CoV-2 individuals at three timepoints after diagnosis: 
+Concerning the INCOV study design, plasma metabolome and proteome analyses were performed in uninfected participants and SARS-CoV-2-infected individuals at three timepoints after diagnosis: 
 acute (median `r (blast(incov$analysis_tbl, timepoint) %>% map_dbl(~round(median(.x$time_po))))['acute']` days), 
@@
 The distribution testing revealed substantial deviations from normality for some SIMMUN study parameters. 
-For this reason, logarithm and square root transformations were applied prior to modeling and statistical hypothesis testing with parametric tools as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__.
+Therefore, logarithm and square root transformations were applied to non-normally distributed variables prior to modeling and statistical hypothesis testing with parametric tools. 
+Transformations of the SIMMUN study variables are specified in __Supplementary Table S\@ref(tab:tab-study-vars)__. 
 Most of the INCOV study variables were non-normally distributed. 
@@
 Differences in characteristic of SIMMUN participants included in the analysis and excluded due to data missingness were assessed by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Supplementary Table S\@ref(tab:tab-excluded)__).
-Differences in characteristic of SIMMUN participants or INCOV study participants divided by the SARS-CoV-2 infection status were investigated by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size metric for numeric and categorical features, respectively (__Tables 1__ and __2__).
+Differences in characteristic of SIMMUN study participants or INCOV study participants divided by the SARS-CoV-2 infection status were investigated by $\chi^2$ test with Cramer's V effect size statistic and Mann-Whitney test with r effect size statistic for numeric and categorical features, respectively (__Tables 1__ and __2__).
 Correlation of metabolite concentrations with age, scoring of mental stress (PSS-4) and markers of inflammation (NEO and NLR) in the SIMMUN cohort was done by Pearson's test (__Supplementary Table S\@ref(tab:tab-univariate-correlation)__).
@@
 
-Effects of age, sex, body mass class, physical and mental disorders, BMI class, smoking and alcohol consumption, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress scoring (PSS-4) on readouts of the kynurenine (TRP, KYN, KYN/TRP) and catecholamine pathway activity (PHE, TYR, PHE/TYR) were assessed by multi-parameter linear regression with backward elimination.
+Effects of age, sex, physical and mental disorders, BMI class, smoking and alcohol consumption, the inflammation markers NEO and NLR, SARS-CoV-2 infection status, anti-RBD IgG titer, clinically relevant symptoms of depression and anxiety (HADS), mental stress scoring (PSS-4) on readouts of the kynurenine (TRP, KYN, KYN/TRP) and catecholamine pathway activity (PHE, TYR, PHE/TYR) were assessed by multi-parameter linear regression with backward elimination.
 Responses and numeric explanatory variables were transformed with log or squared root functions to improve normality as specified in __Supplementary Table S\@ref(tab:tab-study-vars)__ and normalized prior to modeling.
@@
 Significance of the estimates ($p(\beta\neq0)$) was determined by two-tailed T test. 
-Results of the linear modeling in the INCOV cohort are presented in __Figure 2__ and __Supplementary Table S\@ref(tab:tab-modeling-simmun)__.
+Results of the linear modeling in the SIMMUN cohort are presented in __Figure 2__ and __Supplementary Table S\@ref(tab:tab-modeling-simmun)__.
 
@@
 
-Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on systemic neurotransmitter availability were subsequently tested in the publicly available INCOV cohort [@Su2022].
-We modeled normalized plasma levels of serotonin as the product of the serotonin pathway [@Lukic2022], and the major product of systemic dopamine catabolism, dopamine 3-O-sulfate (DA sulfate) [@Goldstein1999; @Meiser2013].
-The explanatory variables were age, sex and BMI class, plasma levels of cytokines markers of inflammation (IL6, IL10, TNF, IFNG), plasma concentrations of metabolites of the serotonin and kynurenine pathway (TRP, KYN, QUIN), plasma concentrations of metabolites of the catecholamine pathway (PHE, TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected control).
+Effects of the key explanatory factors identified by multi-parameter linear modeling in the SIMMUN cohort on kynurenine and catecholamine pathway activity were subsequently tested in the publicly available INCOV cohort [@Su2022].
+We modeled normalized plasma levels of serotonin, and the major product of systemic dopamine catabolism, dopamine 3-O-sulfate (DA sulfate) [@Goldstein1999; @Meiser2013]. 
+The explanatory variables were age, sex and BMI class, plasma levels of cytokines markers of inflammation (IL6, IL10, TNF, IFNG), plasma concentrations of metabolites of the kynurenine (TRP, KYN, QUIN) and catecholamine pathway (PHE, TYR) and timepoint of SARS-CoV-2 infection (acute, sub-acute, recovery versus uninfected individuals).
 Since most numeric features were non-normally distributed, MM algorithm robust linear regression with Huber's psi function was chosen [@Huber2011; @Ripley2022] as a modeling approach.
@@
 
-Differences in normalized blood concentrations of cytokines and metabolites between uninfected controls, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset or acute infection serving as baselines.
+Differences in normalized blood concentrations of cytokines and metabolites between uninfected individuals, acute and sub-acute SARS-CoV-2 infection and recovery were investigated with robust linear modeling with uninfected subset or acute infection serving as baselines.
 Models were constructed with the `make_lm()` function (package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)) employing internally the MM `rlm()` algorithm with Huber's psi function (package _MASS_) [@Huber2011; @Ripley2022].
@@
 
-Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of cytokines (IL6, IL10, TNF and IFNG), metabolites of the serotonin and kynurenine pathways (serotonin, TRP, KYN and QUIN), and metabolites of the catecholamine pathway (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
+Pairwise Spearman's correlation coefficients $\rho$ were calculated for plasma concentrations of cytokines (IL6, IL10, TNF and IFNG), serotonin and metabolites of the kynurenine (TRP, KYN and QUIN), and catecholamine pathway (PHE, TYR and DA sulfate) with the `cor()` function from the _stats_ package at each of the timepoints of SARS-CoV-2 infection (uninfected, acute, sub-acute and recovery).
 The correlation matrices were subsequently scaled into the [0, 1] range with the $0.5\times(x + 1)$ function and converted to undirected force directed graphs with the `graph_from_adjacency_matrix()` function from the _igraph_ package [@Csardi2006].
@@
   footnote(1, 2, 
-           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinic acid; PHE: phenylalanine; 5-HT: serotonin; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'),
+           value = as_paragraph('TNF: tumor necrosis factor-alpha; IFNG: interferon gamma; IL6: interleukin-6; IL10: interleukin-10; QUIN: quinolinic acid; PHE: phenylalanine; TRP: tryptophan; TYR: tyrosine; KYN: kynurenine; DA sulfate: dopamine 3-O-sulfate.'),
            part = 'header', 
@@
 
-```{r tab-modeling-simmun, tab.cap = 'Results of multi-parameter linear modeling of plasma concentrations of tryptophan, kynurenine, tyrosine, and kynurenine - tryptophan, and phenylalanine - tyrosine ratios in the SIMMUN cohort.'}
+```{r tab-modeling-simmun, tab.cap = 'Results of multi-parameter linear modeling of serum concentrations of tryptophan, kynurenine, tyrosine, and kynurenine/tryptophan and phenylalanine/tyrosine ratios in the SIMMUN cohort.'}
 
@@
 
-```{r tab-univariate-correlation, tab.cap = "Correlation of age, perceived mental stress scoring, plasma neopterin levels and neutrophil/leukocyte ratio with plasma metabolite levels in the SIMMUN cohort investigated by Pearson's test."}
+```{r tab-univariate-correlation, tab.cap = "Correlation of age, perceived mental stress scoring, serum neopterin levels and neutrophil/leukocyte ratio with serum levels of metabolites of the kynurenine and catecholamine pathways in the SIMMUN cohort investigated by Pearson's test."}
 
@@
 
-```{r tab-univariate-complarison, tab.cap = "Comparison of plasma metabolite concentrations in SIMMUN study participants split by clinically relevant depression signs and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}
+```{r tab-univariate-complarison, tab.cap = "Comparison of serum metabolite concentrations in SIMMUN study participants split by presence of clinically relevant symptoms of depression and SARS-CoV-2 infection status with two-tailed T test and Cohen's d effect size statistic."}
 
@@
   footnote(1, 1, 
-           value = as_paragraph('5-HT: 5-hydroxy tryptamine/serotonin; DA sulfate: dopamine 3-O-sulfate.'), 
+           value = as_paragraph('DA sulfate: dopamine 3-O-sulfate.'), 
            part = 'header', 
@@
 
-```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of serum levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort.
+```{r tab-time-uninfected, tab.cap = 'Results of robust linear modeling of plasma levels of cytokine markers of inflammation, serotonin, and metabolites of the kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort.
 The uninfected subset served as a baseline.'}
@@
 
-```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of serum levels of cytokine markers of inflammation, and metabolites of the serotonin, kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort.
+```{r tab-time-acute, tab.cap = 'Results of robust linear modeling of plasma levels of cytokine markers of inflammation, serotonin, and metabolites of the kynurenine and catecholamine pathways as a function of SARS-CoV-2 infection timepoint in the INCOV cohort.
 The acute SARS-CoV-2 infection subset served as a baseline.'}
@@
 
-_TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine/tryptophan ratio; PHE/TYR: phenylalanine/tyrosine ratio; NLR: neutrophil - leukocyte ratio; QUIN: quinolinic acid; 5-HT: serotonin; DA: dopamine; Ab: antibody; IL6: interleukin-6; IL10: interleukin-10; TNF: tumor-necrosis factor alpha; IFNG: interferon gamma._
+_TRP: tryptophan; KYN: kynurenine; PHE: phenylalanine; TYR: tyrosine; KYN/TRP: kynurenine/tryptophan ratio; PHE/TYR: phenylalanine/tyrosine ratio; NLR: neutrophil - leukocyte ratio; QUIN: quinolinic acid; DA: dopamine; Ab: antibody; IL6: interleukin-6; IL10: interleukin-10; TNF: tumor-necrosis factor alpha; IFNG: interferon gamma._
 
@@
 
-_Multi-parameter linear regression models of plasma levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by cross-validation (CV, 10 folds).
+_Multi-parameter linear regression models of serum levels of tryptophan (TRP), kynurenine (KYN), kynurenine/tryptophan ratio (KYN/TRP), phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TRP) in the SIMMUN models were optimized by backwards elimination and their reproducibility was tested by cross-validation (CV, 10 folds).
 Values of root mean square error (RMSE, A) and R^2^ (B) in the training data set and CV are plotted._
@@
 
-_Age, systemic inflammatory markers: neopterin (NEO), neutrophil/lymphocyte ratio (NLR), SARS-CoV-2 infection status, clinically relevant symptoms of depression (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of plasma levels of tryptophan (TRP), kynurenine (KYN) and kynurenine/tryptophan ratio (KYN/TRP).
-Their association with plasma concentrations of these metabolites was investigated by correlation analysis and statistical hypothesis testing._
+_Age, systemic inflammatory markers: neopterin (NEO), neutrophil/lymphocyte ratio (NLR), SARS-CoV-2 infection status, clinically relevant symptoms of depression (hospital anxiety and depression scale [HADS] > 8 points) and mental stress scoring (perceived stress scale, 4 item [PSS-4]) were identified as significant determinants of serum levels of tryptophan (TRP), kynurenine (KYN) and kynurenine/tryptophan ratio (KYN/TRP).
+Their association with serum concentrations of these metabolites was investigated by correlation analysis and statistical hypothesis testing._
 
@@
 (A) Correlation coefficients are presented in a bubble plot with point size and color corresponding to values of correlation coefficients (r); points are labeled with their r values, significant effects are highlighted in bold.
-(B) Scatter plots of plasma concentrations of NEO, TRP, KYN and KYN/TRP.
+(B) Scatter plots of serum concentrations of NEO, TRP, KYN and KYN/TRP.
 Points represent single observations, fitted linear trends with 95% confidence intervals are visualized as blue lines with gray ribbons.
@@
 
-_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of plasma levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TYR).
-Their association with plasma concentrations of these metabolites was investigated by correlation analysis and statistical hypothesis testing._
+_Age, the systemic inflammatory marker neopterin (NEO), and SARS-CoV-2 infection status were identified as significant determinants of serum levels of phenylalanine (PHE), tyrosine (TYR) and phenylalanine/tyrosine ratio (PHE/TYR).
+Their association with serum concentrations of these metabolites was investigated by correlation analysis and statistical hypothesis testing._
 
@@
 
-_(C) Levels of the metabolite levels were compared between infected and uninfected participants by two-tailed T test with Cohen's d effect size statistic.
+_(C) Levels of the metabolites were compared between infected and uninfected participants by two-tailed T test with Cohen's d effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
@@
 
-_Plasma levels of metabolites of the serotonin and kynurenine pathways (5-hydroxy tryptamine/serotonin [5-HT], tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]) and metabolites of the catecholamine pathway (phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]) were correlated with plasma levels of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) in uninfected individuals, and during acute (median 10 days), sub-acute (median 14 days) SARS-CoV-2 infection and recovery (median 64 days after SARS-CoV-2 infection diagnosis via PCR).
+_Plasma levels of serotonin and metabolites of the kynurenine (tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]) and catecholamine pathways (phenylalanine [PHE], tyrosine [TYR], dopamine 3-O-sulfate [DA sulfate]) were correlated with plasma levels of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) in uninfected individuals, and during acute (median 10 days), sub-acute (median 14 days) SARS-CoV-2 infection and recovery (median 64 days after SARS-CoV-2 infection diagnosis via PCR).
 Statistical significance was determined by Spearman's rank test corrected for multiple testing with the false discovery rate method.

### 2023-06-11 exploration scripts/cohort.R

@@ -211,7 +211,7 @@
                               cohort$samples,
                               by = 'timepoint') %>%
     set_names(c('Time point',
-                'Days post infection',
+                'Days since positive PCR test',
                 'Sample number'))
 # END -----
 

### 2023-06-11 exploration scripts/cohort_comparison.R

@@ -16,7 +16,7 @@
     c('sex' = 'Sex',
       'age' = 'Age, years',
       'bmi_class' = 'Body mass',
-      'cov' = 'SARS-CoV-2',
+      'cov' = 'SARS-CoV-2 infection',
       'severity' = 'SARS-CoV-2 hospitalization') %>%
     compress(names_to = 'variable',
              values_to = 'label')

### 2023-06-11 exploration scripts/excluded.R

@@ -16,13 +16,13 @@
     c(sex = 'Sex',
       age = 'Age, years',
       bmi_class = 'Body mass class',
-      somatic_comorb = 'Somatic comorbidity',
-      psych_comorb = 'Psychiatric comorbidity',
+      somatic_comorb = 'Physical illness',
+      psych_comorb = 'Mental illness',
       hads_anx_score = 'HADS anxiety score',
       hads_dpr_score = 'HADS depression score',
       hads_signs = 'Depression or anxiety signs, HADS \u2265 8',
       pss_stress_score = 'PSS-4 stress score',
-      cov = 'Infection',
+      cov = 'SARS-CoV-2 infection',
       anti_rbd = 'anti-RBD SARS-CoV-2, IgG, AU',
       severity = 'COVID-19 severity',
       trp = 'TRP, µmol/L',
@@ -49,7 +49,14 @@
                                     c('analyzed', 'excluded'))) %>%
     select(patient_id,
            analysis_status,
-           all_of(excl$var_lexicon$variable))
+           all_of(excl$var_lexicon$variable)) %>%
+    mutate(cov = car::recode(cov,
+                             "'healthy' = 'uninfected'"),
+           cov = factor(cov, c('uninfected', 'SARS-CoV-2')),
+           severity = car::recode(severity,
+                                  "'healthy' = 'uninfected'"),
+           severity = factor(severity,
+                             c('uninfected', 'ambulatory', 'hospitalized')))
 
 # Descriptive stats -----
 

### 2023-06-11 import scripts/incov.R

@@ -52,7 +52,7 @@
                                'phenylalanine' = 'PHE';
                                'tyrosine' = 'TYR';
                                'dopamine 3-O-sulfate' = 'DA sulfate';
-                               'serotonin' = '5-HT';
+                               'serotonin' = 'serotonin';
                                'quinolinate' = 'QUIN'"),
            unit = 'Z-score of log\u2082 concentration',
            axis_lab = paste(label, unit, sep = ', '))

### 2023-06-11 incov scripts/correlation.R

@@ -122,7 +122,7 @@
            map(mutate,
                variable1 = variable2,
                variable2 = timepoint),
-         plot_title = c('5-HT, INCOV', 'DA sulfate, INCOV')) %>%
+         plot_title = c('Serotonin, INCOV', 'DA sulfate, INCOV')) %>%
     pmap(corr_buble,
          signif_only = FALSE) %>%
     map(~.x +

### 2023-06-11 incov scripts/mds.R

@@ -157,7 +157,10 @@
 
   incov_mds$network_plots <-
     list(graph = incov_mds$graph_obj,
-         plot_title =  paste0(c('Uninfected', 'Acute Cov', 'Sub-acute CoV', 'Recovery'),
+         plot_title =  paste0(c('Uninfected',
+                                'Acute SARS-CoV-2 infection',
+                                'Sub-acute SARS-CoV-2 infection',
+                                'Recovery'),
                               ', INCOV') %>%
            rep(2),
          plot_subtitle = paste('n =', map_dbl(incov_mds$analysis_tbl, ncol))) %>%

### 2023-06-11 incov scripts/neuro_modeling.R

@@ -137,10 +137,8 @@
                                 'Explained variance, INCOV')) %>%
     map(~.x +
           scale_y_discrete(limits = rev(names(incov_neuro$caret_models)),
-                           labels = exchange(names(incov_neuro$caret_models),
-                                             dict = incov_neuro$response_lexicon,
-                                             key = 'variable',
-                                             value = 'label')))
+                           labels = c(serotonin = 'serotonin',
+                                      dopamine = 'DA sulfate')))
 
 # Inference --------
 
@@ -160,8 +158,8 @@
 
   incov_neuro$forest_plots <-
     list(x = incov_neuro$inference,
-         plot_title = c('5-HT', 'DA sulfate'),
-         x_lab = paste0(c('5-HT', 'DA sulfate'),
+         plot_title = c('Serotonin', 'DA sulfate'),
+         x_lab = paste0(c('Serotonin', 'DA sulfate'),
                         ', Z-score')) %>%
     pmap(plot_forest,
          variable = 'var_lab',

### 2023-06-11 paper scripts/render.R

@@ -12,7 +12,7 @@
 
   insert_msg('Rendering the issue file')
 
-  render('./paper/markdown/issues.Rmd',
+  render('./paper/markdown/issues_10062023.Rmd',
          output_format = word_document2(number_sections = FALSE,
                                         reference_docx = 'ms_template.docx'),
          output_dir = './paper')

### 2023-06-11 paper scripts/supplement.R

@@ -87,8 +87,7 @@
                       legend.position = 'none') +
                 labs(subtitle = paste('Pearson correlation, n =',
                                       nrow(stigma$analysis_tbl)),
-                     title = paste0(tst_trp$correlation$correlogram$labels$title,
-                                    ', SIMMUN')),
+                     title = 'Kynurenine pathway, SIMMUN'),
               get_legend(tst_trp$correlation$correlogram),
               ncol = 2,
               rel_widths = c(0.65, 0.35),
@@ -151,8 +150,7 @@
                       legend.position = 'none') +
                 labs(subtitle = paste('Pearson correlation, n =',
                                       nrow(stigma$analysis_tbl)),
-                     title = paste0(tst_tyr$correlation$correlogram$labels$title,
-                                    ', SIMMUN')),
+                     title = 'Catecholamine pathway, SIMMUN'),
               get_legend(tst_tyr$correlation$correlogram),
               ncol = 2,
               rel_widths = c(0.55, 0.45),

### 2023-06-11 paper scripts/tables.R

@@ -34,7 +34,7 @@
                            'Mental disorder',
                            'HADS anxiety score',
                            'HADS depression score',
-                           'Clinically relevant signs of depression or anxiety, HADS ≥ 8',
+                           #'Clinically relevant signs of depression or anxiety, HADS ≥ 8',
                            'PSS-4 mental stress score',
                            'anti-RBD SARS-CoV-2, IgG, AU',
                            'COVID-19 severity'))
@@ -103,7 +103,12 @@
 
   insert_msg('Table S2: included and excluded participants, SIMMUN')
 
+  ## the variable 'Clinically relevant signs of depression or anxiety, HADS ≥ 8'
+  ## is redundant and removed from the tables and supplementary tables
+
   suppl_tabs$excluded <- excl$feat_table %>%
+    filter(!Variable %in% c('Clinically relevant signs of depression or anxiety, HADS ≥ 8',
+                            'Depression or anxiety signs, HADS ≥ 8')) %>%
     mdtable(label = 'table_s2_included_excluded',
             ref_name = 'excluded',
             caption = paste('Significant differences between participant',

### 2023-06-11 tools/globals.R

@@ -53,7 +53,7 @@
          sex = c('sex', NA, NA),
          psych_comorb = c('mental disorder', NA, NA),
          somatic_comorb = c('physical disorder', NA, NA),
-         infection = c('SARS-CoV2', NA, NA),
+         infection = c('SARS-CoV-2', NA, NA),
          bmi_class = c('body mass index', NA, NA),
          smoking = c('smoking', NA, NA),
          alcohol = c('alcohol', NA, NA),

### 2023-07-19 paper/markdown/main_manuscript.Rmd
Paper proofs

@@
 SARS-CoV-2-positivity (`r my_beta(mod_eli$inference$kyn_trp, 'infection', show_ci_txt = FALSE)`), 
-mental stress (`r my_beta(mod_eli$inference$kyn_trp, 'pss_stress_score', show_ci_txt = FALSE)`) and age were positively associated with elevated kynurenine/tryptophan ratio. 
+mental stress (`r my_beta(mod_eli$inference$kyn_trp, 'pss_stress_score', show_ci_txt = FALSE)`) and 
+age (`r my_beta(mod_eli$inference$kyn_trp, 'age', show_ci_txt = FALSE)`) were positively associated with elevated kynurenine/tryptophan ratio. 
 Phenylalanine/tyrosine ratio was lower in 
@@
 
-As compared with the SIMMUN cohort, the INCOV cohort had a significantly higher percentage of males (`r comp$test %>% filter(variable == 'sex') %>% .$significance`, 
+As compared with the SIMMUN cohort, the INCOV cohort had a significantly higher percentage of 
+males (`r comp$test %>% filter(variable == 'sex') %>% .$significance`, 
 effect size: `r comp$test %>% filter(variable == 'sex') %>% .$eff_size`), 
 higher age (`r comp$test %>% filter(variable == 'age') %>% .$significance`, 
-effect size: `r comp$test %>% filter(variable == 'age') %>% .$significance`), 
+effect size: `r comp$test %>% filter(variable == 'age') %>% .$eff_size`), 
 higher rates of overweight or obese participants (`r comp$test %>% filter(variable == 'bmi_class') %>% .$significance`, 
@@
 
-# Acknowledgements
+# Acknowledgments
 
@@
 
-_Effects of age, timepoint of SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR), plasma levels of metabolites of the kynurenine (tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]) and catecholamine pathways (phenylalanine [PHE], tyrosine [TYR]), and plasma concentrations of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin and dopamine 3-O-sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
+_Effects of age, sex, body mass index (BMI), timepoint of SARS-CoV-2 infection (acute: median 10 days, sub-acute: median 14 days, recovery: median 64 days after SARS-CoV-2 infection diagnosis via PCR), plasma levels of metabolites of the kynurenine (tryptophan [TRP], kynurenine [KYN], quinolinic acid [QUIN]) and catecholamine pathways (phenylalanine [PHE], tyrosine [TYR]), and plasma concentrations of cytokine markers of inflammation (interleukin-6 [IL6], interleukin-10 [IL10], tumor necrosis factor-alpha [TNF] and interferon-gamma [IFNG]) on plasma concentrations of serotonin and dopamine 3-O-sulfate (DA sulfate) were modeled by multi-parameter robust linear regression._
 

### 2023-07-19 tools/tools.R

@@ -1002,14 +1002,14 @@
       beta_dat <-  beta_dat %>%
         mutate(txt_lab = paste0(signif(estimate, signif_digits),
                                 ' [95% CI: ', signif(lower_ci, signif_digits),
-                                '-', signif(upper_ci, signif_digits), ']'))
+                                ' to ', signif(upper_ci, signif_digits), ']'))
 
     } else {
 
       beta_dat <-  beta_dat %>%
         mutate(txt_lab = paste0(signif(estimate, signif_digits),
                                 ' [', signif(lower_ci, signif_digits),
-                                '-', signif(upper_ci, signif_digits), ']'))
+                                ' to ', signif(upper_ci, signif_digits), ']'))
 
     }
 
